### 2015-11-14 paper.tex
abstraction

%
% LaTeX template for prepartion of submissions to PLDI'16
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'16 web site.
%
\documentclass[pldi]{sigplanconf-pldi16}
% \documentclass[pldi-cameraready]{sigplanconf-pldi16}

%
% the following standard packages may be helpful, but are not required
%
\usepackage{SIunits}            % typset units correctly
\usepackage{courier}            % standard fixed width font
\usepackage[scaled]{helvet}     % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                % format URLs
\usepackage{listings}           % format code
\usepackage{enumitem}           % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI

% \usepackage[utf8]{inputenc}
% \usepackage[english]{babel}
\usepackage{minted}     % code highlighting
\usemintedstyle{tango}

\begin{document}

\title{Catching runtime errors of Redis at compile time}

%
% any author declaration will be ignored  when using 'pldi' option (for double blind review)
%

\authorinfo{Person 1 \and Person 2}
{\makebox{A Department} \\
\makebox{A University}  \\
\makebox{A Place, AS 12345}}
{\{person1,person2\}@cs.auniv.edu}

\maketitle

\begin{abstract}
We often interacts with database systems through dedicated libraries or packages
implemented in some programming languages. People make mistakes, while some of
the syntactic and semantic errors can be prevented by the language at compile
time, or caught by the package at runtime, most semantic errors are just
being ignored, causing problems at the database system.

In this paper, we demonstrates how to prevent those runtime errors at compile
time, and allow users to write reliable database queries without runtime overhead,
by exploiting type-level programming techniques such as parametedzied monad,
type-level literals and type families in Haskell.

The database system and the package we are targeting
are \emph{Redis} and \emph{Hedis} respectively, and our implementation is
available as \emph{Edis} on \emph{Hackage}.

\paragraph{Categories and Subject Descriptors}
\paragraph{Keywords}
type-level programming; Haskell; database query language; Redis;
\end{abstract}

\section{Introduction}

\subsection{Redis}

Redis is an open source, in-memory data structure store, used as
database, cache and message broker. Each value is associated with a binary-safe
string key to identify and manipulate with. Redis supports many different
kind of values, such as strings, hashes, lists and sets.

To manipulate these values, Redis comes with a set of atomic operations,
called \emph{commands}.

For example, if we want to add a bunch of strings to 2 sets, and intersect them,
we could input the following sequence of commands into Redis's client interactively.

\begin{minted}{text}
redis> SADD some-set a b c
(integer) 3
redis> SADD another-set a b
(integer) 2
redis> SINTER some-set another-set
1) "a"
2) "b"
\end{minted}

Keys such as \mintinline{text}{some-set} and \mintinline{text}{another-set}
are created on site, with the command \mintinline{text}{SADD}, which returns
the size of set after completed.

\subsection{Hedis}
Redis can also be used in most programming languages, with 3rd party libraries
or packages that talk with Redis' TCP protocol. In Haskell, the most popular
package people are using is \emph{Hedis}.

The previous example in Redis's client would look something like this with Hedis
in Haskell.
\begin{minted}{haskell}
program :: Redis (Either Reply [ByteString])
program = do
    sadd "some-set" ["a", "b"]
    sadd "another-set" ["a", "b", "C"]
    sinter ["some-set", "another-set"]
\end{minted}

Function \mintinline{haskell}{sadd} takes a key and a list of values and returns an
\mintinline{haskell}{Integer},
wrapped in the context \mintinline{haskell}{Redis (Either Reply Integer)}
to which encapsulates possible failures.

 to indicate it's size. \mintinline{haskell}{sinter}
takes a list of keys and returns

 , both of type
  \mintinline{haskell}{ByteString}, and returns a \mintinline{haskell}{Integer}
  to indicate it's size.

  since keys and values are all just binary strings in Redis.

 and \mintinline{haskell}{sinter} now

\begin{minted}{haskell}
sadd :: ByteString      -- key
    -> [ByteString]     -- values
    -> Redis (Either Reply Integer)
sinter :: [ByteString]  -- keys
    -> Redis (Either Reply [ByteString])
\end{minted}

\end{document}

### 2015-11-14 pldi16-template.tex
abstraction

%
% LaTeX template for prepartion of submissions to PLDI'16
%
% Requires temporary version of sigplanconf style file provided on
% PLDI'16 web site.
%
\documentclass[pldi]{sigplanconf-pldi16}
% \documentclass[pldi-cameraready]{sigplanconf-pldi16}

%
% the following standard packages may be helpful, but are not required
%
\usepackage{SIunits}            % typset units correctly
\usepackage{courier}            % standard fixed width font
\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
\usepackage{url}                  % format URLs
\usepackage{listings}          % format code
\usepackage{enumitem}      % adjust spacing in enums
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI

\begin{document}

\title{Instructions for Submission to PLDI'16}

%
% any author declaration will be ignored  when using 'pldi' option (for double blind review)
%

\authorinfo{Person 1 \and Person 2}
{\makebox{A Department} \\
\makebox{A University}  \\
\makebox{A Place, AS 12345}}
{\{person1,person2\}@cs.auniv.edu}

\maketitle

\begin{abstract}
  This document is intended to serve as a sample for submissions to PLDI'16,
  the 37th Annual ACM SIGPLAN Conference on Programming Language
  Design and Implementation.  We provide some guidelines
  that authors should follow when submitting papers to the conference.
\end{abstract}

\section{Introduction}

This document provides instructions for submitting papers to PLDI'16.
In an effort to respect the efforts of reviewers and in the interest
of fairness to all prospective authors, we request that all
submissions to PLDI'16 follow the formatting and submission rules
detailed below. \textbf{Submissions that violate these instructions
  may not be reviewed}, at the discretion of the
\href{mailto:emery@cs.umass.edu?subject=[PLDI'16]}{program
  chair}, in order to maintain a review process that is fair to all
potential authors.

\paragraph{Paper evaluation objectives}
The committee will make every effort to judge each submitted paper on
its own merits. There will be no target acceptance rate.  We expect to
accept a wide range of papers with appropriate expectations for
evaluation --- while papers that build on significant past work with
strong evaluations are valuable, papers that open new areas with less
rigorous evaluation are equally welcome and especially
encouraged. \textbf{In either case, what is important is that the
  paper's claims are no stronger than what is supported by the
  evaluation.}  Given the wide range of topics covered by PLDI, every
effort will be made to find expert reviewers.

All questions regarding paper formatting and submission should be
directed to the
\href{mailto:emery@cs.umass.edu?subject=[PLDI'16]}{program
  chair}.

\paragraph{Highlights From This Document}
\begin{itemize}[noitemsep]
\item Paper must be submitted in printable PDF format.
\item Text must be in a minimum \textbf{10\,pt font} (\emph{not} 9\,pt).
\item Papers at most \textbf{11 pages}, not including references.
\item There is no page limit for references.
\item Each reference must specify \textbf{all authors} (no \emph{et al.}).
\item The reviewing process will include two phases.
  \begin{itemize}
  \item There will be an initial round of reviews.
  \item Those papers identified as most worthy of further consideration will proceed to the second phase and receive additional reviews.
  \item Authors \emph{will not} be required, nor given the opportunity, to revise their submissions after phase one.
  \end{itemize}
\item Authors of \emph{all} accepted papers will be required to give a
  lightning presentation (about 90\,s) in addition to
  the regular conference talk.
\item Proceedings may appear in the ACM digital library as early as as early as May 30, 2016.
\end{itemize}

\section{Paper Preparation Instructions}

\subsection{Paper Formatting}

Papers must be submitted in printable PDF format and should contain a
\textbf{maximum of 11 pages} of single-spaced two-column text,
\textbf{not including references}.  In-text citations may follow either the
\textbf{ACM / TOPLAS author-year style (`[Smith 1990]')},
\emph{or} \textbf{the numerical style (`[1]')}.  You may include any number of pages for
references.  If you are using
\LaTeX~\cite{Lamport:94} to typeset your paper, then we suggest that
you use the temporary versions of the SIGPLAN
\LaTeX~\href{https://github.com/emeryberger/PLDI-2016/blob/master/sigplanconf-pldi16.cls}{class
  file} and
\href{https://github.com/emeryberger/PLDI-2016/blob/master/pldi16-template.tex}{template}.
If you are using
Microsoft Word, you may wish to use
\href{http://conf.researchr.org/getImage/pldi2015/orig/pldi15-word-template.dot}{this}
template.  \textbf{Whichever tool you use, please ensure you adhere to the
guidelines given in Table~\ref{table:formatting}.}  The conference
submission website will use the
\href{https://www.usenix.org/legacy/event/wowcs08/tech/full_papers/voelker/voelker.pdf}{banal}
format checker to \emph{advise} on formatting compliance.

\begin{table}[t!]
  \centering
{  \sffamily\small % tabular data either 10pt times, or 9pt helvetica
  \begin{tabular}{rl}
    \textbf{Field} & \textbf{Value}\\
    \hline
    File format & PDF \\
    Page limit & 11 pages, \textbf{excluding references}\\
    Paper size & US Letter 8.5in $\times$ 11in\\
    Top margin & 1in\\
    Bottom margin & 1in\\
    Left margin & 0.75in\\
    Right margin & 0.75in\\
    Body & 2-column, single-spaced\\
    Column separation & 0.25in\\
    Body font & 10pt\\
    Tabular font & 10pt Times (or 9pt Helvetica)\\
    Abstract font & 10pt\\
    Section heading  & 12pt, bold\\
    Subsection heading  & 10pt, bold\\
    Caption font & 10pt\\
%    In-text citation & Author-year (`[Smith 1990]' \emph{\textbf{not}} `[3]') \\
    References & 10pt, no page limit, list all authors' names\\
  \end{tabular}
}
  \caption{Formatting guidelines for submission. }
  \label{table:formatting}
\end{table}

For the \emph{convenience of reviewers}, all submissions will be
automatically watermarked by the paper submission site
(\href{http://conf.researchr.org/getImage/pldi2015/orig/pldi15-template-wm.pdf}{example
  here}), so please \textbf{avoid adding any text in the margins or
  headers} such as footers, headers, or banners,
other than the centered page number provided by the template (which is
only for the author's convenience, since the watermarking adds page
numbers).  The watermarking process strips the submitted pdf of all
metadata (thereby removing information which might otherwise identify authors).

\subsection{Content}

\paragraph{Double Blind.}  Reviewing will be double blind; therefore,
please do not include any author names on any submitted documents
except in the space provided on the online submission form.  Please
take time to read the
\href{http://conf.researchr.org/track/pldi-2016/pldi-2016-papers#FAQ-on-Double-Blind-Reviewing}{PLDI
  FAQ on Double Blind Reviewing}, which gives a more comprehensive and
authoritative account than this short paragraph.  If you are improving
upon your prior work, refer to your prior work in the third person and
include a full citation for the work in the bibliography. For example,
if you happened to be Collins and McCarthy, building on \emph{your
  own} prior work, you might say something like: ``While prior work
\cite{Backus:60,Collins:60,McCarthy:60} did X, Y, and Z, this paper
additionally does W, and is therefore much better.''  \textbf{Do NOT omit or
anonymize references for blind review}.

\paragraph{Figures and Tables.} Ensure that the figures and tables
are legible.  Please also ensure that you refer to your figures in the main
text.  Many reviewers print the papers in gray-scale. Therefore, if you use
colors for your figures, ensure that the different colors are highly
distinguishable in gray-scale.

\paragraph{References.}  There is no length limit for references.
\textbf{Each reference must explicitly list all authors of the paper.}
  Papers not meeting this requirement will be rejected. Authors of
NSF proposals should be familiar with this requirement. Knowing all
authors of related work will help find the best reviewers. Since there
is no length limit for the number of pages used for references, there
is no need to save space here.

\section{Paper Submission Instructions}

\subsection{Declaring Authors}

Enter all authors of the paper into the online paper submission tool
upfront. Addition/removal of authors once the paper is accepted will
have to be approved by the \href{mailto:emery@cs.umass.edu?subject=[PLDI'16]}{program chair}, since it potentially
undermines the goal of eliminating conflicts for reviewer assignment.

\subsection{Supplementary Material}

The paper submission website will allow authors to upload an
additional document containing material that supplements the paper
(such as an extended proof or extensive results).  However, it is
essential that authors understand that: a) reviewers are \textbf{not
  obliged} to read the supplement, and b) the supplement must
  be \textbf{fully anonymized.}

\subsection{Areas and Topics}

PLDI is a broad conference.  If you are unsure whether your work falls
within scope for PLDI, please check the
\href{http://conf.researchr.org/track/pldi2016/pldi2016-papers#Call-for-Papers}{call
  for papers} and if still in doubt, check with the
\href{mailto:emery@cs.umass.edu?subject=[PLDI'16]}{program
  chair}.

\subsection{Concurrent Submissions and Workshops}

By submitting a manuscript to PLDI'16, authors guarantee that they
are adhering to the
\href{http://www.sigplan.org/Resources/Policies/Republication/}{SIGPLAN
  Republication Policy}. Please ensure that you are familiar with it.
Violation of any of these conditions will lead to rejection.
As always, if you are in doubt, it is best to contact the
\href{mailto:emery@cs.umass.edu?subject=[PLDI'16]}{program
  chair}.

Finally, we also note that the
\href{http://www.acm.org/publications/policies/plagiarism_policy}{ACM
  Plagiarism Policy} covers a range of ethical issues concerning the
misrepresentation of other works or one's own work.

\section{Early Access in the Digital Library}

The PLDI'16 proceedings may be available on the ACM Digital Library as
early as May 30, 2016. \textbf{Authors must consider any implications
  of this early disclosure of their work \emph{before} submitting
  their papers.}

\section{Acknowledgements}

This document is based primarily on the document created by Steve
Blackburn (PLDI'15), which in turn relies heavily on ones prepared for
previous conferences and we thank their program chairs; in particular,
Sandhya Dwarkadas (ASPLOS'15), Sarita Adve (ASPLOS'14), Steve Keckler
(ISCA'14), Christos Kozyrakis (Micro'13), Margaret Martonosi
(ISCA'13), Onur Mutlu (Micro'12), and Michael L. Scott (ASPLOS'12).

\bibliographystyle{abbrvnat}

% We recommend that you use BibTeX.  The inlined bibitems below are
% used to keep this template to a single file.
\begin{thebibliography}{}
\softraggedright

\bibitem[Backus et~al.(1960)]{Backus:60}
J.~W. Backus, F.~L. Bauer, J.~Green, C.~Katz, J.~McCarthy, A.~J. Perlis,
  H.~Rutishauser, K.~Samelson, B.~Vauquois, J.~H. Wegstein, A.~van Wijngaarden,
  and M.~Woodger.
\newblock Report on the algorithmic language ALGOL 60.
\newblock \emph{Commun. ACM}, 3\penalty0 (5):\penalty0 299--314, May 1960.
\newblock ISSN 0001-0782.
\newblock \doi{10.1145/367236.367262}.

\bibitem[Collins(1960)]{Collins:60}
G.~E. Collins.
\newblock A method for overlapping and erasure of lists.
\newblock \emph{Commun. ACM}, 3\penalty0 (12):\penalty0 655--657, December
  1960.
\newblock \doi{10.1145/367487.367501}

\bibitem[Lamport(1994)]{Lamport:94}
L.~Lamport.
\newblock \emph{{\LaTeX: A Document Preparation System}}.
\newblock Addison-Wesley, Reading, Massachusetts, 2nd edition, 1994.

\bibitem[McCarthy(1960)]{McCarthy:60}
J.~McCarthy.
\newblock Recursive functions of symbolic expressions and their computation by
  machine, part {I}.
\newblock \emph{Commun. ACM}, 3\penalty0 (4):\penalty0 184--195, April 1960.
\newblock \doi{10.1145/367177.367199}

\end{thebibliography}

\end{document}

### 2015-11-14 paper.tex
redis & hedis

@@
 or packages that talk with Redis' TCP protocol. In Haskell, the most popular
-package people are using is \emph{Hedis}.
+package is \emph{Hedis}.
 
-The previous example in Redis's client would look something like this with Hedis
+The previous example in Redis's client would look something like this with \emph{Hedis}
 in Haskell.
@@
 
-Function \mintinline{haskell}{sadd} takes a key and a list of values and returns an
-\mintinline{haskell}{Integer},
-wrapped in the context \mintinline{haskell}{Redis (Either Reply Integer)}
-to which encapsulates possible failures.
-
- to indicate it's size. \mintinline{haskell}{sinter}
-takes a list of keys and returns
-
- , both of type
-  \mintinline{haskell}{ByteString}, and returns a \mintinline{haskell}{Integer}
-  to indicate it's size.
-
-  since keys and values are all just binary strings in Redis.
-
- and \mintinline{haskell}{sinter} now
+Function \mintinline{haskell}{sadd} takes a key and a list of values as arguments
+, and returns an \mintinline{haskell}{Integer}, wrapped in
+\mintinline{haskell}{Either} to indicate possible failures in
+the context \mintinline{haskell}{Redis} of command execution.
 
@@
     -> Redis (Either Reply Integer)
-sinter :: [ByteString]  -- keys
-    -> Redis (Either Reply [ByteString])
 \end{minted}
 
+Keys and values are of type \mintinline{haskell}{ByteString}, since they are all
+just binary strings in Redis. That means if a user wants to store values of
+arbitrary types, he or she will have to encode and decode them as
+ \mintinline{haskell}{ByteString}s.
+
 \end{document}

### 2015-11-15 paper.tex
introduction

@@
 \begin{abstract}
-We often interacts with database systems through dedicated libraries or packages
-implemented in some programming languages. People make mistakes, while some of
-the syntactic and semantic errors can be prevented by the language at compile
-time, or caught by the package at runtime, most semantic errors are just
-being ignored, causing problems at the database system.
+Programmers often interact with database systems, by sending queries through
+libraries or packages in some programming languages. But people make mistakes,
+while some of the syntactic and semantic errors can be prevented by the language
+at compile time, or caught by the package at runtime, most semantic errors are
+just being ignored, causing problems at the database system.
 
 In this paper, we demonstrates how to prevent those runtime errors at compile
-time, and allow users to write reliable database queries without runtime overhead,
-by exploiting type-level programming techniques such as parametedzied monad,
-type-level literals and type families in Haskell.
+time, allowing users to write reliable database queries without runtime
+overhead, by exploiting type-level programming techniques such as parametedzied
+monad, type-level literals and type families in Haskell.
 
@@
 For example, if we want to add a bunch of strings to 2 sets, and intersect them,
-we could input the following sequence of commands into Redis's client interactively.
+we could input the following sequence of commands into Redis's client
+interactively.
 
@@
 
-The previous example in Redis's client would look something like this with \emph{Hedis}
-in Haskell.
+The previous example in Redis's client would look something like this with
+\emph{Hedis} in Haskell.
 \begin{minted}{haskell}
@@
     sadd "some-set" ["a", "b"]
-    sadd "another-set" ["a", "b", "C"]
+    sadd "another-set" ["a", "b", "c"]
     sinter ["some-set", "another-set"]
@@
 
-Function \mintinline{haskell}{sadd} takes a key and a list of values as arguments
-, and returns an \mintinline{haskell}{Integer}, wrapped in
-\mintinline{haskell}{Either} to indicate possible failures in
-the context \mintinline{haskell}{Redis} of command execution.
+Function \mintinline{haskell}{sadd} takes a key and a list of values as
+arguments, and returns an \mintinline{haskell}{Integer}, wrapped in
+\mintinline{haskell}{Either} to indicate possible failures in the context
+\mintinline{haskell}{Redis} of command execution.
 
@@
 Keys and values are of type \mintinline{haskell}{ByteString}, since they are all
-just binary strings in Redis. That means if a user wants to store values of
+just binary strings in Redis. If a user wants to store values of
 arbitrary types, he or she will have to encode and decode them as
@@
 
+\subsection{Semantics errors}
+
+Although everything in Redis is binary string, they are treated diffenrently.
+Redis supports many different kind of data structures, such as strings, hashes,
+ lists, etc. These values are of different data types, and some commands only
+ work on certain type of values.
+
+\paragraph{Problem 1} Consider the following example. \mintinline{text}{SET} is a command that only
+works on strings. The key \mintinline{text}{some-string} is now related with a
+string. If we treat it as a set and add an element to it with
+\mintinline{text}{SADD}, a runtime error arises.
+
+\begin{minted}{text}
+redis> SET some-string foo
+OK
+redis> SADD some-string bar
+(error) WRONGTYPE Operation against a key
+ holding the wrong kind of value
+\end{minted}
+
+\paragraph{Problem 2} Even worse, not all strings are equal! \mintinline{text}{INCR} parses the
+string value as an integer, increments it by one, and store it back as a string.
+If a string value can't be parse as an integer, another runtime error arises
+
+\begin{minted}{text}
+redis> SET some-string foo
+OK
+redis> INCR some-string
+(error) ERR value is not an integer or out
+ of range
+\end{minted}
+
+\paragraph{In Hedis} The same goes for Hedis, since it's only a simple wrapping on top of Redis's
+protocol written in Haskell. No semantic checking is done, neither
+dynamically nor statically.
+
+\begin{minted}{haskell}
+program :: Redis (Either Reply Integer)
+program = do
+    set "some-string" "foo"
+    sadd "some-string" ["a"]
+\end{minted}
+
+Which yields the same error as in Redis's client.
+
+\begin{minted}{haskell}
+Left (Error "WRONGTYPE Operation against a
+ key holding the wrong kind of value")
+\end{minted}
+
+\subsection{Hedis as an embedded DSL}
+
+Haskell makes it easy to build and use Domain Specific Languages (DSLs).
+Hedis can be regardred as one DSL that embeds in Haskell, that has variable
+bindings (between keys and values), but with very little or none semantic
+checkings.
+
+We started out from making Hedis a dynamically checked embedded DSL, with a
+runtime type checker that keeps track of everything. But then we found that,
+things can be a lot easier, by laveraging the host language's type checker,
+and make it statically checked!
+
+\subsection{Contributions}
+
+To summerize our contributions:
+
+\begin{itemize}[noitemsep]
+\item We model variable bindings of an embedded DSL with language extensions
+like type-level literals and data kinds.
+\item We provide (yet another) an example of how to encode effects and
+constraints of an action in types, with indexed monad and other language
+ extensions such as closed type-families and constraints kinds.
+\item Edis, a package we built for programmers. This package helps programmers
+to write more reliable Redis codes, and makes Redis polymorphic by converting
+back and forth from values of any types and boring ByteStrings.
+\end{itemize}
+
+\section{Implementation}
+\subsection{Model bindings with Type-level Dictionaries}
+\paragraph{Type-level literals}
+\subsection{Bookkeeping with Type families}
+\subsection{Capture effects with Parameterised Monad}
+\paragraph{Type-level literals}
+
+%
+% Values have types, but keys don't. In other words, if we treat keys like
+% variables, then Redis behaves likes a dynamically and strongly typed language.
+
+% We can only reuse this key as a \emph{set} after it is removed with \mintinline{text}{DEL}
+% explicitly.
+%
+% \begin{minted}{text}
+% redis> DEL some-string
+% (integer) 1
+% redis> SADD some-string bar
+% (integer) 1
+% \end{minted}
+
 \end{document}

### 2015-11-17 paper.tex
Section 3: Indexed Monad

@@
 %
-% LaTeX template for prepartion of submissions to PLDI'16
+% LaTeX template for preparation of submissions to PLDI'16
 %
@@
 
-% \usepackage[utf8]{inputenc}
-% \usepackage[english]{babel}
-\usepackage{minted}     % code highlighting
+% code highlighting
+\usepackage{minted}
 \usemintedstyle{tango}
+
+% no red boxes on parser error:
+\AtBeginEnvironment{minted}{%
+  \renewcommand{\fcolorbox}[4][]{#4}}
+
+% quotes
+\usepackage{dirtytalk}
+
+% for symbols
+\usepackage[T1]{fontenc}
+\usepackage{lmodern}
+\usepackage{upquote}
+
+% use arbitrary fonts
+\usepackage{fontspec}
+\setmonofont{Inconsolata}
 
@@
 
-\title{Catching runtime errors of Redis at compile time}
+\title{Preventing runtime errors of Redis at compile time}
 
@@
 In this paper, we demonstrates how to prevent those runtime errors at compile
-time, allowing users to write reliable database queries without runtime
-overhead, by exploiting type-level programming techniques such as parametedzied
-monad, type-level literals and type families in Haskell.
+time, allowing users to write more reliable database queries without runtime
+overhead, by exploiting type-level programming techniques such as indexed
+monad, type-level literals and closed type families in Haskell.
 
@@
 
-\begin{minted}{text}
+\begin{minted}[xleftmargin=1em]{text}
 redis> SADD some-set a b c
@@
 or packages that talk with Redis' TCP protocol. In Haskell, the most popular
-package is \emph{Hedis}.
+package is Hedis.
 
 The previous example in Redis's client would look something like this with
-\emph{Hedis} in Haskell.
-\begin{minted}{haskell}
+Hedis in Haskell.
+\begin{minted}[xleftmargin=1em]{haskell}
 program :: Redis (Either Reply [ByteString])
@@
 \mintinline{haskell}{Either} to indicate possible failures in the context
-\mintinline{haskell}{Redis} of command execution.
+\mintinline{haskell}{Redis}\footnotemark of command execution.
 
-\begin{minted}{haskell}
+\begin{minted}[xleftmargin=1em]{haskell}
 sadd :: ByteString      -- key
@@
 
-\subsection{Semantics errors}
+\footnotetext{Hedis provides another kind of context, \mintinline{haskell}{RedisTx}, for \emph{transactions}, united with \mintinline{haskell}{Redis} under the class of \mintinline{haskell}{RedisCtx}.
+We use \mintinline{haskell}{Redis} only here for brevity.
+}
 
-Although everything in Redis is binary string, they are treated diffenrently.
-Redis supports many different kind of data structures, such as strings, hashes,
- lists, etc. These values are of different data types, and some commands only
- work on certain type of values.
+\subsection{Problems}
 
-\paragraph{Problem 1} Consider the following example. \mintinline{text}{SET} is a command that only
-works on strings. The key \mintinline{text}{some-string} is now related with a
-string. If we treat it as a set and add an element to it with
+\say{All binary strings are equal, but some binary strings are more equal than others.}
+
+Although everything in Redis is essentially a binary string, they are treated
+differently. Redis supports many different kind of data structures, such as
+strings, hashes, lists, etc. These values are of different data types, and some
+commands only work on certain types of them, much like how language C treats
+a piece of data.
+
+\paragraph{Problem 1} Consider the following example. \mintinline{text}{SET} is
+a command that only works on strings. The key \mintinline{text}{some-string} is
+now related with a string. If we treat it as a set and add an element to it with
 \mintinline{text}{SADD}, a runtime error arises.
 
-\begin{minted}{text}
+\begin{minted}[xleftmargin=1em]{text}
 redis> SET some-string foo
@@
 
-\paragraph{Problem 2} Even worse, not all strings are equal! \mintinline{text}{INCR} parses the
-string value as an integer, increments it by one, and store it back as a string.
-If a string value can't be parse as an integer, another runtime error arises
+\paragraph{Problem 2} Even worse, not all strings are equal!
+\mintinline{text}{INCR} parses the string value as an integer, increments it by
+one, and store it back as a string. If a string value can't be parse as an
+integer, another runtime error arises
 
-\begin{minted}{text}
+\begin{minted}[xleftmargin=1em]{text}
 redis> SET some-string foo
@@
 
-\paragraph{In Hedis} The same goes for Hedis, since it's only a simple wrapping on top of Redis's
-protocol written in Haskell. No semantic checking is done, neither
-dynamically nor statically.
+\paragraph{In Hedis} The same goes for Hedis, since it's only a simple wrapping
+ on top of Redis's protocol written in Haskell.
 
-\begin{minted}{haskell}
+\begin{minted}[xleftmargin=1em]{haskell}
 program :: Redis (Either Reply Integer)
@@
 
-Which yields the same error as in Redis's client.
+It yields the same error as in Redis's client.
 
-\begin{minted}{haskell}
+\begin{minted}[xleftmargin=1em]{haskell}
 Left (Error "WRONGTYPE Operation against a
@@
 
+\paragraph{The Cause} These problems arise from the absence of type checking,
+with respects to \textbf{the type of a value that associates with a key}.
+
 \subsection{Hedis as an embedded DSL}
 
-Haskell makes it easy to build and use Domain Specific Languages (DSLs).
-Hedis can be regardred as one DSL that embeds in Haskell, that has variable
-bindings (between keys and values), but with very little or none semantic
-checkings.
+Haskell makes it easy to build and use Domain Specific Languages (DSLs),
+and Hedis can be regarded as one of them. What makes Hedis special is that,
+it has variable bindings (between keys and values), but with very
+little or no semantic checking, neither dynamically nor statically.
 
-We started out from making Hedis a dynamically checked embedded DSL, with a
-runtime type checker that keeps track of everything. But then we found that,
-things can be a lot easier, by laveraging the host language's type checker,
-and make it statically checked!
+We began with making Hedis a dynamically checked embedded DSL, and implemented a
+runtime type checker that keeps track of everything. But then we found that
+things can be a lot easier, by leveraging the host language's type checker,
+which also make it statically type-checked!
 
@@
 
-To summerize our contributions:
+To summarize our contributions:
 
 \begin{itemize}[noitemsep]
-\item We model variable bindings of an embedded DSL with language extensions
-like type-level literals and data kinds.
-\item We provide (yet another) an example of how to encode effects and
-constraints of an action in types, with indexed monad and other language
- extensions such as closed type-families and constraints kinds.
+\item We make Hedis statically type-checked, without runtime overhead.
+\item We demonstrates how to model variable bindings of an embedded DSL with
+language extensions like type-level literals and data kinds.
+\item We provide (yet another) an example of encoding effects and
+ constraints of an action in types, with indexed monad\cite{indexedmonad}
+  and other language extensions such as closed
+  type-families\cite{closedtypefamilies} and constraints
+  kinds\cite{constraintskinds}.
 \item Edis, a package we built for programmers. This package helps programmers
-to write more reliable Redis codes, and makes Redis polymorphic by converting
-back and forth from values of any types and boring ByteStrings.
+ to write more reliable Redis codes, and also makes Redis polymorphic by
+ automatically converting back and forth from values of any types and boring
+ ByteStrings.
 \end{itemize}
 
-\section{Implementation}
-\subsection{Model bindings with Type-level Dictionaries}
-\paragraph{Type-level literals}
-\subsection{Bookkeeping with Type families}
-\subsection{Capture effects with Parameterised Monad}
-\paragraph{Type-level literals}
+\section{Type-level Dictionary}
+
+To check the bindings between keys and values, we need a \emph{dictionary-like}
+structure, and encode it as a \emph{type} somehow.
+
+\subsection{Datatype promotion}
+Normally, at the term level, we could express data type of the dictionary with
+\emph{type synonym} like this.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+type Key = String
+type Dictionary = [(Key, TypeRep)]
+\end{minted}
+
+To encode this in the type level, everything has to be
+\emph{promoted}\cite{promotion} one level up.
+From terms to types, and from types to kinds.
+
+Luckily, with recently added GHC extension \emph{data kinds}. With data kinds
+we can define our own kinds just like we can with datatypes, and every suitable
+datatype will be promoted to be a kind, and its value constructors promoted
+to be type constructors, automatically.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+data List a = Nil | Cons a (List a)
+\end{minted}
+
+Give rise to the following kinds and type constructors:\footnotemark\footnotemark
+
+\footnotetext{To distinguish between types and promoted constructors that have
+ ambiguous names, prefix promoted constructor with a single quote like
+ \mintinline{text}{'Nil} and \mintinline{text}{'Cons}}
+\footnotetext{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}
+
+\begin{minted}[xleftmargin=1em]{haskell}
+List k :: BOX
+Nil  :: List k
+Cons :: k -> List k -> List k
+\end{minted}
+
+Haskell sugars lists \mintinline{haskell}{[1, 2, 3]} and
+tuples \mintinline{text}{(1, 'a')} with brackets and parentheses.
+Promoted lists and tuples are also sugared with single quote prefixed. For
+example \mintinline{text}{'}\mintinline{haskell}{[Int, Char]},
+\mintinline{text}{'}\mintinline{haskell}{(Int, Char)}.
+
+\subsection{Type-level literals}
+
+Now we have type-level lists and tuples to construct the dictionary.
+For keys, \mintinline{haskell}{String} also has a type-level correspondence:
+\mintinline{haskell}{Symbol}.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+data Symbol
+\end{minted}
+
+Symbol is defined without a value constructor, because it's
+intended to be used as a promoted kind.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+"this is a type-level string literal" :: Symbol
+\end{minted}
+
+Nonetheless, it's still useful to have a term-level value that links with a
+Symbol, when we want to retreive type-level informations at runtime (but not the
+ other way around!).
+
+\subsection{Putting everything together}
+
+With all of these ingrediants ready, let's build some dictionaries!
+
+\begin{minted}[xleftmargin=1em]{haskell}
+type DictEmpty = '[]
+type Dict0 = '[ '("key", Bool) ]
+type Dict1 = '[ '("A", Int), '("B", "A") ]
+\end{minted}
+
+They are defined with \emph{type synonym}, since they are \emph{types}
+, not \emph{terms}. If we ask \mintinline{text}{GHCi} what is the kind of
+\mintinline{haskell}{Dict1}, we get \mintinline{haskell}{Dict1 :: [ (Symbol, *) ]}
+
+The star \mintinline{haskell}{*} is a kind, it stands for any kind of types,
+such as \mintinline{haskell}{Int}, \mintinline{haskell}{Char} or even a symbol
+\mintinline{haskell}{"symbol"}, while \mintinline{haskell}{Symbol} stands only
+for all symbols.
+
+\section{Indexed Monad}
+
+Redis commands are \emph{actions}.
+We could capture the effects caused by an action, by expressing the states it
+affects, before and after. That is, the \emph{preconditions} and
+\emph{postconditions} of an action. In such way, we could also impose
+constraints on the preconditions.
+
+\emph{Indexed monads} (or \emph{monadish},
+\emph{parameterised monad})\cite{indexedmonad}
+can be used\cite{typefun}\cite{staticresources} to model such preconditions and
+postconditions in types. An indexed monad is a type constructor that takes three
+arguments: an intitial state, a final state, and the type of a value that an
+action computes, which can be read like a Hoare triple\cite{kleisli}.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+class IMonad m where
+    unit :: a -> m p p a
+    bind :: m p q a -> (a -> m q r b) -> m p r b
+\end{minted}
+
+Class \mintinline{haskell}{IMonad} comes with two operations:
+\mintinline{haskell}{unit} for identities and \mintinline{haskell}{bind} for
+compositions, as in Monad.
+
+We define a new datatype \mintinline{haskell}{Edis}, which is basically just
+\mintinline{haskell}{Redis} indexed with more informations in types. And we make
+ it an instance of \mintinline{haskell}{IMonad}.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+newtype Edis p q a = Edis { unEdis :: Redis a }
+
+instance IMonad Edis where
+    unit = Edis . return
+    bind m f = Edis (unEdis m >>= unEdis . f )
+\end{minted}
+
+The first and second argument of type \mintinline{haskell}{Edis} is where the
+dictionaries going to stay.
+
+\subsection{The first attempt}
+
+In Redis, \mintinline{text}{PING} does nothing but replies with
+\mintinline{text}{PONG} if the connection is alive. In Hedis,
+\mintinline{text}{PING} has type:
+
+\begin{minted}[xleftmargin=1em]{haskell}
+ping :: Redis (Either Reply Status)
+\end{minted}
+
+Now we have \mintinline{haskell}{Edis}, let's make our own version of
+\mintinline{haskell}{ping}\footnotemark
+
+\begin{minted}[xleftmargin=1em]{haskell}
+ping :: Edis xs xs (Either Reply Status)
+ping = Edis Hedis.ping
+\end{minted}
+
+Dicionary \mintinline{haskell}{xs} in the type remain unaffected after the
+action, because \mintinline{haskell}{ping} does not affect any key-value
+bindings. To encode other commands that alters key-value bindings, we need
+ type-level functions to annotate those effects on the dictionary.
+
+\section{Bookkeeping with type families}
+\subsection{Closed type families}
+
+Type families have a wide variaty of applications. They can appear inside type
+classes\cite{tfclass}\cite{tfsynonym}, or at toplevel. Toplevel type families
+can be used to compute over types, they come in open\cite{tfopen} or closed
+forms\cite{tfclosed}. We are using \emph{closed type families} since we need
+none of the extensibility provided by open type families, and we may need to
+overlap instances, like what we do in defining function cases.
+
+%  are
+% permitted as well.
+%
+%
+%  open type synonym families, and
+% closed type synonym families
+% One of which is to compute types, we are using closed type families, since all
+% we want is to use it as a function on the type level, no extensibility needed,
+% everything is defined in one-piece
+
+\footnotetext{\mintinline{haskell}{ping} from Hedis is qualified with
+\mintinline{haskell}{Hedis} to prevent function name clashing}
+%
+% \mintinline{haskell}{IMonad} \mintinline{haskell}{IMonad} takes three arguments: an intitial state \mintinline{haskell}{p}, a final
+%  state, and computes a value of type a
+
+\subsection{Proxies and Kind polymorphism}
+
+Since Haskell has a phase distinction\cite{phasedistinction} between
+
+To link a symbol with a term-level value
+
+What we need is a \emph{singleton}\cite{singletons}, a type
+that has only one instance at the term level.
+
+With \emph{kind polymorphism}\cite{promotion}, we have parametric polymorphism
+at the kind level.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+Proxy :: Proxy "a"
+\end{minted}
+
+\subsection{Closed type families}
+
+\subsection{Conclusions and Related Work}
+
+% references
+\bibliographystyle{abbrvnat}
+\bibliography{cites}
+
+% \subsection{Model bindings with Type-level Dictionaries}
+% \subsection{Bookkeeping with Type families}
+% \subsection{Capture effects with Index Monad}
+% \paragraph{Type-level literals}
 

### 2015-11-17 paper.tex
4.1: Closed type families

@@
 classes\cite{tfclass}\cite{tfsynonym}, or at toplevel. Toplevel type families
-can be used to compute over types, they come in open\cite{tfopen} or closed
-forms\cite{tfclosed}. We are using \emph{closed type families} since we need
-none of the extensibility provided by open type families, and we may need to
-overlap instances, like what we do in defining function cases.
+can be used to compute over types, they come in two forms: open\cite{tfopen} and
+closed \cite{tfclosed}.
+
+We choose \emph{closed type families}, because it allows overlapping instances,
+and we need none of the extensibility provided by open type families.
+For example, consider both term-level and type-level \mintinline{haskell}{&&}:
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+(&&) :: Bool -> Bool -> Bool
+True && True = True
+a    && b    = False
+
+type family And (a :: Bool) (b :: Bool) :: Bool where
+    And True True = True
+    And a    b    = False
+\end{minted}
+
+The first instance of \mintinline{haskell}{And} could be subsumed
+under the more general instance \mintinline{haskell}{And a b}.
+But the closedness allows these instances to be resoluted in order, just like
+how cases are resoluted in term-level functions. Also notice that how much
+\mintinline{haskell}{And} resembles to it's term-level brother.
 
@@
 
-Since Haskell has a phase distinction\cite{phasedistinction} between
+% Since Haskell has a phase distinction\cite{phasedistinction} between
+%
+% To link a symbol with a term-level value
+%
+% What we need is a \emph{singleton}\cite{singletons}, a type
+% that has only one instance at the term level.
 
-To link a symbol with a term-level value
-
-What we need is a \emph{singleton}\cite{singletons}, a type
-that has only one instance at the term level.
-
-With \emph{kind polymorphism}\cite{promotion}, we have parametric polymorphism
-at the kind level.
+% With \emph{kind polymorphism}\cite{promotion}, we have parametric polymorphism
+% at the kind level.
 
@@
 
-% \subsection{Model bindings with Type-level Dictionaries}
-% \subsection{Bookkeeping with Type families}
-% \subsection{Capture effects with Index Monad}
-% \paragraph{Type-level literals}
-
-%
-% Values have types, but keys don't. In other words, if we treat keys like
-% variables, then Redis behaves likes a dynamically and strongly typed language.
-
-% We can only reuse this key as a \emph{set} after it is removed with \mintinline{text}{DEL}
-% explicitly.
-%
-% \begin{minted}{text}
-% redis> DEL some-string
-% (integer) 1
-% redis> SADD some-string bar
-% (integer) 1
-% \end{minted}
-
 \end{document}

### 2015-11-17 paper.tex
4.2 Functions on type-level dictionaries

@@
 Normally, at the term level, we could express data type of the dictionary with
-\emph{type synonym} like this.
+\emph{type synonym} like this.\footnotemark
 
@@
 \end{minted}
+
+\footnotetext{\mintinline{haskell}{TypeRep} here represents term-level
+representations of datatypes, available in \mintinline{haskell}{Data.Typeable}}
 
@@
 
+\footnotetext{\mintinline{haskell}{ping} from Hedis is qualified with
+\mintinline{haskell}{Hedis} to prevent function name clashing}
+
 \begin{minted}[xleftmargin=1em]{haskell}
@@
 how cases are resoluted in term-level functions. Also notice that how much
-\mintinline{haskell}{And} resembles to it's term-level brother.
+\mintinline{haskell}{And} resembles to it's term-level sibling.
+
+\subsection{Functions on type-level dictionaries}
+
+With closed type families, we could define functions on the type level.
+Let's begin with dictionary lookup.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+type family Get
+    (xs :: [(Symbol, *)])   -- dictionary
+    (s :: Symbol)           -- key
+    :: * where              -- type
+
+    Get (’(s, x) ’: xs) s = x
+    Get (’(t, x) ’: xs) s = Get xs s
+\end{minted}
+
+Another benifit of closed type families is that type-level equality can be
+expressed by unifying type variables with the same name.
+\mintinline{haskell}{Get} takes two type arguments, a dictionary and a symbol.
+If the key we are looking for unifies with the symbol of an entry, then
+ \mintinline{haskell}{Get} returns the corresponding type, else it keeps
+ searching down the rest of the dictionary.
+
+\mintinline{haskell}{Get }\mintinline{text}{ '}\mintinline{haskell}{[ }
+\mintinline{text}{'}\mintinline{haskell}{("A", Int) ] "A"} evaluates to
+\mintinline{haskell}{Int}.
+
+But \mintinline{haskell}{Get }
+\mintinline{text}{'}\mintinline{haskell}{[ }
+\mintinline{text}{'}\mintinline{haskell}{("A", Int) ] "B"} would get stuck.
+That's because \mintinline{haskell}{Get} is a partial function on types,
+ and these types are computed at compile-time. It wouldn't make
+much sense for type checkers to crash and throw a "Non-exhaustive" error or
+be non-terminating.
+
+We could make \mintinline{haskell}{Get} total, as we would at the term level,
+with \mintinline{haskell}{Maybe}.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+type family Get
+    (xs :: [(Symbol, *)])   -- dictionary
+    (s :: Symbol)           -- key
+    :: Maybe * where        -- type
+
+    Get ’[]             s = Nothing
+    Get (’(s, x) ’: xs) s = Just x
+    Get (’(t, x) ’: xs) s = Get xs s
+\end{minted}
+
+Other dictionary-related functions are defined in a similar way.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+-- inserts or updates an entry
+type family Set
+    (xs :: [(Symbol, *)])   -- old dictionary
+    (s :: Symbol)           -- key
+    (x :: *)                -- type
+    :: [(Symbol, *)] where  -- new dictionary
+
+    Set ’[]             s x = ’[ ’(s, x) ]
+    Set (’(s, y) ’: xs) s x = (’(s, x) ’: xs)
+    Set (’(t, y) ’: xs) s x =
+        ’(t, y) ’: (Set xs s x)
+
+-- removes an entry
+type family Del
+    (xs :: [(Symbol, *)])   -- old dictionary
+    (s :: Symbol)           -- key
+    :: [(Symbol, *)] where  -- new dictionary
+
+    Del ’[] s             = ’[]
+    Del (’(s, y) ’: xs) s = xs
+    Del (’(t, y) ’: xs) s = ’(t, y) ’: (Del xs s)
+
+-- membership
+type family Member
+    (xs :: [(Symbol, *)])   -- dictionary
+    (s :: Symbol)           -- key
+    :: Bool where           -- exists?
+
+    Member ’[]             s = False
+    Member (’(s, x) ’: xs) s = True
+    Member (’(t, x) ’: xs) s = Member xs s
+\end{minted}
+
+\begin{minted}[xleftmargin=1em]{haskell}
+type family Get
+    (xs :: [(Symbol, *)])   -- dictionary
+    (s :: Symbol)           -- key
+    :: Maybe * where        -- type
+
+    Get ’[]             s = Nothing
+    Get (’(s, x) ’: xs) s = Just x
+    Get (’(t, x) ’: xs) s = Get xs s
+\end{minted}
+
+% \begin{minted}[xleftmargin=1em]{haskell}
+% Get '['("ha", Int)]
+% \end{minted}
 
@@
 
-\footnotetext{\mintinline{haskell}{ping} from Hedis is qualified with
-\mintinline{haskell}{Hedis} to prevent function name clashing}
 %

### 2015-11-17 paper.tex
5. Making Redis polymorphic

@@
 differently. Redis supports many different kind of data structures, such as
-strings, hashes, lists, etc. These values are of different data types, and some
-commands only work on certain types of them, much like how language C treats
+strings, hashes, lists, etc. These values have different datatypes, and most
+commands only work with certain types of them, much like how C language treats
 a piece of data.
@@
 
-\section{Type-level Dictionary}
+\section{Type-level dictionaries}
 
@@
 \subsection{Datatype promotion}
-Normally, at the term level, we could express data type of the dictionary with
+Normally, at the term level, we could express the datatype of dictionary with
 \emph{type synonym} like this.\footnotemark
@@
 
-\section{Indexed Monad}
+\section{Indexed monads}
 
@@
 
-\subsection{The first attempt}
+\subsection{\mintinline{text}{PING}: the first attempt}
 
@@
 
-\section{Bookkeeping with type families}
+\section{Type-level functions}
 \subsection{Closed type families}
@@
 
+\subsection{Proxies and singleton types}
+
+With functions on dictionaries, now we could annotate the effects of a command
+in types. \mintinline{text}{DEL} removes a key from the current database,
+regardless of its type.
+
 \begin{minted}[xleftmargin=1em]{haskell}
-type family Get
-    (xs :: [(Symbol, *)])   -- dictionary
-    (s :: Symbol)           -- key
-    :: Maybe * where        -- type
-
-    Get ’[]             s = Nothing
-    Get (’(s, x) ’: xs) s = Just x
-    Get (’(t, x) ’: xs) s = Get xs s
+del :: KnownSymbol s
+    => Proxy s
+    -> Edis xs (Del xs s) (Either Reply Integer)
+del key = Edis $ Hedis.del (symbolVal key)
 \end{minted}
 
-% \begin{minted}[xleftmargin=1em]{haskell}
-% Get '['("ha", Int)]
-% \end{minted}
+\mintinline{haskell}{KnownSymbol} is a class that gives the string associated
+with a concrete type-level symbol, which can be retreived with
+\mintinline{haskell}{symbolVal}.\footnotemark
 
-%  are
-% permitted as well.
-%
-%
-%  open type synonym families, and
-% closed type synonym families
-% One of which is to compute types, we are using closed type families, since all
-% we want is to use it as a function on the type level, no extensibility needed,
-% everything is defined in one-piece
+\footnotetext{They are defined in \mintinline{haskell}{GHC.TypeLits}.}
 
-%
-% \mintinline{haskell}{IMonad} \mintinline{haskell}{IMonad} takes three arguments: an intitial state \mintinline{haskell}{p}, a final
-%  state, and computes a value of type a
+Since Haskell has a \emph{phase distinction}\cite{phasedistinction}, types are
+erased before runtime. It's impossible to obtain informations directly from
+types, we can only do this indirectly, with \emph{singleton types}
+\cite{singletons}.
 
-\subsection{Proxies and Kind polymorphism}
+A singleton type is a type that has only one instance, and the instance can be
+think of as the representative of the type at the realm of runtime values.
 
-% Since Haskell has a phase distinction\cite{phasedistinction} between
-%
-% To link a symbol with a term-level value
-%
-% What we need is a \emph{singleton}\cite{singletons}, a type
-% that has only one instance at the term level.
-
-% With \emph{kind polymorphism}\cite{promotion}, we have parametric polymorphism
-% at the kind level.
+\mintinline{haskell}{Proxy}, as its name would suggest, can be used as
+singletons. It's a phantom type that could be indexed with any type.
 
 \begin{minted}[xleftmargin=1em]{haskell}
-Proxy :: Proxy "a"
+data Proxy t = Proxy
 \end{minted}
 
-\subsection{Closed type families}
+In the type of \mintinline{haskell}{del}, the type varabile
+\mintinline{haskell}{s} would be a \mintinline{haskell}{Symbol} that is decided
+ by the argument of type \mintinline{haskell}{Proxy s}.
+To use \mintinline{haskell}{del}, we would have to apply it with a clunky
+ term-level proxy like this:
 
-\subsection{Conclusions and Related Work}
+\begin{minted}[xleftmargin=1em]{haskell}
+del (Proxy :: Proxy "A")
+\end{minted}
+
+\section{Making Redis polymorphic}
+
+So far, the datatypes we are dealing with, are all of those built-in Redis data
+structures, such as \emph{strings}, \emph{lists (of strings)},
+\emph{sets (of strings)}, etc.
+But in the real world, these raw strings are hardly useful, people would
+usually serialize their data into strings before storing them, and deserialize
+them back when in need.
+
+Instead of letting users writing these boilerplates, we will be doing these
+serializations/deserializations for them. With the help from
+\mintinline{text}{cereal}, a binary serialization library.
+
+\mintinline{text}{cereal} comes with these two functions:
+\begin{minted}[xleftmargin=1em]{haskell}
+encode :: Serialize a
+       => a -> ByteString Source
+
+decode :: Serialize a
+       => ByteString -> Either String a
+\end{minted}
+
+Which would do all the works for us, as long as the datatype it's handling is
+an instance of class \mintinline{haskell}{Serialize}.\footnotemark
+
+\footnotetext{The methods of \mintinline{haskell}{Serialize} will have default
+generic implementations for all datatypes with some language extenstions
+enabled}
+
+\section{Imposing constraints}
+
+To rule out programs with undesired properties, certain constraints must be
+imposed, on what arguments they can take, and what preconditions they must hold.
+These constraints are best encoded in types.
+
+\subsection{Redis datatypes}
+
+Redis supports many different datatypes, and most commands only work with
+certain types of them. For example: \mintinline{haskell}{SET} with strings or
+\mintinline{haskell}{SADD} with sets.
+
+In Edis we implemented five of them.
+
+\subsection{Imposing constraints}
+
+\section{Conclusions and Related Work}
 

### 2015-11-18 paper.tex
6. Making Redis Polymorphic

@@
 
+To execute a \mintinline{haskell}{Edis} program, simply apply it to
+ \mintinline{haskell}{unEdis} to get an ordinary
+ \mintinline{haskell}{Hedis} program, with type informations erased.
+
 \subsection{\mintinline{text}{PING}: the first attempt}
@@
 
-With functions on dictionaries, now we could annotate the effects of a command
-in types. \mintinline{text}{DEL} removes a key from the current database,
+Now we could annotate the effects of a command in types.
+\mintinline{text}{DEL} removes a key from the current database,
 regardless of its type.
@@
  by the argument of type \mintinline{haskell}{Proxy s}.
-To use \mintinline{haskell}{del}, we would have to apply it with a clunky
+To use \mintinline{haskell}{del}, we would have to apply it with a clumsy
  term-level proxy like this:
@@
 
-So far, the datatypes we are dealing with, are all of those built-in Redis data
-structures, such as \emph{strings}, \emph{lists (of strings)},
-\emph{sets (of strings)}, etc.
-But in the real world, these raw strings are hardly useful, people would
+Redis supports many different datatypes, these datatypes as can be viewed as
+\emph{containers} of strings. For example, lists (of strings),
+sets (of strings), and strings themselves.
+
+\section{Denoting containers}
+Most Redis commands only work with a certain type of these containers. To
+annotate what container a key is associated with, we introduce these container
+types.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+data Strings
+data Lists
+data Sets
+...
+\end{minted}
+
+\mintinline{text}{SET} stores a string, regardless the datatype the key was
+ associated with. Now we could implement \mintinline{text}{SET} like this:
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+set :: KnownSymbol s
+    => Proxy s
+    -> ByteString       -- data to store
+    -> Edis xs (Set xs s Strings) (Either Reply Status)
+set key val = Edis $ Hedis.set (symbolVal key) val
+\end{minted}
+
+After \mintinline{text}{SET}, the key will be associated with
+\mintinline{haskell}{Strings} in the dictionary, indicating that it's a string.
+%
+% So far, the datatypes we are dealing with, are all of those built-in Redis data
+% structures, such as \emph{strings}, \emph{lists (of strings)},
+% \emph{sets (of strings)}, etc.
+
+\subsection{Automatic data serialization}
+But in the real world, raw binary strings are hardly useful, people would
 usually serialize their data into strings before storing them, and deserialize
@@
 \mintinline{text}{cereal}, a binary serialization library.
+\mintinline{text}{cereal} comes with these two functions:
 
-\mintinline{text}{cereal} comes with these two functions:
 \begin{minted}[xleftmargin=1em]{haskell}
@@
 
+\subsection{Extending container types}
+We rename container types and extend it with an extra type arguement,
+to indicate what kind of data it's holding.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+data StringOf x
+data ListOf x
+data SetOf x
+...
+\end{minted}
+
+\mintinline{haskell}{set} reimplemented with extended container types:
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+set :: (KnownSymbol s, Serialize x)
+    => Proxy s
+    -> x       -- can be anything, as long as it's serializable
+    -> Edis xs (Set xs s (StringOf x)) (Either Reply Status)
+set key val = Edis $ Hedis.set (symbolVal key) (encode val)
+\end{minted}
+
+For example, if we execute \mintinline{haskell}{set (Proxy :: Proxy "A") True},
+a new entry \mintinline{text}{'}\mintinline{haskell}{("A", StringOf Bool)} will
+be inserted to the dictionary.
+
 \footnotetext{The methods of \mintinline{haskell}{Serialize} will have default
 generic implementations for all datatypes with some language extenstions
-enabled}
+enabled, no sweat!}
 
-\section{Imposing constraints}
+\subsection{Imposing constraints}
 
@@
 
-In Edis we implemented five of them.
-
-\subsection{Imposing constraints}
-
+\section{Discussions}
 \section{Conclusions and Related Work}

### 2015-11-18 paper.tex
6. Imposing constraints

@@
 \setmonofont{Inconsolata}
+
+% for clickable footnotes
+\usepackage{hyperref}
 
@@
 
-\section{Denoting containers}
+\subsection{Denoting containers}
 Most Redis commands only work with a certain type of these containers. To
@@
 
+\footnotetext{The methods of \mintinline{haskell}{Serialize} will have default
+generic implementations for all datatypes with some language extenstions
+enabled, no sweat!}
+
 \subsection{Extending container types}
 We rename container types and extend it with an extra type arguement,
-to indicate what kind of data it's holding.
+ to indicate what kind of encoded value it's holding.
 
@@
 
-\footnotetext{The methods of \mintinline{haskell}{Serialize} will have default
-generic implementations for all datatypes with some language extenstions
-enabled, no sweat!}
-
-\subsection{Imposing constraints}
+\section{Imposing constraints}
 
 To rule out programs with undesired properties, certain constraints must be
-imposed, on what arguments they can take, and what preconditions they must hold.
-These constraints are best encoded in types.
+imposed, on what arguments they can take, or what preconditions they must hold.
 
-\subsection{Redis datatypes}
+Consider the following example: \mintinline{text}{LLEN} retrurns the length of
+the list associated with a key, else raises a type error.
 
-Redis supports many different datatypes, and most commands only work with
-certain types of them. For example: \mintinline{haskell}{SET} with strings or
-\mintinline{haskell}{SADD} with sets.
+\begin{minted}[xleftmargin=1em]{text}
+redis> LPUSH some-list bar
+(integer) 1
+redis> LLEN some-list
+(integer) 1
+redis> SET some-string foo
+OK
+redis> LLEN some-string
+(error) WRONGTYPE Operation against a key
+holding the wrong kind of value
+\end{minted}
+
+Such constraint could be expressed in types with
+\emph{equality constraints}\cite{typeeq}.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+llen :: (KnownSymbol s, Get xs s ~ Just (ListOf x))
+        => Proxy s
+        -> Edis xs xs (Either Reply Integer)
+llen key = Edis $ Hedis.llen (symbolVal key)
+\end{minted}
+
+Where \mintinline{haskell}{(~)} denotes that \mintinline{haskell}{Get xs s}
+and \mintinline{haskell}{Just (ListOf x)} needs to be the same.
+
+The semantics of \mintinline{text}{LLEN} defined above is actually not
+complete. \mintinline{text}{LLEN} also accepts keys that do not exist and
+ returns 0. We will need \emph{}
+
+\begin{minted}[xleftmargin=1em]{text}
+redis> LLEN nonexistent
+(integer) 0
+\end{minted}
+
+In other words, we require that the key to be associated with a list,
+ \textbf{or} does not exist at all.
+
+\subsection{Expressing constraint disjunctions}
+
+Unfortunately, expressing disjunctions in constraints is much more difficult
+than expressing conjunctions, since the latter could be easily done by placing
+ constraints in a tuple (at the left side of \mintinline{haskell}{=>}).
+
+But there are at least three ways to express type-level constraints
+\cite{singletons}. Luckily we could express constraint disjunctions with type
+ families in a modular way.
+
+The semantics we want could be expressed informally like this:
+
+\mintinline{haskell}{Get xs s} $\equiv$ \mintinline{haskell}{Just (ListOf x)}
+$\vee$ $\neg$ \mintinline{haskell}{(Member xs s)}
+
+We could achieve this simply by translating the semantics we want to the
+domain of Booleans, with type-level boolean functions such as
+\mintinline{haskell}{(&&)},
+\mintinline{haskell}{(||)}, \mintinline{haskell}{Not},
+\mintinline{haskell}{(==)}, etc.\footnotemark To avoid
+
+\footnotetext{Available in \mintinline{text}{Data.Type.Bool} and
+ \mintinline{text}{Data.Type.Equality}}
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+Get xs s == Just (ListOf x) || Not (Member xs s)
+\end{minted}
+
+To avoid adressing the type of value (as it may not exist at all), we defined an
+ auxiliary predicate \mintinline{haskell}{IsList :: Maybe * -> Bool} to replace
+ the former part.
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+IsList (Get xs s) || Not (Member xs s)
+\end{minted}
+
+The type expression above has kind \mintinline{haskell}{Bool}, we could make it
+ a type constraint by asserting equality.
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+(IsList (Get xs s) || Not (Member xs s)) ~ True
+\end{minted}
+
+With \emph{constraint kind}, a recent addition to GHC. Type constraints now has
+its own kind: \mintinline{haskell}{Constraint}. That means type constraints
+are not restricted to the left side of a \mintinline{haskell}{=>} anymore,
+they could appear in anywhere that accepts something of kind
+\mintinline{haskell}{Constraint}, and any type that has kind
+\mintinline{haskell}{Constraint} can also be used as a type constraint.\footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}
+We could give the lengthy constraint above a type synonym.
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+ListOrNX xs s =
+    (IsList (Get xs s) || Not (Member xs s)) ~ True
+\end{minted}
 
 \section{Discussions}
+
 \section{Conclusions and Related Work}

### 2015-11-18 paper.tex
draft

@@
 Nonetheless, it's still useful to have a term-level value that links with a
-Symbol, when we want to retreive type-level informations at runtime (but not the
+Symbol, when we want to retrieve type-level information at runtime (but not the
  other way around!).
@@
 
-With all of these ingrediants ready, let's build some dictionaries!
+With all of these ingredients ready, let's build some dictionaries!
 
@@
 postconditions in types. An indexed monad is a type constructor that takes three
-arguments: an intitial state, a final state, and the type of a value that an
+arguments: an initial state, a final state, and the type of a value that an
 action computes, which can be read like a Hoare triple\cite{kleisli}.
@@
 We define a new datatype \mintinline{haskell}{Edis}, which is basically just
-\mintinline{haskell}{Redis} indexed with more informations in types. And we make
+\mintinline{haskell}{Redis} indexed with more information in types. And we make
  it an instance of \mintinline{haskell}{IMonad}.
@@
  \mintinline{haskell}{unEdis} to get an ordinary
- \mintinline{haskell}{Hedis} program, with type informations erased.
+ \mintinline{haskell}{Hedis} program, with type information erased.
 
@@
 
-Dicionary \mintinline{haskell}{xs} in the type remain unaffected after the
+Dictionary \mintinline{haskell}{xs} in the type remain unaffected after the
 action, because \mintinline{haskell}{ping} does not affect any key-value
@@
 
-Type families have a wide variaty of applications. They can appear inside type
+Type families have a wide variety of applications. They can appear inside type
 classes\cite{tfclass}\cite{tfsynonym}, or at toplevel. Toplevel type families
@@
 under the more general instance \mintinline{haskell}{And a b}.
-But the closedness allows these instances to be resoluted in order, just like
-how cases are resoluted in term-level functions. Also notice that how much
+But the closedness allows these instances to be resolved in order, just like
+how cases are resolved in term-level functions. Also notice that how much
 \mintinline{haskell}{And} resembles to it's term-level sibling.
@@
 
-Another benifit of closed type families is that type-level equality can be
+Another benefit of closed type families is that type-level equality can be
 expressed by unifying type variables with the same name.
@@
 \mintinline{haskell}{KnownSymbol} is a class that gives the string associated
-with a concrete type-level symbol, which can be retreived with
+with a concrete type-level symbol, which can be retrieved with
 \mintinline{haskell}{symbolVal}.\footnotemark
@@
 Since Haskell has a \emph{phase distinction}\cite{phasedistinction}, types are
-erased before runtime. It's impossible to obtain informations directly from
+erased before runtime. It's impossible to obtain information directly from
 types, we can only do this indirectly, with \emph{singleton types}
@@
 
-In the type of \mintinline{haskell}{del}, the type varabile
+In the type of \mintinline{haskell}{del}, the type variable
 \mintinline{haskell}{s} would be a \mintinline{haskell}{Symbol} that is decided
@@
 \mintinline{haskell}{Strings} in the dictionary, indicating that it's a string.
-%
-% So far, the datatypes we are dealing with, are all of those built-in Redis data
-% structures, such as \emph{strings}, \emph{lists (of strings)},
-% \emph{sets (of strings)}, etc.
 
@@
 \footnotetext{The methods of \mintinline{haskell}{Serialize} will have default
-generic implementations for all datatypes with some language extenstions
+generic implementations for all datatypes with some language extensions
 enabled, no sweat!}
@@
 \subsection{Extending container types}
-We rename container types and extend it with an extra type arguement,
+We rename container types and extend it with an extra type argument,
  to indicate what kind of encoded value it's holding.
@@
 
+\subsection{Handling \mintinline{text}{INCR}}
+
+Commands such as \mintinline{text}{INCR}, \mintinline{text}{INCRBYFLOAT}, are
+ not only container-specific, there are also some requirements on what types of
+ value they could operate on.
+
+We handle this by mapping Redis's strings of integers and floats to Haskell's
+\mintinline{text}{Integer} and \mintinline{text}{Double}.
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+incr :: (KnownSymbol s, Get xs s ~ Just (StringOf Integer))
+     => Proxy s -> Edis xs xs (Either Reply Integer)
+incr key = Edis $ Hedis.incr (symbolVal key)
+
+incrbyfloat :: (KnownSymbol s, Get xs s ~ Just (StringOf Double))
+     => Proxy s -> Double -> Edis xs xs (Either Reply Double)
+incrbyfloat key n = Edis $ Hedis.incrbyfloat (symbolVal key) n
+\end{minted}
+
 \section{Imposing constraints}
@@
 
-Consider the following example: \mintinline{text}{LLEN} retrurns the length of
+Consider the following example: \mintinline{text}{LLEN} returns the length of
 the list associated with a key, else raises a type error.
@@
 We could achieve this simply by translating the semantics we want to the
-domain of Booleans, with type-level boolean functions such as
+domain of Boolean, with type-level boolean functions such as
 \mintinline{haskell}{(&&)},
@@
 
-To avoid adressing the type of value (as it may not exist at all), we defined an
+To avoid addressing the type of value (as it may not exist at all), we defined an
  auxiliary predicate \mintinline{haskell}{IsList :: Maybe * -> Bool} to replace
@@
 \mintinline{haskell}{Constraint} can also be used as a type constraint.\footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}
-We could give the lengthy constraint above a type synonym.
+
+As many other list-related commands also have this "List or Nothing" semantics,
+we could abstract the lengthy type constraint above and give it an alias with
+type synonym.
 
@@
 
+The complete implementation of \mintinline{text}{LLEN} with
+\mintinline{haskell}{ListOrNX} would become:
+
+\begin{minted}[xleftmargin=1em]{haskell}
+llen :: (KnownSymbol s, ListOrNX xs s)
+        => Proxy s
+        -> Edis xs xs (Either Reply Integer)
+llen key = Edis $ Hedis.llen (symbolVal key)
+\end{minted}
+
+\section{Assertions}
+
+Users may need to make assertions about the status of certain key-type bindings
+ in a Redis program. For example, declaring the existence of a key and the type
+ of its associating value. We provide these functions, which do nothing but
+ fiddling with types.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+declare :: (KnownSymbol s, Member xs s ~ False)
+        => Proxy s
+        -> Proxy x  -- type of value
+        -> Edis xs (Set xs s x) ()
+declare s x = Edis $ return ()
+
+renounce :: (KnownSymbol s, Member xs s ~ True)
+        => Proxy s -> Edis xs (Del xs s) ()
+renounce s = Edis $ return ()
+
+-- empty precondition
+start :: Edis ’[] ’[] ()
+start = Edis $ return ()
+\end{minted}
+
+\subsection{A complete example}
+
+The following program increases the value of "A" as an integer,
+and push the result of increment to list "L", and then pops it out.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+main :: IO ()
+main = do
+    conn <- connect defaultConnectInfo
+    result <- runRedis conn $ unEdis $ start
+        `bind` \_ ->  declare
+                        (Proxy :: Proxy "A")
+                        (Proxy :: Proxy Integer)
+        `bind` \_ ->  incr     (Proxy :: Proxy "A")
+        `bind` \n ->  case n of
+            Left  err -> lpush (Proxy :: Proxy "L") 0
+            Right n   -> lpush (Proxy :: Proxy "L") n
+        `bind` \_ ->  lpop     (Proxy :: Proxy "L")
+    print result
+\end{minted}
+
+The syntax is pretty heavy, like the old days when there's no
+do-notation\cite{history}. But if we don't need any variable bindings between
+operations, we could compose these commands with a sequencing operator
+\mintinline{haskell}{(>>>)}.
+
+\begin{minted}[xleftmargin=1em]{haskell}
+(>>>) :: IMonad m => m p q a -> m q r b -> m p r b
+
+program = start
+    >>> declare
+            (Proxy :: Proxy "A")
+            (Proxy :: Proxy Integer)
+    >>> incr    (Proxy :: Proxy "A")
+    >>> lpush   (Proxy :: Proxy "L") 0
+    >>> lpop    (Proxy :: Proxy "L")
+\end{minted}
+
+We could even get rid of \mintinline{haskell}{(>>>)} with
+ \emph{rebindable syntax}\footnote{See \url{https://downloads.haskell.org/~ghc/latest/docs/html/users_guide/syntax-extns.html#rebindable-syntax}.},
+if we are so inclined.
+
 \section{Discussions}
-
-\section{Conclusions and Related Work}
+\paragraph{Syntax}
+\paragraph{Commands with multiple inputs or outputs}
+\paragraph{Redis Transactions}
+\section{Conclusion and Related Work}
+\paragraph{Acknowledgements}
 

### 2015-11-19 paper.tex
brushed up

@@
  libraries or packages in some programming languages. But people make mistakes,
-while some of the syntactic and semantic errors can be prevented by the language
-at compile time, or caught by the package at runtime, most semantic errors are
-just being ignored, causing problems at the database system.
+ while some of the syntactic and semantic errors can be prevented by the
+ language at compile time, or caught by the package at runtime, most semantic
+ errors are just being ignored, causing problems at the database system.
 
-In this paper, we demonstrates how to prevent those runtime errors at compile
-time, allowing users to write more reliable database queries without runtime
-overhead, by exploiting type-level programming techniques such as indexed
-monad, type-level literals and closed type families in Haskell.
+In this paper, we demonstrate how to prevent those runtime errors at compile
+ time, thus allowing users to write more reliable database queries without
+ runtime overhead, by exploiting type-level programming techniques such as
+ indexed monad, type-level literals and closed type families in Haskell.
 
-The database system and the package we are targeting
-are \emph{Redis} and \emph{Hedis} respectively, and our implementation is
-available as \emph{Edis} on \emph{Hackage}.
+The database system and the package we are targeting are \emph{Redis} and
+ \emph{Hedis} respectively, and our implementation is available as \emph{Edis}
+ on \emph{Hackage}.
 
@@
 \paragraph{Keywords}
-type-level programming; Haskell; database query language; Redis;
+Haskell; type-level programming; database query language; Redis;
 \end{abstract}
@@
  string key to identify and manipulate with. Redis supports many different
-kind of values, such as strings, hashes, lists and sets.
+ kind of values, such as strings, hashes, lists, sets, etc.
 
-To manipulate these values, Redis comes with a set of atomic operations,
-called \emph{commands}.
+To manipulate these values, Redis comes with a set of atomic operations, called
+ \emph{commands}.
 
@@
  Hedis in Haskell.
+
 \begin{minted}[xleftmargin=1em]{haskell}
@@
 
-Function \mintinline{haskell}{sadd} takes a key and a list of values as
-arguments, and returns an \mintinline{haskell}{Integer}, wrapped in
+\mintinline{haskell}{sadd} takes a key and a list of values as arguments, and
+ returns an \mintinline{haskell}{Integer}, wrapped in
  \mintinline{haskell}{Either} to indicate possible failures in the context
@@
 \footnotetext{Hedis provides another kind of context, \mintinline{haskell}{RedisTx}, for \emph{transactions}, united with \mintinline{haskell}{Redis} under the class of \mintinline{haskell}{RedisCtx}.
-We use \mintinline{haskell}{Redis} only here for brevity.
+We demonstrate \mintinline{haskell}{Redis} only for brevity.
 }
@@
 
-Although everything in Redis is essentially a binary string, they are treated
-differently. Redis supports many different kind of data structures, such as
-strings, hashes, lists, etc. These values have different datatypes, and most
-commands only work with certain types of them, much like how C language treats
-a piece of data.
+Although everything in Redis is essentially a binary string, these strings are
+ treated differently. Redis supports many different kind of data structures,
+ such as strings, hashes, lists, etc. These values have different datatypes,
+ and most commands only work with certain types of them, much like how C
+ language treats a piece of data.
 
-\paragraph{Problem 1} Consider the following example. \mintinline{text}{SET} is
-a command that only works on strings. The key \mintinline{text}{some-string} is
-now related with a string. If we treat it as a set and add an element to it with
-\mintinline{text}{SADD}, a runtime error arises.
+\paragraph{Problem 1} Consider the following case. \mintinline{text}{SET} is
+ a command that only works with strings, it associates the key
+ \mintinline{text}{some-string} with a value of string. If we treat the like a
+ set and add an element to it with \mintinline{text}{SADD}, a runtime error
+ arises.
 
@@
  and Hedis can be regarded as one of them. What makes Hedis special is that,
-it has variable bindings (between keys and values), but with very
+ it has \emph{variable bindings} (between keys and values), but with very
  little or no semantic checking, neither dynamically nor statically.
@@
  language extensions like type-level literals and data kinds.
-\item We provide (yet another) an example of encoding effects and
- constraints of an action in types, with indexed monad\cite{indexedmonad}
-  and other language extensions such as closed
-  type-families\cite{closedtypefamilies} and constraints
-  kinds\cite{constraintskinds}.
+\item We provide (yet another) an example of encoding effects and constraints of
+ an action in types, with indexed monad\cite{indexedmonad} and other language
+ extensions such as closed type-families\cite{closedtypefamilies} and
+ constraints kinds\cite{constraintskinds}.
 \item Edis, a package we built for programmers. This package helps programmers
- to write more reliable Redis codes, and also makes Redis polymorphic by
- automatically converting back and forth from values of any types and boring
- ByteStrings.
+ to write more reliable Redis programs, and also makes Redis polymorphic by
+ automatically converting back and forth from values of arbitrary types and
+ boring ByteStrings.
 \end{itemize}
@@
 
-\footnotetext{\mintinline{haskell}{TypeRep} here represents term-level
-representations of datatypes, available in \mintinline{haskell}{Data.Typeable}}
+\footnotetext{\mintinline{haskell}{TypeRep} supports term-level representations
+ of datatypes, available in \mintinline{haskell}{Data.Typeable}}
 
@@
 
-Luckily, with recently added GHC extension \emph{data kinds}. With data kinds
-we can define our own kinds just like we can with datatypes, and every suitable
-datatype will be promoted to be a kind, and its value constructors promoted
-to be type constructors, automatically.
+Luckily, with recently added GHC extension \emph{data kinds}, suitable
+ datatype will be automatically promoted to be a kind, and its value
+ constructors to be type constructors.
 
@@
 
-Haskell sugars lists \mintinline{haskell}{[1, 2, 3]} and
-tuples \mintinline{text}{(1, 'a')} with brackets and parentheses.
-Promoted lists and tuples are also sugared with single quote prefixed. For
-example \mintinline{text}{'}\mintinline{haskell}{[Int, Char]},
+Haskell sugars lists \mintinline{haskell}{[1, 2, 3]} and tuples
+ \mintinline{text}{(1, 'a')} with brackets and parentheses.
+ We could also express promoted lists and tuples in types like this with
+ a single quote prefixed. For example:
+ \mintinline{text}{'}\mintinline{haskell}{[Int, Char]},
  \mintinline{text}{'}\mintinline{haskell}{(Int, Char)}.
@@
 
-Symbol is defined without a value constructor, because it's
-intended to be used as a promoted kind.
+Symbol is defined without a value constructor, because it's intended to be used
+ as a promoted kind.
 
@@
 \end{minted}
-
-Nonetheless, it's still useful to have a term-level value that links with a
-Symbol, when we want to retrieve type-level information at runtime (but not the
- other way around!).
+%
+% Nonetheless, it's still useful to have a term-level value that links with a
+%  Symbol, when we want to retrieve type-level information at runtime (but not the
+%  other way around!).
 
@@
 
-They are defined with \emph{type synonym}, since they are \emph{types}
-, not \emph{terms}. If we ask \mintinline{text}{GHCi} what is the kind of
-\mintinline{haskell}{Dict1}, we get \mintinline{haskell}{Dict1 :: [ (Symbol, *) ]}
+These dictionaries are defined with \emph{type synonym}, since they are
+ \emph{types}, not \emph{terms}. If we ask \mintinline{text}{GHCi} what is the
+ kind of \mintinline{haskell}{Dict1}, we will get
+ \mintinline{haskell}{Dict1 :: [ (Symbol, *) ]}
 
-The star \mintinline{haskell}{*} is a kind, it stands for any kind of types,
-such as \mintinline{haskell}{Int}, \mintinline{haskell}{Char} or even a symbol
-\mintinline{haskell}{"symbol"}, while \mintinline{haskell}{Symbol} stands only
-for all symbols.
+The kind \mintinline{haskell}{*} (pronounced "star") stands for the set of all
+ concrete type expressions, such as \mintinline{haskell}{Int},
+ \mintinline{haskell}{Char} or even a symbol \mintinline{haskell}{"symbol"},
+ while \mintinline{haskell}{Symbol} is restricted to all symbols only.
 
@@
  can be used\cite{typefun}\cite{staticresources} to model such preconditions and
-postconditions in types. An indexed monad is a type constructor that takes three
-arguments: an initial state, a final state, and the type of a value that an
-action computes, which can be read like a Hoare triple\cite{kleisli}.
+ postconditions in types. An indexed monad is a type constructor that takes
+ three arguments: an initial state, a final state, and the type of a value that
+ an action computes, which can be read like a Hoare triple\cite{kleisli}.
 
@@
  \mintinline{haskell}{unit} for identities and \mintinline{haskell}{bind} for
-compositions, as in Monad.
+ compositions, as in monads.
 
 We define a new datatype \mintinline{haskell}{Edis}, which is basically just
-\mintinline{haskell}{Redis} indexed with more information in types. And we make
- it an instance of \mintinline{haskell}{IMonad}.
+ the context \mintinline{haskell}{Redis} indexed with more information in types.
+ We make it an instance of \mintinline{haskell}{IMonad}.
 
@@
 To execute a \mintinline{haskell}{Edis} program, simply apply it to
- \mintinline{haskell}{unEdis} to get an ordinary
- \mintinline{haskell}{Hedis} program, with type information erased.
+ \mintinline{haskell}{unEdis} to get an ordinary program of type
+ \mintinline{haskell}{Redis}, with type information erased.
 
@@
 \footnotetext{\mintinline{haskell}{ping} from Hedis is qualified with
-\mintinline{haskell}{Hedis} to prevent function name clashing}
+\mintinline{haskell}{Hedis} to prevent function name clashing in our code.}
 
@@
 
-Dictionary \mintinline{haskell}{xs} in the type remain unaffected after the
+The dictionary \mintinline{haskell}{xs} in the type remains unaffected after the
  action, because \mintinline{haskell}{ping} does not affect any key-value
-bindings. To encode other commands that alters key-value bindings, we need
+ bindings. To encode other commands that modifies key-value bindings, we need
  type-level functions to annotate those effects on the dictionary.
@@
  classes\cite{tfclass}\cite{tfsynonym}, or at toplevel. Toplevel type families
-can be used to compute over types, they come in two forms: open\cite{tfopen} and
-closed \cite{tfclosed}.
+ can be used to compute over types, they come in two forms: open\cite{tfopen}
+ and closed \cite{tfclosed}.
 
@@
 
-The first instance of \mintinline{haskell}{And} could be subsumed
-under the more general instance \mintinline{haskell}{And a b}.
+The first instance of \mintinline{haskell}{And} could be subsumed under a more
+ general instance, \mintinline{haskell}{And a b}.
 But the closedness allows these instances to be resolved in order, just like
@@
  and these types are computed at compile-time. It wouldn't make
-much sense for type checkers to crash and throw a "Non-exhaustive" error or
+ much sense for a type checker to crash and throw a "Non-exhaustive" error or
  be non-terminating.
@@
 \end{minted}
-
-Other dictionary-related functions are defined in a similar way.
+%
+Other dictionary-related functions are defined in a similar fashion.
 
@@
 
-Now we could annotate the effects of a command in types.
-\mintinline{text}{DEL} removes a key from the current database,
-regardless of its type.
+Now we could annotate the effects of a command in types. \mintinline{text}{DEL}
+ removes a key from the current database, regardless of its type.
 
@@
  erased before runtime. It's impossible to obtain information directly from
-types, we can only do this indirectly, with \emph{singleton types}
-\cite{singletons}.
+ types, we can only do this indirectly, with
+ \emph{singleton types}\cite{singletons}.
 
@@
 In the type of \mintinline{haskell}{del}, the type variable
-\mintinline{haskell}{s} would be a \mintinline{haskell}{Symbol} that is decided
- by the argument of type \mintinline{haskell}{Proxy s}.
+ \mintinline{haskell}{s} is a \mintinline{haskell}{Symbol} that is decided by
+ the argument of type \mintinline{haskell}{Proxy s}.
  To use \mintinline{haskell}{del}, we would have to apply it with a clumsy
@@
 
-Commands such as \mintinline{text}{INCR}, \mintinline{text}{INCRBYFLOAT}, are
- not only container-specific, there are also some requirements on what types of
+Commands such as \mintinline{text}{INCR} and \mintinline{text}{INCRBYFLOAT}, are
+ not only container-specific, they also have some requirements on what types of
  value they could operate on.
 
-We handle this by mapping Redis's strings of integers and floats to Haskell's
-\mintinline{text}{Integer} and \mintinline{text}{Double}.
+We could handle this by mapping Redis's strings of integers and floats to
+ Haskell's \mintinline{text}{Integer} and \mintinline{text}{Double}.
 
@@
 The semantics of \mintinline{text}{LLEN} defined above is actually not
-complete. \mintinline{text}{LLEN} also accepts keys that do not exist and
- returns 0. We will need \emph{}
+complete. \mintinline{text}{LLEN} also accepts keys that do not exist, and
+ replies with \mintinline{text}{0}.
 
@@
 In other words, we require that the key to be associated with a list,
- \textbf{or} does not exist at all.
+ textbf{unless} it doesn't exist at all.
 
@@
 
-But there are at least three ways to express type-level constraints
+There are at least three ways to express type-level constraints
 \cite{singletons}. Luckily we could express constraint disjunctions with type
@@
 
-To avoid addressing the type of value (as it may not exist at all), we defined an
- auxiliary predicate \mintinline{haskell}{IsList :: Maybe * -> Bool} to replace
- the former part.
+To avoid addressing the type of value (as it may not exist at all), we defined
+ an auxiliary predicate \mintinline{haskell}{IsList :: Maybe * -> Bool} to
+ replace the former part.
 
@@
  \mintinline{haskell}{Constraint}, and any type that has kind
-\mintinline{haskell}{Constraint} can also be used as a type constraint.\footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}
+ \mintinline{haskell}{Constraint} can also be used as a type constraint.
+ \footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}
 
@@
 
-Users may need to make assertions about the status of certain key-type bindings
- in a Redis program. For example, declaring the existence of a key and the type
+Users may need to make assertions about the status of some key-type bindings in
+ a Redis program. For example, declaring the existence of a key and the type
  of its associating value. We provide these functions, which do nothing but
@@
 
-The following program increases the value of "A" as an integer,
-and push the result of increment to list "L", and then pops it out.
+The following program increases the value of "A" as an integer, push the result
+ of the increment to list "L", and then pops it out.
 
@@
 The syntax is pretty heavy, like the old days when there's no
-do-notation\cite{history}. But if we don't need any variable bindings between
-operations, we could compose these commands with a sequencing operator
+ \emph{do-notation}\cite{history}. But if we don't need any variable bindings
+ between operations, we could compose these commands with a sequencing operator
  \mintinline{haskell}{(>>>)}.
@@
 
-We could even get rid of \mintinline{haskell}{(>>>)} with
- \emph{rebindable syntax}\footnote{See \url{https://downloads.haskell.org/~ghc/latest/docs/html/users_guide/syntax-extns.html#rebindable-syntax}.},
-if we are so inclined.
-
 \section{Discussions}

### 2015-11-19 paper.tex
8. Discussions

@@
     -> Edis xs (Del xs s) (Either Reply Integer)
-del key = Edis $ Hedis.del (symbolVal key)
+del key = Edis $ Hedis.del (encodeKey key)
 \end{minted}
@@
  \mintinline{haskell}{symbolVal}.\footnotemark
+ Where \mintinline{haskell}{encodeKey} converts \mintinline{haskell}{Proxy s} to
+ \mintinline{haskell}{ByteString}.
+\footnotetext{They are defined in \mintinline{haskell}{GHC.TypeLits}.}
 
-\footnotetext{They are defined in \mintinline{haskell}{GHC.TypeLits}.}
+\begin{minted}[xleftmargin=1em]{haskell}
+encodeKey :: KnownSymbol s => Proxy s -> ByteString
+encodeKey = encode . symbolVal
+\end{minted}
 
@@
     -> Edis xs (Set xs s Strings) (Either Reply Status)
-set key val = Edis $ Hedis.set (symbolVal key) val
+set key val = Edis $ Hedis.set (encodeKey key) val
 \end{minted}
@@
     -> Edis xs (Set xs s (StringOf x)) (Either Reply Status)
-set key val = Edis $ Hedis.set (symbolVal key) (encode val)
+set key val = Edis $ Hedis.set (encodeKey key) (encode val)
 \end{minted}
@@
      => Proxy s -> Edis xs xs (Either Reply Integer)
-incr key = Edis $ Hedis.incr (symbolVal key)
+incr key = Edis $ Hedis.incr (encodeKey key)
 
@@
      => Proxy s -> Double -> Edis xs xs (Either Reply Double)
-incrbyfloat key n = Edis $ Hedis.incrbyfloat (symbolVal key) n
+incrbyfloat key n = Edis $ Hedis.incrbyfloat (encodeKey key) n
 \end{minted}
@@
         -> Edis xs xs (Either Reply Integer)
-llen key = Edis $ Hedis.llen (symbolVal key)
+llen key = Edis $ Hedis.llen (encodeKey key)
 \end{minted}
@@
 
-As many other list-related commands also have this "List or Nothing" semantics,
+As many other list-related commands also have this "List or nothing" semantics,
  we could abstract the lengthy type constraint above and give it an alias with
@@
 \paragraph{Syntax}
+No one could ignore the glaring shortcoming of the syntax, which occurs mainly
+ in two places: symbol singletons and indexed monad. We are hoping that these
+ issues could be solved with future syntactic extensions.
+
+\paragraph{Returns only determined datatypes}
+
+All other data structures in Redis also follow the semantics similar to
+ "List or nothing" as \mintinline{text}{LLEN} has. Take the case of
+ \mintinline{text}{GET}, which has a similar semantics that should be typed:
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+get :: (KnownSymbol s, Serialize x, StringOrNX xs s)
+    => Proxy s -> Edis xs xs (Either Reply (Maybe x))
+\end{minted}
+
+Since the key may not exist, we don't know what \mintinline{haskell}{x} would
+be. We could left \mintinline{haskell}{x} ambiguous, and let it be decided by
+ the caller. But users will then be forced to spell out the complete
+type signature of everything, including the dictionaries, only to specify
+the desired resulting type.
+
+Instead of allowing the key to be non-existent, we require that the key must
+ exist and it's associating type to be determined at compile time. So our
+ version of \mintinline{haskell}{get} has a stricter semantics:
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+get :: (KnownSymbol s, Serialize x
+     , Just (StringOf x) ~ Get xs s)
+    => Proxy s -> Edis xs xs (Either Reply (Maybe x))
+\end{minted}
+
 \paragraph{Commands with multiple inputs or outputs}
+
+Some command may take a variable number of arguments as inputs, and returns more
+ than one value as outputs. To illustrate this, consider
+ \mintinline{haskell}{sinter} in \mintinline{haskell}{Hedis}:
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+Hedis.sinter :: [ByteString]                      -- keys
+             -> Redis (Either Reply [ByteString]) -- values
+\end{minted}
+
+In Hedis such command could easily be expressed with lists of
+ \mintinline{haskell}{ByteString}s. But in Edis, things escalate quickly, as
+ the keys and values will have to be expressed with heterogeneous
+ lists\cite{hetero}, which would be pratically infeasible, considering the cost,
+ if not impossible.
+
+Most importantly, the keys will all have to be constrained by
+ \mintinline{haskell}{KnownSymbol}, which enforces these type literals to be
+ concrete and known at compile time.
+ It's still unclear whether this is possible.
+
+So instead, we are offering commands that only has a single input and output.
+
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+sinter :: ByteString                      -- single key
+       -> Redis (Either Reply ByteString) -- single value
+\end{minted}
+
 \paragraph{Redis Transactions}
+
 \section{Conclusion and Related Work}

### 2015-11-19 paper.tex
8. Discussions: Redis Transactions

@@
 \footnotetext{Hedis provides another kind of context, \mintinline{haskell}{RedisTx}, for \emph{transactions}, united with \mintinline{haskell}{Redis} under the class of \mintinline{haskell}{RedisCtx}.
-We demonstrate \mintinline{haskell}{Redis} only for brevity.
+We are demonstrating \mintinline{haskell}{Redis} only for brevity.
 }
@@
 
+Redis provides transactions, another context for executing commands.
+Redis trasanctions are atomic in the sense that, all commands in a transaction
+ will executed sequentially, and no other requests issued by other clients will
+ be served in the middle.\footnotemark In contrast, we cannot make such
+ guarantee in the ordinary context, which may destroy the assertions we made
+ in types.
+
+At the point of writing, transactions are not supported in our implementation.
+ We are planning to add it in the future, and we are expecting that there
+ wouldn't be much difficulties, since we've implemented a runtime type checker
+ specifically targeting Redis transactions once, before we moved on to the
+ types.
+
+\footnotetext{\url{http://redis.io/topics/transactions}}
+
 \section{Conclusion and Related Work}

### 2015-11-19 paper.tex
Polish some of the wordings, up to Section 1.3.

--- 2015-11-19 paper.tex
@@
 
+\newcommand{\todo}[2]{{\bf [#1]}: #2}
+
 \begin{document}
 
-\title{Preventing runtime errors of Redis at compile time}
+\title{Preventing Runtime Errors of Redis at Compile Time}
 
@@
 \begin{abstract}
-Programmers often interact with database systems, by sending queries through
- libraries or packages in some programming languages. But people make mistakes,
- while some of the syntactic and semantic errors can be prevented by the
+Programmers often interact with database systems by sending queries through libraries or packages in some programming languages. People, however, make mistakes.
+While some of the syntactic and semantic errors can be prevented by the
  language at compile time, or caught by the package at runtime, most semantic
@@
 
-Redis is an open source, in-memory data structure store, used as
- database, cache and message broker. Each value is associated with a binary-safe
- string key to identify and manipulate with. Redis supports many different
- kind of values, such as strings, hashes, lists, sets, etc.
+\todo{scm}{Cite some references for Redis?}
 
-To manipulate these values, Redis comes with a set of atomic operations, called
- \emph{commands}.
+Redis is an open source, in-memory data structure store, often used as
+database, cache and message broker.
+A Redis datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with.
+Redis supports many different kind of values, such as strings, hashes, lists, and sets, etc, and provides a collection of of atomic \emph{commands} to manipulate these values.
 
-For example, if we want to add a bunch of strings to 2 sets, and intersect them,
- we could input the following sequence of commands into Redis's client
- interactively.
-
+For an example, consider the following sequence of commands, entered through the interactive interface of Redis. The keys
+\mintinline{text}{some-set} abd \mintinline{text}{another-set} are
+both associated to a set. The two call to command \mintinline{text}{SADD} respectively adds three and two values to the two sets, before \mintinline{text}{SINTER} takes their intersection:
 \begin{minted}[xleftmargin=1em]{text}
@@
 
-Keys such as \mintinline{text}{some-set} and \mintinline{text}{another-set}
- are created on site, with the command \mintinline{text}{SADD}, which returns
- the size of set after completed.
+Note that the keys \mintinline{text}{some-set} and \mintinline{text}{another-set}, if not existing before the call to \mintinline{text}{SADD}, are created on site. The call to
+\mintinline{text}{SADD} returns the size of the set after
+completion of the command.
 
 \subsection{Hedis}
-Redis can also be used in most programming languages, with 3rd party libraries
- or packages that talk with Redis' TCP protocol. In Haskell, the most popular
- package is Hedis.
 
-The previous example in Redis's client would look something like this with
- Hedis in Haskell.
+Many third party library exist to allow general purpose programmings
+to access Redis databases through its TCP protocol. The most popular
+such library for Haskell is Hedis.
+%
+\todo{scm}{Citation for Hedis?}
+%
+The following program is how the previous example looks like in
+Hedis:
 
@@
 
-\mintinline{haskell}{sadd} takes a key and a list of values as arguments, and
+The function \mintinline{haskell}{sadd} takes a key and a list of values as arguments, and
  returns an \mintinline{haskell}{Integer}, wrapped in
@@
  \mintinline{haskell}{Redis}\footnotemark of command execution.
+\todo{scm}{Explain Reply in one sentence?}
 
@@
 
-Keys and values are of type \mintinline{haskell}{ByteString}, since they are all
- just binary strings in Redis. If a user wants to store values of
- arbitrary types, he or she will have to encode and decode them as
- \mintinline{haskell}{ByteString}s.
+Note that keys and values, being nothing but binary strings in Redis,
+are represented using Haskell \mintinline{haskell}{ByteString}.
+Values of other types must be encoded as \mintinline{haskell}{ByteString}s before being written to the database,
+and decoded after being read back.
 
 \footnotetext{Hedis provides another kind of context, \mintinline{haskell}{RedisTx}, for \emph{transactions}, united with \mintinline{haskell}{Redis} under the class of \mintinline{haskell}{RedisCtx}.
-We demonstrate \mintinline{haskell}{Redis} only for brevity.
+For brevity, we demonstrate only \mintinline{haskell}{Redis} in this paper.
 }
@@
 
-Although everything in Redis is essentially a binary string, these strings are
- treated differently. Redis supports many different kind of data structures,
- such as strings, hashes, lists, etc. These values have different datatypes,
- and most commands only work with certain types of them, much like how C
- language treats a piece of data.
+\todo{scm}{Is this a quote? If so please cite the source. If not,
+we shouldn't use the double quotes.}
 
-\paragraph{Problem 1} Consider the following case. \mintinline{text}{SET} is
- a command that only works with strings, it associates the key
- \mintinline{text}{some-string} with a value of string. If we treat the like a
- set and add an element to it with \mintinline{text}{SADD}, a runtime error
- arises.
+Although everything in Redis is essentially a binary string, these strings are treated differently. Redis supports many different
+kind of data structures, such as strings, hashes, lists, etc. While
+they are all encoded as binary strings before being written to the databse, most commands, much like how the C language treats a piece of data, only works with data of certain types.
+
+\paragraph{Problem 1} The command \mintinline{text}{SET}, by
+its definition, associates a key to a string. In the following
+example, the key \mintinline{text}{some-string} is associated
+to string \mintinline{text}{foo}. Subsequent calls to
+\mintinline{text}{SADD} causes runtime errors, since the value
+of \mintinline{text}{some-string} is not a set, but a string.
 
@@
 \paragraph{Problem 2} Even worse, not all strings are equal!
- \mintinline{text}{INCR} parses the string value as an integer, increments it by
- one, and store it back as a string. If a string value can't be parse as an
- integer, another runtime error arises
+The call \mintinline{text}{INCR some-string} parses the string
+associated with key \mintinline{text}{some-string}
+to an integer, increments it by one, and store it back as a string.
+If the string can not be parse as an integer, a runtime error
+is raised.
 
@@
 
-\paragraph{In Hedis} The same goes for Hedis, since it's only a simple wrapping
- on top of Redis's protocol written in Haskell.
+\paragraph{In Hedis} Hedis, being only a simple wrapper on top
+of the TCP protocol of Redis, inherits all the problems mentioned
+above. The following program yields the same error as that in
+the Redis client.
 
@@
 \end{minted}
-
-It yields the same error as in Redis's client.
-
 \begin{minted}[xleftmargin=1em]{haskell}

### 2015-11-19 paper.tex
Added one simple notice.
draft

--- 2015-11-19 paper.tex
@@
 
-\authorinfo{Person 1 \and Person 2}
-{\makebox{A Department} \\
-\makebox{A University}  \\
-\makebox{A Place, AS 12345}}
-{\{person1,person2\}@cs.auniv.edu}
+\authorinfo{Ting-Yan Lai}
+{\makebox{Institute of Computer Science and Engineering} \\
+\makebox{National Chiao Tung University, Taiwan}}
+{banacorn.cs03g@nctu.edu.tw}
+
+\authorinfo{Tyng-Ruey Chuang \and Shin-Cheng Mu}
+{\makebox{Institute of Information Science} \\
+\makebox{Academia Sinica, Taiwan}}
+{\{trc,scm\}@iis.sinica.edu.tw}
 
@@
 
-\subsection{Problems}
+\section{Motivation}
 
@@
 No one could ignore the glaring shortcoming of the syntax, which occurs mainly
- in two places: symbol singletons and indexed monad. We are hoping that these
- issues could be solved with future syntactic extensions.
+ in two places: \emph{symbol singletons} and \emph{indexed monad}. We are hoping
+ that these issues could be solved with future syntactic extensions.
 
@@
  \mintinline{haskell}{ByteString}s. But in Edis, things escalate quickly, as
- the keys and values will have to be expressed with heterogeneous
- lists\cite{hetero}, which would be pratically infeasible, considering the cost,
+ the keys and values will have to be expressed with \emph{heterogeneous
+ lists}\cite{hetero}, which would be pratically infeasible, considering the cost,
  if not impossible.
@@
 
-Redis provides transactions, another context for executing commands.
+Redis also provides \emph{transactions}, another context for executing commands.
 Redis trasanctions are atomic in the sense that, all commands in a transaction
  will executed sequentially, and no other requests issued by other clients will
- be served in the middle.\footnotemark In contrast, we cannot make such
- guarantee in the ordinary context, which may destroy the assertions we made
+ be served \textbf{in the middle}.\footnotemark In contrast, we cannot make such
+ a guarantee in the ordinary context, which may destroy the assertions we made
  in types.
@@
 
-\section{Conclusion and Related Work}
+\section{Related Work}
+
+While Redis can be viewed as a non-relational database system (although Redis
+ seems reluctant to admit this in recent years), there also similar goals on
+ relational database systems, trying to achieve a safer database interface by
+ making them statically checked.
+ \emph{HaskellDB}\cite{haskelldb}\cite{haskelldbimproved} expresses quries
+ with relational algebra-like combinators, wrapped in phantom types. Or as
+ examples, at first to demonstrate the power of dependent types in
+ Agda\cite{pi}, and then with singletons in Haskell\cite{singletons}.
+
+\section{Conclusions}
+
+We have shown that how a embedded DSL with variable-bindings could be expressed
+ in types and statically checked. With recent extensions to the language,
+ Haskell is becoming more and more capable of dependently-typed programming, and
+ it's not that scary as many (including me) would have thought.
+
 \paragraph{Acknowledgements}
+This material is based upon work supported by Institute of Information Science,
+Academia Sinica under the Summer Internship Program of 2015.
 

### 2015-11-20 paper.tex
Comments up to section 2.

--- 2015-11-19 paper.tex
@@
 
+\todo{scm}{say a bit more.}
+
 \subsection{Hedis as an embedded DSL}
@@
 
-We began with making Hedis a dynamically checked embedded DSL, and implemented a
- runtime type checker that keeps track of everything. But then we found that
- things can be a lot easier, by leveraging the host language's type checker,
+\todo{scm}{Is this a good thing? Being ``special'' sounds positive.}
+
+We began with making Hedis a dynamically checked embedded DSL, and implemented a runtime type checker that keeps track of types of
+each variable. But then we found that things can be a lot easier,
+by leveraging the host language's type checker,
 which also make it statically type-checked!
+
+\todo{scm}{Elaborate on how the latter approach is easier.}
+\todo{scm}{Briefly summarise the approach you use. E.g, ``monadish'',
+type of variables are recorded in a list, type level list, etc, such
+that readers familiar with the topic will have a rough idea even
+without reading on.}
 
@@
 
+\todo{scm}{I think we have to somehow mention that
+not all Hedis programs can be typechecked (even if they might
+turn out to be type safe), as a cost of static type checking.}
+
 \section{Type-level dictionaries}
+
+\todo{scm}{At this point the readers should have some idea
+what we are going to do, thus the readers is motivated to
+use type-level lists.}
 

### 2015-11-20 paper.tex

--- 2015-11-19 paper.tex
@@
  with respects to \textbf{the type of a value that associates with a key}.
+
+\todo{scm}{say a bit more.}
 
@@
 
-\section{Conclusion and Related Work}
+Redis also provides \emph{transactions}, another context for executing commands.
+Redis trasanctions are atomic in the sense that, all commands in a transaction
+ will executed sequentially, and no other requests issued by other clients will
+ be served \textbf{in the middle}.\footnotemark In contrast, we cannot make such
+ a guarantee in the ordinary context, which may destroy the assertions we made
+ in types.
+
+At the point of writing, transactions are not supported in our implementation.
+ We are planning to add it in the future, and we are expecting that there
+ wouldn't be much difficulties, since we've implemented a runtime type checker
+ specifically targeting Redis transactions once, before we moved on to the
+ types.
+
+\section{Related Work}
+
+While Redis can be viewed as a non-relational database system (although Redis
+ seems reluctant to admit this in recent years), there also similar goals on
+ relational database systems, trying to achieve a safer database interface by
+ making them statically checked.
+ \emph{HaskellDB}\cite{haskelldb}\cite{haskelldbimproved} expresses quries
+ with relational algebra-like combinators, wrapped in phantom types. Or as
+ examples, at first to demonstrate the power of dependent types in
+ Agda\cite{pi}, and then with singletons in Haskell\cite{singletons}.
+
+\section{Conclusions}
+
+We have shown that how a embedded DSL with variable-bindings could be expressed
+ in types and statically checked. With recent extensions to the language,
+ Haskell is becoming more and more capable of dependently-typed programming, and
+ it's not that scary as many (including me) would have thought.
+
 \paragraph{Acknowledgements}
+This material is based upon work supported by Institute of Information Science,
+Academia Sinica under the Summer Internship Program of 2015.
 

### 2015-11-20 paper.tex
fix all \todos

--- 2015-11-20 paper.tex
@@
 
-\todo{scm}{Cite some references for Redis?}
-
-Redis is an open source, in-memory data structure store, often used as
+Redis\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as
 database, cache and message broker.
@@
 to access Redis databases through its TCP protocol. The most popular
-such library for Haskell is Hedis.
-%
-\todo{scm}{Citation for Hedis?}
-%
+such library for Haskell is Hedis\footnote{\url{https://hackage.haskell.org/package/hedis}}.
+
 The following program is how the previous example looks like in
@@
 
-The function \mintinline{haskell}{sadd} takes a key and a list of values as arguments, and
- returns an \mintinline{haskell}{Integer}, wrapped in
- \mintinline{haskell}{Either} to indicate possible failures in the context
- \mintinline{haskell}{Redis}\footnotemark of command execution.
-\todo{scm}{Explain Reply in one sentence?}
+The function \mintinline{haskell}{sadd} takes a key and a list of values as
+ arguments, and returns an \mintinline{haskell}{Integer} on success, or
+ returns a \mintinline{haskell}{Reply}, a low-level representation of replies
+ from the Redis server, in case of failues. All wrapped in
+ \mintinline{haskell}{Redis}\footnotemark, the context of command execution.
 
@@
 
-\say{All binary strings are equal, but some binary strings are more equal than others.}
-
-\todo{scm}{Is this a quote? If so please cite the source. If not,
-we shouldn't use the double quotes.}
+All binary strings are equal, but some binary strings are more equal than others.
 
@@
 
-\paragraph{The Cause} These problems arise from the absence of type checking,
- with respects to \textbf{the type of a value that associates with a key}.
-
-\todo{scm}{say a bit more.}
+\paragraph{The Cause} Every key is associated with a value, and every value has
+ it's own type. But Most commands in Redis only work with a certain type of
+ value. When a command is used on a wrong type of key, a runtime error occurs.
+ The problems illustrated above arise from the absence of type checking, with
+ respects to \textbf{the type of a value that associates with a key}.
+ These problems could have been avoided, if we could know the type every key
+ associates with in advance, and prevent programs with invalid commands from
+ executing.
 
@@
 Haskell makes it easy to build and use Domain Specific Languages (DSLs),
- and Hedis can be regarded as one of them. What makes Hedis special is that,
+ and Hedis can be regarded as one of them. What makes Hedis peculiar is that,
  it has \emph{variable bindings} (between keys and values), but with very
@@
 We began with making Hedis a dynamically checked embedded DSL, and implemented a
- runtime type checker that keeps track of everything. But then we found that
- things can be a lot easier, by leveraging the host language's type checker,
- which also make it statically type-checked!
+ runtime type checker that keeps track of types of each variable. But then we
+ found that things can be a lot easier, by leveraging the host language's type
+ checker. We encode variable bindings with \emph{type-level lists} and
+ \emph{strings}, and control the effects on the bindings with
+ \emph{indexed monad}. In contrast to the former approach, we \textbf{embedded}
+ our type checker into Haskell's type system, without having to build a
+ \textbf{standalone} one on the term level.
 
@@
 
+Notice that \textbf{not all Hedis programs can be typechecked} (even if they
+ might turn out to be type safe). We opted for type safety instead of
+ expressiveness.
+
 \paragraph{Redis Transactions}
@@
 
-We have shown that how a embedded DSL with variable-bindings could be expressed
- in types and statically checked. With recent extensions to the language,
- Haskell is becoming more and more capable of dependently-typed programming, and
- it's not that scary as many (including me) would have thought.
+We have demonstrated how a DSL with variable-bindings could be embedded in types.
+With more and more extensions added to the language, Haskell is gradually
+ becoming a dependently-typed language,\footnote{Why not be dependently typed? \url{http://stackoverflow.com/questions/12961651/why-not-be-dependently-typed}}
+ but it's not that dreadful as many (including us) would have thought.
 

### 2015-11-20 paper.tex
codename for the purposes of double-blind reviewing

@@
 The database system and the package we are targeting are \emph{Redis} and
- \emph{Hedis} respectively, and our implementation is available as \emph{Edis}
+ \emph{Hedis} respectively, and our implementation is available as \emph{Popcorn}\footnotemark
  on \emph{Hackage}.
+
+\footnotetext{Popcorn, a temporary codename for the purposes of double-blind reviewing}
 
@@
  constraints kinds\cite{constraintskinds}.
-\item Edis, a package we built for programmers. This package helps programmers
+\item Popcorn, a package we built for programmers. This package helps programmers
  to write more reliable Redis programs, and also makes Redis polymorphic by
@@
 
-We define a new datatype \mintinline{haskell}{Edis}, which is basically just
+We define a new datatype \mintinline{haskell}{Popcorn}, which is basically just
  the context \mintinline{haskell}{Redis} indexed with more information in types.
@@
 \begin{minted}[xleftmargin=1em]{haskell}
-newtype Edis p q a = Edis { unEdis :: Redis a }
+newtype Popcorn p q a = Popcorn { unPopcorn :: Redis a }
 
-instance IMonad Edis where
-    unit = Edis . return
-    bind m f = Edis (unEdis m >>= unEdis . f )
+instance IMonad Popcorn where
+    unit = Popcorn . return
+    bind m f = Popcorn (unPopcorn m >>= unPopcorn . f )
 \end{minted}
 
-The first and second argument of type \mintinline{haskell}{Edis} is where the
+The first and second argument of type \mintinline{haskell}{Popcorn} is where the
 dictionaries going to stay.
 
-To execute a \mintinline{haskell}{Edis} program, simply apply it to
- \mintinline{haskell}{unEdis} to get an ordinary program of type
+To execute a \mintinline{haskell}{Popcorn} program, simply apply it to
+ \mintinline{haskell}{unPopcorn} to get an ordinary program of type
  \mintinline{haskell}{Redis}, with type information erased.
@@
 
-Now we have \mintinline{haskell}{Edis}, let's make our own version of
+Now we have \mintinline{haskell}{Popcorn}, let's make our own version of
 \mintinline{haskell}{ping}\footnotemark
@@
 \begin{minted}[xleftmargin=1em]{haskell}
-ping :: Edis xs xs (Either Reply Status)
-ping = Edis Hedis.ping
+ping :: Popcorn xs xs (Either Reply Status)
+ping = Popcorn Hedis.ping
 \end{minted}
@@
     => Proxy s
-    -> Edis xs (Del xs s) (Either Reply Integer)
-del key = Edis $ Hedis.del (encodeKey key)
+    -> Popcorn xs (Del xs s) (Either Reply Integer)
+del key = Popcorn $ Hedis.del (encodeKey key)
 \end{minted}
@@
     -> ByteString       -- data to store
-    -> Edis xs (Set xs s Strings) (Either Reply Status)
-set key val = Edis $ Hedis.set (encodeKey key) val
+    -> Popcorn xs (Set xs s Strings) (Either Reply Status)
+set key val = Popcorn $ Hedis.set (encodeKey key) val
 \end{minted}
@@
     -> x       -- can be anything, as long as it's serializable
-    -> Edis xs (Set xs s (StringOf x)) (Either Reply Status)
-set key val = Edis $ Hedis.set (encodeKey key) (encode val)
+    -> Popcorn xs (Set xs s (StringOf x)) (Either Reply Status)
+set key val = Popcorn $ Hedis.set (encodeKey key) (encode val)
 \end{minted}
@@
 incr :: (KnownSymbol s, Get xs s ~ Just (StringOf Integer))
-     => Proxy s -> Edis xs xs (Either Reply Integer)
-incr key = Edis $ Hedis.incr (encodeKey key)
+     => Proxy s -> Popcorn xs xs (Either Reply Integer)
+incr key = Popcorn $ Hedis.incr (encodeKey key)
 
 incrbyfloat :: (KnownSymbol s, Get xs s ~ Just (StringOf Double))
-     => Proxy s -> Double -> Edis xs xs (Either Reply Double)
-incrbyfloat key n = Edis $ Hedis.incrbyfloat (encodeKey key) n
+     => Proxy s -> Double -> Popcorn xs xs (Either Reply Double)
+incrbyfloat key n = Popcorn $ Hedis.incrbyfloat (encodeKey key) n
 \end{minted}
@@
         => Proxy s
-        -> Edis xs xs (Either Reply Integer)
-llen key = Edis $ Hedis.llen (encodeKey key)
+        -> Popcorn xs xs (Either Reply Integer)
+llen key = Popcorn $ Hedis.llen (encodeKey key)
 \end{minted}
@@
         -> Proxy x  -- type of value
-        -> Edis xs (Set xs s x) ()
-declare s x = Edis $ return ()
+        -> Popcorn xs (Set xs s x) ()
+declare s x = Popcorn $ return ()
 
 renounce :: (KnownSymbol s, Member xs s ~ True)
-        => Proxy s -> Edis xs (Del xs s) ()
-renounce s = Edis $ return ()
+        => Proxy s -> Popcorn xs (Del xs s) ()
+renounce s = Popcorn $ return ()
 
 -- empty precondition
-start :: Edis ’[] ’[] ()
-start = Edis $ return ()
+start :: Popcorn ’[] ’[] ()
+start = Popcorn $ return ()
 \end{minted}
@@
     conn <- connect defaultConnectInfo
-    result <- runRedis conn $ unEdis $ start
+    result <- runRedis conn $ unPopcorn $ start
         `bind` \_ ->  declare
@@
 get :: (KnownSymbol s, Serialize x, StringOrNX xs s)
-    => Proxy s -> Edis xs xs (Either Reply (Maybe x))
+    => Proxy s -> Popcorn xs xs (Either Reply (Maybe x))
 \end{minted}
@@
      , Just (StringOf x) ~ Get xs s)
-    => Proxy s -> Edis xs xs (Either Reply (Maybe x))
+    => Proxy s -> Popcorn xs xs (Either Reply (Maybe x))
 \end{minted}
@@
 In Hedis such command could easily be expressed with lists of
- \mintinline{haskell}{ByteString}s. But in Edis, things escalate quickly, as
+ \mintinline{haskell}{ByteString}s. But in Popcorn, things escalate quickly, as
  the keys and values will have to be expressed with \emph{heterogeneous

### 2015-11-20 paper.tex
Categories and Subject Descriptors & Keywords

@@
 
-\paragraph{Categories and Subject Descriptors}
-\paragraph{Keywords}
-Haskell; type-level programming; database query language; Redis;
+\category{D.3.3}{Programming Languages}{Language Constructs and Features}
+\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Specification techniques}
+\category{H.2.3}{Database Management}{Languages}
+\keywords{Haskell; type families; type-level literals; indexed monad; Redis; database query language}
 \end{abstract}
@@
  returns a \mintinline{haskell}{Reply}, a low-level representation of replies
- from the Redis server, in case of failues. All wrapped in
+ from the Redis server, in case of failures. All wrapped in
  \mintinline{haskell}{Redis}\footnotemark, the context of command execution.
@@
  \emph{parameterised monad})\cite{indexedmonad}
- can be used\cite{typefun}\cite{staticresources} to model such preconditions and
+ can be used\cite{typefun, staticresources} to model such preconditions and
  postconditions in types. An indexed monad is a type constructor that takes
@@
 Type families have a wide variety of applications. They can appear inside type
- classes\cite{tfclass}\cite{tfsynonym}, or at toplevel. Toplevel type families
+ classes\cite{tfclass, tfsynonym}, or at toplevel. Toplevel type families
  can be used to compute over types, they come in two forms: open\cite{tfopen}
@@
 
-Since Haskell has a \emph{phase distinction}\cite{phasedistinction}, types are
+Since Haskell has a \emph{phase distinction, phasedistinction}, types are
  erased before runtime. It's impossible to obtain information directly from
  types, we can only do this indirectly, with
- \emph{singleton types}\cite{singletons}.
+ \emph{singleton types, singletons}.
 
@@
 Redis also provides \emph{transactions}, another context for executing commands.
-Redis trasanctions are atomic in the sense that, all commands in a transaction
+Redis transactions are atomic in the sense that, all commands in a transaction
  will executed sequentially, and no other requests issued by other clients will
@@
  making them statically checked.
- \emph{HaskellDB}\cite{haskelldb}\cite{haskelldbimproved} expresses quries
+ \emph{HaskellDB}\cite{haskelldb, haskelldbimproved} expresses quires
  with relational algebra-like combinators, wrapped in phantom types. Or as

### 2015-11-20 paper.tex
errata

@@
 \paragraph{The Cause} Every key is associated with a value, and every value has
- it's own type. But Most commands in Redis only work with a certain type of
+ it's own type. But most commands in Redis only work with a certain type of
  value. When a command is used on a wrong type of key, a runtime error occurs.
@@
 
-We began with making Hedis a dynamically checked embedded DSL, and implemented a
- runtime type checker that keeps track of types of each variable. But then we
+We began with making Hedis a dynamically typechecked embedded DSL, and implemented a
+ runtime type checker that keeps track of types of all the variable. But then we
  found that things can be a lot easier, by leveraging the host language's type
@@
  datatype will be automatically promoted to be a kind, and its value
- constructors to be type constructors.
+ constructors to be type constructors. The following type {\mintinline{haskell}{List}
 
@@
 
-Give rise to the following kinds and type constructors:\footnotemark\footnotemark
-
-\footnotetext{To distinguish between types and promoted constructors that have
+Give rise to the following kinds and type constructors:\footnote{To distinguish
+ between types and promoted constructors that have
  ambiguous names, prefix promoted constructor with a single quote like
  \mintinline{text}{'Nil} and \mintinline{text}{'Cons}}
-\footnotetext{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}
+\footnote{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}
 
@@
 
-Instead of letting users writing these boilerplates, we will be doing these
+Instead of letting users writing these boilerplates, we can do these
  serializations/deserializations for them. With the help from
@@
 In other words, we require that the key to be associated with a list,
- textbf{unless} it doesn't exist at all.
+ \textbf{unless} it doesn't exist at all.
 
@@
 
-With \emph{constraint kind}, a recent addition to GHC. Type constraints now has
+With \emph{constraint kind}, a recent addition to GHC, type constraints now has
  its own kind: \mintinline{haskell}{Constraint}. That means type constraints
@@
 
-At the point of writing, transactions are not supported in our implementation.
+At this point of writing, transactions are not supported in our implementation.
  We are planning to add it in the future, and we are expecting that there
- wouldn't be much difficulties, since we've implemented a runtime type checker
+ wouldn't be much difficulty, since we've implemented a runtime type checker
  specifically targeting Redis transactions once, before we moved on to the
@@
 While Redis can be viewed as a non-relational database system (although Redis
- seems reluctant to admit this in recent years), there also similar goals on
+ seems reluctant to admit this in recent years), there also are similar goals on
  relational database systems, trying to achieve a safer database interface by
@@
 \paragraph{Acknowledgements}
-This material is based upon work supported by Institute of Information Science,
-Academia Sinica under the Summer Internship Program of 2015.
+\footnote{Hidden for the purposes of double-blind reviewing}
 

### 2015-11-20 paper.tex
ready to go

@@
 
-\begin{minted}[xleftmargin=1em]{haskell}
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
 newtype Popcorn p q a = Popcorn { unPopcorn :: Redis a }
@@
 
-\begin{minted}[xleftmargin=1em]{haskell}
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
 del :: KnownSymbol s
@@
 
-\begin{minted}[xleftmargin=1em]{haskell}
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
 encodeKey :: KnownSymbol s => Proxy s -> ByteString
@@
 Most Redis commands only work with a certain type of these containers. To
- annotate what container a key is associated with, we introduce these container
- types.
+ annotate what container a key is associated with, we introduce these types for
+ the universe of containers.
 
@@
 
-incrbyfloat :: (KnownSymbol s, Get xs s ~ Just (StringOf Double))
-     => Proxy s -> Double -> Popcorn xs xs (Either Reply Double)
-incrbyfloat key n = Popcorn $ Hedis.incrbyfloat (encodeKey key) n
+incrbyfloat :: (KnownSymbol s
+    , Get xs s ~ Just (StringOf Double))
+    => Proxy s
+    -> Double
+    -> Popcorn xs xs (Either Reply Double)
+incrbyfloat key n =
+    Popcorn $ Hedis.incrbyfloat (encodeKey key) n
 \end{minted}
@@
 \begin{minted}[xleftmargin=1em]{haskell}
-llen :: (KnownSymbol s, Get xs s ~ Just (ListOf x))
+llen :: (KnownSymbol s
+      , Get xs s ~ Just (ListOf x))
      => Proxy s
      -> Popcorn xs xs (Either Reply Integer)
-llen key = Popcorn $ Hedis.llen (encodeKey key)
+llen key =
+     Popcorn $ Hedis.llen (encodeKey key)
 \end{minted}
@@
 Users may need to make assertions about the status of some key-type bindings in
- a Redis program. For example, declaring the existence of a key and the type
- of its associating value. We provide these functions, which do nothing but
- fiddling with types.
+ a Redis program. Specifically, when declaring or renouncing the existence of
+ a key and the type of its associating value. We provide these functions,
+ which do nothing but fiddle with types.
 
@@
 
--- empty precondition
+-- to be used at the beginning of a program
 start :: Popcorn ’[] ’[] ()
@@
 
-\begin{minted}[xleftmargin=1em]{haskell}
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
 main :: IO ()
@@
 
+\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
+(>>>) :: IMonad m => m p q a -> m q r b -> m p r b
+\end{minted}
 \begin{minted}[xleftmargin=1em]{haskell}
-(>>>) :: IMonad m => m p q a -> m q r b -> m p r b
-
 program = start
@@
  in two places: \emph{symbol singletons} and \emph{indexed monad}. We are hoping
- that these issues could be solved with future syntactic extensions.
+ that these issues could be resolved with future syntactic extensions.
 
@@
 All other data structures in Redis also follow the semantics similar to
- "List or nothing" as \mintinline{text}{LLEN} has. Take the case of
- \mintinline{text}{GET}, which has a similar semantics that should be typed:
+ "List or nothing". Take the case of \mintinline{text}{GET}, which also shares a
+ "String or nothing" semantics that should be typed:
 
@@
  than one value as outputs. To illustrate this, consider
- \mintinline{haskell}{sinter} in \mintinline{haskell}{Hedis}:
+ \mintinline{haskell}{sinter} in Hedis:
 
@@
 
-Notice that \textbf{not all Hedis programs can be typechecked} (even if they
- might turn out to be type safe). We opted for type safety instead of
+\textbf{Not all Redis programs can be typechecked} (even if they
+ might turn out to be type safe). We opted for type safety rather than
  expressiveness.
@@
 
-Redis also provides \emph{transactions}, another context for executing commands.
+Redis has \emph{transactions}, another context for executing commands.
 Redis transactions are atomic in the sense that, all commands in a transaction
@@
 
-We have demonstrated how a DSL with variable-bindings could be embedded in types.
+We have demonstrated how a DSL with variable bindings could be embedded in types.
 With more and more extensions added to the language, Haskell is gradually

### 2015-11-21 paper.tex
submitted

@@
 
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-newtype Popcorn p q a = Popcorn { unPopcorn :: Redis a }
+\begin{minted}[xleftmargin=1em]{haskell}
+newtype Popcorn p q a =
+    Popcorn { unPopcorn :: Redis a }
 
@@
     unit = Popcorn . return
-    bind m f = Popcorn (unPopcorn m >>= unPopcorn . f )
+    bind m f =
+        Popcorn (unPopcorn m >>= unPopcorn . f )
 \end{minted}
@@
 The first and second argument of type \mintinline{haskell}{Popcorn} is where the
-dictionaries going to stay.
+dictionaries going to be.
 
 To execute a \mintinline{haskell}{Popcorn} program, simply apply it to
- \mintinline{haskell}{unPopcorn} to get an ordinary program of type
- \mintinline{haskell}{Redis}, with type information erased.
+ \mintinline{haskell}{unPopcorn} to get an ordinary Hedis program back,
+ with type information erased.
 
@@
  \mintinline{text}{PONG} if the connection is alive. In Hedis,
- \mintinline{text}{PING} has type:
+ \mintinline{haskell}{ping} has type:
 
@@
 
-Now we have \mintinline{haskell}{Popcorn}, let's make our own version of
+Now with \mintinline{haskell}{Popcorn}, we could make our own version of
 \mintinline{haskell}{ping}\footnotemark
@@
 The dictionary \mintinline{haskell}{xs} in the type remains unaffected after the
- action, because \mintinline{haskell}{ping} does not affect any key-value
- bindings. To encode other commands that modifies key-value bindings, we need
+ action, because \mintinline{haskell}{ping} does not affect any key-type
+ bindings. To encode other commands that modifies key-type bindings, we need
  type-level functions to annotate those effects on the dictionary.
@@
  not only container-specific, they also have some requirements on what types of
- value they could operate on.
+ value they could operate with.
 
@@
 
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-incr :: (KnownSymbol s, Get xs s ~ Just (StringOf Integer))
-     => Proxy s -> Popcorn xs xs (Either Reply Integer)
+\begin{minted}[xleftmargin=1em]{haskell}
+incr :: (KnownSymbol s
+      , Get xs s ~ Just (StringOf Integer))
+     => Proxy s
+     -> Popcorn xs xs (Either Reply Integer)
 incr key = Popcorn $ Hedis.incr (encodeKey key)

### 2016-06-11 aplas16/main-aplas.tex
Created a directory for aplas16.

--- 2015-11-20 paper.tex
@@
-%
-% LaTeX template for preparation of submissions to PLDI'16
-%
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'16 web site.
-%
-\documentclass[pldi]{sigplanconf-pldi16}
-% \documentclass[pldi-cameraready]{sigplanconf-pldi16}
+\documentclass{llncs}
 
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet}     % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                % format URLs
-\usepackage{listings}           % format code
-\usepackage{enumitem}           % adjust spacing in enums
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
+\usepackage{amsmath}
+%\usepackage{amsfonts}
+%\usepackage{amsthm}
+\usepackage{url}
+\usepackage{mathptmx}
+\usepackage{doubleequals}
+\usepackage{makeidx}  % allows for indexgeneration
 
-% code highlighting
-\usepackage{minted}
-\usemintedstyle{tango}
-
-% no red boxes on parser error:
-\AtBeginEnvironment{minted}{%
-  \renewcommand{\fcolorbox}[4][]{#4}}
-
-% quotes
-\usepackage{dirtytalk}
-
-% for symbols
-\usepackage[T1]{fontenc}
-\usepackage{lmodern}
-\usepackage{upquote}
-
-% use arbitrary fonts
-\usepackage{fontspec}
-\setmonofont{Inconsolata}
-
-% for clickable footnotes
-\usepackage{hyperref}
-
-\newcommand{\todo}[2]{{\bf [#1]}: #2}
+\input{Preamble}
+\input{macro}
 
@@
 \title{Preventing Runtime Errors of Redis at Compile Time}
-
-%
-% any author declaration will be ignored  when using 'pldi' option (for double blind review)
-%
-
-\authorinfo{Ting-Yan Lai}
-{\makebox{Institute of Computer Science and Engineering} \\
-\makebox{National Chiao Tung University, Taiwan}}
-{banacorn.cs03g@nctu.edu.tw}
-
-\authorinfo{Tyng-Ruey Chuang \and Shin-Cheng Mu}
-{\makebox{Institute of Information Science} \\
-\makebox{Academia Sinica, Taiwan}}
-{\{trc,scm\}@iis.sinica.edu.tw}
+\author{Ting-Yan Lai \inst{1}\inst{2}\and Tyng-Ruey Chuang \inst{1}
+  \and Shin-Cheng Mu \inst{1}}
+\institute{Institute of Information Science, Academia Sinica, Taiwan
+\and Institute of Computer Science and Engineering,
+National Chiao Tung University, Taiwan}
 
@@
 The database system and the package we are targeting are \emph{Redis} and
- \emph{Hedis} respectively, and our implementation is available as \emph{Popcorn}\footnotemark
+ \emph{Hedis} respectively, and our implementation is available as \emph{Popcorn}
  on \emph{Hackage}.
-
-\footnotetext{Popcorn, a temporary codename for the purposes of double-blind reviewing}
-
-\category{D.3.3}{Programming Languages}{Language Constructs and Features}
-\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Specification techniques}
-\category{H.2.3}{Database Management}{Languages}
-\keywords{Haskell; type families; type-level literals; indexed monad; Redis; database query language}
 \end{abstract}
 
-\section{Introduction}
+\input{sections/Introduction}
+\input{sections/Motivation}
+\input{sections/TypeLevelDict}
+\input{sections/IndexedMonads}
+\input{sections/TypeLevelFun}
+\input{sections/PolyRedis}
+\input{sections/Constraints}
+\input{sections/Assertions}
+\input{sections/Discussions}
+\input{sections/Related}
+\input{sections/Conclusions}
 
-\subsection{Redis}
-
-Redis\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as
-database, cache and message broker.
-A Redis datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with.
-Redis supports many different kind of values, such as strings, hashes, lists, and sets, etc, and provides a collection of of atomic \emph{commands} to manipulate these values.
-
-For an example, consider the following sequence of commands, entered through the interactive interface of Redis. The keys
-\mintinline{text}{some-set} abd \mintinline{text}{another-set} are
-both associated to a set. The two call to command \mintinline{text}{SADD} respectively adds three and two values to the two sets, before \mintinline{text}{SINTER} takes their intersection:
-\begin{minted}[xleftmargin=1em]{text}
-redis> SADD some-set a b c
-(integer) 3
-redis> SADD another-set a b
-(integer) 2
-redis> SINTER some-set another-set
-1) "a"
-2) "b"
-\end{minted}
-
-Note that the keys \mintinline{text}{some-set} and \mintinline{text}{another-set}, if not existing before the call to \mintinline{text}{SADD}, are created on site. The call to
-\mintinline{text}{SADD} returns the size of the set after
-completion of the command.
-
-\subsection{Hedis}
-
-Many third party library exist to allow general purpose programmings
-to access Redis databases through its TCP protocol. The most popular
-such library for Haskell is Hedis\footnote{\url{https://hackage.haskell.org/package/hedis}}.
-
-The following program is how the previous example looks like in
-Hedis:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-program :: Redis (Either Reply [ByteString])
-program = do
-    sadd "some-set" ["a", "b"]
-    sadd "another-set" ["a", "b", "c"]
-    sinter ["some-set", "another-set"]
-\end{minted}
-
-The function \mintinline{haskell}{sadd} takes a key and a list of values as
- arguments, and returns an \mintinline{haskell}{Integer} on success, or
- returns a \mintinline{haskell}{Reply}, a low-level representation of replies
- from the Redis server, in case of failures. All wrapped in
- \mintinline{haskell}{Redis}\footnotemark, the context of command execution.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-sadd :: ByteString      -- key
-    -> [ByteString]     -- values
-    -> Redis (Either Reply Integer)
-\end{minted}
-
-Note that keys and values, being nothing but binary strings in Redis,
-are represented using Haskell \mintinline{haskell}{ByteString}.
-Values of other types must be encoded as \mintinline{haskell}{ByteString}s before being written to the database,
-and decoded after being read back.
-
-\footnotetext{Hedis provides another kind of context, \mintinline{haskell}{RedisTx}, for \emph{transactions}, united with \mintinline{haskell}{Redis} under the class of \mintinline{haskell}{RedisCtx}.
-For brevity, we demonstrate only \mintinline{haskell}{Redis} in this paper.
-}
-
-\section{Motivation}
-
-All binary strings are equal, but some binary strings are more equal than others.
-
-Although everything in Redis is essentially a binary string, these strings are treated differently. Redis supports many different
-kind of data structures, such as strings, hashes, lists, etc. While
-they are all encoded as binary strings before being written to the databse, most commands, much like how the C language treats a piece of data, only works with data of certain types.
-
-\paragraph{Problem 1} The command \mintinline{text}{SET}, by
-its definition, associates a key to a string. In the following
-example, the key \mintinline{text}{some-string} is associated
-to string \mintinline{text}{foo}. Subsequent calls to
-\mintinline{text}{SADD} causes runtime errors, since the value
-of \mintinline{text}{some-string} is not a set, but a string.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> SET some-string foo
-OK
-redis> SADD some-string bar
-(error) WRONGTYPE Operation against a key
- holding the wrong kind of value
-\end{minted}
-
-\paragraph{Problem 2} Even worse, not all strings are equal!
-The call \mintinline{text}{INCR some-string} parses the string
-associated with key \mintinline{text}{some-string}
-to an integer, increments it by one, and store it back as a string.
-If the string can not be parse as an integer, a runtime error
-is raised.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> SET some-string foo
-OK
-redis> INCR some-string
-(error) ERR value is not an integer or out
- of range
-\end{minted}
-
-\paragraph{In Hedis} Hedis, being only a simple wrapper on top
-of the TCP protocol of Redis, inherits all the problems mentioned
-above. The following program yields the same error as that in
-the Redis client.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-program :: Redis (Either Reply Integer)
-program = do
-    set "some-string" "foo"
-    sadd "some-string" ["a"]
-\end{minted}
-\begin{minted}[xleftmargin=1em]{haskell}
-Left (Error "WRONGTYPE Operation against a
- key holding the wrong kind of value")
-\end{minted}
-
-\paragraph{The Cause} Every key is associated with a value, and every value has
- it's own type. But most commands in Redis only work with a certain type of
- value. When a command is used on a wrong type of key, a runtime error occurs.
- The problems illustrated above arise from the absence of type checking, with
- respects to \textbf{the type of a value that associates with a key}.
- These problems could have been avoided, if we could know the type every key
- associates with in advance, and prevent programs with invalid commands from
- executing.
-
-\subsection{Hedis as an embedded DSL}
-
-Haskell makes it easy to build and use Domain Specific Languages (DSLs),
- and Hedis can be regarded as one of them. What makes Hedis peculiar is that,
- it has \emph{variable bindings} (between keys and values), but with very
- little or no semantic checking, neither dynamically nor statically.
-
-We began with making Hedis a dynamically typechecked embedded DSL, and implemented a
- runtime type checker that keeps track of types of all the variable. But then we
- found that things can be a lot easier, by leveraging the host language's type
- checker. We encode variable bindings with \emph{type-level lists} and
- \emph{strings}, and control the effects on the bindings with
- \emph{indexed monad}. In contrast to the former approach, we \textbf{embedded}
- our type checker into Haskell's type system, without having to build a
- \textbf{standalone} one on the term level.
-
-\subsection{Contributions}
-
-To summarize our contributions:
-
-\begin{itemize}[noitemsep]
-\item We make Hedis statically type-checked, without runtime overhead.
-\item We demonstrates how to model variable bindings of an embedded DSL with
- language extensions like type-level literals and data kinds.
-\item We provide (yet another) an example of encoding effects and constraints of
- an action in types, with indexed monad\cite{indexedmonad} and other language
- extensions such as closed type-families\cite{closedtypefamilies} and
- constraints kinds\cite{constraintskinds}.
-\item Popcorn, a package we built for programmers. This package helps programmers
- to write more reliable Redis programs, and also makes Redis polymorphic by
- automatically converting back and forth from values of arbitrary types and
- boring ByteStrings.
-\end{itemize}
-
-\section{Type-level dictionaries}
-
-To check the bindings between keys and values, we need a \emph{dictionary-like}
- structure, and encode it as a \emph{type} somehow.
-
-\subsection{Datatype promotion}
-Normally, at the term level, we could express the datatype of dictionary with
-\emph{type synonym} like this.\footnotemark
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type Key = String
-type Dictionary = [(Key, TypeRep)]
-\end{minted}
-
-\footnotetext{\mintinline{haskell}{TypeRep} supports term-level representations
- of datatypes, available in \mintinline{haskell}{Data.Typeable}}
-
-To encode this in the type level, everything has to be
- \emph{promoted}\cite{promotion} one level up.
- From terms to types, and from types to kinds.
-
-Luckily, with recently added GHC extension \emph{data kinds}, suitable
- datatype will be automatically promoted to be a kind, and its value
- constructors to be type constructors. The following type {\mintinline{haskell}{List}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data List a = Nil | Cons a (List a)
-\end{minted}
-
-Give rise to the following kinds and type constructors:\footnote{To distinguish
- between types and promoted constructors that have
- ambiguous names, prefix promoted constructor with a single quote like
- \mintinline{text}{'Nil} and \mintinline{text}{'Cons}}
-\footnote{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-List k :: BOX
-Nil  :: List k
-Cons :: k -> List k -> List k
-\end{minted}
-
-Haskell sugars lists \mintinline{haskell}{[1, 2, 3]} and tuples
- \mintinline{text}{(1, 'a')} with brackets and parentheses.
- We could also express promoted lists and tuples in types like this with
- a single quote prefixed. For example:
- \mintinline{text}{'}\mintinline{haskell}{[Int, Char]},
- \mintinline{text}{'}\mintinline{haskell}{(Int, Char)}.
-
-\subsection{Type-level literals}
-
-Now we have type-level lists and tuples to construct the dictionary.
-For keys, \mintinline{haskell}{String} also has a type-level correspondence:
-\mintinline{haskell}{Symbol}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data Symbol
-\end{minted}
-
-Symbol is defined without a value constructor, because it's intended to be used
- as a promoted kind.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-"this is a type-level string literal" :: Symbol
-\end{minted}
-%
-% Nonetheless, it's still useful to have a term-level value that links with a
-%  Symbol, when we want to retrieve type-level information at runtime (but not the
-%  other way around!).
-
-\subsection{Putting everything together}
-
-With all of these ingredients ready, let's build some dictionaries!
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type DictEmpty = '[]
-type Dict0 = '[ '("key", Bool) ]
-type Dict1 = '[ '("A", Int), '("B", "A") ]
-\end{minted}
-
-These dictionaries are defined with \emph{type synonym}, since they are
- \emph{types}, not \emph{terms}. If we ask \mintinline{text}{GHCi} what is the
- kind of \mintinline{haskell}{Dict1}, we will get
- \mintinline{haskell}{Dict1 :: [ (Symbol, *) ]}
-
-The kind \mintinline{haskell}{*} (pronounced "star") stands for the set of all
- concrete type expressions, such as \mintinline{haskell}{Int},
- \mintinline{haskell}{Char} or even a symbol \mintinline{haskell}{"symbol"},
- while \mintinline{haskell}{Symbol} is restricted to all symbols only.
-
-\section{Indexed monads}
-
-Redis commands are \emph{actions}.
- We could capture the effects caused by an action, by expressing the states it
- affects, before and after. That is, the \emph{preconditions} and
- \emph{postconditions} of an action. In such way, we could also impose
- constraints on the preconditions.
-
-\emph{Indexed monads} (or \emph{monadish},
- \emph{parameterised monad})\cite{indexedmonad}
- can be used\cite{typefun, staticresources} to model such preconditions and
- postconditions in types. An indexed monad is a type constructor that takes
- three arguments: an initial state, a final state, and the type of a value that
- an action computes, which can be read like a Hoare triple\cite{kleisli}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-class IMonad m where
-    unit :: a -> m p p a
-    bind :: m p q a -> (a -> m q r b) -> m p r b
-\end{minted}
-
-Class \mintinline{haskell}{IMonad} comes with two operations:
- \mintinline{haskell}{unit} for identities and \mintinline{haskell}{bind} for
- compositions, as in monads.
-
-We define a new datatype \mintinline{haskell}{Popcorn}, which is basically just
- the context \mintinline{haskell}{Redis} indexed with more information in types.
- We make it an instance of \mintinline{haskell}{IMonad}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-newtype Popcorn p q a = Popcorn { unPopcorn :: Redis a }
-
-instance IMonad Popcorn where
-    unit = Popcorn . return
-    bind m f = Popcorn (unPopcorn m >>= unPopcorn . f )
-\end{minted}
-
-The first and second argument of type \mintinline{haskell}{Popcorn} is where the
-dictionaries going to stay.
-
-To execute a \mintinline{haskell}{Popcorn} program, simply apply it to
- \mintinline{haskell}{unPopcorn} to get an ordinary program of type
- \mintinline{haskell}{Redis}, with type information erased.
-
-\subsection{\mintinline{text}{PING}: the first attempt}
-
-In Redis, \mintinline{text}{PING} does nothing but replies with
- \mintinline{text}{PONG} if the connection is alive. In Hedis,
- \mintinline{text}{PING} has type:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-ping :: Redis (Either Reply Status)
-\end{minted}
-
-Now we have \mintinline{haskell}{Popcorn}, let's make our own version of
-\mintinline{haskell}{ping}\footnotemark
-
-\footnotetext{\mintinline{haskell}{ping} from Hedis is qualified with
-\mintinline{haskell}{Hedis} to prevent function name clashing in our code.}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-ping :: Popcorn xs xs (Either Reply Status)
-ping = Popcorn Hedis.ping
-\end{minted}
-
-The dictionary \mintinline{haskell}{xs} in the type remains unaffected after the
- action, because \mintinline{haskell}{ping} does not affect any key-value
- bindings. To encode other commands that modifies key-value bindings, we need
- type-level functions to annotate those effects on the dictionary.
-
-\section{Type-level functions}
-\subsection{Closed type families}
-
-Type families have a wide variety of applications. They can appear inside type
- classes\cite{tfclass, tfsynonym}, or at toplevel. Toplevel type families
- can be used to compute over types, they come in two forms: open\cite{tfopen}
- and closed \cite{tfclosed}.
-
-We choose \emph{closed type families}, because it allows overlapping instances,
- and we need none of the extensibility provided by open type families.
-For example, consider both term-level and type-level \mintinline{haskell}{&&}:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-(&&) :: Bool -> Bool -> Bool
-True && True = True
-a    && b    = False
-
-type family And (a :: Bool) (b :: Bool) :: Bool where
-    And True True = True
-    And a    b    = False
-\end{minted}
-
-The first instance of \mintinline{haskell}{And} could be subsumed under a more
- general instance, \mintinline{haskell}{And a b}.
-But the closedness allows these instances to be resolved in order, just like
- how cases are resolved in term-level functions. Also notice that how much
- \mintinline{haskell}{And} resembles to it's term-level sibling.
-
-\subsection{Functions on type-level dictionaries}
-
-With closed type families, we could define functions on the type level.
- Let's begin with dictionary lookup.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type family Get
-    (xs :: [(Symbol, *)])   -- dictionary
-    (s :: Symbol)           -- key
-    :: * where              -- type
-
-    Get (’(s, x) ’: xs) s = x
-    Get (’(t, x) ’: xs) s = Get xs s
-\end{minted}
-
-Another benefit of closed type families is that type-level equality can be
-expressed by unifying type variables with the same name.
-\mintinline{haskell}{Get} takes two type arguments, a dictionary and a symbol.
-If the key we are looking for unifies with the symbol of an entry, then
- \mintinline{haskell}{Get} returns the corresponding type, else it keeps
- searching down the rest of the dictionary.
-
-\mintinline{haskell}{Get }\mintinline{text}{ '}\mintinline{haskell}{[ }
-\mintinline{text}{'}\mintinline{haskell}{("A", Int) ] "A"} evaluates to
-\mintinline{haskell}{Int}.
-
-But \mintinline{haskell}{Get }
-\mintinline{text}{'}\mintinline{haskell}{[ }
-\mintinline{text}{'}\mintinline{haskell}{("A", Int) ] "B"} would get stuck.
-That's because \mintinline{haskell}{Get} is a partial function on types,
- and these types are computed at compile-time. It wouldn't make
- much sense for a type checker to crash and throw a "Non-exhaustive" error or
- be non-terminating.
-
-We could make \mintinline{haskell}{Get} total, as we would at the term level,
- with \mintinline{haskell}{Maybe}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type family Get
-    (xs :: [(Symbol, *)])   -- dictionary
-    (s :: Symbol)           -- key
-    :: Maybe * where        -- type
-
-    Get ’[]             s = Nothing
-    Get (’(s, x) ’: xs) s = Just x
-    Get (’(t, x) ’: xs) s = Get xs s
-\end{minted}
-%
-Other dictionary-related functions are defined in a similar fashion.
-
-\begin{minted}[xleftmargin=1em]{haskell}
--- inserts or updates an entry
-type family Set
-    (xs :: [(Symbol, *)])   -- old dictionary
-    (s :: Symbol)           -- key
-    (x :: *)                -- type
-    :: [(Symbol, *)] where  -- new dictionary
-
-    Set ’[]             s x = ’[ ’(s, x) ]
-    Set (’(s, y) ’: xs) s x = (’(s, x) ’: xs)
-    Set (’(t, y) ’: xs) s x =
-        ’(t, y) ’: (Set xs s x)
-
--- removes an entry
-type family Del
-    (xs :: [(Symbol, *)])   -- old dictionary
-    (s :: Symbol)           -- key
-    :: [(Symbol, *)] where  -- new dictionary
-
-    Del ’[] s             = ’[]
-    Del (’(s, y) ’: xs) s = xs
-    Del (’(t, y) ’: xs) s = ’(t, y) ’: (Del xs s)
-
--- membership
-type family Member
-    (xs :: [(Symbol, *)])   -- dictionary
-    (s :: Symbol)           -- key
-    :: Bool where           -- exists?
-
-    Member ’[]             s = False
-    Member (’(s, x) ’: xs) s = True
-    Member (’(t, x) ’: xs) s = Member xs s
-\end{minted}
-
-\subsection{Proxies and singleton types}
-
-Now we could annotate the effects of a command in types. \mintinline{text}{DEL}
- removes a key from the current database, regardless of its type.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-del :: KnownSymbol s
-    => Proxy s
-    -> Popcorn xs (Del xs s) (Either Reply Integer)
-del key = Popcorn $ Hedis.del (encodeKey key)
-\end{minted}
-
-\mintinline{haskell}{KnownSymbol} is a class that gives the string associated
- with a concrete type-level symbol, which can be retrieved with
- \mintinline{haskell}{symbolVal}.\footnotemark
- Where \mintinline{haskell}{encodeKey} converts \mintinline{haskell}{Proxy s} to
- \mintinline{haskell}{ByteString}.
-\footnotetext{They are defined in \mintinline{haskell}{GHC.TypeLits}.}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-encodeKey :: KnownSymbol s => Proxy s -> ByteString
-encodeKey = encode . symbolVal
-\end{minted}
-
-Since Haskell has a \emph{phase distinction, phasedistinction}, types are
- erased before runtime. It's impossible to obtain information directly from
- types, we can only do this indirectly, with
- \emph{singleton types, singletons}.
-
-A singleton type is a type that has only one instance, and the instance can be
- think of as the representative of the type at the realm of runtime values.
-
-\mintinline{haskell}{Proxy}, as its name would suggest, can be used as
- singletons. It's a phantom type that could be indexed with any type.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data Proxy t = Proxy
-\end{minted}
-
-In the type of \mintinline{haskell}{del}, the type variable
- \mintinline{haskell}{s} is a \mintinline{haskell}{Symbol} that is decided by
- the argument of type \mintinline{haskell}{Proxy s}.
- To use \mintinline{haskell}{del}, we would have to apply it with a clumsy
- term-level proxy like this:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-del (Proxy :: Proxy "A")
-\end{minted}
-
-\section{Making Redis polymorphic}
-
-Redis supports many different datatypes, these datatypes as can be viewed as
- \emph{containers} of strings. For example, lists (of strings),
- sets (of strings), and strings themselves.
-
-\subsection{Denoting containers}
-Most Redis commands only work with a certain type of these containers. To
- annotate what container a key is associated with, we introduce these container
- types.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data Strings
-data Lists
-data Sets
-...
-\end{minted}
-
-\mintinline{text}{SET} stores a string, regardless the datatype the key was
- associated with. Now we could implement \mintinline{text}{SET} like this:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-set :: KnownSymbol s
-    => Proxy s
-    -> ByteString       -- data to store
-    -> Popcorn xs (Set xs s Strings) (Either Reply Status)
-set key val = Popcorn $ Hedis.set (encodeKey key) val
-\end{minted}
-
-After \mintinline{text}{SET}, the key will be associated with
- \mintinline{haskell}{Strings} in the dictionary, indicating that it's a string.
-
-\subsection{Automatic data serialization}
-But in the real world, raw binary strings are hardly useful, people would
- usually serialize their data into strings before storing them, and deserialize
- them back when in need.
-
-Instead of letting users writing these boilerplates, we can do these
- serializations/deserializations for them. With the help from
- \mintinline{text}{cereal}, a binary serialization library.
- \mintinline{text}{cereal} comes with these two functions:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-encode :: Serialize a
-       => a -> ByteString Source
-
-decode :: Serialize a
-       => ByteString -> Either String a
-\end{minted}
-
-Which would do all the works for us, as long as the datatype it's handling is
- an instance of class \mintinline{haskell}{Serialize}.\footnotemark
-
-\footnotetext{The methods of \mintinline{haskell}{Serialize} will have default
- generic implementations for all datatypes with some language extensions
- enabled, no sweat!}
-
-\subsection{Extending container types}
-We rename container types and extend it with an extra type argument,
- to indicate what kind of encoded value it's holding.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data StringOf x
-data ListOf x
-data SetOf x
-...
-\end{minted}
-
-\mintinline{haskell}{set} reimplemented with extended container types:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-set :: (KnownSymbol s, Serialize x)
-    => Proxy s
-    -> x       -- can be anything, as long as it's serializable
-    -> Popcorn xs (Set xs s (StringOf x)) (Either Reply Status)
-set key val = Popcorn $ Hedis.set (encodeKey key) (encode val)
-\end{minted}
-
-For example, if we execute \mintinline{haskell}{set (Proxy :: Proxy "A") True},
- a new entry \mintinline{text}{'}\mintinline{haskell}{("A", StringOf Bool)} will
- be inserted to the dictionary.
-
-\subsection{Handling \mintinline{text}{INCR}}
-
-Commands such as \mintinline{text}{INCR} and \mintinline{text}{INCRBYFLOAT}, are
- not only container-specific, they also have some requirements on what types of
- value they could operate on.
-
-We could handle this by mapping Redis's strings of integers and floats to
- Haskell's \mintinline{text}{Integer} and \mintinline{text}{Double}.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-incr :: (KnownSymbol s, Get xs s ~ Just (StringOf Integer))
-     => Proxy s -> Popcorn xs xs (Either Reply Integer)
-incr key = Popcorn $ Hedis.incr (encodeKey key)
-
-incrbyfloat :: (KnownSymbol s, Get xs s ~ Just (StringOf Double))
-     => Proxy s -> Double -> Popcorn xs xs (Either Reply Double)
-incrbyfloat key n = Popcorn $ Hedis.incrbyfloat (encodeKey key) n
-\end{minted}
-
-\section{Imposing constraints}
-
-To rule out programs with undesired properties, certain constraints must be
- imposed, on what arguments they can take, or what preconditions they must hold.
-
-Consider the following example: \mintinline{text}{LLEN} returns the length of
- the list associated with a key, else raises a type error.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> LPUSH some-list bar
-(integer) 1
-redis> LLEN some-list
-(integer) 1
-redis> SET some-string foo
-OK
-redis> LLEN some-string
-(error) WRONGTYPE Operation against a key
-holding the wrong kind of value
-\end{minted}
-
-Such constraint could be expressed in types with
-\emph{equality constraints}\cite{typeeq}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-llen :: (KnownSymbol s, Get xs s ~ Just (ListOf x))
-        => Proxy s
-        -> Popcorn xs xs (Either Reply Integer)
-llen key = Popcorn $ Hedis.llen (encodeKey key)
-\end{minted}
-
-Where \mintinline{haskell}{(~)} denotes that \mintinline{haskell}{Get xs s}
-and \mintinline{haskell}{Just (ListOf x)} needs to be the same.
-
-The semantics of \mintinline{text}{LLEN} defined above is actually not
-complete. \mintinline{text}{LLEN} also accepts keys that do not exist, and
- replies with \mintinline{text}{0}.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> LLEN nonexistent
-(integer) 0
-\end{minted}
-
-In other words, we require that the key to be associated with a list,
- \textbf{unless} it doesn't exist at all.
-
-\subsection{Expressing constraint disjunctions}
-
-Unfortunately, expressing disjunctions in constraints is much more difficult
- than expressing conjunctions, since the latter could be easily done by placing
- constraints in a tuple (at the left side of \mintinline{haskell}{=>}).
-
-There are at least three ways to express type-level constraints
-\cite{singletons}. Luckily we could express constraint disjunctions with type
- families in a modular way.
-
-The semantics we want could be expressed informally like this:
-
-\mintinline{haskell}{Get xs s} $\equiv$ \mintinline{haskell}{Just (ListOf x)}
-$\vee$ $\neg$ \mintinline{haskell}{(Member xs s)}
-
-We could achieve this simply by translating the semantics we want to the
- domain of Boolean, with type-level boolean functions such as
-\mintinline{haskell}{(&&)},
-\mintinline{haskell}{(||)}, \mintinline{haskell}{Not},
-\mintinline{haskell}{(==)}, etc.\footnotemark To avoid
-
-\footnotetext{Available in \mintinline{text}{Data.Type.Bool} and
- \mintinline{text}{Data.Type.Equality}}
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-Get xs s == Just (ListOf x) || Not (Member xs s)
-\end{minted}
-
-To avoid addressing the type of value (as it may not exist at all), we defined
- an auxiliary predicate \mintinline{haskell}{IsList :: Maybe * -> Bool} to
- replace the former part.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-IsList (Get xs s) || Not (Member xs s)
-\end{minted}
-
-The type expression above has kind \mintinline{haskell}{Bool}, we could make it
- a type constraint by asserting equality.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-(IsList (Get xs s) || Not (Member xs s)) ~ True
-\end{minted}
-
-With \emph{constraint kind}, a recent addition to GHC, type constraints now has
- its own kind: \mintinline{haskell}{Constraint}. That means type constraints
- are not restricted to the left side of a \mintinline{haskell}{=>} anymore,
- they could appear in anywhere that accepts something of kind
- \mintinline{haskell}{Constraint}, and any type that has kind
- \mintinline{haskell}{Constraint} can also be used as a type constraint.
- \footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}
-
-As many other list-related commands also have this "List or nothing" semantics,
- we could abstract the lengthy type constraint above and give it an alias with
- type synonym.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-ListOrNX xs s =
-    (IsList (Get xs s) || Not (Member xs s)) ~ True
-\end{minted}
-
-The complete implementation of \mintinline{text}{LLEN} with
-\mintinline{haskell}{ListOrNX} would become:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-llen :: (KnownSymbol s, ListOrNX xs s)
-        => Proxy s
-        -> Popcorn xs xs (Either Reply Integer)
-llen key = Popcorn $ Hedis.llen (encodeKey key)
-\end{minted}
-
-\section{Assertions}
-
-Users may need to make assertions about the status of some key-type bindings in
- a Redis program. For example, declaring the existence of a key and the type
- of its associating value. We provide these functions, which do nothing but
- fiddling with types.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-declare :: (KnownSymbol s, Member xs s ~ False)
-        => Proxy s
-        -> Proxy x  -- type of value
-        -> Popcorn xs (Set xs s x) ()
-declare s x = Popcorn $ return ()
-
-renounce :: (KnownSymbol s, Member xs s ~ True)
-        => Proxy s -> Popcorn xs (Del xs s) ()
-renounce s = Popcorn $ return ()
-
--- empty precondition
-start :: Popcorn ’[] ’[] ()
-start = Popcorn $ return ()
-\end{minted}
-
-\subsection{A complete example}
-
-The following program increases the value of "A" as an integer, push the result
- of the increment to list "L", and then pops it out.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-main :: IO ()
-main = do
-    conn <- connect defaultConnectInfo
-    result <- runRedis conn $ unPopcorn $ start
-        `bind` \_ ->  declare
-                        (Proxy :: Proxy "A")
-                        (Proxy :: Proxy Integer)
-        `bind` \_ ->  incr     (Proxy :: Proxy "A")
-        `bind` \n ->  case n of
-            Left  err -> lpush (Proxy :: Proxy "L") 0
-            Right n   -> lpush (Proxy :: Proxy "L") n
-        `bind` \_ ->  lpop     (Proxy :: Proxy "L")
-    print result
-\end{minted}
-
-The syntax is pretty heavy, like the old days when there's no
- \emph{do-notation}\cite{history}. But if we don't need any variable bindings
- between operations, we could compose these commands with a sequencing operator
- \mintinline{haskell}{(>>>)}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-(>>>) :: IMonad m => m p q a -> m q r b -> m p r b
-
-program = start
-    >>> declare
-            (Proxy :: Proxy "A")
-            (Proxy :: Proxy Integer)
-    >>> incr    (Proxy :: Proxy "A")
-    >>> lpush   (Proxy :: Proxy "L") 0
-    >>> lpop    (Proxy :: Proxy "L")
-\end{minted}
-
-\section{Discussions}
-\paragraph{Syntax}
-No one could ignore the glaring shortcoming of the syntax, which occurs mainly
- in two places: \emph{symbol singletons} and \emph{indexed monad}. We are hoping
- that these issues could be solved with future syntactic extensions.
-
-\paragraph{Returns only determined datatypes}
-
-All other data structures in Redis also follow the semantics similar to
- "List or nothing" as \mintinline{text}{LLEN} has. Take the case of
- \mintinline{text}{GET}, which has a similar semantics that should be typed:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-get :: (KnownSymbol s, Serialize x, StringOrNX xs s)
-    => Proxy s -> Popcorn xs xs (Either Reply (Maybe x))
-\end{minted}
-
-Since the key may not exist, we don't know what \mintinline{haskell}{x} would
-be. We could left \mintinline{haskell}{x} ambiguous, and let it be decided by
- the caller. But users will then be forced to spell out the complete
-type signature of everything, including the dictionaries, only to specify
-the desired resulting type.
-
-Instead of allowing the key to be non-existent, we require that the key must
- exist and it's associating type to be determined at compile time. So our
- version of \mintinline{haskell}{get} has a stricter semantics:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-get :: (KnownSymbol s, Serialize x
-     , Just (StringOf x) ~ Get xs s)
-    => Proxy s -> Popcorn xs xs (Either Reply (Maybe x))
-\end{minted}
-
-\paragraph{Commands with multiple inputs or outputs}
-
-Some command may take a variable number of arguments as inputs, and returns more
- than one value as outputs. To illustrate this, consider
- \mintinline{haskell}{sinter} in \mintinline{haskell}{Hedis}:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-Hedis.sinter :: [ByteString]                      -- keys
-             -> Redis (Either Reply [ByteString]) -- values
-\end{minted}
-
-In Hedis such command could easily be expressed with lists of
- \mintinline{haskell}{ByteString}s. But in Popcorn, things escalate quickly, as
- the keys and values will have to be expressed with \emph{heterogeneous
- lists}\cite{hetero}, which would be pratically infeasible, considering the cost,
- if not impossible.
-
-Most importantly, the keys will all have to be constrained by
- \mintinline{haskell}{KnownSymbol}, which enforces these type literals to be
- concrete and known at compile time.
- It's still unclear whether this is possible.
-
-So instead, we are offering commands that only has a single input and output.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-sinter :: ByteString                      -- single key
-       -> Redis (Either Reply ByteString) -- single value
-\end{minted}
-
-Notice that \textbf{not all Hedis programs can be typechecked} (even if they
- might turn out to be type safe). We opted for type safety instead of
- expressiveness.
-
-\paragraph{Redis Transactions}
-
-Redis also provides \emph{transactions}, another context for executing commands.
-Redis transactions are atomic in the sense that, all commands in a transaction
- will executed sequentially, and no other requests issued by other clients will
- be served \textbf{in the middle}.\footnotemark In contrast, we cannot make such
- a guarantee in the ordinary context, which may destroy the assertions we made
- in types.
-
-At this point of writing, transactions are not supported in our implementation.
- We are planning to add it in the future, and we are expecting that there
- wouldn't be much difficulty, since we've implemented a runtime type checker
- specifically targeting Redis transactions once, before we moved on to the
- types.
-
-\section{Related Work}
-
-While Redis can be viewed as a non-relational database system (although Redis
- seems reluctant to admit this in recent years), there also are similar goals on
- relational database systems, trying to achieve a safer database interface by
- making them statically checked.
- \emph{HaskellDB}\cite{haskelldb, haskelldbimproved} expresses quires
- with relational algebra-like combinators, wrapped in phantom types. Or as
- examples, at first to demonstrate the power of dependent types in
- Agda\cite{pi}, and then with singletons in Haskell\cite{singletons}.
-
-\section{Conclusions}
-
-We have demonstrated how a DSL with variable-bindings could be embedded in types.
-With more and more extensions added to the language, Haskell is gradually
- becoming a dependently-typed language,\footnote{Why not be dependently typed? \url{http://stackoverflow.com/questions/12961651/why-not-be-dependently-typed}}
- but it's not that dreadful as many (including us) would have thought.
-
-\paragraph{Acknowledgements}
-\footnote{Hidden for the purposes of double-blind reviewing}
-
-% references
-\bibliographystyle{abbrvnat}
+\bibliographystyle{splncs03}
 \bibliography{cites}

### 2016-06-11 aplas16/sections/Conclusions.tex
Created a directory for aplas16.

--- 2015-11-21 paper.tex
@@
-%
-% LaTeX template for preparation of submissions to PLDI'16
-%
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'16 web site.
-%
-\documentclass[pldi]{sigplanconf-pldi16}
-% \documentclass[pldi-cameraready]{sigplanconf-pldi16}
-
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet}     % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                % format URLs
-\usepackage{listings}           % format code
-\usepackage{enumitem}           % adjust spacing in enums
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
-
-% code highlighting
-\usepackage{minted}
-\usemintedstyle{tango}
-
-% no red boxes on parser error:
-\AtBeginEnvironment{minted}{%
-  \renewcommand{\fcolorbox}[4][]{#4}}
-
-% quotes
-\usepackage{dirtytalk}
-
-% for symbols
-\usepackage[T1]{fontenc}
-\usepackage{lmodern}
-\usepackage{upquote}
-
-% use arbitrary fonts
-\usepackage{fontspec}
-\setmonofont{Inconsolata}
-
-% for clickable footnotes
-\usepackage{hyperref}
-
-\newcommand{\todo}[2]{{\bf [#1]}: #2}
-
-\begin{document}
-
-\title{Preventing Runtime Errors of Redis at Compile Time}
-
-%
-% any author declaration will be ignored  when using 'pldi' option (for double blind review)
-%
-
-\authorinfo{Ting-Yan Lai}
-{\makebox{Institute of Computer Science and Engineering} \\
-\makebox{National Chiao Tung University, Taiwan}}
-{banacorn.cs03g@nctu.edu.tw}
-
-\authorinfo{Tyng-Ruey Chuang \and Shin-Cheng Mu}
-{\makebox{Institute of Information Science} \\
-\makebox{Academia Sinica, Taiwan}}
-{\{trc,scm\}@iis.sinica.edu.tw}
-
-\maketitle
-
-\begin{abstract}
-Programmers often interact with database systems by sending queries through libraries or packages in some programming languages. People, however, make mistakes.
-While some of the syntactic and semantic errors can be prevented by the
- language at compile time, or caught by the package at runtime, most semantic
- errors are just being ignored, causing problems at the database system.
-
-In this paper, we demonstrate how to prevent those runtime errors at compile
- time, thus allowing users to write more reliable database queries without
- runtime overhead, by exploiting type-level programming techniques such as
- indexed monad, type-level literals and closed type families in Haskell.
-
-The database system and the package we are targeting are \emph{Redis} and
- \emph{Hedis} respectively, and our implementation is available as \emph{Popcorn}\footnotemark
- on \emph{Hackage}.
-
-\footnotetext{Popcorn, a temporary codename for the purposes of double-blind reviewing}
-
-\category{D.3.3}{Programming Languages}{Language Constructs and Features}
-\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Specification techniques}
-\category{H.2.3}{Database Management}{Languages}
-\keywords{Haskell; type families; type-level literals; indexed monad; Redis; database query language}
-\end{abstract}
-
-\section{Introduction}
-
-\subsection{Redis}
-
-Redis\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as
-database, cache and message broker.
-A Redis datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with.
-Redis supports many different kind of values, such as strings, hashes, lists, and sets, etc, and provides a collection of of atomic \emph{commands} to manipulate these values.
-
-For an example, consider the following sequence of commands, entered through the interactive interface of Redis. The keys
-\mintinline{text}{some-set} abd \mintinline{text}{another-set} are
-both associated to a set. The two call to command \mintinline{text}{SADD} respectively adds three and two values to the two sets, before \mintinline{text}{SINTER} takes their intersection:
-\begin{minted}[xleftmargin=1em]{text}
-redis> SADD some-set a b c
-(integer) 3
-redis> SADD another-set a b
-(integer) 2
-redis> SINTER some-set another-set
-1) "a"
-2) "b"
-\end{minted}
-
-Note that the keys \mintinline{text}{some-set} and \mintinline{text}{another-set}, if not existing before the call to \mintinline{text}{SADD}, are created on site. The call to
-\mintinline{text}{SADD} returns the size of the set after
-completion of the command.
-
-\subsection{Hedis}
-
-Many third party library exist to allow general purpose programmings
-to access Redis databases through its TCP protocol. The most popular
-such library for Haskell is Hedis\footnote{\url{https://hackage.haskell.org/package/hedis}}.
-
-The following program is how the previous example looks like in
-Hedis:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-program :: Redis (Either Reply [ByteString])
-program = do
-    sadd "some-set" ["a", "b"]
-    sadd "another-set" ["a", "b", "c"]
-    sinter ["some-set", "another-set"]
-\end{minted}
-
-The function \mintinline{haskell}{sadd} takes a key and a list of values as
- arguments, and returns an \mintinline{haskell}{Integer} on success, or
- returns a \mintinline{haskell}{Reply}, a low-level representation of replies
- from the Redis server, in case of failures. All wrapped in
- \mintinline{haskell}{Redis}\footnotemark, the context of command execution.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-sadd :: ByteString      -- key
-    -> [ByteString]     -- values
-    -> Redis (Either Reply Integer)
-\end{minted}
-
-Note that keys and values, being nothing but binary strings in Redis,
-are represented using Haskell \mintinline{haskell}{ByteString}.
-Values of other types must be encoded as \mintinline{haskell}{ByteString}s before being written to the database,
-and decoded after being read back.
-
-\footnotetext{Hedis provides another kind of context, \mintinline{haskell}{RedisTx}, for \emph{transactions}, united with \mintinline{haskell}{Redis} under the class of \mintinline{haskell}{RedisCtx}.
-For brevity, we demonstrate only \mintinline{haskell}{Redis} in this paper.
-}
-
-\section{Motivation}
-
-All binary strings are equal, but some binary strings are more equal than others.
-
-Although everything in Redis is essentially a binary string, these strings are treated differently. Redis supports many different
-kind of data structures, such as strings, hashes, lists, etc. While
-they are all encoded as binary strings before being written to the databse, most commands, much like how the C language treats a piece of data, only works with data of certain types.
-
-\paragraph{Problem 1} The command \mintinline{text}{SET}, by
-its definition, associates a key to a string. In the following
-example, the key \mintinline{text}{some-string} is associated
-to string \mintinline{text}{foo}. Subsequent calls to
-\mintinline{text}{SADD} causes runtime errors, since the value
-of \mintinline{text}{some-string} is not a set, but a string.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> SET some-string foo
-OK
-redis> SADD some-string bar
-(error) WRONGTYPE Operation against a key
- holding the wrong kind of value
-\end{minted}
-
-\paragraph{Problem 2} Even worse, not all strings are equal!
-The call \mintinline{text}{INCR some-string} parses the string
-associated with key \mintinline{text}{some-string}
-to an integer, increments it by one, and store it back as a string.
-If the string can not be parse as an integer, a runtime error
-is raised.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> SET some-string foo
-OK
-redis> INCR some-string
-(error) ERR value is not an integer or out
- of range
-\end{minted}
-
-\paragraph{In Hedis} Hedis, being only a simple wrapper on top
-of the TCP protocol of Redis, inherits all the problems mentioned
-above. The following program yields the same error as that in
-the Redis client.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-program :: Redis (Either Reply Integer)
-program = do
-    set "some-string" "foo"
-    sadd "some-string" ["a"]
-\end{minted}
-\begin{minted}[xleftmargin=1em]{haskell}
-Left (Error "WRONGTYPE Operation against a
- key holding the wrong kind of value")
-\end{minted}
-
-\paragraph{The Cause} Every key is associated with a value, and every value has
- it's own type. But most commands in Redis only work with a certain type of
- value. When a command is used on a wrong type of key, a runtime error occurs.
- The problems illustrated above arise from the absence of type checking, with
- respects to \textbf{the type of a value that associates with a key}.
- These problems could have been avoided, if we could know the type every key
- associates with in advance, and prevent programs with invalid commands from
- executing.
-
-\subsection{Hedis as an embedded DSL}
-
-Haskell makes it easy to build and use Domain Specific Languages (DSLs),
- and Hedis can be regarded as one of them. What makes Hedis peculiar is that,
- it has \emph{variable bindings} (between keys and values), but with very
- little or no semantic checking, neither dynamically nor statically.
-
-We began with making Hedis a dynamically typechecked embedded DSL, and implemented a
- runtime type checker that keeps track of types of all the variable. But then we
- found that things can be a lot easier, by leveraging the host language's type
- checker. We encode variable bindings with \emph{type-level lists} and
- \emph{strings}, and control the effects on the bindings with
- \emph{indexed monad}. In contrast to the former approach, we \textbf{embedded}
- our type checker into Haskell's type system, without having to build a
- \textbf{standalone} one on the term level.
-
-\subsection{Contributions}
-
-To summarize our contributions:
-
-\begin{itemize}[noitemsep]
-\item We make Hedis statically type-checked, without runtime overhead.
-\item We demonstrates how to model variable bindings of an embedded DSL with
- language extensions like type-level literals and data kinds.
-\item We provide (yet another) an example of encoding effects and constraints of
- an action in types, with indexed monad\cite{indexedmonad} and other language
- extensions such as closed type-families\cite{closedtypefamilies} and
- constraints kinds\cite{constraintskinds}.
-\item Popcorn, a package we built for programmers. This package helps programmers
- to write more reliable Redis programs, and also makes Redis polymorphic by
- automatically converting back and forth from values of arbitrary types and
- boring ByteStrings.
-\end{itemize}
-
-\section{Type-level dictionaries}
-
-To check the bindings between keys and values, we need a \emph{dictionary-like}
- structure, and encode it as a \emph{type} somehow.
-
-\subsection{Datatype promotion}
-Normally, at the term level, we could express the datatype of dictionary with
-\emph{type synonym} like this.\footnotemark
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type Key = String
-type Dictionary = [(Key, TypeRep)]
-\end{minted}
-
-\footnotetext{\mintinline{haskell}{TypeRep} supports term-level representations
- of datatypes, available in \mintinline{haskell}{Data.Typeable}}
-
-To encode this in the type level, everything has to be
- \emph{promoted}\cite{promotion} one level up.
- From terms to types, and from types to kinds.
-
-Luckily, with recently added GHC extension \emph{data kinds}, suitable
- datatype will be automatically promoted to be a kind, and its value
- constructors to be type constructors. The following type {\mintinline{haskell}{List}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data List a = Nil | Cons a (List a)
-\end{minted}
-
-Give rise to the following kinds and type constructors:\footnote{To distinguish
- between types and promoted constructors that have
- ambiguous names, prefix promoted constructor with a single quote like
- \mintinline{text}{'Nil} and \mintinline{text}{'Cons}}
-\footnote{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-List k :: BOX
-Nil  :: List k
-Cons :: k -> List k -> List k
-\end{minted}
-
-Haskell sugars lists \mintinline{haskell}{[1, 2, 3]} and tuples
- \mintinline{text}{(1, 'a')} with brackets and parentheses.
- We could also express promoted lists and tuples in types like this with
- a single quote prefixed. For example:
- \mintinline{text}{'}\mintinline{haskell}{[Int, Char]},
- \mintinline{text}{'}\mintinline{haskell}{(Int, Char)}.
-
-\subsection{Type-level literals}
-
-Now we have type-level lists and tuples to construct the dictionary.
-For keys, \mintinline{haskell}{String} also has a type-level correspondence:
-\mintinline{haskell}{Symbol}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data Symbol
-\end{minted}
-
-Symbol is defined without a value constructor, because it's intended to be used
- as a promoted kind.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-"this is a type-level string literal" :: Symbol
-\end{minted}
-%
-% Nonetheless, it's still useful to have a term-level value that links with a
-%  Symbol, when we want to retrieve type-level information at runtime (but not the
-%  other way around!).
-
-\subsection{Putting everything together}
-
-With all of these ingredients ready, let's build some dictionaries!
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type DictEmpty = '[]
-type Dict0 = '[ '("key", Bool) ]
-type Dict1 = '[ '("A", Int), '("B", "A") ]
-\end{minted}
-
-These dictionaries are defined with \emph{type synonym}, since they are
- \emph{types}, not \emph{terms}. If we ask \mintinline{text}{GHCi} what is the
- kind of \mintinline{haskell}{Dict1}, we will get
- \mintinline{haskell}{Dict1 :: [ (Symbol, *) ]}
-
-The kind \mintinline{haskell}{*} (pronounced "star") stands for the set of all
- concrete type expressions, such as \mintinline{haskell}{Int},
- \mintinline{haskell}{Char} or even a symbol \mintinline{haskell}{"symbol"},
- while \mintinline{haskell}{Symbol} is restricted to all symbols only.
-
-\section{Indexed monads}
-
-Redis commands are \emph{actions}.
- We could capture the effects caused by an action, by expressing the states it
- affects, before and after. That is, the \emph{preconditions} and
- \emph{postconditions} of an action. In such way, we could also impose
- constraints on the preconditions.
-
-\emph{Indexed monads} (or \emph{monadish},
- \emph{parameterised monad})\cite{indexedmonad}
- can be used\cite{typefun, staticresources} to model such preconditions and
- postconditions in types. An indexed monad is a type constructor that takes
- three arguments: an initial state, a final state, and the type of a value that
- an action computes, which can be read like a Hoare triple\cite{kleisli}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-class IMonad m where
-    unit :: a -> m p p a
-    bind :: m p q a -> (a -> m q r b) -> m p r b
-\end{minted}
-
-Class \mintinline{haskell}{IMonad} comes with two operations:
- \mintinline{haskell}{unit} for identities and \mintinline{haskell}{bind} for
- compositions, as in monads.
-
-We define a new datatype \mintinline{haskell}{Popcorn}, which is basically just
- the context \mintinline{haskell}{Redis} indexed with more information in types.
- We make it an instance of \mintinline{haskell}{IMonad}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-newtype Popcorn p q a =
-    Popcorn { unPopcorn :: Redis a }
-
-instance IMonad Popcorn where
-    unit = Popcorn . return
-    bind m f =
-        Popcorn (unPopcorn m >>= unPopcorn . f )
-\end{minted}
-
-The first and second argument of type \mintinline{haskell}{Popcorn} is where the
-dictionaries going to be.
-
-To execute a \mintinline{haskell}{Popcorn} program, simply apply it to
- \mintinline{haskell}{unPopcorn} to get an ordinary Hedis program back,
- with type information erased.
-
-\subsection{\mintinline{text}{PING}: the first attempt}
-
-In Redis, \mintinline{text}{PING} does nothing but replies with
- \mintinline{text}{PONG} if the connection is alive. In Hedis,
- \mintinline{haskell}{ping} has type:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-ping :: Redis (Either Reply Status)
-\end{minted}
-
-Now with \mintinline{haskell}{Popcorn}, we could make our own version of
-\mintinline{haskell}{ping}\footnotemark
-
-\footnotetext{\mintinline{haskell}{ping} from Hedis is qualified with
-\mintinline{haskell}{Hedis} to prevent function name clashing in our code.}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-ping :: Popcorn xs xs (Either Reply Status)
-ping = Popcorn Hedis.ping
-\end{minted}
-
-The dictionary \mintinline{haskell}{xs} in the type remains unaffected after the
- action, because \mintinline{haskell}{ping} does not affect any key-type
- bindings. To encode other commands that modifies key-type bindings, we need
- type-level functions to annotate those effects on the dictionary.
-
-\section{Type-level functions}
-\subsection{Closed type families}
-
-Type families have a wide variety of applications. They can appear inside type
- classes\cite{tfclass, tfsynonym}, or at toplevel. Toplevel type families
- can be used to compute over types, they come in two forms: open\cite{tfopen}
- and closed \cite{tfclosed}.
-
-We choose \emph{closed type families}, because it allows overlapping instances,
- and we need none of the extensibility provided by open type families.
-For example, consider both term-level and type-level \mintinline{haskell}{&&}:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-(&&) :: Bool -> Bool -> Bool
-True && True = True
-a    && b    = False
-
-type family And (a :: Bool) (b :: Bool) :: Bool where
-    And True True = True
-    And a    b    = False
-\end{minted}
-
-The first instance of \mintinline{haskell}{And} could be subsumed under a more
- general instance, \mintinline{haskell}{And a b}.
-But the closedness allows these instances to be resolved in order, just like
- how cases are resolved in term-level functions. Also notice that how much
- \mintinline{haskell}{And} resembles to it's term-level sibling.
-
-\subsection{Functions on type-level dictionaries}
-
-With closed type families, we could define functions on the type level.
- Let's begin with dictionary lookup.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type family Get
-    (xs :: [(Symbol, *)])   -- dictionary
-    (s :: Symbol)           -- key
-    :: * where              -- type
-
-    Get (’(s, x) ’: xs) s = x
-    Get (’(t, x) ’: xs) s = Get xs s
-\end{minted}
-
-Another benefit of closed type families is that type-level equality can be
-expressed by unifying type variables with the same name.
-\mintinline{haskell}{Get} takes two type arguments, a dictionary and a symbol.
-If the key we are looking for unifies with the symbol of an entry, then
- \mintinline{haskell}{Get} returns the corresponding type, else it keeps
- searching down the rest of the dictionary.
-
-\mintinline{haskell}{Get }\mintinline{text}{ '}\mintinline{haskell}{[ }
-\mintinline{text}{'}\mintinline{haskell}{("A", Int) ] "A"} evaluates to
-\mintinline{haskell}{Int}.
-
-But \mintinline{haskell}{Get }
-\mintinline{text}{'}\mintinline{haskell}{[ }
-\mintinline{text}{'}\mintinline{haskell}{("A", Int) ] "B"} would get stuck.
-That's because \mintinline{haskell}{Get} is a partial function on types,
- and these types are computed at compile-time. It wouldn't make
- much sense for a type checker to crash and throw a "Non-exhaustive" error or
- be non-terminating.
-
-We could make \mintinline{haskell}{Get} total, as we would at the term level,
- with \mintinline{haskell}{Maybe}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type family Get
-    (xs :: [(Symbol, *)])   -- dictionary
-    (s :: Symbol)           -- key
-    :: Maybe * where        -- type
-
-    Get ’[]             s = Nothing
-    Get (’(s, x) ’: xs) s = Just x
-    Get (’(t, x) ’: xs) s = Get xs s
-\end{minted}
-%
-Other dictionary-related functions are defined in a similar fashion.
-
-\begin{minted}[xleftmargin=1em]{haskell}
--- inserts or updates an entry
-type family Set
-    (xs :: [(Symbol, *)])   -- old dictionary
-    (s :: Symbol)           -- key
-    (x :: *)                -- type
-    :: [(Symbol, *)] where  -- new dictionary
-
-    Set ’[]             s x = ’[ ’(s, x) ]
-    Set (’(s, y) ’: xs) s x = (’(s, x) ’: xs)
-    Set (’(t, y) ’: xs) s x =
-        ’(t, y) ’: (Set xs s x)
-
--- removes an entry
-type family Del
-    (xs :: [(Symbol, *)])   -- old dictionary
-    (s :: Symbol)           -- key
-    :: [(Symbol, *)] where  -- new dictionary
-
-    Del ’[] s             = ’[]
-    Del (’(s, y) ’: xs) s = xs
-    Del (’(t, y) ’: xs) s = ’(t, y) ’: (Del xs s)
-
--- membership
-type family Member
-    (xs :: [(Symbol, *)])   -- dictionary
-    (s :: Symbol)           -- key
-    :: Bool where           -- exists?
-
-    Member ’[]             s = False
-    Member (’(s, x) ’: xs) s = True
-    Member (’(t, x) ’: xs) s = Member xs s
-\end{minted}
-
-\subsection{Proxies and singleton types}
-
-Now we could annotate the effects of a command in types. \mintinline{text}{DEL}
- removes a key from the current database, regardless of its type.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-del :: KnownSymbol s
-    => Proxy s
-    -> Popcorn xs (Del xs s) (Either Reply Integer)
-del key = Popcorn $ Hedis.del (encodeKey key)
-\end{minted}
-
-\mintinline{haskell}{KnownSymbol} is a class that gives the string associated
- with a concrete type-level symbol, which can be retrieved with
- \mintinline{haskell}{symbolVal}.\footnotemark
- Where \mintinline{haskell}{encodeKey} converts \mintinline{haskell}{Proxy s} to
- \mintinline{haskell}{ByteString}.
-\footnotetext{They are defined in \mintinline{haskell}{GHC.TypeLits}.}
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-encodeKey :: KnownSymbol s => Proxy s -> ByteString
-encodeKey = encode . symbolVal
-\end{minted}
-
-Since Haskell has a \emph{phase distinction, phasedistinction}, types are
- erased before runtime. It's impossible to obtain information directly from
- types, we can only do this indirectly, with
- \emph{singleton types, singletons}.
-
-A singleton type is a type that has only one instance, and the instance can be
- think of as the representative of the type at the realm of runtime values.
-
-\mintinline{haskell}{Proxy}, as its name would suggest, can be used as
- singletons. It's a phantom type that could be indexed with any type.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data Proxy t = Proxy
-\end{minted}
-
-In the type of \mintinline{haskell}{del}, the type variable
- \mintinline{haskell}{s} is a \mintinline{haskell}{Symbol} that is decided by
- the argument of type \mintinline{haskell}{Proxy s}.
- To use \mintinline{haskell}{del}, we would have to apply it with a clumsy
- term-level proxy like this:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-del (Proxy :: Proxy "A")
-\end{minted}
-
-\section{Making Redis polymorphic}
-
-Redis supports many different datatypes, these datatypes as can be viewed as
- \emph{containers} of strings. For example, lists (of strings),
- sets (of strings), and strings themselves.
-
-\subsection{Denoting containers}
-Most Redis commands only work with a certain type of these containers. To
- annotate what container a key is associated with, we introduce these types for
- the universe of containers.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data Strings
-data Lists
-data Sets
-...
-\end{minted}
-
-\mintinline{text}{SET} stores a string, regardless the datatype the key was
- associated with. Now we could implement \mintinline{text}{SET} like this:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-set :: KnownSymbol s
-    => Proxy s
-    -> ByteString       -- data to store
-    -> Popcorn xs (Set xs s Strings) (Either Reply Status)
-set key val = Popcorn $ Hedis.set (encodeKey key) val
-\end{minted}
-
-After \mintinline{text}{SET}, the key will be associated with
- \mintinline{haskell}{Strings} in the dictionary, indicating that it's a string.
-
-\subsection{Automatic data serialization}
-But in the real world, raw binary strings are hardly useful, people would
- usually serialize their data into strings before storing them, and deserialize
- them back when in need.
-
-Instead of letting users writing these boilerplates, we can do these
- serializations/deserializations for them. With the help from
- \mintinline{text}{cereal}, a binary serialization library.
- \mintinline{text}{cereal} comes with these two functions:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-encode :: Serialize a
-       => a -> ByteString Source
-
-decode :: Serialize a
-       => ByteString -> Either String a
-\end{minted}
-
-Which would do all the works for us, as long as the datatype it's handling is
- an instance of class \mintinline{haskell}{Serialize}.\footnotemark
-
-\footnotetext{The methods of \mintinline{haskell}{Serialize} will have default
- generic implementations for all datatypes with some language extensions
- enabled, no sweat!}
-
-\subsection{Extending container types}
-We rename container types and extend it with an extra type argument,
- to indicate what kind of encoded value it's holding.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data StringOf x
-data ListOf x
-data SetOf x
-...
-\end{minted}
-
-\mintinline{haskell}{set} reimplemented with extended container types:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-set :: (KnownSymbol s, Serialize x)
-    => Proxy s
-    -> x       -- can be anything, as long as it's serializable
-    -> Popcorn xs (Set xs s (StringOf x)) (Either Reply Status)
-set key val = Popcorn $ Hedis.set (encodeKey key) (encode val)
-\end{minted}
-
-For example, if we execute \mintinline{haskell}{set (Proxy :: Proxy "A") True},
- a new entry \mintinline{text}{'}\mintinline{haskell}{("A", StringOf Bool)} will
- be inserted to the dictionary.
-
-\subsection{Handling \mintinline{text}{INCR}}
-
-Commands such as \mintinline{text}{INCR} and \mintinline{text}{INCRBYFLOAT}, are
- not only container-specific, they also have some requirements on what types of
- value they could operate with.
-
-We could handle this by mapping Redis's strings of integers and floats to
- Haskell's \mintinline{text}{Integer} and \mintinline{text}{Double}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-incr :: (KnownSymbol s
-      , Get xs s ~ Just (StringOf Integer))
-     => Proxy s
-     -> Popcorn xs xs (Either Reply Integer)
-incr key = Popcorn $ Hedis.incr (encodeKey key)
-
-incrbyfloat :: (KnownSymbol s
-     , Get xs s ~ Just (StringOf Double))
-    => Proxy s
-    -> Double
-    -> Popcorn xs xs (Either Reply Double)
-incrbyfloat key n =
-    Popcorn $ Hedis.incrbyfloat (encodeKey key) n
-\end{minted}
-
-\section{Imposing constraints}
-
-To rule out programs with undesired properties, certain constraints must be
- imposed, on what arguments they can take, or what preconditions they must hold.
-
-Consider the following example: \mintinline{text}{LLEN} returns the length of
- the list associated with a key, else raises a type error.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> LPUSH some-list bar
-(integer) 1
-redis> LLEN some-list
-(integer) 1
-redis> SET some-string foo
-OK
-redis> LLEN some-string
-(error) WRONGTYPE Operation against a key
-holding the wrong kind of value
-\end{minted}
-
-Such constraint could be expressed in types with
-\emph{equality constraints}\cite{typeeq}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-llen :: (KnownSymbol s
-      , Get xs s ~ Just (ListOf x))
-     => Proxy s
-     -> Popcorn xs xs (Either Reply Integer)
-llen key =
-     Popcorn $ Hedis.llen (encodeKey key)
-\end{minted}
-
-Where \mintinline{haskell}{(~)} denotes that \mintinline{haskell}{Get xs s}
-and \mintinline{haskell}{Just (ListOf x)} needs to be the same.
-
-The semantics of \mintinline{text}{LLEN} defined above is actually not
-complete. \mintinline{text}{LLEN} also accepts keys that do not exist, and
- replies with \mintinline{text}{0}.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> LLEN nonexistent
-(integer) 0
-\end{minted}
-
-In other words, we require that the key to be associated with a list,
- \textbf{unless} it doesn't exist at all.
-
-\subsection{Expressing constraint disjunctions}
-
-Unfortunately, expressing disjunctions in constraints is much more difficult
- than expressing conjunctions, since the latter could be easily done by placing
- constraints in a tuple (at the left side of \mintinline{haskell}{=>}).
-
-There are at least three ways to express type-level constraints
-\cite{singletons}. Luckily we could express constraint disjunctions with type
- families in a modular way.
-
-The semantics we want could be expressed informally like this:
-
-\mintinline{haskell}{Get xs s} $\equiv$ \mintinline{haskell}{Just (ListOf x)}
-$\vee$ $\neg$ \mintinline{haskell}{(Member xs s)}
-
-We could achieve this simply by translating the semantics we want to the
- domain of Boolean, with type-level boolean functions such as
-\mintinline{haskell}{(&&)},
-\mintinline{haskell}{(||)}, \mintinline{haskell}{Not},
-\mintinline{haskell}{(==)}, etc.\footnotemark To avoid
-
-\footnotetext{Available in \mintinline{text}{Data.Type.Bool} and
- \mintinline{text}{Data.Type.Equality}}
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-Get xs s == Just (ListOf x) || Not (Member xs s)
-\end{minted}
-
-To avoid addressing the type of value (as it may not exist at all), we defined
- an auxiliary predicate \mintinline{haskell}{IsList :: Maybe * -> Bool} to
- replace the former part.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-IsList (Get xs s) || Not (Member xs s)
-\end{minted}
-
-The type expression above has kind \mintinline{haskell}{Bool}, we could make it
- a type constraint by asserting equality.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-(IsList (Get xs s) || Not (Member xs s)) ~ True
-\end{minted}
-
-With \emph{constraint kind}, a recent addition to GHC, type constraints now has
- its own kind: \mintinline{haskell}{Constraint}. That means type constraints
- are not restricted to the left side of a \mintinline{haskell}{=>} anymore,
- they could appear in anywhere that accepts something of kind
- \mintinline{haskell}{Constraint}, and any type that has kind
- \mintinline{haskell}{Constraint} can also be used as a type constraint.
- \footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}
-
-As many other list-related commands also have this "List or nothing" semantics,
- we could abstract the lengthy type constraint above and give it an alias with
- type synonym.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-ListOrNX xs s =
-    (IsList (Get xs s) || Not (Member xs s)) ~ True
-\end{minted}
-
-The complete implementation of \mintinline{text}{LLEN} with
-\mintinline{haskell}{ListOrNX} would become:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-llen :: (KnownSymbol s, ListOrNX xs s)
-        => Proxy s
-        -> Popcorn xs xs (Either Reply Integer)
-llen key = Popcorn $ Hedis.llen (encodeKey key)
-\end{minted}
-
-\section{Assertions}
-
-Users may need to make assertions about the status of some key-type bindings in
- a Redis program. Specifically, when declaring or renouncing the existence of
- a key and the type of its associating value. We provide these functions,
- which do nothing but fiddle with types.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-declare :: (KnownSymbol s, Member xs s ~ False)
-        => Proxy s
-        -> Proxy x  -- type of value
-        -> Popcorn xs (Set xs s x) ()
-declare s x = Popcorn $ return ()
-
-renounce :: (KnownSymbol s, Member xs s ~ True)
-        => Proxy s -> Popcorn xs (Del xs s) ()
-renounce s = Popcorn $ return ()
-
--- to be used at the beginning of a program
-start :: Popcorn ’[] ’[] ()
-start = Popcorn $ return ()
-\end{minted}
-
-\subsection{A complete example}
-
-The following program increases the value of "A" as an integer, push the result
- of the increment to list "L", and then pops it out.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-main :: IO ()
-main = do
-    conn <- connect defaultConnectInfo
-    result <- runRedis conn $ unPopcorn $ start
-        `bind` \_ ->  declare
-                        (Proxy :: Proxy "A")
-                        (Proxy :: Proxy Integer)
-        `bind` \_ ->  incr     (Proxy :: Proxy "A")
-        `bind` \n ->  case n of
-            Left  err -> lpush (Proxy :: Proxy "L") 0
-            Right n   -> lpush (Proxy :: Proxy "L") n
-        `bind` \_ ->  lpop     (Proxy :: Proxy "L")
-    print result
-\end{minted}
-
-The syntax is pretty heavy, like the old days when there's no
- \emph{do-notation}\cite{history}. But if we don't need any variable bindings
- between operations, we could compose these commands with a sequencing operator
- \mintinline{haskell}{(>>>)}.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-(>>>) :: IMonad m => m p q a -> m q r b -> m p r b
-\end{minted}
-\begin{minted}[xleftmargin=1em]{haskell}
-program = start
-    >>> declare
-            (Proxy :: Proxy "A")
-            (Proxy :: Proxy Integer)
-    >>> incr    (Proxy :: Proxy "A")
-    >>> lpush   (Proxy :: Proxy "L") 0
-    >>> lpop    (Proxy :: Proxy "L")
-\end{minted}
-
-\section{Discussions}
-\paragraph{Syntax}
-No one could ignore the glaring shortcoming of the syntax, which occurs mainly
- in two places: \emph{symbol singletons} and \emph{indexed monad}. We are hoping
- that these issues could be resolved with future syntactic extensions.
-
-\paragraph{Returns only determined datatypes}
-
-All other data structures in Redis also follow the semantics similar to
- "List or nothing". Take the case of \mintinline{text}{GET}, which also shares a
- "String or nothing" semantics that should be typed:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-get :: (KnownSymbol s, Serialize x, StringOrNX xs s)
-    => Proxy s -> Popcorn xs xs (Either Reply (Maybe x))
-\end{minted}
-
-Since the key may not exist, we don't know what \mintinline{haskell}{x} would
-be. We could left \mintinline{haskell}{x} ambiguous, and let it be decided by
- the caller. But users will then be forced to spell out the complete
-type signature of everything, including the dictionaries, only to specify
-the desired resulting type.
-
-Instead of allowing the key to be non-existent, we require that the key must
- exist and it's associating type to be determined at compile time. So our
- version of \mintinline{haskell}{get} has a stricter semantics:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-get :: (KnownSymbol s, Serialize x
-     , Just (StringOf x) ~ Get xs s)
-    => Proxy s -> Popcorn xs xs (Either Reply (Maybe x))
-\end{minted}
-
-\paragraph{Commands with multiple inputs or outputs}
-
-Some command may take a variable number of arguments as inputs, and returns more
- than one value as outputs. To illustrate this, consider
- \mintinline{haskell}{sinter} in Hedis:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-Hedis.sinter :: [ByteString]                      -- keys
-             -> Redis (Either Reply [ByteString]) -- values
-\end{minted}
-
-In Hedis such command could easily be expressed with lists of
- \mintinline{haskell}{ByteString}s. But in Popcorn, things escalate quickly, as
- the keys and values will have to be expressed with \emph{heterogeneous
- lists}\cite{hetero}, which would be pratically infeasible, considering the cost,
- if not impossible.
-
-Most importantly, the keys will all have to be constrained by
- \mintinline{haskell}{KnownSymbol}, which enforces these type literals to be
- concrete and known at compile time.
- It's still unclear whether this is possible.
-
-So instead, we are offering commands that only has a single input and output.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-sinter :: ByteString                      -- single key
-       -> Redis (Either Reply ByteString) -- single value
-\end{minted}
-
-\textbf{Not all Redis programs can be typechecked} (even if they
- might turn out to be type safe). We opted for type safety rather than
- expressiveness.
-
-\paragraph{Redis Transactions}
-
-Redis has \emph{transactions}, another context for executing commands.
-Redis transactions are atomic in the sense that, all commands in a transaction
- will executed sequentially, and no other requests issued by other clients will
- be served \textbf{in the middle}.\footnotemark In contrast, we cannot make such
- a guarantee in the ordinary context, which may destroy the assertions we made
- in types.
-
-At this point of writing, transactions are not supported in our implementation.
- We are planning to add it in the future, and we are expecting that there
- wouldn't be much difficulty, since we've implemented a runtime type checker
- specifically targeting Redis transactions once, before we moved on to the
- types.
-
-\section{Related Work}
-
-While Redis can be viewed as a non-relational database system (although Redis
- seems reluctant to admit this in recent years), there also are similar goals on
- relational database systems, trying to achieve a safer database interface by
- making them statically checked.
- \emph{HaskellDB}\cite{haskelldb, haskelldbimproved} expresses quires
- with relational algebra-like combinators, wrapped in phantom types. Or as
- examples, at first to demonstrate the power of dependent types in
- Agda\cite{pi}, and then with singletons in Haskell\cite{singletons}.
-
 \section{Conclusions}
+\label{sec:conclusions}
 
@@
  but it's not that dreadful as many (including us) would have thought.
-
-\paragraph{Acknowledgements}
-\footnote{Hidden for the purposes of double-blind reviewing}
-
-% references
-\bibliographystyle{abbrvnat}
-\bibliography{cites}
-
-\end{document}

### 2016-06-11 aplas16/sections/IndexedMonads.tex
Created a directory for aplas16.

%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%
%
%
%
%
%
% This package provides two environments suitable to take the place
% of hscode, called "plainhscode" and "arrayhscode". 
%
% The plain environment surrounds each code block by vertical space,
% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
% similar to formulas. Note that if these dimensions are changed,
% the spacing around displayed math formulas changes as well.
% All code is indented using \leftskip.
%
% Changed 19.08.2004 to reflect changes in colorcode. Should work with
% CodeGroup.sty.
%
\ReadOnlyOnce{polycode.fmt}%
\makeatletter

\newcommand{\hsnewpar}[1]%
  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}

% can be used, for instance, to redefine the code size, by setting the
% command to \small or something alike
\newcommand{\hscodestyle}{}

% The command \sethscode can be used to switch the code formatting
% behaviour by mapping the hscode environment in the subst directive
% to a new LaTeX environment.

\newcommand{\sethscode}[1]%
  {\expandafter\let\expandafter\hscode\csname #1\endcsname
   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}

% "compatibility" mode restores the non-polycode.fmt layout.

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

\newcommand{\compaths}{\sethscode{compathscode}}

% "plain" mode is the proposed default.
% It should now work with \centering.
% This required some changes. The old version
% is still available for reference as oldplainhscode.

\newenvironment{plainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newenvironment{oldplainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

% Here, we make plainhscode the default environment.

\newcommand{\plainhs}{\sethscode{plainhscode}}
\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
\plainhs

% The arrayhscode is like plain, but makes use of polytable's
% parray environment which disallows page breaks in code blocks.

\newenvironment{arrayhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\parray}%
  {\endparray\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newcommand{\arrayhs}{\sethscode{arrayhscode}}

% The mathhscode environment also makes use of polytable's parray 
% environment. It is supposed to be used only inside math mode 
% (I used it to typeset the type rules in my thesis).

\newenvironment{mathhscode}%
  {\parray}{\endparray}

\newcommand{\mathhs}{\sethscode{mathhscode}}

% texths is similar to mathhs, but works in text mode.

\newenvironment{texthscode}%
  {\(\parray}{\endparray\)}

\newcommand{\texths}{\sethscode{texthscode}}

% The framed environment places code in a framed box.

\def\codeframewidth{\arrayrulewidth}
\RequirePackage{calc}

\newenvironment{framedhscode}%
  {\parskip=\abovedisplayskip\par\noindent
   \hscodestyle
   \arrayrulewidth=\codeframewidth
   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
   \hline\framedhslinecorrect\\{-1.5ex}%
   \let\endoflinesave=\\
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \framedhslinecorrect\endoflinesave{.5ex}\hline
   \endtabular
   \parskip=\belowdisplayskip\par\noindent
   \ignorespacesafterend}

\newcommand{\framedhslinecorrect}[2]%
  {#1[#2]}

\newcommand{\framedhs}{\sethscode{framedhscode}}

% The inlinehscode environment is an experimental environment
% that can be used to typeset displayed code inline.

\newenvironment{inlinehscode}%
  {\(\def\column##1##2{}%
   \let\>\undefined\let\<\undefined\let\\\undefined
   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
   \def\fromto##1##2##3{##3}%
   \def\nextline{}}{\) }%

\newcommand{\inlinehs}{\sethscode{inlinehscode}}

% The joincode environment is a separate environment that
% can be used to surround and thereby connect multiple code
% blocks.

\newenvironment{joincode}%
  {\let\orighscode=\hscode
   \let\origendhscode=\endhscode
   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
   %\let\SaveRestoreHook=\empty
   %\let\ColumnHook=\empty
   %\let\resethooks=\empty
   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
  {\origendhscode
   \global\let\hscode=\orighscode
   \global\let\endhscode=\origendhscode}%

\makeatother
\EndFmtInput
%

\ReadOnlyOnce{Formatting.fmt}%
\makeatletter

\let\Varid\mathit
\let\Conid\mathsf

\def\commentbegin{\quad\{\ }
\def\commentend{\}}

\newcommand{\ty}[1]{\Conid{#1}}
\newcommand{\con}[1]{\Conid{#1}}
\newcommand{\id}[1]{\Varid{#1}}
\newcommand{\cl}[1]{\Varid{#1}}
\newcommand{\opsym}[1]{\mathrel{#1}}

\newcommand\Keyword[1]{\textbf{\textsf{#1}}}
\newcommand\Hide{\mathbin{\downarrow}}
\newcommand\Reveal{\mathbin{\uparrow}}

%% Paper-specific keywords

\makeatother
\EndFmtInput

\section{Indexed Monads}
\label{sec:indexed-monads}

Redis commands are \emph{actions}.
 We could capture the effects caused by an action, by expressing the states it
 affects, before and after. That is, the \emph{preconditions} and
 \emph{postconditions} of an action. In such way, we could also impose
 constraints on the preconditions.

\emph{Indexed monads} (or \emph{monadish},
 \emph{parameterised monad})\cite{indexedmonad}
 can be used\cite{typefun,staticresources} to model such preconditions and
 postconditions in types. An indexed monad is a type constructor that takes
 three arguments: an initial state, a final state, and the type of a value that
 an action computes, which can be read like a Hoare triple\cite{kleisli}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Class \ensuremath{\Conid{IMonad}} comes with two operations: \ensuremath{\Varid{unit}} for identities and \ensuremath{\Varid{bind}} for
compositions, as in monads.

We define a new datatype \ensuremath{\Conid{Popcorn}}, which is basically just
 the context \ensuremath{\Conid{Redis}} indexed with more information in types.
 We make it an instance of \ensuremath{\Conid{IMonad}}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The first and second argument of type \ensuremath{\Conid{Popcorn}} is where the
dictionaries going to be.

To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to
 \ensuremath{\Varid{unPopcorn}} to get an ordinary Hedis program back,
 with type information erased.

\subsection{\text{PING}: the first attempt}

In Redis, \text{PING} does nothing but replies with
 \text{PONG} if the connection is alive. In Hedis,
 \ensuremath{\Varid{ping}} has type:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Now with \ensuremath{\Conid{Popcorn}}, we could make our own version of
\ensuremath{\Varid{ping}}\footnotemark

\footnotetext{\ensuremath{\Varid{ping}} from Hedis is qualified with
\ensuremath{\Conid{Hedis}} to prevent function name clashing in our code.}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Popcorn}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\\
\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The dictionary \ensuremath{\Varid{xs}} in the type remains unaffected after the
 action, because \ensuremath{\Varid{ping}} does not affect any key-type
 bindings. To encode other commands that modifies key-type bindings, we need
 type-level functions to annotate those effects on the dictionary.

### 2016-06-11 aplas16/sections/Introduction.tex
Created a directory for aplas16.

--- 2016-06-11 aplas16/sections/IndexedMonads.tex
@@
 
-\section{Indexed Monads}
-\label{sec:indexed-monads}
+\section{Introduction}
+\label{sec:introduction}
 
-Redis commands are \emph{actions}.
- We could capture the effects caused by an action, by expressing the states it
- affects, before and after. That is, the \emph{preconditions} and
- \emph{postconditions} of an action. In such way, we could also impose
- constraints on the preconditions.
+\subsection{Redis}
 
-\emph{Indexed monads} (or \emph{monadish},
- \emph{parameterised monad})\cite{indexedmonad}
- can be used\cite{typefun,staticresources} to model such preconditions and
- postconditions in types. An indexed monad is a type constructor that takes
- three arguments: an initial state, a final state, and the type of a value that
- an action computes, which can be read like a Hoare triple\cite{kleisli}.
+Redis\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as
+database, cache and message broker.
+A Redis datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with.
+Redis supports many different kind of values, such as strings, hashes, lists, and sets, etc, and provides a collection of of atomic \emph{commands} to manipulate these values.
+
+For an example, consider the following sequence of commands, entered through the interactive interface of Redis. The keys
+\text{some-set} abd \text{another-set} are
+both associated to a set. The two call to command \text{SADD} respectively adds three and two values to the two sets, before \text{SINTER} takes their intersection:
+\begin{tabbing}\tt
+~redis\char62{}~SADD~some\char45{}set~a~b~c\\
+\tt ~\char40{}integer\char41{}~3\\
+\tt ~redis\char62{}~SADD~another\char45{}set~a~b\\
+\tt ~\char40{}integer\char41{}~2\\
+\tt ~redis\char62{}~SINTER~some\char45{}set~another\char45{}set\\
+\tt ~1\char41{}~\char34{}a\char34{}\\
+\tt ~2\char41{}~\char34{}b\char34{}
+\end{tabbing}
+
+Note that the keys \text{some-set} and \text{another-set}, if not existing before the call to \text{SADD}, are created on site. The call to
+\text{SADD} returns the size of the set after
+completion of the command.
+
+\subsection{Hedis}
+
+Many third party library exist to allow general purpose programmings
+to access Redis databases through its TCP protocol. The most popular
+such library for Haskell is Hedis\footnote{\url{https://hackage.haskell.org/package/hedis}}.
+
+The following program is how the previous example looks like in
+Hedis:
 
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
+\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]){}\<[E]%
+\\
+\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
+\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34}\mskip1.5mu]{}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}{}\<[E]%
+\>[5]{}\Varid{sadd}\;\text{\tt \char34 another-set\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34},\text{\tt \char34 b\char34},\text{\tt \char34 c\char34}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]{}\<[E]%
 \ColumnHook
@@
 
-Class \ensuremath{\Conid{IMonad}} comes with two operations: \ensuremath{\Varid{unit}} for identities and \ensuremath{\Varid{bind}} for
-compositions, as in monads.
-
-We define a new datatype \ensuremath{\Conid{Popcorn}}, which is basically just
- the context \ensuremath{\Conid{Redis}} indexed with more information in types.
- We make it an instance of \ensuremath{\Conid{IMonad}}.
+The function \ensuremath{\Varid{sadd}} takes a key and a list of values as
+ arguments, and returns an \ensuremath{\Conid{Integer}} on success, or
+ returns a \ensuremath{\Conid{Reply}}, a low-level representation of replies
+ from the Redis server, in case of failures. All wrapped in
+ \ensuremath{\Conid{Redis}}\footnotemark, the context of command execution.
 
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}{}\<[E]%
+\>[B]{}\Varid{sadd}{}\<[7]%
+\>[7]{}\mathbin{::}\Conid{ByteString}{}\<[27]%
+\>[27]{}\mbox{\onelinecomment  key}{}\<[E]%
 \\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
+\>[7]{}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]{}\<[27]%
+\>[27]{}\mbox{\onelinecomment  values}{}\<[E]%
 \\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[5]{}\hsindent{4}{}\<[9]%
-\>[9]{}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f}){}\<[E]%
+\>[7]{}\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
 \ColumnHook
@@
 
-The first and second argument of type \ensuremath{\Conid{Popcorn}} is where the
-dictionaries going to be.
+Note that keys and values, being nothing but binary strings in Redis,
+are represented using Haskell \ensuremath{\Conid{ByteString}}.
+Values of other types must be encoded as \ensuremath{\Conid{ByteString}}s before being written to the database,
+and decoded after being read back.
 
-To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to
- \ensuremath{\Varid{unPopcorn}} to get an ordinary Hedis program back,
- with type information erased.
-
-\subsection{\text{PING}: the first attempt}
-
-In Redis, \text{PING} does nothing but replies with
- \text{PONG} if the connection is alive. In Hedis,
- \ensuremath{\Varid{ping}} has type:
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ping}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-Now with \ensuremath{\Conid{Popcorn}}, we could make our own version of
-\ensuremath{\Varid{ping}}\footnotemark
-
-\footnotetext{\ensuremath{\Varid{ping}} from Hedis is qualified with
-\ensuremath{\Conid{Hedis}} to prevent function name clashing in our code.}
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ping}\mathbin{::}\Conid{Popcorn}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
-\\
-\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-The dictionary \ensuremath{\Varid{xs}} in the type remains unaffected after the
- action, because \ensuremath{\Varid{ping}} does not affect any key-type
- bindings. To encode other commands that modifies key-type bindings, we need
- type-level functions to annotate those effects on the dictionary.
+\footnotetext{Hedis provides another kind of context, \ensuremath{\Conid{RedisTx}}, for \emph{transactions}, united with \ensuremath{\Conid{Redis}} under the class of \ensuremath{\Conid{RedisCtx}}.
+For brevity, we demonstrate only \ensuremath{\Conid{Redis}} in this paper.
+}

### 2016-06-11 aplas16/sections/Motivation.tex
Created a directory for aplas16.

--- 2016-06-11 aplas16/sections/IndexedMonads.tex
@@
 
-\section{Indexed Monads}
-\label{sec:indexed-monads}
+\section{Motivation}
+\label{sec:motivation}
 
-Redis commands are \emph{actions}.
- We could capture the effects caused by an action, by expressing the states it
- affects, before and after. That is, the \emph{preconditions} and
- \emph{postconditions} of an action. In such way, we could also impose
- constraints on the preconditions.
+All binary strings are equal, but some binary strings are more equal than others.
 
-\emph{Indexed monads} (or \emph{monadish},
- \emph{parameterised monad})\cite{indexedmonad}
- can be used\cite{typefun,staticresources} to model such preconditions and
- postconditions in types. An indexed monad is a type constructor that takes
- three arguments: an initial state, a final state, and the type of a value that
- an action computes, which can be read like a Hoare triple\cite{kleisli}.
+Although everything in Redis is essentially a binary string, these strings are treated differently. Redis supports many different
+kind of data structures, such as strings, hashes, lists, etc. While
+they are all encoded as binary strings before being written to the databse, most commands, much like how the C language treats a piece of data, only works with data of certain types.
+
+\paragraph{Problem 1} The command \text{SET}, by
+its definition, associates a key to a string. In the following
+example, the key \text{some-string} is associated
+to string \text{foo}. Subsequent calls to
+\text{SADD} causes runtime errors, since the value
+of \text{some-string} is not a set, but a string.
+
+\begin{tabbing}\tt
+~redis\char62{}~SET~some\char45{}string~foo\\
+\tt ~OK\\
+\tt ~redis\char62{}~SADD~some\char45{}string~bar\\
+\tt ~\char40{}error\char41{}~WRONGTYPE~Operation~against~a~key\\
+\tt ~~holding~the~wrong~kind~of~value
+\end{tabbing}
+
+\paragraph{Problem 2} Even worse, not all strings are equal!
+The call \text{INCR some-string} parses the string
+associated with key \text{some-string}
+to an integer, increments it by one, and store it back as a string.
+If the string can not be parse as an integer, a runtime error
+is raised.
+
+\begin{tabbing}\tt
+~redis\char62{}~SET~some\char45{}string~foo\\
+\tt ~OK\\
+\tt ~redis\char62{}~INCR~some\char45{}string\\
+\tt ~\char40{}error\char41{}~ERR~value~is~not~an~integer~or~out\\
+\tt ~~of~range
+\end{tabbing}
+
+\paragraph{In Hedis} Hedis, being only a simple wrapper on top
+of the TCP protocol of Redis, inherits all the problems mentioned
+above. The following program yields the same error as that in
+the Redis client.
 
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
+\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
+\\
+\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
+\>[5]{}\Varid{set}\;\text{\tt \char34 some-string\char34}\;\text{\tt \char34 foo\char34}{}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}{}\<[E]%
+\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-string\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34}\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Conid{Left}\;(\Conid{Error}\;\text{\tt \char34 WRONGTYPE~Operation~against~a\;~key~holding~the~wrong~kind~of~value\char34}{}\<[E]%
+\\
+\>[B]{}){}\<[E]%
 \ColumnHook
@@
 
-Class \ensuremath{\Conid{IMonad}} comes with two operations: \ensuremath{\Varid{unit}} for identities and \ensuremath{\Varid{bind}} for
-compositions, as in monads.
+\paragraph{The Cause} Every key is associated with a value, and every value has
+ it's own type. But most commands in Redis only work with a certain type of
+ value. When a command is used on a wrong type of key, a runtime error occurs.
+ The problems illustrated above arise from the absence of type checking, with
+ respects to \textbf{the type of a value that associates with a key}.
+ These problems could have been avoided, if we could know the type every key
+ associates with in advance, and prevent programs with invalid commands from
+ executing.
 
-We define a new datatype \ensuremath{\Conid{Popcorn}}, which is basically just
- the context \ensuremath{\Conid{Redis}} indexed with more information in types.
- We make it an instance of \ensuremath{\Conid{IMonad}}.
+\subsection{Hedis as an embedded DSL}
 
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[5]{}\hsindent{4}{}\<[9]%
-\>[9]{}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
+Haskell makes it easy to build and use Domain Specific Languages (DSLs),
+ and Hedis can be regarded as one of them. What makes Hedis peculiar is that,
+ it has \emph{variable bindings} (between keys and values), but with very
+ little or no semantic checking, neither dynamically nor statically.
 
-The first and second argument of type \ensuremath{\Conid{Popcorn}} is where the
-dictionaries going to be.
+We began with making Hedis a dynamically typechecked embedded DSL, and implemented a
+ runtime type checker that keeps track of types of all the variable. But then we
+ found that things can be a lot easier, by leveraging the host language's type
+ checker. We encode variable bindings with \emph{type-level lists} and
+ \emph{strings}, and control the effects on the bindings with
+ \emph{indexed monad}. In contrast to the former approach, we \textbf{embedded}
+ our type checker into Haskell's type system, without having to build a
+ \textbf{standalone} one on the term level.
 
-To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to
- \ensuremath{\Varid{unPopcorn}} to get an ordinary Hedis program back,
- with type information erased.
+\subsection{Contributions}
 
-\subsection{\text{PING}: the first attempt}
+To summarize our contributions:
 
-In Redis, \text{PING} does nothing but replies with
- \text{PONG} if the connection is alive. In Hedis,
- \ensuremath{\Varid{ping}} has type:
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ping}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-Now with \ensuremath{\Conid{Popcorn}}, we could make our own version of
-\ensuremath{\Varid{ping}}\footnotemark
-
-\footnotetext{\ensuremath{\Varid{ping}} from Hedis is qualified with
-\ensuremath{\Conid{Hedis}} to prevent function name clashing in our code.}
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ping}\mathbin{::}\Conid{Popcorn}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
-\\
-\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-The dictionary \ensuremath{\Varid{xs}} in the type remains unaffected after the
- action, because \ensuremath{\Varid{ping}} does not affect any key-type
- bindings. To encode other commands that modifies key-type bindings, we need
- type-level functions to annotate those effects on the dictionary.
+\begin{itemize}
+\item We make Hedis statically type-checked, without runtime overhead.
+\item We demonstrates how to model variable bindings of an embedded DSL with
+ language extensions like type-level literals and data kinds.
+\item We provide (yet another) an example of encoding effects and constraints of
+ an action in types, with indexed monad\cite{indexedmonad} and other language
+ extensions such as closed type-families\cite{closedtypefamilies} and
+ constraints kinds\cite{constraintskinds}.
+\item Popcorn, a package we built for programmers. This package helps programmers
+ to write more reliable Redis programs, and also makes Redis polymorphic by
+ automatically converting back and forth from values of arbitrary types and
+ boring ByteStrings.
+\end{itemize}

### 2016-06-11 aplas16/sections/Related.tex
Created a directory for aplas16.

--- 2015-11-20 paper.tex
@@
-%
-% LaTeX template for preparation of submissions to PLDI'16
-%
-% Requires temporary version of sigplanconf style file provided on
-% PLDI'16 web site.
-%
-\documentclass[pldi]{sigplanconf-pldi16}
-% \documentclass[pldi-cameraready]{sigplanconf-pldi16}
-
-%
-% the following standard packages may be helpful, but are not required
-%
-\usepackage{SIunits}            % typset units correctly
-\usepackage{courier}            % standard fixed width font
-\usepackage[scaled]{helvet}     % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{url}                % format URLs
-\usepackage{listings}           % format code
-\usepackage{enumitem}           % adjust spacing in enums
-\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
-% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
-\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}   % print a hyperlinked DOI
-
-% code highlighting
-\usepackage{minted}
-\usemintedstyle{tango}
-
-% no red boxes on parser error:
-\AtBeginEnvironment{minted}{%
-  \renewcommand{\fcolorbox}[4][]{#4}}
-
-% quotes
-\usepackage{dirtytalk}
-
-% for symbols
-\usepackage[T1]{fontenc}
-\usepackage{lmodern}
-\usepackage{upquote}
-
-% use arbitrary fonts
-\usepackage{fontspec}
-\setmonofont{Inconsolata}
-
-% for clickable footnotes
-\usepackage{hyperref}
-
-\newcommand{\todo}[2]{{\bf [#1]}: #2}
-
-\begin{document}
-
-\title{Preventing Runtime Errors of Redis at Compile Time}
-
-%
-% any author declaration will be ignored  when using 'pldi' option (for double blind review)
-%
-
-\authorinfo{Ting-Yan Lai}
-{\makebox{Institute of Computer Science and Engineering} \\
-\makebox{National Chiao Tung University, Taiwan}}
-{banacorn.cs03g@nctu.edu.tw}
-
-\authorinfo{Tyng-Ruey Chuang \and Shin-Cheng Mu}
-{\makebox{Institute of Information Science} \\
-\makebox{Academia Sinica, Taiwan}}
-{\{trc,scm\}@iis.sinica.edu.tw}
-
-\maketitle
-
-\begin{abstract}
-Programmers often interact with database systems by sending queries through libraries or packages in some programming languages. People, however, make mistakes.
-While some of the syntactic and semantic errors can be prevented by the
- language at compile time, or caught by the package at runtime, most semantic
- errors are just being ignored, causing problems at the database system.
-
-In this paper, we demonstrate how to prevent those runtime errors at compile
- time, thus allowing users to write more reliable database queries without
- runtime overhead, by exploiting type-level programming techniques such as
- indexed monad, type-level literals and closed type families in Haskell.
-
-The database system and the package we are targeting are \emph{Redis} and
- \emph{Hedis} respectively, and our implementation is available as \emph{Popcorn}\footnotemark
- on \emph{Hackage}.
-
-\footnotetext{Popcorn, a temporary codename for the purposes of double-blind reviewing}
-
-\category{D.3.3}{Programming Languages}{Language Constructs and Features}
-\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Specification techniques}
-\category{H.2.3}{Database Management}{Languages}
-\keywords{Haskell; type families; type-level literals; indexed monad; Redis; database query language}
-\end{abstract}
-
-\section{Introduction}
-
-\subsection{Redis}
-
-Redis\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as
-database, cache and message broker.
-A Redis datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with.
-Redis supports many different kind of values, such as strings, hashes, lists, and sets, etc, and provides a collection of of atomic \emph{commands} to manipulate these values.
-
-For an example, consider the following sequence of commands, entered through the interactive interface of Redis. The keys
-\mintinline{text}{some-set} abd \mintinline{text}{another-set} are
-both associated to a set. The two call to command \mintinline{text}{SADD} respectively adds three and two values to the two sets, before \mintinline{text}{SINTER} takes their intersection:
-\begin{minted}[xleftmargin=1em]{text}
-redis> SADD some-set a b c
-(integer) 3
-redis> SADD another-set a b
-(integer) 2
-redis> SINTER some-set another-set
-1) "a"
-2) "b"
-\end{minted}
-
-Note that the keys \mintinline{text}{some-set} and \mintinline{text}{another-set}, if not existing before the call to \mintinline{text}{SADD}, are created on site. The call to
-\mintinline{text}{SADD} returns the size of the set after
-completion of the command.
-
-\subsection{Hedis}
-
-Many third party library exist to allow general purpose programmings
-to access Redis databases through its TCP protocol. The most popular
-such library for Haskell is Hedis\footnote{\url{https://hackage.haskell.org/package/hedis}}.
-
-The following program is how the previous example looks like in
-Hedis:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-program :: Redis (Either Reply [ByteString])
-program = do
-    sadd "some-set" ["a", "b"]
-    sadd "another-set" ["a", "b", "c"]
-    sinter ["some-set", "another-set"]
-\end{minted}
-
-The function \mintinline{haskell}{sadd} takes a key and a list of values as
- arguments, and returns an \mintinline{haskell}{Integer} on success, or
- returns a \mintinline{haskell}{Reply}, a low-level representation of replies
- from the Redis server, in case of failures. All wrapped in
- \mintinline{haskell}{Redis}\footnotemark, the context of command execution.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-sadd :: ByteString      -- key
-    -> [ByteString]     -- values
-    -> Redis (Either Reply Integer)
-\end{minted}
-
-Note that keys and values, being nothing but binary strings in Redis,
-are represented using Haskell \mintinline{haskell}{ByteString}.
-Values of other types must be encoded as \mintinline{haskell}{ByteString}s before being written to the database,
-and decoded after being read back.
-
-\footnotetext{Hedis provides another kind of context, \mintinline{haskell}{RedisTx}, for \emph{transactions}, united with \mintinline{haskell}{Redis} under the class of \mintinline{haskell}{RedisCtx}.
-For brevity, we demonstrate only \mintinline{haskell}{Redis} in this paper.
-}
-
-\section{Motivation}
-
-All binary strings are equal, but some binary strings are more equal than others.
-
-Although everything in Redis is essentially a binary string, these strings are treated differently. Redis supports many different
-kind of data structures, such as strings, hashes, lists, etc. While
-they are all encoded as binary strings before being written to the databse, most commands, much like how the C language treats a piece of data, only works with data of certain types.
-
-\paragraph{Problem 1} The command \mintinline{text}{SET}, by
-its definition, associates a key to a string. In the following
-example, the key \mintinline{text}{some-string} is associated
-to string \mintinline{text}{foo}. Subsequent calls to
-\mintinline{text}{SADD} causes runtime errors, since the value
-of \mintinline{text}{some-string} is not a set, but a string.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> SET some-string foo
-OK
-redis> SADD some-string bar
-(error) WRONGTYPE Operation against a key
- holding the wrong kind of value
-\end{minted}
-
-\paragraph{Problem 2} Even worse, not all strings are equal!
-The call \mintinline{text}{INCR some-string} parses the string
-associated with key \mintinline{text}{some-string}
-to an integer, increments it by one, and store it back as a string.
-If the string can not be parse as an integer, a runtime error
-is raised.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> SET some-string foo
-OK
-redis> INCR some-string
-(error) ERR value is not an integer or out
- of range
-\end{minted}
-
-\paragraph{In Hedis} Hedis, being only a simple wrapper on top
-of the TCP protocol of Redis, inherits all the problems mentioned
-above. The following program yields the same error as that in
-the Redis client.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-program :: Redis (Either Reply Integer)
-program = do
-    set "some-string" "foo"
-    sadd "some-string" ["a"]
-\end{minted}
-\begin{minted}[xleftmargin=1em]{haskell}
-Left (Error "WRONGTYPE Operation against a
- key holding the wrong kind of value")
-\end{minted}
-
-\paragraph{The Cause} Every key is associated with a value, and every value has
- it's own type. But most commands in Redis only work with a certain type of
- value. When a command is used on a wrong type of key, a runtime error occurs.
- The problems illustrated above arise from the absence of type checking, with
- respects to \textbf{the type of a value that associates with a key}.
- These problems could have been avoided, if we could know the type every key
- associates with in advance, and prevent programs with invalid commands from
- executing.
-
-\subsection{Hedis as an embedded DSL}
-
-Haskell makes it easy to build and use Domain Specific Languages (DSLs),
- and Hedis can be regarded as one of them. What makes Hedis peculiar is that,
- it has \emph{variable bindings} (between keys and values), but with very
- little or no semantic checking, neither dynamically nor statically.
-
-We began with making Hedis a dynamically typechecked embedded DSL, and implemented a
- runtime type checker that keeps track of types of all the variable. But then we
- found that things can be a lot easier, by leveraging the host language's type
- checker. We encode variable bindings with \emph{type-level lists} and
- \emph{strings}, and control the effects on the bindings with
- \emph{indexed monad}. In contrast to the former approach, we \textbf{embedded}
- our type checker into Haskell's type system, without having to build a
- \textbf{standalone} one on the term level.
-
-\subsection{Contributions}
-
-To summarize our contributions:
-
-\begin{itemize}[noitemsep]
-\item We make Hedis statically type-checked, without runtime overhead.
-\item We demonstrates how to model variable bindings of an embedded DSL with
- language extensions like type-level literals and data kinds.
-\item We provide (yet another) an example of encoding effects and constraints of
- an action in types, with indexed monad\cite{indexedmonad} and other language
- extensions such as closed type-families\cite{closedtypefamilies} and
- constraints kinds\cite{constraintskinds}.
-\item Popcorn, a package we built for programmers. This package helps programmers
- to write more reliable Redis programs, and also makes Redis polymorphic by
- automatically converting back and forth from values of arbitrary types and
- boring ByteStrings.
-\end{itemize}
-
-\section{Type-level dictionaries}
-
-To check the bindings between keys and values, we need a \emph{dictionary-like}
- structure, and encode it as a \emph{type} somehow.
-
-\subsection{Datatype promotion}
-Normally, at the term level, we could express the datatype of dictionary with
-\emph{type synonym} like this.\footnotemark
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type Key = String
-type Dictionary = [(Key, TypeRep)]
-\end{minted}
-
-\footnotetext{\mintinline{haskell}{TypeRep} supports term-level representations
- of datatypes, available in \mintinline{haskell}{Data.Typeable}}
-
-To encode this in the type level, everything has to be
- \emph{promoted}\cite{promotion} one level up.
- From terms to types, and from types to kinds.
-
-Luckily, with recently added GHC extension \emph{data kinds}, suitable
- datatype will be automatically promoted to be a kind, and its value
- constructors to be type constructors. The following type {\mintinline{haskell}{List}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data List a = Nil | Cons a (List a)
-\end{minted}
-
-Give rise to the following kinds and type constructors:\footnote{To distinguish
- between types and promoted constructors that have
- ambiguous names, prefix promoted constructor with a single quote like
- \mintinline{text}{'Nil} and \mintinline{text}{'Cons}}
-\footnote{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-List k :: BOX
-Nil  :: List k
-Cons :: k -> List k -> List k
-\end{minted}
-
-Haskell sugars lists \mintinline{haskell}{[1, 2, 3]} and tuples
- \mintinline{text}{(1, 'a')} with brackets and parentheses.
- We could also express promoted lists and tuples in types like this with
- a single quote prefixed. For example:
- \mintinline{text}{'}\mintinline{haskell}{[Int, Char]},
- \mintinline{text}{'}\mintinline{haskell}{(Int, Char)}.
-
-\subsection{Type-level literals}
-
-Now we have type-level lists and tuples to construct the dictionary.
-For keys, \mintinline{haskell}{String} also has a type-level correspondence:
-\mintinline{haskell}{Symbol}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data Symbol
-\end{minted}
-
-Symbol is defined without a value constructor, because it's intended to be used
- as a promoted kind.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-"this is a type-level string literal" :: Symbol
-\end{minted}
-%
-% Nonetheless, it's still useful to have a term-level value that links with a
-%  Symbol, when we want to retrieve type-level information at runtime (but not the
-%  other way around!).
-
-\subsection{Putting everything together}
-
-With all of these ingredients ready, let's build some dictionaries!
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type DictEmpty = '[]
-type Dict0 = '[ '("key", Bool) ]
-type Dict1 = '[ '("A", Int), '("B", "A") ]
-\end{minted}
-
-These dictionaries are defined with \emph{type synonym}, since they are
- \emph{types}, not \emph{terms}. If we ask \mintinline{text}{GHCi} what is the
- kind of \mintinline{haskell}{Dict1}, we will get
- \mintinline{haskell}{Dict1 :: [ (Symbol, *) ]}
-
-The kind \mintinline{haskell}{*} (pronounced "star") stands for the set of all
- concrete type expressions, such as \mintinline{haskell}{Int},
- \mintinline{haskell}{Char} or even a symbol \mintinline{haskell}{"symbol"},
- while \mintinline{haskell}{Symbol} is restricted to all symbols only.
-
-\section{Indexed monads}
-
-Redis commands are \emph{actions}.
- We could capture the effects caused by an action, by expressing the states it
- affects, before and after. That is, the \emph{preconditions} and
- \emph{postconditions} of an action. In such way, we could also impose
- constraints on the preconditions.
-
-\emph{Indexed monads} (or \emph{monadish},
- \emph{parameterised monad})\cite{indexedmonad}
- can be used\cite{typefun, staticresources} to model such preconditions and
- postconditions in types. An indexed monad is a type constructor that takes
- three arguments: an initial state, a final state, and the type of a value that
- an action computes, which can be read like a Hoare triple\cite{kleisli}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-class IMonad m where
-    unit :: a -> m p p a
-    bind :: m p q a -> (a -> m q r b) -> m p r b
-\end{minted}
-
-Class \mintinline{haskell}{IMonad} comes with two operations:
- \mintinline{haskell}{unit} for identities and \mintinline{haskell}{bind} for
- compositions, as in monads.
-
-We define a new datatype \mintinline{haskell}{Popcorn}, which is basically just
- the context \mintinline{haskell}{Redis} indexed with more information in types.
- We make it an instance of \mintinline{haskell}{IMonad}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-newtype Popcorn p q a = Popcorn { unPopcorn :: Redis a }
-
-instance IMonad Popcorn where
-    unit = Popcorn . return
-    bind m f = Popcorn (unPopcorn m >>= unPopcorn . f )
-\end{minted}
-
-The first and second argument of type \mintinline{haskell}{Popcorn} is where the
-dictionaries going to stay.
-
-To execute a \mintinline{haskell}{Popcorn} program, simply apply it to
- \mintinline{haskell}{unPopcorn} to get an ordinary program of type
- \mintinline{haskell}{Redis}, with type information erased.
-
-\subsection{\mintinline{text}{PING}: the first attempt}
-
-In Redis, \mintinline{text}{PING} does nothing but replies with
- \mintinline{text}{PONG} if the connection is alive. In Hedis,
- \mintinline{text}{PING} has type:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-ping :: Redis (Either Reply Status)
-\end{minted}
-
-Now we have \mintinline{haskell}{Popcorn}, let's make our own version of
-\mintinline{haskell}{ping}\footnotemark
-
-\footnotetext{\mintinline{haskell}{ping} from Hedis is qualified with
-\mintinline{haskell}{Hedis} to prevent function name clashing in our code.}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-ping :: Popcorn xs xs (Either Reply Status)
-ping = Popcorn Hedis.ping
-\end{minted}
-
-The dictionary \mintinline{haskell}{xs} in the type remains unaffected after the
- action, because \mintinline{haskell}{ping} does not affect any key-value
- bindings. To encode other commands that modifies key-value bindings, we need
- type-level functions to annotate those effects on the dictionary.
-
-\section{Type-level functions}
-\subsection{Closed type families}
-
-Type families have a wide variety of applications. They can appear inside type
- classes\cite{tfclass, tfsynonym}, or at toplevel. Toplevel type families
- can be used to compute over types, they come in two forms: open\cite{tfopen}
- and closed \cite{tfclosed}.
-
-We choose \emph{closed type families}, because it allows overlapping instances,
- and we need none of the extensibility provided by open type families.
-For example, consider both term-level and type-level \mintinline{haskell}{&&}:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-(&&) :: Bool -> Bool -> Bool
-True && True = True
-a    && b    = False
-
-type family And (a :: Bool) (b :: Bool) :: Bool where
-    And True True = True
-    And a    b    = False
-\end{minted}
-
-The first instance of \mintinline{haskell}{And} could be subsumed under a more
- general instance, \mintinline{haskell}{And a b}.
-But the closedness allows these instances to be resolved in order, just like
- how cases are resolved in term-level functions. Also notice that how much
- \mintinline{haskell}{And} resembles to it's term-level sibling.
-
-\subsection{Functions on type-level dictionaries}
-
-With closed type families, we could define functions on the type level.
- Let's begin with dictionary lookup.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type family Get
-    (xs :: [(Symbol, *)])   -- dictionary
-    (s :: Symbol)           -- key
-    :: * where              -- type
-
-    Get (’(s, x) ’: xs) s = x
-    Get (’(t, x) ’: xs) s = Get xs s
-\end{minted}
-
-Another benefit of closed type families is that type-level equality can be
-expressed by unifying type variables with the same name.
-\mintinline{haskell}{Get} takes two type arguments, a dictionary and a symbol.
-If the key we are looking for unifies with the symbol of an entry, then
- \mintinline{haskell}{Get} returns the corresponding type, else it keeps
- searching down the rest of the dictionary.
-
-\mintinline{haskell}{Get }\mintinline{text}{ '}\mintinline{haskell}{[ }
-\mintinline{text}{'}\mintinline{haskell}{("A", Int) ] "A"} evaluates to
-\mintinline{haskell}{Int}.
-
-But \mintinline{haskell}{Get }
-\mintinline{text}{'}\mintinline{haskell}{[ }
-\mintinline{text}{'}\mintinline{haskell}{("A", Int) ] "B"} would get stuck.
-That's because \mintinline{haskell}{Get} is a partial function on types,
- and these types are computed at compile-time. It wouldn't make
- much sense for a type checker to crash and throw a "Non-exhaustive" error or
- be non-terminating.
-
-We could make \mintinline{haskell}{Get} total, as we would at the term level,
- with \mintinline{haskell}{Maybe}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-type family Get
-    (xs :: [(Symbol, *)])   -- dictionary
-    (s :: Symbol)           -- key
-    :: Maybe * where        -- type
-
-    Get ’[]             s = Nothing
-    Get (’(s, x) ’: xs) s = Just x
-    Get (’(t, x) ’: xs) s = Get xs s
-\end{minted}
-%
-Other dictionary-related functions are defined in a similar fashion.
-
-\begin{minted}[xleftmargin=1em]{haskell}
--- inserts or updates an entry
-type family Set
-    (xs :: [(Symbol, *)])   -- old dictionary
-    (s :: Symbol)           -- key
-    (x :: *)                -- type
-    :: [(Symbol, *)] where  -- new dictionary
-
-    Set ’[]             s x = ’[ ’(s, x) ]
-    Set (’(s, y) ’: xs) s x = (’(s, x) ’: xs)
-    Set (’(t, y) ’: xs) s x =
-        ’(t, y) ’: (Set xs s x)
-
--- removes an entry
-type family Del
-    (xs :: [(Symbol, *)])   -- old dictionary
-    (s :: Symbol)           -- key
-    :: [(Symbol, *)] where  -- new dictionary
-
-    Del ’[] s             = ’[]
-    Del (’(s, y) ’: xs) s = xs
-    Del (’(t, y) ’: xs) s = ’(t, y) ’: (Del xs s)
-
--- membership
-type family Member
-    (xs :: [(Symbol, *)])   -- dictionary
-    (s :: Symbol)           -- key
-    :: Bool where           -- exists?
-
-    Member ’[]             s = False
-    Member (’(s, x) ’: xs) s = True
-    Member (’(t, x) ’: xs) s = Member xs s
-\end{minted}
-
-\subsection{Proxies and singleton types}
-
-Now we could annotate the effects of a command in types. \mintinline{text}{DEL}
- removes a key from the current database, regardless of its type.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-del :: KnownSymbol s
-    => Proxy s
-    -> Popcorn xs (Del xs s) (Either Reply Integer)
-del key = Popcorn $ Hedis.del (encodeKey key)
-\end{minted}
-
-\mintinline{haskell}{KnownSymbol} is a class that gives the string associated
- with a concrete type-level symbol, which can be retrieved with
- \mintinline{haskell}{symbolVal}.\footnotemark
- Where \mintinline{haskell}{encodeKey} converts \mintinline{haskell}{Proxy s} to
- \mintinline{haskell}{ByteString}.
-\footnotetext{They are defined in \mintinline{haskell}{GHC.TypeLits}.}
-
-\begin{minted}[xleftmargin=1em]{haskell}
-encodeKey :: KnownSymbol s => Proxy s -> ByteString
-encodeKey = encode . symbolVal
-\end{minted}
-
-Since Haskell has a \emph{phase distinction, phasedistinction}, types are
- erased before runtime. It's impossible to obtain information directly from
- types, we can only do this indirectly, with
- \emph{singleton types, singletons}.
-
-A singleton type is a type that has only one instance, and the instance can be
- think of as the representative of the type at the realm of runtime values.
-
-\mintinline{haskell}{Proxy}, as its name would suggest, can be used as
- singletons. It's a phantom type that could be indexed with any type.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data Proxy t = Proxy
-\end{minted}
-
-In the type of \mintinline{haskell}{del}, the type variable
- \mintinline{haskell}{s} is a \mintinline{haskell}{Symbol} that is decided by
- the argument of type \mintinline{haskell}{Proxy s}.
- To use \mintinline{haskell}{del}, we would have to apply it with a clumsy
- term-level proxy like this:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-del (Proxy :: Proxy "A")
-\end{minted}
-
-\section{Making Redis polymorphic}
-
-Redis supports many different datatypes, these datatypes as can be viewed as
- \emph{containers} of strings. For example, lists (of strings),
- sets (of strings), and strings themselves.
-
-\subsection{Denoting containers}
-Most Redis commands only work with a certain type of these containers. To
- annotate what container a key is associated with, we introduce these container
- types.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data Strings
-data Lists
-data Sets
-...
-\end{minted}
-
-\mintinline{text}{SET} stores a string, regardless the datatype the key was
- associated with. Now we could implement \mintinline{text}{SET} like this:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-set :: KnownSymbol s
-    => Proxy s
-    -> ByteString       -- data to store
-    -> Popcorn xs (Set xs s Strings) (Either Reply Status)
-set key val = Popcorn $ Hedis.set (encodeKey key) val
-\end{minted}
-
-After \mintinline{text}{SET}, the key will be associated with
- \mintinline{haskell}{Strings} in the dictionary, indicating that it's a string.
-
-\subsection{Automatic data serialization}
-But in the real world, raw binary strings are hardly useful, people would
- usually serialize their data into strings before storing them, and deserialize
- them back when in need.
-
-Instead of letting users writing these boilerplates, we can do these
- serializations/deserializations for them. With the help from
- \mintinline{text}{cereal}, a binary serialization library.
- \mintinline{text}{cereal} comes with these two functions:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-encode :: Serialize a
-       => a -> ByteString Source
-
-decode :: Serialize a
-       => ByteString -> Either String a
-\end{minted}
-
-Which would do all the works for us, as long as the datatype it's handling is
- an instance of class \mintinline{haskell}{Serialize}.\footnotemark
-
-\footnotetext{The methods of \mintinline{haskell}{Serialize} will have default
- generic implementations for all datatypes with some language extensions
- enabled, no sweat!}
-
-\subsection{Extending container types}
-We rename container types and extend it with an extra type argument,
- to indicate what kind of encoded value it's holding.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-data StringOf x
-data ListOf x
-data SetOf x
-...
-\end{minted}
-
-\mintinline{haskell}{set} reimplemented with extended container types:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-set :: (KnownSymbol s, Serialize x)
-    => Proxy s
-    -> x       -- can be anything, as long as it's serializable
-    -> Popcorn xs (Set xs s (StringOf x)) (Either Reply Status)
-set key val = Popcorn $ Hedis.set (encodeKey key) (encode val)
-\end{minted}
-
-For example, if we execute \mintinline{haskell}{set (Proxy :: Proxy "A") True},
- a new entry \mintinline{text}{'}\mintinline{haskell}{("A", StringOf Bool)} will
- be inserted to the dictionary.
-
-\subsection{Handling \mintinline{text}{INCR}}
-
-Commands such as \mintinline{text}{INCR} and \mintinline{text}{INCRBYFLOAT}, are
- not only container-specific, they also have some requirements on what types of
- value they could operate on.
-
-We could handle this by mapping Redis's strings of integers and floats to
- Haskell's \mintinline{text}{Integer} and \mintinline{text}{Double}.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-incr :: (KnownSymbol s, Get xs s ~ Just (StringOf Integer))
-     => Proxy s -> Popcorn xs xs (Either Reply Integer)
-incr key = Popcorn $ Hedis.incr (encodeKey key)
-
-incrbyfloat :: (KnownSymbol s, Get xs s ~ Just (StringOf Double))
-     => Proxy s -> Double -> Popcorn xs xs (Either Reply Double)
-incrbyfloat key n = Popcorn $ Hedis.incrbyfloat (encodeKey key) n
-\end{minted}
-
-\section{Imposing constraints}
-
-To rule out programs with undesired properties, certain constraints must be
- imposed, on what arguments they can take, or what preconditions they must hold.
-
-Consider the following example: \mintinline{text}{LLEN} returns the length of
- the list associated with a key, else raises a type error.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> LPUSH some-list bar
-(integer) 1
-redis> LLEN some-list
-(integer) 1
-redis> SET some-string foo
-OK
-redis> LLEN some-string
-(error) WRONGTYPE Operation against a key
-holding the wrong kind of value
-\end{minted}
-
-Such constraint could be expressed in types with
-\emph{equality constraints}\cite{typeeq}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-llen :: (KnownSymbol s, Get xs s ~ Just (ListOf x))
-        => Proxy s
-        -> Popcorn xs xs (Either Reply Integer)
-llen key = Popcorn $ Hedis.llen (encodeKey key)
-\end{minted}
-
-Where \mintinline{haskell}{(~)} denotes that \mintinline{haskell}{Get xs s}
-and \mintinline{haskell}{Just (ListOf x)} needs to be the same.
-
-The semantics of \mintinline{text}{LLEN} defined above is actually not
-complete. \mintinline{text}{LLEN} also accepts keys that do not exist, and
- replies with \mintinline{text}{0}.
-
-\begin{minted}[xleftmargin=1em]{text}
-redis> LLEN nonexistent
-(integer) 0
-\end{minted}
-
-In other words, we require that the key to be associated with a list,
- \textbf{unless} it doesn't exist at all.
-
-\subsection{Expressing constraint disjunctions}
-
-Unfortunately, expressing disjunctions in constraints is much more difficult
- than expressing conjunctions, since the latter could be easily done by placing
- constraints in a tuple (at the left side of \mintinline{haskell}{=>}).
-
-There are at least three ways to express type-level constraints
-\cite{singletons}. Luckily we could express constraint disjunctions with type
- families in a modular way.
-
-The semantics we want could be expressed informally like this:
-
-\mintinline{haskell}{Get xs s} $\equiv$ \mintinline{haskell}{Just (ListOf x)}
-$\vee$ $\neg$ \mintinline{haskell}{(Member xs s)}
-
-We could achieve this simply by translating the semantics we want to the
- domain of Boolean, with type-level boolean functions such as
-\mintinline{haskell}{(&&)},
-\mintinline{haskell}{(||)}, \mintinline{haskell}{Not},
-\mintinline{haskell}{(==)}, etc.\footnotemark To avoid
-
-\footnotetext{Available in \mintinline{text}{Data.Type.Bool} and
- \mintinline{text}{Data.Type.Equality}}
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-Get xs s == Just (ListOf x) || Not (Member xs s)
-\end{minted}
-
-To avoid addressing the type of value (as it may not exist at all), we defined
- an auxiliary predicate \mintinline{haskell}{IsList :: Maybe * -> Bool} to
- replace the former part.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-IsList (Get xs s) || Not (Member xs s)
-\end{minted}
-
-The type expression above has kind \mintinline{haskell}{Bool}, we could make it
- a type constraint by asserting equality.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-(IsList (Get xs s) || Not (Member xs s)) ~ True
-\end{minted}
-
-With \emph{constraint kind}, a recent addition to GHC, type constraints now has
- its own kind: \mintinline{haskell}{Constraint}. That means type constraints
- are not restricted to the left side of a \mintinline{haskell}{=>} anymore,
- they could appear in anywhere that accepts something of kind
- \mintinline{haskell}{Constraint}, and any type that has kind
- \mintinline{haskell}{Constraint} can also be used as a type constraint.
- \footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}
-
-As many other list-related commands also have this "List or nothing" semantics,
- we could abstract the lengthy type constraint above and give it an alias with
- type synonym.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-ListOrNX xs s =
-    (IsList (Get xs s) || Not (Member xs s)) ~ True
-\end{minted}
-
-The complete implementation of \mintinline{text}{LLEN} with
-\mintinline{haskell}{ListOrNX} would become:
-
-\begin{minted}[xleftmargin=1em]{haskell}
-llen :: (KnownSymbol s, ListOrNX xs s)
-        => Proxy s
-        -> Popcorn xs xs (Either Reply Integer)
-llen key = Popcorn $ Hedis.llen (encodeKey key)
-\end{minted}
-
-\section{Assertions}
-
-Users may need to make assertions about the status of some key-type bindings in
- a Redis program. For example, declaring the existence of a key and the type
- of its associating value. We provide these functions, which do nothing but
- fiddling with types.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-declare :: (KnownSymbol s, Member xs s ~ False)
-        => Proxy s
-        -> Proxy x  -- type of value
-        -> Popcorn xs (Set xs s x) ()
-declare s x = Popcorn $ return ()
-
-renounce :: (KnownSymbol s, Member xs s ~ True)
-        => Proxy s -> Popcorn xs (Del xs s) ()
-renounce s = Popcorn $ return ()
-
--- empty precondition
-start :: Popcorn ’[] ’[] ()
-start = Popcorn $ return ()
-\end{minted}
-
-\subsection{A complete example}
-
-The following program increases the value of "A" as an integer, push the result
- of the increment to list "L", and then pops it out.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-main :: IO ()
-main = do
-    conn <- connect defaultConnectInfo
-    result <- runRedis conn $ unPopcorn $ start
-        `bind` \_ ->  declare
-                        (Proxy :: Proxy "A")
-                        (Proxy :: Proxy Integer)
-        `bind` \_ ->  incr     (Proxy :: Proxy "A")
-        `bind` \n ->  case n of
-            Left  err -> lpush (Proxy :: Proxy "L") 0
-            Right n   -> lpush (Proxy :: Proxy "L") n
-        `bind` \_ ->  lpop     (Proxy :: Proxy "L")
-    print result
-\end{minted}
-
-The syntax is pretty heavy, like the old days when there's no
- \emph{do-notation}\cite{history}. But if we don't need any variable bindings
- between operations, we could compose these commands with a sequencing operator
- \mintinline{haskell}{(>>>)}.
-
-\begin{minted}[xleftmargin=1em]{haskell}
-(>>>) :: IMonad m => m p q a -> m q r b -> m p r b
-
-program = start
-    >>> declare
-            (Proxy :: Proxy "A")
-            (Proxy :: Proxy Integer)
-    >>> incr    (Proxy :: Proxy "A")
-    >>> lpush   (Proxy :: Proxy "L") 0
-    >>> lpop    (Proxy :: Proxy "L")
-\end{minted}
-
-\section{Discussions}
-\paragraph{Syntax}
-No one could ignore the glaring shortcoming of the syntax, which occurs mainly
- in two places: \emph{symbol singletons} and \emph{indexed monad}. We are hoping
- that these issues could be solved with future syntactic extensions.
-
-\paragraph{Returns only determined datatypes}
-
-All other data structures in Redis also follow the semantics similar to
- "List or nothing" as \mintinline{text}{LLEN} has. Take the case of
- \mintinline{text}{GET}, which has a similar semantics that should be typed:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-get :: (KnownSymbol s, Serialize x, StringOrNX xs s)
-    => Proxy s -> Popcorn xs xs (Either Reply (Maybe x))
-\end{minted}
-
-Since the key may not exist, we don't know what \mintinline{haskell}{x} would
-be. We could left \mintinline{haskell}{x} ambiguous, and let it be decided by
- the caller. But users will then be forced to spell out the complete
-type signature of everything, including the dictionaries, only to specify
-the desired resulting type.
-
-Instead of allowing the key to be non-existent, we require that the key must
- exist and it's associating type to be determined at compile time. So our
- version of \mintinline{haskell}{get} has a stricter semantics:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-get :: (KnownSymbol s, Serialize x
-     , Just (StringOf x) ~ Get xs s)
-    => Proxy s -> Popcorn xs xs (Either Reply (Maybe x))
-\end{minted}
-
-\paragraph{Commands with multiple inputs or outputs}
-
-Some command may take a variable number of arguments as inputs, and returns more
- than one value as outputs. To illustrate this, consider
- \mintinline{haskell}{sinter} in \mintinline{haskell}{Hedis}:
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-Hedis.sinter :: [ByteString]                      -- keys
-             -> Redis (Either Reply [ByteString]) -- values
-\end{minted}
-
-In Hedis such command could easily be expressed with lists of
- \mintinline{haskell}{ByteString}s. But in Popcorn, things escalate quickly, as
- the keys and values will have to be expressed with \emph{heterogeneous
- lists}\cite{hetero}, which would be pratically infeasible, considering the cost,
- if not impossible.
-
-Most importantly, the keys will all have to be constrained by
- \mintinline{haskell}{KnownSymbol}, which enforces these type literals to be
- concrete and known at compile time.
- It's still unclear whether this is possible.
-
-So instead, we are offering commands that only has a single input and output.
-
-\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
-sinter :: ByteString                      -- single key
-       -> Redis (Either Reply ByteString) -- single value
-\end{minted}
-
-Notice that \textbf{not all Hedis programs can be typechecked} (even if they
- might turn out to be type safe). We opted for type safety instead of
- expressiveness.
-
-\paragraph{Redis Transactions}
-
-Redis also provides \emph{transactions}, another context for executing commands.
-Redis transactions are atomic in the sense that, all commands in a transaction
- will executed sequentially, and no other requests issued by other clients will
- be served \textbf{in the middle}.\footnotemark In contrast, we cannot make such
- a guarantee in the ordinary context, which may destroy the assertions we made
- in types.
-
-At this point of writing, transactions are not supported in our implementation.
- We are planning to add it in the future, and we are expecting that there
- wouldn't be much difficulty, since we've implemented a runtime type checker
- specifically targeting Redis transactions once, before we moved on to the
- types.
-
 \section{Related Work}
+\label{sec:related}
 
@@
  Agda\cite{pi}, and then with singletons in Haskell\cite{singletons}.
-
-\section{Conclusions}
-
-We have demonstrated how a DSL with variable-bindings could be embedded in types.
-With more and more extensions added to the language, Haskell is gradually
- becoming a dependently-typed language,\footnote{Why not be dependently typed? \url{http://stackoverflow.com/questions/12961651/why-not-be-dependently-typed}}
- but it's not that dreadful as many (including us) would have thought.
-
-\paragraph{Acknowledgements}
-\footnote{Hidden for the purposes of double-blind reviewing}
-
-% references
-\bibliographystyle{abbrvnat}
-\bibliography{cites}
-
-\end{document}

### 2016-06-11 aplas16/sections/TypeLevelDict.tex
Created a directory for aplas16.

--- 2016-06-11 aplas16/sections/IndexedMonads.tex
@@
 
-\section{Indexed Monads}
-\label{sec:indexed-monads}
+\section{Type-Level Dictionaries}
+\label{sec:type-level-dict}
 
-Redis commands are \emph{actions}.
- We could capture the effects caused by an action, by expressing the states it
- affects, before and after. That is, the \emph{preconditions} and
- \emph{postconditions} of an action. In such way, we could also impose
- constraints on the preconditions.
+To check the bindings between keys and values, we need a \emph{dictionary-like}
+ structure, and encode it as a \emph{type} somehow.
 
-\emph{Indexed monads} (or \emph{monadish},
- \emph{parameterised monad})\cite{indexedmonad}
- can be used\cite{typefun,staticresources} to model such preconditions and
- postconditions in types. An indexed monad is a type constructor that takes
- three arguments: an initial state, a final state, and the type of a value that
- an action computes, which can be read like a Hoare triple\cite{kleisli}.
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-Class \ensuremath{\Conid{IMonad}} comes with two operations: \ensuremath{\Varid{unit}} for identities and \ensuremath{\Varid{bind}} for
-compositions, as in monads.
-
-We define a new datatype \ensuremath{\Conid{Popcorn}}, which is basically just
- the context \ensuremath{\Conid{Redis}} indexed with more information in types.
- We make it an instance of \ensuremath{\Conid{IMonad}}.
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[5]{}\hsindent{4}{}\<[9]%
-\>[9]{}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-The first and second argument of type \ensuremath{\Conid{Popcorn}} is where the
-dictionaries going to be.
-
-To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to
- \ensuremath{\Varid{unPopcorn}} to get an ordinary Hedis program back,
- with type information erased.
-
-\subsection{\text{PING}: the first attempt}
-
-In Redis, \text{PING} does nothing but replies with
- \text{PONG} if the connection is alive. In Hedis,
- \ensuremath{\Varid{ping}} has type:
+\subsection{Datatype promotion}
+Normally, at the term level, we could express the datatype of dictionary with
+\emph{type synonym} like this.\footnotemark
 
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ping}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
+\>[B]{}\mathbf{type}\;\Conid{Key}\mathrel{=}\Conid{String}{}\<[E]%
+\\
+\>[B]{}\mathbf{type}\;\Conid{Dictionary}\mathrel{=}[\mskip1.5mu (\Conid{Key},\Conid{TypeRep})\mskip1.5mu]{}\<[E]%
 \ColumnHook
@@
 
-Now with \ensuremath{\Conid{Popcorn}}, we could make our own version of
-\ensuremath{\Varid{ping}}\footnotemark
+\footnotetext{\ensuremath{\Conid{TypeRep}} supports term-level representations
+ of datatypes, available in \ensuremath{\Conid{\Conid{Data}.Typeable}}}
 
-\footnotetext{\ensuremath{\Varid{ping}} from Hedis is qualified with
-\ensuremath{\Conid{Hedis}} to prevent function name clashing in our code.}
+To encode this in the type level, everything has to be
+ \emph{promoted}\cite{promotion} one level up.
+ From terms to types, and from types to kinds.
+
+Luckily, with recently added GHC extension \emph{data kinds}, suitable
+ datatype will be automatically promoted to be a kind, and its value
+ constructors to be type constructors. The following type \ensuremath{\Conid{List}}
 
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ping}\mathbin{::}\Conid{Popcorn}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
-\\
-\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}{}\<[E]%
+\>[B]{}\mathbf{data}\;\Conid{List}\;\Varid{a}\mathrel{=}\Conid{Nil}\mid \thinspace'\Varid{Cons}\;\Varid{a}\;(\Conid{List}\;\Varid{a}){}\<[E]%
 \ColumnHook
@@
 
-The dictionary \ensuremath{\Varid{xs}} in the type remains unaffected after the
- action, because \ensuremath{\Varid{ping}} does not affect any key-type
- bindings. To encode other commands that modifies key-type bindings, we need
- type-level functions to annotate those effects on the dictionary.
+Give rise to the following kinds and type constructors:\footnote{To distinguish
+ between types and promoted constructors that have
+ ambiguous names, prefix promoted constructor with a single quote like
+ \ensuremath{\thinspace'\Varid{Nil}} and \ensuremath{\Conid{CONS}}}
+\footnote{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Conid{List}\;\Varid{k}\mathbin{::}\Conid{BOX}{}\<[E]%
+\\
+\>[B]{}\Conid{Nil}{}\<[6]%
+\>[6]{}\mathbin{::}\Conid{List}\;\Varid{k}{}\<[E]%
+\\
+\>[B]{}\thinspace'\Varid{Cons}\mathbin{::}\Varid{k}\to \Conid{List}\;\Varid{k}\to \Conid{List}\;\Varid{k}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Haskell sugars lists \ensuremath{[\mskip1.5mu \mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu]} and tuples
+ \ensuremath{(\mathrm{1},\text{\tt 'a'})} with brackets and parentheses.
+ We could also express promoted lists and tuples in types like this with
+ a single quote prefixed. For example:
+ \ensuremath{\thinspace'[\!\;\Conid{Int},\Conid{Char}\;\!]}, \ensuremath{\thinspace'(\!\;\Conid{Int},\Conid{Char}\;\!)}.
+
+\subsection{Type-level literals}
+
+Now we have type-level lists and tuples to construct the dictionary.
+For keys, \ensuremath{\Conid{String}} also has a type-level correspondence:
+\ensuremath{\Conid{Symbol}}.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{data}\;\Conid{Symbol}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Symbol is defined without a value constructor, because it's intended to be used
+ as a promoted kind.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\text{\tt \char34 this~is~a~type-level~string~literal\char34}\mathbin{::}\Conid{Symbol}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+% Nonetheless, it's still useful to have a term-level value that links with a
+%  Symbol, when we want to retrieve type-level information at runtime (but not the
+%  other way around!).
+
+\subsection{Putting Everything Together}
+
+With all of these ingredients ready, let's build some dictionaries!
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{type}\;\Conid{DictEmpty}\mathrel{=}\text{\tt '[]\;type~Dict0~=~'}{}\<[E]%
+\\
+\>[B]{}[\mskip1.5mu \text{\tt '(\char34 key\char34 ,~Bool)~]\;type~Dict1~=~'}{}\<[E]%
+\\
+\>[B]{}[\mskip1.5mu \text{\tt '(\char34 A\char34 ,~Int),~'}\;(\text{\tt \char34 B\char34},\text{\tt \char34 A\char34})\mskip1.5mu]{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+These dictionaries are defined with \emph{type synonym}, since they are
+ \emph{types}, not \emph{terms}. If we ask \text{GHCi} what is the
+ kind of \ensuremath{\Conid{Dict1}}, we will get \ensuremath{\Conid{Dict1}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}
+
+The kind \ensuremath{\mathbin{*}} (pronounced ``star'') stands for the set of all
+ concrete type expressions, such as \ensuremath{\Conid{Int}},
+ \ensuremath{\Conid{Char}} or even a symbol \ensuremath{\text{\tt \char34 symbol\char34}},
+ while \ensuremath{\Conid{Symbol}} is restricted to all symbols only.

### 2016-06-11 aplas16/main-aplas.tex
Revision up to Section 2.

@@
 %\usepackage{amsthm}
+\usepackage{color}
 \usepackage{url}
@@
 
-\title{Preventing Runtime Errors of Redis at Compile Time}
+\title{Safe Redis Queries}
+\subtitle{A Case Study of Type-Level Programming}
 \author{Ting-Yan Lai \inst{1}\inst{2}\and Tyng-Ruey Chuang \inst{1}
@@
 \begin{abstract}
-Programmers often interact with database systems by sending queries through libraries or packages in some programming languages. People, however, make mistakes.
-While some of the syntactic and semantic errors can be prevented by the
- language at compile time, or caught by the package at runtime, most semantic
- errors are just being ignored, causing problems at the database system.
-
-In this paper, we demonstrate how to prevent those runtime errors at compile
- time, thus allowing users to write more reliable database queries without
- runtime overhead, by exploiting type-level programming techniques such as
- indexed monad, type-level literals and closed type families in Haskell.
-
-The database system and the package we are targeting are \emph{Redis} and
- \emph{Hedis} respectively, and our implementation is available as \emph{Popcorn}
- on \emph{Hackage}.
+\Redis{} is an in-memory data strcture store, often used as a database, with a
+Haskell interface \Hedis{}. \Redis{} is dynamically typed --- a value associated
+with a key can be overwritten by a value of a different type, and a command,
+when fetching a value of a type it does not expect, signals a runtime error. We
+develop a domain-specific language that, by exploiting Haskell type-level
+programming techniques including indexed monad, type-level literals and closed
+type families, keeps track of types of values in the database and statically
+guarantees that type errors cannot happen.
 \end{abstract}
@@
 \input{sections/Introduction}
-\input{sections/Motivation}
+%\input{sections/Motivation}
 \input{sections/TypeLevelDict}

### 2016-06-11 aplas16/sections/Introduction.tex
Revision up to Section 2.

@@
 
-\subsection{Redis}
+\Redis{}\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as database, cache and message broker.
+A \Redis{} datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with.
+\Redis{} allows values of various types, including strings, hashes, lists, and sets, etc, to be stored, and provides a collection of of atomic \emph{commands} to manipulate these values.
 
-Redis\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as
-database, cache and message broker.
-A Redis datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with.
-Redis supports many different kind of values, such as strings, hashes, lists, and sets, etc, and provides a collection of of atomic \emph{commands} to manipulate these values.
-
-For an example, consider the following sequence of commands, entered through the interactive interface of Redis. The keys
-\text{some-set} abd \text{another-set} are
-both associated to a set. The two call to command \text{SADD} respectively adds three and two values to the two sets, before \text{SINTER} takes their intersection:
+For an example, consider the following sequence of commands, entered through the interactive interface of \Redis{}. The keys \texttt{some-set} and \texttt{another-set}
+are both associated to a set. The two call to command \texttt{SADD} respectively
+adds three and two values to the two sets, before \texttt{SINTER} takes their intersection:
 \begin{tabbing}\tt
@@
 
-Note that the keys \text{some-set} and \text{another-set}, if not existing before the call to \text{SADD}, are created on site. The call to
-\text{SADD} returns the size of the set after
-completion of the command.
+Note that the keys \texttt{some-set} and \texttt{another-set}, if not existing before the call to \texttt{SADD}, are created on site. The call to
+\texttt{SADD} returns the size of the set after completion of the command.
 
-\subsection{Hedis}
-
-Many third party library exist to allow general purpose programmings
-to access Redis databases through its TCP protocol. The most popular
-such library for Haskell is Hedis\footnote{\url{https://hackage.haskell.org/package/hedis}}.
-
-The following program is how the previous example looks like in
-Hedis:
-
+Many third party libraries provide interfaces that allow general purpose programming languages to access \Redis{} through its TCP protocol.
+For Haskell, the most popular library is \Hedis{}\footnote{\url{https://hackage.haskell.org/package/hedis}}.
+The following program implements the previous example:
 \begin{hscode}\SaveRestoreHook
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]{}\<[E]%
+\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The function \ensuremath{\Varid{sadd}\mathbin{::}\Conid{ByteString}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer})} takes a key and a list of values as arguments, and returns
+an \ensuremath{\Conid{Integer}} on success, or returns a \ensuremath{\Conid{Reply}}, a low-level representation of
+replies from the Redis server, in case of failures. All wrapped in the monad
+\ensuremath{\Conid{Redis}}, the context of command execution.\footnotemark
+
+Note that keys and values, being nothing but binary strings in Redis, are
+represented using Haskell \ensuremath{\Conid{ByteString}}. Values of other types must be encoded
+as \ensuremath{\Conid{ByteString}}s before being written to the database, and decoded after being
+read back.
+
+\footnotetext{\Hedis{} provides another kind of context, \ensuremath{\Conid{RedisTx}}, for \emph{transactions}, united with \ensuremath{\Conid{Redis}} under the class \ensuremath{\Conid{RedisCtx}}. We
+demonstrate only \ensuremath{\Conid{Redis}} in this paper.}
+
+%\paragraph{Motivation}
+%All binary strings are equal, but some binary strings are more equal than others.
+%While everything in \Redis{} is essentially a binary string, these strings
+%are treated differently.
+\Redis{} supports many different kind of data
+structures, such as strings, hashes, lists, etc. While they are all encoded as
+binary strings before being written to the database, most commands only works
+with data of certain types. In the following example, the key
+\texttt{some-string} is associated to string \texttt{foo} --- the command
+\texttt{SET} always associates a key to a string. The subsequent call to \texttt{SADD}, which adds a value to a set, thus causes a runtime error.
+\begin{tabbing}\tt
+~redis\char62{}~SET~some\char45{}string~foo\\
+\tt ~OK\\
+\tt ~redis\char62{}~SADD~some\char45{}string~bar\\
+\tt ~\char40{}error\char41{}~WRONGTYPE~Operation~against~a~key~holding~the~wrong\\
+\tt ~kind~of~value
+\end{tabbing}
+
+%\paragraph{Example 2} Even worse, not all strings are equal!
+% The call \texttt{INCR some-string} parses the string associated with key
+% \texttt{some-string} to an integer, increments it by one, and store it back as
+% a string. If the string can not be parse as an integer, a runtime error
+% is raised.
+% \begin{verbatim}
+% redis> SET some-string foo
+% OK
+% redis> INCR some-string
+% (error) ERR value is not an integer or out of range
+% \end{verbatim}
+
+Being a simple wrapper on top of the TCP protocol of \Redis{}, \Hedis{}
+inherits the problem. Executing following program yields the same error
+wrapped in Haskell: \ensuremath{\Conid{Left}\;(\Conid{Error}} \texttt{"WRONGTYPE Operation against a
+key holding the wrong kind of value"}\ensuremath{)}.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
+\\
+\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{set}\;\text{\tt \char34 some-string\char34}\;\text{\tt \char34 foo\char34}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-string\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34}\mskip1.5mu]~~.{}\<[E]%
 \ColumnHook
@@
 
-The function \ensuremath{\Varid{sadd}} takes a key and a list of values as
- arguments, and returns an \ensuremath{\Conid{Integer}} on success, or
- returns a \ensuremath{\Conid{Reply}}, a low-level representation of replies
- from the Redis server, in case of failures. All wrapped in
- \ensuremath{\Conid{Redis}}\footnotemark, the context of command execution.
+% \paragraph{The Cause} Every key is associated with a value, and every value has
+% its own type. But most commands in \Redis{} only work with a certain type of
+%  value. When a command is used on a wrong type of key, a runtime error occurs.
+%  The problems illustrated above arise from the absence of type checking, with
+%  respects to \textbf{the type of a value that associates with a key}.
+%  These problems could have been avoided, if we could know the type every key
+%  associates with in advance, and prevent programs with invalid commands from
+%  executing.
+%
+% \paragraph{Hedis as an embedded DSL}
+% Haskell makes it easy to build and use {\em domain specific embedded languages} (DSELs), and \Hedis{} can be regarded as one of them. What makes \Hedis{} peculiar is that,
+%  it has \emph{variable bindings} (between keys and values), but with very
+%  little or no semantic checking, neither dynamically nor statically.
+%
+Such a programming model is certainly very error-prone. Working within Haskell,
+a host language with a strong typing system, one naturally wishes to build a
+a domain-specific embedded language (DSEL) that exploits the rich type system
+of Haskell to ensure absence of \Redis{} type errors.
 
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{sadd}{}\<[7]%
-\>[7]{}\mathbin{::}\Conid{ByteString}{}\<[27]%
-\>[27]{}\mbox{\onelinecomment  key}{}\<[E]%
-\\
-\>[7]{}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]{}\<[27]%
-\>[27]{}\mbox{\onelinecomment  values}{}\<[E]%
-\\
-\>[7]{}\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-Note that keys and values, being nothing but binary strings in Redis,
-are represented using Haskell \ensuremath{\Conid{ByteString}}.
-Values of other types must be encoded as \ensuremath{\Conid{ByteString}}s before being written to the database,
-and decoded after being read back.
-
-\footnotetext{Hedis provides another kind of context, \ensuremath{\Conid{RedisTx}}, for \emph{transactions}, united with \ensuremath{\Conid{Redis}} under the class of \ensuremath{\Conid{RedisCtx}}.
-For brevity, we demonstrate only \ensuremath{\Conid{Redis}} in this paper.
-}
+This paper discusses the techniques we used and experiences we learned from building such a language, nicknamed \Popcorn{}. We constructed an {\em indexed
+monad}, on top of the monad \ensuremath{\Conid{Redis}}, which is indexed by a dictionary that
+maintains the set of currently defined keys and their types. To represent
+the dictionary, we need to encode variable binds with {\em type-level} lists
+and strings. To summarize our contributions:
+\begin{itemize}
+\item We present \Popcorn{}, a statically typed domain-specific language embedded in Haskell and built on \Hedis{}.
+% also makes Redis polymorphic by automatically converting back and forth from values of arbitrary types and boring ByteStrings.
+%
+\item We demonstrate how to model variable bindings of an embedded DSL using
+ language extensions including type-level literals and data kinds.
+%
+\item We provide (yet another) an example of encoding effects and constraints of
+in types, with indexed monad~\cite{indexedmonad}, closed type-families~\cite{closedtypefamilies} and constraints kinds~\cite{constraintskinds}.
+\end{itemize}
+\todo{Phrase this better.}

### 2016-06-11 aplas16/sections/TypeLevelDict.tex
Revision up to Section 2.

@@
 
-To check the bindings between keys and values, we need a \emph{dictionary-like}
- structure, and encode it as a \emph{type} somehow.
+One of the challenges of statically ensuring type correctness of \Redis{},
+which also presents in other stateful languages, is that the type of the value
+associated to a key can be altered after updating. To ensure type correctness,
+we have to keep track of the (\Redis{}) types of all existing keys in a
+{\em dictionary} --- conceptually, a list of pairs of keys and \Redis{} types.
+Each \Redis{} command is embedded in \Popcorn{} as a monadic computation. The
+monad, to be presented in Section \ref{sec:indexed-monads}, is indexed by
+the dictionaries before and after the computation. In a dependently typed
+programming language (without the so-called ``phase distinction'' ---
+separation between types and terms), this would pose no problem. In Haskell
+however, the dictionaries, to index a monad, has to be a Haskell type as well.
 
-\subsection{Datatype promotion}
+In this section we describe how to construct a type-level dictionary, to be
+used with the indexed monad in Section \ref{sec:indexed-monads}. More operations
+on the dictionary will be presented in Section \ref{sec:type-level-fun}.
+
+\subsection{Datatype Promotion}
+
 Normally, at the term level, we could express the datatype of dictionary with

### 2016-06-11 aplas16/sections/TypeLevelDict.tex
Tweaks in Section 2.

@@
 Each \Redis{} command is embedded in \Popcorn{} as a monadic computation. The
-monad, to be presented in Section \ref{sec:indexed-monads}, is indexed by
+monad, to be presented in Section~\ref{sec:indexed-monads}, is indexed by
 the dictionaries before and after the computation. In a dependently typed
@@
 In this section we describe how to construct a type-level dictionary, to be
-used with the indexed monad in Section \ref{sec:indexed-monads}. More operations
-on the dictionary will be presented in Section \ref{sec:type-level-fun}.
+used with the indexed monad in Section~\ref{sec:indexed-monads}. More operations
+on the dictionary will be presented in Section~\ref{sec:type-level-fun}.
 
@@
 
-Normally, at the term level, we could express the datatype of dictionary with
-\emph{type synonym} like this.\footnotemark
+% Normally, at the term level, we could express the datatype of dictionary with
+% \emph{type synonym} like this.\footnotemark
+%
+% \begin{spec}
+% type Key = String
+% type Dictionary = [(Key, TypeRep)]
+% \end{spec}
+%
+% \footnotetext{|TypeRep| supports term-level representations
+%  of datatypes, available in |Data.Typeable|}
 
+A datatype definition such as the one below:
 \begin{hscode}\SaveRestoreHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{type}\;\Conid{Key}\mathrel{=}\Conid{String}{}\<[E]%
-\\
-\>[B]{}\mathbf{type}\;\Conid{Dictionary}\mathrel{=}[\mskip1.5mu (\Conid{Key},\Conid{TypeRep})\mskip1.5mu]{}\<[E]%
+\>[B]{}\mathbf{data}\;\Conid{List}\;\Varid{a}\mathrel{=}\Conid{Nil}\mid \Conid{Cons}\;\Varid{a}\;(\Conid{List}\;\Varid{a})~~,{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
+is usually understood as having defined a type constructor \ensuremath{\Conid{List}}, and two value
+constructors \ensuremath{\Conid{Nil}} and \ensuremath{\Conid{Cons}}. For all lifted types \ensuremath{\Varid{a}}, \ensuremath{\Conid{List}\;\Varid{a}} is also a
+lifted type. Thus the {\em kind} of \ensuremath{\Conid{List}} is \ensuremath{\mathbin{*}\to \mathbin{*}}, where \ensuremath{\mathbin{*}} is the kind of
+all {\em lifted types} in Haskell. The two value constructors respectively
+have types \ensuremath{\Conid{Nil}\mathbin{::}\Conid{List}\;\Varid{a}} and \ensuremath{\Conid{Cons}\mathbin{::}\Varid{a}\to \Conid{List}\;\Varid{a}\to \Conid{List}\;\Varid{a}}.
 
-\footnotetext{\ensuremath{\Conid{TypeRep}} supports term-level representations
- of datatypes, available in \ensuremath{\Conid{\Conid{Data}.Typeable}}}
+The GHC extension \emph{data kinds}~\cite{promotion}, however, automatically
+promotes certain ``suitable'' types to kinds. With the extension, the \ensuremath{\mathbf{data}}
+definition has an alternative reading: for all kind \ensuremath{\Varid{k}}, \ensuremath{\Conid{List}\;\Varid{k}} is also a
+kind. \ensuremath{\Conid{Nil}} is a type, whose kind is \ensuremath{\Conid{List}\;\Varid{k}} for some \ensuremath{\Varid{k}}. Given a type \ensuremath{\Varid{x}}
+of kind \ensuremath{\Varid{k}} and a type \ensuremath{\Varid{xs}} of kind \ensuremath{\Conid{List}\;\Varid{k}}, \ensuremath{\Conid{Cons}\;\Varid{x}\;\Varid{xs}} is again a type of
+kind \ensuremath{\Conid{List}\;\Varid{k}}.
+Formally, for kind \ensuremath{\Varid{k}}, \ensuremath{\Conid{Nil}\mathbin{::}\Conid{List}\;\Varid{k}} and \ensuremath{\Conid{Cons}\mathbin{::}\Varid{k}\to \Conid{List}\;\Varid{k}\to \Conid{List}\;\Varid{k}}.
+Whether a constructor is promoted can often be inferred
+from the context. To be more specific, prefixing a contructor with
+a single quote, such as in \ensuremath{\thinspace'\Conid{Nil}} and \ensuremath{\thinspace'\Conid{Cons}}, denotes that it is promoted.
 
-To encode this in the type level, everything has to be
- \emph{promoted}\cite{promotion} one level up.
- From terms to types, and from types to kinds.
+The built-in lists in Haskell has two constructors \ensuremath{[\mskip1.5mu \mskip1.5mu]} and \ensuremath{(\mathbin{:})}, and
+\ensuremath{[\mskip1.5mu \mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu]} is take as a syntax sugar for \ensuremath{\mathrm{1}\mathbin{:}(\mathrm{2}\mathbin{:}(\mathrm{3}\mathbin{:}[\mskip1.5mu \mskip1.5mu]))}.
+The type and
+data constructors can also be promoted. Given kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]} is also a kind.
+The type constructor 
+For example, since \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, etc., all have kind \ensuremath{\mathbin{*}}, \ensuremath{\Conid{Int}\mathbin{\thinspace':}(\Conid{Char}\mathbin{\thinspace':}(\Conid{Bool}\mathbin{\thinspace':}\thinspace'[]))} is a type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} -- a list of types.
+The same list can be denoted by a syntax
+sugar \ensuremath{\thinspace'[\!\;\Conid{Int},\Conid{Char},\Conid{Bool}\;\!]}.
 
-Luckily, with recently added GHC extension \emph{data kinds}, suitable
- datatype will be automatically promoted to be a kind, and its value
- constructors to be type constructors. The following type \ensuremath{\Conid{List}}
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{data}\;\Conid{List}\;\Varid{a}\mathrel{=}\Conid{Nil}\mid \thinspace'\Varid{Cons}\;\Varid{a}\;(\Conid{List}\;\Varid{a}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-Give rise to the following kinds and type constructors:\footnote{To distinguish
- between types and promoted constructors that have
- ambiguous names, prefix promoted constructor with a single quote like
- \ensuremath{\thinspace'\Varid{Nil}} and \ensuremath{\Conid{CONS}}}
-\footnote{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Conid{List}\;\Varid{k}\mathbin{::}\Conid{BOX}{}\<[E]%
-\\
-\>[B]{}\Conid{Nil}{}\<[6]%
-\>[6]{}\mathbin{::}\Conid{List}\;\Varid{k}{}\<[E]%
-\\
-\>[B]{}\thinspace'\Varid{Cons}\mathbin{::}\Varid{k}\to \Conid{List}\;\Varid{k}\to \Conid{List}\;\Varid{k}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
+sugar for \ensuremath{\mathrm{1}\mathbin{\thinspace':}(\mathrm{2}\mathbin{\thinspace':}(\mathrm{3}\mathbin{\thinspace':}\thinspace'[]))}
 Haskell sugars lists \ensuremath{[\mskip1.5mu \mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu]} and tuples

### 2016-06-12 aplas16/sections/TypeLevelDict.tex
Section 2 done.

--- 2016-06-11 aplas16/sections/TypeLevelDict.tex
@@
 In this section we describe how to construct a type-level dictionary, to be
-used with the indexed monad in Section \ref{sec:indexed-monads}. More operations
-on the dictionary will be presented in Section \ref{sec:type-level-fun}.
+used with the indexed monad in Section~\ref{sec:indexed-monads}. More operations
+on the dictionary will be presented in Section~\ref{sec:type-level-fun}.
 
-\subsection{Datatype Promotion}
-
-Normally, at the term level, we could express the datatype of dictionary with
-\emph{type synonym} like this.\footnotemark
-
+Haskell maintains the distinction between values, types, and kinds: values are
+categorized by types, and types are categorized by kinds. The kinds are relatively simple: \ensuremath{\mathbin{*}} is the kind of all {\em lifted} types, while type
+constructors have kinds such as \ensuremath{\mathbin{*}\to \mathbin{*}}, \ensuremath{\mathbin{*}\to \mathbin{*}\to \mathbin{*}}, etc. Consider the
+datatype definitions below:
 \begin{hscode}\SaveRestoreHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{type}\;\Conid{Key}\mathrel{=}\Conid{String}{}\<[E]%
-\\
-\>[B]{}\mathbf{type}\;\Conid{Dictionary}\mathrel{=}[\mskip1.5mu (\Conid{Key},\Conid{TypeRep})\mskip1.5mu]{}\<[E]%
+\>[B]{}\mathbf{data}\;\Conid{Nat}\mathrel{=}\Conid{Zero}\mid \Conid{Suc}\;\Conid{Nat}~~,\qquad\;\mathbf{data}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\mathrel{=}[\mskip1.5mu \mskip1.5mu]\mid \Varid{a}\mathbin{:}[\mskip1.5mu \Varid{a}\mskip1.5mu]~~.{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
+The lefthand side is usually seen as having defined a type \ensuremath{\Conid{Nat}\mathbin{::}\mathbin{*}},
+and two value constructors \ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}}. The righthand
+side is how Haskell lists are understood. The {\em kind} of \ensuremath{[\mskip1.5mu \mathbin{\cdot}\mskip1.5mu]} is \ensuremath{\mathbin{*}\to \mathbin{*}},
+since it takes a lifted type \ensuremath{\Varid{a}} to a lifted type \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]}.
+The two value constructors respectively have types \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{(\mathbin{:})\mathbin{::}\Varid{a}\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]}, for all type \ensuremath{\Varid{a}}.
 
-\footnotetext{\ensuremath{\Conid{TypeRep}} supports term-level representations
- of datatypes, available in \ensuremath{\Conid{\Conid{Data}.Typeable}}}
+The GHC extension \emph{data kinds}~\cite{promotion}, however, automatically
+promotes certain ``suitable'' types to kinds.\footnote{It is only informally
+described in the GHC manual what types are ``suitable''.} With the extension,
+the \ensuremath{\mathbf{data}} definitions above has an alternative reading: \ensuremath{\Conid{Nat}} is a new kind,
+\ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} is a type having kind \ensuremath{\Conid{Nat}}, and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}} is a type
+constructor, taking a type in kind \ensuremath{\Conid{Nat}} to another type in \ensuremath{\Conid{Nat}}. Whether a
+constructor is promoted can often be inferred from the context. To be more specific, prefixing a constructor with a single quote, such as in \ensuremath{\mbox{\textquotesingle}\Conid{Zero}} and
+\ensuremath{\mbox{\textquotesingle}\Conid{Suc}}, denotes that it is promoted.
 
-To encode this in the type level, everything has to be
- \emph{promoted}\cite{promotion} one level up.
- From terms to types, and from types to kinds.
+The situation of lists is similar: for all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]} is also a kind. For
+all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{k}\mskip1.5mu]} is a type. Given a type \ensuremath{\Varid{x}}
+of kind \ensuremath{\Varid{k}} and a type \ensuremath{\Varid{xs}} of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}, \ensuremath{\Varid{x}\mathbin{:}\Varid{xs}} is again a type of
+kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}.
+Formally, \ensuremath{(\mathbin{:})\mathbin{::}\Varid{k}\to [\mskip1.5mu \Varid{k}\mskip1.5mu]\to [\mskip1.5mu \Varid{k}\mskip1.5mu]}.
+For example,
+\ensuremath{\Conid{Int}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Char}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[]))} is a type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} --- a list of
+(lifted) types. The optional quote denotes that the constructors are promoted.
+The same list can be denoted by a syntax sugar \ensuremath{\mbox{\textquotesingle}[\Conid{Int},\Conid{Char},\Conid{Bool}]}.
 
-Luckily, with recently added GHC extension \emph{data kinds}, suitable
- datatype will be automatically promoted to be a kind, and its value
- constructors to be type constructors. The following type \ensuremath{\Conid{List}}
+Tuples are also promoted. Thus we may put two types in a pair to form another
+type, such as in \ensuremath{\mbox{\textquotesingle}(\Conid{Int},\Conid{Char})}, a type having kind \ensuremath{(\mathbin{*},\mathbin{*})}.
 
+Strings in Haskell are nothing but lists of \ensuremath{\Conid{Char}}s. Regarding promotion,
+however, a \ensuremath{\Conid{String}} can be promoted to a type having kind \ensuremath{\Conid{Symbol}}. \ensuremath{\Conid{Symbol}} is
+a type without a constructor: \ensuremath{\mathbf{data}\;\Conid{Symbol}},
+intended to be used as a promoted kind. In the expression:
 \begin{hscode}\SaveRestoreHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{data}\;\Conid{List}\;\Varid{a}\mathrel{=}\Conid{Nil}\mid \thinspace'\Varid{Cons}\;\Varid{a}\;(\Conid{List}\;\Varid{a}){}\<[E]%
+\>[B]{}\text{\tt \char34 this~is~a~type-level~string~literal\char34}\mathbin{::}\Conid{Symbol}~~,{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
+the string on the lefthand side of \ensuremath{(\mathbin{::})} is a type, having kind \ensuremath{\Conid{Symbol}}.
 
-Give rise to the following kinds and type constructors:\footnote{To distinguish
- between types and promoted constructors that have
- ambiguous names, prefix promoted constructor with a single quote like
- \ensuremath{\thinspace'\Varid{Nil}} and \ensuremath{\Conid{CONS}}}
-\footnote{All kinds have \emph{sort} BOX in Haskell\cite{sorts}}
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Conid{List}\;\Varid{k}\mathbin{::}\Conid{BOX}{}\<[E]%
-\\
-\>[B]{}\Conid{Nil}{}\<[6]%
-\>[6]{}\mathbin{::}\Conid{List}\;\Varid{k}{}\<[E]%
-\\
-\>[B]{}\thinspace'\Varid{Cons}\mathbin{::}\Varid{k}\to \Conid{List}\;\Varid{k}\to \Conid{List}\;\Varid{k}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-Haskell sugars lists \ensuremath{[\mskip1.5mu \mathrm{1},\mathrm{2},\mathrm{3}\mskip1.5mu]} and tuples
- \ensuremath{(\mathrm{1},\text{\tt 'a'})} with brackets and parentheses.
- We could also express promoted lists and tuples in types like this with
- a single quote prefixed. For example:
- \ensuremath{\thinspace'[\!\;\Conid{Int},\Conid{Char}\;\!]}, \ensuremath{\thinspace'(\!\;\Conid{Int},\Conid{Char}\;\!)}.
-
-\subsection{Type-level literals}
-
-Now we have type-level lists and tuples to construct the dictionary.
-For keys, \ensuremath{\Conid{String}} also has a type-level correspondence:
-\ensuremath{\Conid{Symbol}}.
-
+With all of these ingredients, we are ready to build our dictionaries, or
+type-level associate lists:
 \begin{hscode}\SaveRestoreHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{data}\;\Conid{Symbol}{}\<[E]%
+\>[B]{}\mathbf{type}\;\Conid{DictEmpty}\mathrel{=}\mbox{\textquotesingle}[]~~,{}\<[E]%
+\\
+\>[B]{}\mathbf{type}\;\Conid{Dict0}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 key\char34},\Conid{Bool})]~~,{}\<[E]%
+\\
+\>[B]{}\mathbf{type}\;\Conid{Dict1}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int}),\mbox{\textquotesingle}(\text{\tt \char34 B\char34},\text{\tt \char34 A\char34})]~~.{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
-
-Symbol is defined without a value constructor, because it's intended to be used
- as a promoted kind.
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\text{\tt \char34 this~is~a~type-level~string~literal\char34}\mathbin{::}\Conid{Symbol}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-%
-% Nonetheless, it's still useful to have a term-level value that links with a
-%  Symbol, when we want to retrieve type-level information at runtime (but not the
-%  other way around!).
-
-\subsection{Putting Everything Together}
-
-With all of these ingredients ready, let's build some dictionaries!
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{type}\;\Conid{DictEmpty}\mathrel{=}\text{\tt '[]\;type~Dict0~=~'}{}\<[E]%
-\\
-\>[B]{}[\mskip1.5mu \text{\tt '(\char34 key\char34 ,~Bool)~]\;type~Dict1~=~'}{}\<[E]%
-\\
-\>[B]{}[\mskip1.5mu \text{\tt '(\char34 A\char34 ,~Int),~'}\;(\text{\tt \char34 B\char34},\text{\tt \char34 A\char34})\mskip1.5mu]{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-These dictionaries are defined with \emph{type synonym}, since they are
- \emph{types}, not \emph{terms}. If we ask \text{GHCi} what is the
- kind of \ensuremath{\Conid{Dict1}}, we will get \ensuremath{\Conid{Dict1}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}
-
-The kind \ensuremath{\mathbin{*}} (pronounced ``star'') stands for the set of all
- concrete type expressions, such as \ensuremath{\Conid{Int}},
- \ensuremath{\Conid{Char}} or even a symbol \ensuremath{\text{\tt \char34 symbol\char34}},
- while \ensuremath{\Conid{Symbol}} is restricted to all symbols only.
+All the entities defined above are types, where \ensuremath{\Conid{Dict0}} and \ensuremath{\Conid{Dict1}}
+have kind \ensuremath{[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}.

### 2016-06-12 aplas16/sections/IndexedMonads.tex
Halfway editing the section in indexed monads.

@@
 
-Redis commands are \emph{actions}.
- We could capture the effects caused by an action, by expressing the states it
- affects, before and after. That is, the \emph{preconditions} and
- \emph{postconditions} of an action. In such way, we could also impose
- constraints on the preconditions.
+Stateful computations are often reasoned in a Hoare-logic style: each command
+is labelled by a \emph{precondition} and a \emph{postcondition}. If the former
+is satisfied before the command is executed, the latter is guaranteed to hold
+afterwards.
 
-\emph{Indexed monads} (or \emph{monadish},
- \emph{parameterised monad})\cite{indexedmonad}
- can be used\cite{typefun,staticresources} to model such preconditions and
- postconditions in types. An indexed monad is a type constructor that takes
- three arguments: an initial state, a final state, and the type of a value that
- an action computes, which can be read like a Hoare triple\cite{kleisli}.
-
+In Haskell, stateful computations are represented by monads. In order to
+reason about their behaviors within the type system, we wish to label a state
+monad with its pre and postcondition. An \emph{indexed
+monad}~\cite{indexedmonad} (also called \emph{monadish} or
+\emph{parameterised monad}) is a monad that, in addition to the type of value
+it computes, takes two more type arguments representing an initial state and
+a final state, to be interpreted like a Hoare triple~\cite{kleisli}:
 \begin{hscode}\SaveRestoreHook
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}{}\<[E]%
+\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}~~.{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
+The intention is that a computation of type \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is a stateful computation
+such that if it starts execution in a state satisfying \ensuremath{\Varid{p}} and terminates, it
+yields a value of type \ensuremath{\Varid{a}}, and the new state satisfies \ensuremath{\Varid{q}}.
+The operator \ensuremath{\Varid{unit}} lifts a pure computation to a stateful computation that
+does not alter the state. In \ensuremath{\Varid{x}\mathbin{`\Varid{bind}`}\Varid{f}}, a computation \ensuremath{\Varid{x}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} must
+be chained before \ensuremath{\Varid{f}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b}}, which expects a value of type \ensuremath{\Varid{a}} and
+a state satisfying \ensuremath{\Varid{q}} and, if terminates, ends in a state satisfying \ensuremath{\Varid{r}}.
+The result is a monad \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}} --- a computation that, if executed in a state
+satisfying \ensuremath{\Varid{p}} and terminates, yields a value \ensuremath{\Varid{b}} and a state satisfying \ensuremath{\Varid{r}}.
+Indexed monads have been used ~\cite{typefun,staticresources} ... \todo{for what? Some discriptions here to properly cite them.}
 
-Class \ensuremath{\Conid{IMonad}} comes with two operations: \ensuremath{\Varid{unit}} for identities and \ensuremath{\Varid{bind}} for
-compositions, as in monads.
-
-We define a new datatype \ensuremath{\Conid{Popcorn}}, which is basically just
- the context \ensuremath{\Conid{Redis}} indexed with more information in types.
- We make it an instance of \ensuremath{\Conid{IMonad}}.
-
+We define a new indexed monad \ensuremath{\Conid{Popcorn}} which, at term level, merely wraps
+\ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the pre and
+postconditions at type level:
 \begin{hscode}\SaveRestoreHook
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}{}\<[E]%
-\\
-\>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
+\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}~~,{}\<[E]%
 \\[\blanklineskip]%
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}{}\<[E]%
-\\
-\>[5]{}\hsindent{4}{}\<[9]%
-\>[9]{}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f}){}\<[E]%
+\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f})~~.{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
+To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to \ensuremath{\Varid{unPopcorn}} to erase the additional type information and get back an ordinary \Hedis{} program.
 
-The first and second argument of type \ensuremath{\Conid{Popcorn}} is where the
-dictionaries going to be.
-
-To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to
- \ensuremath{\Varid{unPopcorn}} to get an ordinary Hedis program back,
- with type information erased.
-
-\subsection{\text{PING}: the first attempt}
-
-In Redis, \text{PING} does nothing but replies with
+\paragraph{\text{PING}: A First Example}
+In \Redis{}, \text{PING} does nothing but replies with
  \text{PONG} if the connection is alive. In Hedis,

### 2016-06-12 aplas16/sections/IndexedMonads.tex
Half of Section 3.3.

@@
 
-Stateful computations are often reasoned in a Hoare-logic style: each command
-is labelled by a \emph{precondition} and a \emph{postcondition}. If the former
-is satisfied before the command is executed, the latter is guaranteed to hold
-afterwards.
+Stateful computations are often reasoned using Hoare logic. A {\em Hoare triple}
+$\{P\} S \{Q\}$ denotes such a proposition: if the statement $S$ is executed in
+a state satisfying prediate $P$, when it terminates, the state must satisfy
+predicate $Q$. Predicates $P$ and $Q$ are respectively is called the
+\emph{precondition} and the \emph{postcondition} of the Hoare triple.
 
@@
 reason about their behaviors within the type system, we wish to label a state
-monad with its pre and postcondition. An \emph{indexed
-monad}~\cite{indexedmonad} (also called \emph{monadish} or
-\emph{parameterised monad}) is a monad that, in addition to the type of value
-it computes, takes two more type arguments representing an initial state and
-a final state, to be interpreted like a Hoare triple~\cite{kleisli}:
+monad with its pre and postcondition. An \emph{indexed monad}~%
+\cite{indexedmonad} (also called \emph{monadish} or \emph{parameterised monad})
+is a monad that, in addition to the type of value it computes, takes two more
+type arguments representing an initial state and a final state, to be
+interpreted like a Hoare triple~\cite{kleisli}:
 \begin{hscode}\SaveRestoreHook
@@
 The intention is that a computation of type \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is a stateful computation
-such that if it starts execution in a state satisfying \ensuremath{\Varid{p}} and terminates, it
-yields a value of type \ensuremath{\Varid{a}}, and the new state satisfies \ensuremath{\Varid{q}}.
-The operator \ensuremath{\Varid{unit}} lifts a pure computation to a stateful computation that
-does not alter the state. In \ensuremath{\Varid{x}\mathbin{`\Varid{bind}`}\Varid{f}}, a computation \ensuremath{\Varid{x}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} must
-be chained before \ensuremath{\Varid{f}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b}}, which expects a value of type \ensuremath{\Varid{a}} and
-a state satisfying \ensuremath{\Varid{q}} and, if terminates, ends in a state satisfying \ensuremath{\Varid{r}}.
-The result is a monad \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}} --- a computation that, if executed in a state
-satisfying \ensuremath{\Varid{p}} and terminates, yields a value \ensuremath{\Varid{b}} and a state satisfying \ensuremath{\Varid{r}}.
+such that, if it starts execution in a state satisfying \ensuremath{\Varid{p}} and terminates, it
+yields a value of type \ensuremath{\Varid{a}}, and the new state satisfies \ensuremath{\Varid{q}}. The operator \ensuremath{\Varid{unit}}
+lifts a pure computation to a stateful computation that does not alter the
+state. In \ensuremath{\Varid{x}\mathbin{`\Varid{bind}`}\Varid{f}}, a computation \ensuremath{\Varid{x}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is followed by
+\ensuremath{\Varid{f}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b}} --- the postcondition of \ensuremath{\Varid{x}} matches the precondition of
+the computation returned by \ensuremath{\Varid{f}}. The result is a monad \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}}.
 Indexed monads have been used ~\cite{typefun,staticresources} ... \todo{for what? Some discriptions here to properly cite them.}
@@
 We define a new indexed monad \ensuremath{\Conid{Popcorn}} which, at term level, merely wraps
-\ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the pre and
-postconditions at type level:
+\ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the
+pre/postconditions at type level:
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to \ensuremath{\Varid{unPopcorn}} to erase the additional type information and get back an ordinary \Hedis{} program.
+At term level, the \ensuremath{\Varid{unit}} and \ensuremath{\Varid{bind}} methods are not interesting: they merely
+make calls to \ensuremath{\Varid{return}} and \ensuremath{(\bind )} of \ensuremath{\Conid{Redis}}, and extracts and re-apply the constructor \ensuremath{\Conid{Popcorn}} when necessary.
+With \ensuremath{\Conid{Popcorn}} being a \ensuremath{\mathbf{newtype}}, they
+can be optimized away in runtime. The interesting bits happen in compile type,
+on the added type information.
 
-\paragraph{\text{PING}: A First Example}
-In \Redis{}, \text{PING} does nothing but replies with
- \text{PONG} if the connection is alive. In Hedis,
- \ensuremath{\Varid{ping}} has type:
+The properties of the state we care about are the set of currently allocated
+keys and their associated types. We will present, in Section~\ref{sec:type-level-dict}, techniques that allow us to specify
+properties such as ``the keys in the database are \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}},
+respectively associated to values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and \ensuremath{\Conid{Bool}}.''
+For now, however, let us look at the simplest \Redis{} command.
 
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ping}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-Now with \ensuremath{\Conid{Popcorn}}, we could make our own version of
-\ensuremath{\Varid{ping}}\footnotemark
-
-\footnotetext{\ensuremath{\Varid{ping}} from Hedis is qualified with
-\ensuremath{\Conid{Hedis}} to prevent function name clashing in our code.}
-
+The command \text{PING} in \Redis{} does nothing but replies a message
+\text{PONG} if the connection is alive. In \Hedis{}, \ensuremath{\Varid{ping}} has type
+\ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status})}. The \Popcorn{} version of \ensuremath{\Varid{ping}} simply
+applys an additional constructor (functions from \Hedis{} are qualified with
+\ensuremath{\Conid{Hedis}} to prevent name clashing):
 \begin{hscode}\SaveRestoreHook
@@
 \\
-\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}{}\<[E]%
+\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}~~.{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
-
-The dictionary \ensuremath{\Varid{xs}} in the type remains unaffected after the
- action, because \ensuremath{\Varid{ping}} does not affect any key-type
- bindings. To encode other commands that modifies key-type bindings, we need
- type-level functions to annotate those effects on the dictionary.
+Since \ensuremath{\Varid{ping}} does not alter the database, the postcondition and precondition
+are the same. Commands that are more interesting will be introduced after
+we present our type-level encoding of constraints on states.

### 2016-06-12 aplas16/sections/TypeLevelDict.tex
Half of Section 3.3.

@@
 associated to a key can be altered after updating. To ensure type correctness,
-we have to keep track of the (\Redis{}) types of all existing keys in a
-{\em dictionary} --- conceptually, a list of pairs of keys and \Redis{} types.
-Each \Redis{} command is embedded in \Popcorn{} as a monadic computation. The
-monad, to be presented in Section~\ref{sec:indexed-monads}, is indexed by
-the dictionaries before and after the computation. In a dependently typed
+we keep track of the types of all existing keys in a {\em dictionary} ---
+conceptually, an associate list, or a list of pairs of keys and \Redis{} types.
+For example, the dictionary \ensuremath{[\mskip1.5mu (\text{\tt \char34 A\char34},\Conid{Int}),(\text{\tt \char34 B\char34},\Conid{Char}),(\text{\tt \char34 C\char34},\Conid{Bool})\mskip1.5mu]} represents
+a predicate stating that ``the keys \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}} are respectively
+associated to values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and \ensuremath{\Conid{Bool}}.''
+
+The dictionary mixes values (strings such as \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}) and types. Further
+more, as mentioned in Section~\ref{sec:indexed-monads}, the dictionaries are
+to be used parameters to the indexed monad \ensuremath{\Conid{Popcorn}}. In a dependently typed
 programming language (without the so-called ``phase distinction'' ---
 separation between types and terms), this would pose no problem. In Haskell
-however, the dictionaries, to index a monad, has to be a Haskell type as well.
+however, the dictionaries, to index a monad, has to be a type as well.
 
 In this section we describe how to construct a type-level dictionary, to be
-used with the indexed monad in Section~\ref{sec:indexed-monads}. More operations
-on the dictionary will be presented in Section~\ref{sec:type-level-fun}.
+used with the indexed monad in Section~\ref{sec:indexed-monads}.
+
+\subsection{Datatype Promotion}
 
@@
 For example,
-\ensuremath{\Conid{Int}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Char}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[]))} is a type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} --- a list of
+\ensuremath{\Conid{Int}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Char}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[~]))} is a type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} --- a list of
 (lifted) types. The optional quote denotes that the constructors are promoted.
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{type}\;\Conid{DictEmpty}\mathrel{=}\mbox{\textquotesingle}[]~~,{}\<[E]%
+\>[B]{}\mathbf{type}\;\Conid{DictEmpty}\mathrel{=}\mbox{\textquotesingle}[~]~~,{}\<[E]%
 \\
@@
 have kind \ensuremath{[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}.
+
+\subsection{Type-Level Functions}
+
+Now that we can represent dictionaries as types, the next step is to define
+operations on them. A function that inserts an entry to a dictionary, for
+example, is a function from a type to a type. While it was shown that it is
+possible to simulate type-level functions using Haskell type
+classes~\cite{McBride:02:Faking}, in recent versions of GHC, {\em indexed type
+families}, or type families for short, are considered a cleaner solution.
+
+For example, compare conjunction \ensuremath{(\mathrel{\wedge})} and its type-level
+counterpart \ensuremath{\Conid{And}}:\\
+\noindent{\centering %\small
+\begin{minipage}[b]{0.35\linewidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}c<{\hspost}@{}}%
+\column{7E}{@{}l@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathrel{\wedge})\mathbin{::}\Conid{Bool}\to \Conid{Bool}\to \Conid{Bool}{}\<[E]%
+\\
+\>[B]{}\Conid{True}{}\<[7]%
+\>[7]{}\mathrel{\wedge}{}\<[7E]%
+\>[11]{}\Conid{True}{}\<[17]%
+\>[17]{}\mathrel{=}\Conid{True}{}\<[E]%
+\\
+\>[B]{}\Varid{a}{}\<[7]%
+\>[7]{}\mathrel{\wedge}{}\<[7E]%
+\>[11]{}\Varid{b}{}\<[17]%
+\>[17]{}\mathrel{=}\Conid{False}~~,{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\begin{minipage}[b]{0.55\linewidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{And}\;(\Varid{a}\mathbin{::}\Conid{Bool})\;(\Varid{b}\mathbin{::}\Conid{Bool})\mathbin{::}\Conid{Bool}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathbf{where}\;{}\<[10]%
+\>[10]{}\Conid{And}\;{}\<[15]%
+\>[15]{}\Conid{True}\;{}\<[21]%
+\>[21]{}\Conid{True}{}\<[27]%
+\>[27]{}\mathrel{=}\Conid{True}{}\<[E]%
+\\
+\>[10]{}\Conid{And}\;{}\<[15]%
+\>[15]{}\Varid{a}\;{}\<[21]%
+\>[21]{}\Varid{b}{}\<[27]%
+\>[27]{}\mathrel{=}\Conid{False}~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+}\\
+The lefthand side is a typical definition of \ensuremath{(\mathrel{\wedge})} by pattern matching.
+On the righthand side, \ensuremath{\Conid{Bool}} is not a type, but a type lifted to a kind,
+while \ensuremath{\Conid{True}} and \ensuremath{\Conid{False}} are types of kind \ensuremath{\Conid{Bool}}. The declaration says
+that \ensuremath{\Conid{And}} is a family of types, indexed by two parameters \ensuremath{\Varid{a}} and \ensuremath{\Varid{b}} of
+kind \ensuremath{\Conid{Bool}}. The type with index \ensuremath{\Conid{True}} and \ensuremath{\Conid{True}} is \ensuremath{\Conid{True}}, and all
+other indices lead to \ensuremath{\Conid{False}}.
+For our purpose, we can read \ensuremath{\Conid{And}} as a type-level function.
+Observe how it resembles the term-level \ensuremath{(\mathrel{\wedge})}.
+
+Note that type families in Haskell come in many flavors. Families can be
+defined for \ensuremath{\mathbf{data}} and \ensuremath{\mathbf{type}} synonym. They can appear inside type
+classes~\cite{tfclass,tfsynonym} or at toplevel. Toplevel type families
+can be open~\cite{tfopen} or closed~\cite{tfclosed}. The flavor we chose
+is top-level, closed type synonym family, since it allows overlapping
+instances, and we need none of the extensibility provided by open type
+families. Notice that the instance \ensuremath{\Conid{And}\;\Conid{True}\;\Conid{True}} could be subsumed under
+the more general instance, \ensuremath{\Conid{And}\;\Varid{a}\;\Varid{b}}. In a closed type family we may resolve
+the overlapping in order, just like how cases overlapping is resolved in term-level functions.
+
+%\subsection{Functions on Type-Level Dictionaries}
+
+We are now able to define operations on type-level dictionaries.
+Let's begin with dictionary lookup.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{32}{@{}>{\hspre}c<{\hspost}@{}}%
+\column{32E}{@{}l@{}}%
+\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Get}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\mathbin{::}\mathbin{*}\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{s}{}\<[32]%
+\>[32]{}\mathrel{=}{}\<[32E]%
+\>[35]{}\Varid{x}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{s}{}\<[32]%
+\>[32]{}\mathrel{=}{}\<[32E]%
+\>[35]{}\Conid{Get}\;\Varid{xs}\;\Varid{s}~~{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The type-level function \ensuremath{\Conid{Get}} returns the entry associated with key \ensuremath{\Varid{s}} in the
+dictionary \ensuremath{\Varid{xs}}. Notice, in the first case, how type-level equality can be
+expressed by unifying type variables with the same name.
+Note also that \ensuremath{\Conid{Get}} is a partial function on types:
+\ensuremath{\Conid{Get}\;\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int})]\;\text{\tt \char34 A\char34}} evaluates to \ensuremath{\Conid{Int}}, but
+\ensuremath{\Conid{Get}\;\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int})]\;\text{\tt \char34 B\char34}} gets stuck.
+... and these types are computed at compile-time. It wouldn't make
+much sense for a type checker to crash and throw a ``Non-exhaustive'' error or
+be non-terminating.
+\todo{So, what exactly happens when we do the second?}
+
+We could make \ensuremath{\Conid{Get}} total, as we would at the term level, with \ensuremath{\Conid{Maybe}}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Get}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\mathbin{::}\Conid{Maybe}\mathbin{*}\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Get}\;\mbox{\textquotesingle}[~]\;{}\<[29]%
+\>[29]{}\Varid{s}\mathrel{=}\Conid{Nothing}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
+\>[29]{}\Varid{s}\mathrel{=}\Conid{Just}\;\Varid{x}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
+\>[29]{}\Varid{s}\mathrel{=}\Conid{Get}\;\Varid{xs}\;\Varid{s}~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+%
+Some other dictionary-related functions are defined in a similar fashion
+in Figure \ref{fig:dict-operations}. The function \ensuremath{\Conid{Set}} either updates an
+existing entry or inserts a new entry, \ensuremath{\Conid{Del}} removes an entry matching
+a given key, while \ensuremath{\Conid{Member}} checks whether a given key exists in the
+dictionary.
+
+\begin{figure}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mbox{\onelinecomment  inserts or updates an entry}{}\<[E]%
+\\
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Set}\;{}\<[18]%
+\>[18]{}(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\;(\Varid{x}\mathbin{::}\mathbin{*})\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Set}\;\mbox{\textquotesingle}[~]\;{}\<[29]%
+\>[29]{}\Varid{s}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\Varid{s},\Varid{x})]{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Set}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
+\>[29]{}\Varid{s}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}(\Varid{s},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Set}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
+\>[29]{}\Varid{s}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Conid{Set}\;\Varid{xs}\;\Varid{s}\;\Varid{x}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mbox{\onelinecomment  removes an entry}{}\<[E]%
+\\
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Del}\;{}\<[18]%
+\>[18]{}(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Del}\;\Conid{Nil}\;{}\<[29]%
+\>[29]{}\Varid{s}{}\<[32]%
+\>[32]{}\mathrel{=}\Conid{Nil}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Del}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{s}{}\<[32]%
+\>[32]{}\mathrel{=}\Varid{xs}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Del}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{s}{}\<[32]%
+\>[32]{}\mathrel{=}\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Conid{Del}\;\Varid{xs}\;\Varid{s}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mbox{\onelinecomment  membership}{}\<[E]%
+\\
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Member}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\mathbin{::}\Conid{Bool}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Member}\;\Conid{Nil}\;{}\<[33]%
+\>[33]{}\Varid{s}\mathrel{=}\Conid{False}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Member}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[33]%
+\>[33]{}\Varid{s}\mathrel{=}\Conid{True}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Member}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[33]%
+\>[33]{}\Varid{s}\mathrel{=}\Conid{Member}\;\Varid{xs}\;\Varid{s}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{Some operations on type-level dictionaries.}
+\label{fig:dict-operations}
+\end{figure}
+
+\subsection{Proxies and Singleton Types}
+
+The \Hedis{} function \ensuremath{\Varid{del}\mathbin{::}[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Either}\;\Conid{Reply}\;\Conid{Integer}} takes a list
+of keys (encoded to \ensuremath{\Conid{ByteString}}) and removes the entries having those keys in
+the database. For simplicity, we consider creating a \Popcorn{} counterpart
+that takes only one key. A first attempt may lead to something like the
+following:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{del}\mathbin{::}\Conid{String}\to \Conid{Popcorn}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\mathbin{?})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
+\\
+\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Popcorn}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;[\mskip1.5mu \Varid{encodeKey}\;\Varid{key}\mskip1.5mu]~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+At term-level, our \ensuremath{\Varid{del}} merely calls \ensuremath{\Varid{\Conid{Hedis}.del}} (assuming that \ensuremath{\Varid{encodeKey}}
+converts \ensuremath{\Varid{key}} to a \ensuremath{\Conid{ByteString}}). At type-level, if the status of the database
+before \ensuremath{\Varid{del}} is called is specified by \ensuremath{\Varid{xs}}, the status after the execution
+should be specified by \ensuremath{\Conid{Del}\;\Varid{xs}\mathbin{?}}. The question, however, is what to fill in
+the question mark. The string \ensuremath{\Varid{key}} is a runtime value. How do we pass it to
+the type level?
+
+\todo{revised up to here.}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{del}\mathbin{::}\Conid{KnownSymbol}\;\Varid{s}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Rightarrow \Conid{Proxy}\;\Varid{s}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\to \Conid{Popcorn}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\;\Varid{s})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
+\\
+\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Popcorn}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;(\Varid{encodeKey}\;\Varid{key}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\ensuremath{\Conid{KnownSymbol}} is a class that gives the string associated
+ with a concrete type-level symbol, which can be retrieved with
+ \ensuremath{\Varid{symbolVal}}.\footnotemark
+ Where \ensuremath{\Varid{encodeKey}} converts \ensuremath{\Conid{Proxy}\;\Varid{s}} to
+ \ensuremath{\Conid{ByteString}}.
+\footnotetext{They are defined in \ensuremath{\Conid{\Conid{GHC}.TypeLits}}.}
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{encodeKey}\mathbin{::}\Conid{KnownSymbol}\;\Varid{s}\Rightarrow \Conid{Proxy}\;\Varid{s}\to \Conid{ByteString}{}\<[E]%
+\\
+\>[B]{}\Varid{encodeKey}\mathrel{=}\Varid{encode}\mathbin{\cdot}\Varid{symbolVal}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Since Haskell has a \emph{phase distinction}~\cite{phasedistinction}, types are
+ erased before runtime. It's impossible to obtain information directly from
+ types, we can only do this indirectly, with
+ \emph{singleton types, singletons}.
+
+A singleton type is a type that has only one instance, and the instance can be
+ think of as the representative of the type at the realm of runtime values.
+
+\ensuremath{\Conid{Proxy}}, as its name would suggest, can be used as
+ singletons. It's a phantom type that could be indexed with any type.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{data}\;\Conid{Proxy}\;\Varid{t}\mathrel{=}\Conid{Proxy}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+In the type of \ensuremath{\Varid{del}}, the type variable
+ \ensuremath{\Varid{s}} is a \ensuremath{\Conid{Symbol}} that is decided by
+ the argument of type \ensuremath{\Conid{Proxy}\;\Varid{s}}.
+ To use \ensuremath{\Varid{del}}, we would have to apply it with a clumsy
+ term-level proxy like this:
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{del}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 A\char34}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks

### 2016-06-12 aplas16/sections/IndexedMonads.tex
Popcorn -> Edis.

@@
 
-We define a new indexed monad \ensuremath{\Conid{Popcorn}} which, at term level, merely wraps
+We define a new indexed monad \ensuremath{\Conid{Edis}} which, at term level, merely wraps
 \ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}~~,{}\<[E]%
+\>[B]{}\mathbf{newtype}\;\Conid{Edis}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}\Conid{Edis}\;\{\mskip1.5mu \Varid{unEdis}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}~~,{}\<[E]%
 \\[\blanklineskip]%
-\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
+\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Edis}\;\mathbf{where}{}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
+\>[5]{}\Varid{unit}\mathrel{=}\Conid{Edis}\mathbin{\cdot}\Varid{return}{}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f})~~.{}\<[E]%
+\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}\Conid{Edis}\;(\Varid{unEdis}\;\Varid{m}\bind \Varid{unEdis}\mathbin{\cdot}\Varid{f})~~.{}\<[E]%
 \ColumnHook
@@
 At term level, the \ensuremath{\Varid{unit}} and \ensuremath{\Varid{bind}} methods are not interesting: they merely
-make calls to \ensuremath{\Varid{return}} and \ensuremath{(\bind )} of \ensuremath{\Conid{Redis}}, and extracts and re-apply the constructor \ensuremath{\Conid{Popcorn}} when necessary.
-With \ensuremath{\Conid{Popcorn}} being a \ensuremath{\mathbf{newtype}}, they
+make calls to \ensuremath{\Varid{return}} and \ensuremath{(\bind )} of \ensuremath{\Conid{Redis}}, and extracts and re-apply the constructor \ensuremath{\Conid{Edis}} when necessary.
+With \ensuremath{\Conid{Edis}} being a \ensuremath{\mathbf{newtype}}, they
 can be optimized away in runtime. The interesting bits happen in compile type,
@@
 \text{PONG} if the connection is alive. In \Hedis{}, \ensuremath{\Varid{ping}} has type
-\ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status})}. The \Popcorn{} version of \ensuremath{\Varid{ping}} simply
+\ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status})}. The \Edis{} version of \ensuremath{\Varid{ping}} simply
 applys an additional constructor (functions from \Hedis{} are qualified with
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ping}\mathbin{::}\Conid{Popcorn}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
+\>[B]{}\Varid{ping}\mathbin{::}\Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
 \\
-\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}~~.{}\<[E]%
+\>[B]{}\Varid{ping}\mathrel{=}\Conid{Edis}\;\Varid{\Conid{Hedis}.ping}~~.{}\<[E]%
 \ColumnHook

### 2016-06-12 aplas16/sections/Introduction.tex
Popcorn -> Edis.

@@
 
-This paper discusses the techniques we used and experiences we learned from building such a language, nicknamed \Popcorn{}. We constructed an {\em indexed
+This paper discusses the techniques we used and experiences we learned from building such a language, nicknamed \Edis{}. We constructed an {\em indexed
 monad}, on top of the monad \ensuremath{\Conid{Redis}}, which is indexed by a dictionary that
@@
 \begin{itemize}
-\item We present \Popcorn{}, a statically typed domain-specific language embedded in Haskell and built on \Hedis{}.
+\item We present \Edis{}, a statically typed domain-specific language embedded in Haskell and built on \Hedis{}.
 % also makes Redis polymorphic by automatically converting back and forth from values of arbitrary types and boring ByteStrings.

### 2016-06-12 aplas16/sections/TypeLevelDict.tex
Popcorn -> Edis.

@@
 more, as mentioned in Section~\ref{sec:indexed-monads}, the dictionaries are
-to be used parameters to the indexed monad \ensuremath{\Conid{Popcorn}}. In a dependently typed
+to be used parameters to the indexed monad \ensuremath{\Conid{Edis}}. In a dependently typed
 programming language (without the so-called ``phase distinction'' ---
@@
 of keys (encoded to \ensuremath{\Conid{ByteString}}) and removes the entries having those keys in
-the database. For simplicity, we consider creating a \Popcorn{} counterpart
+the database. For simplicity, we consider creating a \Edis{} counterpart
 that takes only one key. A first attempt may lead to something like the
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{del}\mathbin{::}\Conid{String}\to \Conid{Popcorn}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\mathbin{?})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
+\>[B]{}\Varid{del}\mathbin{::}\Conid{String}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\mathbin{?})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
 \\
-\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Popcorn}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;[\mskip1.5mu \Varid{encodeKey}\;\Varid{key}\mskip1.5mu]~~.{}\<[E]%
+\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;[\mskip1.5mu \Varid{encode}\;\Varid{key}\mskip1.5mu]~~.{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
-At term-level, our \ensuremath{\Varid{del}} merely calls \ensuremath{\Varid{\Conid{Hedis}.del}} (assuming that \ensuremath{\Varid{encodeKey}}
-converts \ensuremath{\Varid{key}} to a \ensuremath{\Conid{ByteString}}). At type-level, if the status of the database
-before \ensuremath{\Varid{del}} is called is specified by \ensuremath{\Varid{xs}}, the status after the execution
+At term-level, our \ensuremath{\Varid{del}} merely calls \ensuremath{\Varid{\Conid{Hedis}.del}}, with the help of \ensuremath{\Varid{encode}}
+that converts \ensuremath{\Varid{key}} to a \ensuremath{\Conid{ByteString}}.
+At type-level, if the status of the database before \ensuremath{\Varid{del}} is called is specified by \ensuremath{\Varid{xs}}, the status afterwards
 should be specified by \ensuremath{\Conid{Del}\;\Varid{xs}\mathbin{?}}. The question, however, is what to fill in
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\to \Conid{Popcorn}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\;\Varid{s})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
+\>[5]{}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\;\Varid{s})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
 \\
-\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Popcorn}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;(\Varid{encodeKey}\;\Varid{key}){}\<[E]%
+\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;(\Varid{encodeKey}\;\Varid{key}){}\<[E]%
 \ColumnHook

### 2016-06-14 aplas16/sections/Embedding.tex
Moved some examples to Introduction.

%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%
%
%
%
%
%
% This package provides two environments suitable to take the place
% of hscode, called "plainhscode" and "arrayhscode". 
%
% The plain environment surrounds each code block by vertical space,
% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
% similar to formulas. Note that if these dimensions are changed,
% the spacing around displayed math formulas changes as well.
% All code is indented using \leftskip.
%
% Changed 19.08.2004 to reflect changes in colorcode. Should work with
% CodeGroup.sty.
%
\ReadOnlyOnce{polycode.fmt}%
\makeatletter

\newcommand{\hsnewpar}[1]%
  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}

% can be used, for instance, to redefine the code size, by setting the
% command to \small or something alike
\newcommand{\hscodestyle}{}

% The command \sethscode can be used to switch the code formatting
% behaviour by mapping the hscode environment in the subst directive
% to a new LaTeX environment.

\newcommand{\sethscode}[1]%
  {\expandafter\let\expandafter\hscode\csname #1\endcsname
   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}

% "compatibility" mode restores the non-polycode.fmt layout.

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

\newcommand{\compaths}{\sethscode{compathscode}}

% "plain" mode is the proposed default.
% It should now work with \centering.
% This required some changes. The old version
% is still available for reference as oldplainhscode.

\newenvironment{plainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newenvironment{oldplainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

% Here, we make plainhscode the default environment.

\newcommand{\plainhs}{\sethscode{plainhscode}}
\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
\plainhs

% The arrayhscode is like plain, but makes use of polytable's
% parray environment which disallows page breaks in code blocks.

\newenvironment{arrayhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\parray}%
  {\endparray\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newcommand{\arrayhs}{\sethscode{arrayhscode}}

% The mathhscode environment also makes use of polytable's parray 
% environment. It is supposed to be used only inside math mode 
% (I used it to typeset the type rules in my thesis).

\newenvironment{mathhscode}%
  {\parray}{\endparray}

\newcommand{\mathhs}{\sethscode{mathhscode}}

% texths is similar to mathhs, but works in text mode.

\newenvironment{texthscode}%
  {\(\parray}{\endparray\)}

\newcommand{\texths}{\sethscode{texthscode}}

% The framed environment places code in a framed box.

\def\codeframewidth{\arrayrulewidth}
\RequirePackage{calc}

\newenvironment{framedhscode}%
  {\parskip=\abovedisplayskip\par\noindent
   \hscodestyle
   \arrayrulewidth=\codeframewidth
   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
   \hline\framedhslinecorrect\\{-1.5ex}%
   \let\endoflinesave=\\
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \framedhslinecorrect\endoflinesave{.5ex}\hline
   \endtabular
   \parskip=\belowdisplayskip\par\noindent
   \ignorespacesafterend}

\newcommand{\framedhslinecorrect}[2]%
  {#1[#2]}

\newcommand{\framedhs}{\sethscode{framedhscode}}

% The inlinehscode environment is an experimental environment
% that can be used to typeset displayed code inline.

\newenvironment{inlinehscode}%
  {\(\def\column##1##2{}%
   \let\>\undefined\let\<\undefined\let\\\undefined
   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
   \def\fromto##1##2##3{##3}%
   \def\nextline{}}{\) }%

\newcommand{\inlinehs}{\sethscode{inlinehscode}}

% The joincode environment is a separate environment that
% can be used to surround and thereby connect multiple code
% blocks.

\newenvironment{joincode}%
  {\let\orighscode=\hscode
   \let\origendhscode=\endhscode
   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
   %\let\SaveRestoreHook=\empty
   %\let\ColumnHook=\empty
   %\let\resethooks=\empty
   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
  {\origendhscode
   \global\let\hscode=\orighscode
   \global\let\endhscode=\origendhscode}%

\makeatother
\EndFmtInput
%

\ReadOnlyOnce{Formatting.fmt}%
\makeatletter

\let\Varid\mathit
\let\Conid\mathsf

\def\commentbegin{\quad\{\ }
\def\commentend{\}}

\newcommand{\ty}[1]{\Conid{#1}}
\newcommand{\con}[1]{\Conid{#1}}
\newcommand{\id}[1]{\Varid{#1}}
\newcommand{\cl}[1]{\Varid{#1}}
\newcommand{\opsym}[1]{\mathrel{#1}}

\newcommand\Keyword[1]{\textbf{\textsf{#1}}}
\newcommand\Hide{\mathbin{\downarrow}}
\newcommand\Reveal{\mathbin{\uparrow}}

%% Paper-specific keywords

\makeatother
\EndFmtInput

\section{Embedding \Hedis{} Commands}
\label{sec:embedding-commands}

Having the indexed monads and type-level dictionaries, in this section we
present our embedding of \Hedis{} commands into \Edis{}, while introducing
necessary concepts when they are used.

\subsection{Proxies and Singleton Types}

The \Hedis{} function \ensuremath{\Varid{del}\mathbin{::}[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Either}\;\Conid{Reply}\;\Conid{Integer}} takes a list
of keys (encoded to \ensuremath{\Conid{ByteString}}) and removes the entries having those keys in
the database. For simplicity, we consider creating a \Edis{} counterpart
that takes only one key. A first attempt may lead to something like the
following:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{del}\mathbin{::}\Conid{String}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}~\mathbin{?})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;[\mskip1.5mu \Varid{encode}\;\Varid{key}\mskip1.5mu]~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where the function \ensuremath{\Varid{encode}} converts \ensuremath{\Conid{String}} to \ensuremath{\Conid{ByteString}}. At term-level,
our \ensuremath{\Varid{del}} merely calls \ensuremath{\Varid{\Conid{Hedis}.del}}. At
type-level, if the status of the database before \ensuremath{\Varid{del}} is called is specified
by the dictionary \ensuremath{\Varid{xs}}, the status afterwards should be specified by
\ensuremath{\Conid{Del}\;\Varid{xs}~\mathbin{?}}. The question, however, is what to fill in place of the
question mark. It cannot be \ensuremath{\Conid{Del}\;\Varid{xs}\;\Varid{key}}, since \ensuremath{\Varid{key}} is a runtime value and
not a type. How do we smuggle a runtime value to type-level?

In a language with phase distinction like Haskell, it is certainly impossible
to pass the value of \ensuremath{\Varid{key}} to the type checker if it truly is a runtime value,
for example, a string read from the user. If the value of \ensuremath{\Varid{key}} can be
determined statically, however, {\em singleton types} can be used to represent a
type as a value, thus build a connection between the two realms.

A singleton type is a type that has only one term. When the term is built, it
carries a type that can be inspected by the type checker. The term can be think
of as a representative of its type at the realm of runtime values. For our
purpose, we will use the following type \ensuremath{\Conid{Proxy}}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Proxy}\;\Varid{t}\mathrel{=}\Conid{Proxy}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
For every type \ensuremath{\Varid{t}}, \ensuremath{\Conid{Proxy}\;\Varid{t}} is a type that has only one term: \ensuremath{\Conid{Proxy}}.
\footnote{While giving the same name to both the type and the term can be very
confusing, it is unfortunately a common practice in the Haskell community.}
To call \ensuremath{\Varid{del}}, instead of passing a key as a \ensuremath{\Conid{String}}, we give it a proxy with
a specified type:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{del}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 A\char34})~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where \ensuremath{\text{\tt \char34 A\char34}} is not a value, but a string lifted to a type (of kind \ensuremath{\Conid{Symbol}}).
Now that the type checker has access to the key, the type of \ensuremath{\Varid{del}} could be
something alone the line of \ensuremath{\Varid{del}\mathbin{::}\Conid{Proxy}\;\Varid{s}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\;\Varid{s})\mathbin{...}}.

The next problem is that, \ensuremath{\Varid{del}}, at term level, gets only a value constructor
\ensuremath{\Conid{Proxy}} without further information, while it needs to pass a \ensuremath{\Conid{ByteString}} key
to \ensuremath{\Varid{\Conid{Hedis}.del}}. Every concrete string literal lifted to a type, for example
\ensuremath{\text{\tt \char34 A\char34}}, belongs to a type class \ensuremath{\Conid{KnownSymbol}}. For all type \ensuremath{\Varid{n}} in \ensuremath{\Conid{KnownSymbol}},
the function \ensuremath{\Varid{symbolVal}}:\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Varid{symbolVal}\mathbin{::}\Conid{KnownSymbol}\;\Varid{n}\Rightarrow \Varid{proxy}\;\Varid{n}\to \Conid{String}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
retrieves the string associated with a type-level literal that is known at
compile time. In summary, \ensuremath{\Varid{del}} can be implemented as:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{45}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{del}{}\<[6]%
\>[6]{}\mathbin{::}\Conid{KnownSymbol}\;\Varid{s}{}\<[E]%
\\
\>[6]{}\Rightarrow \Conid{Proxy}\;\Varid{s}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\;\Varid{s})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.del}\;[\mskip1.5mu \Varid{encodeKey}\;\Varid{key}\mskip1.5mu]){}\<[45]%
\>[45]{}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where \ensuremath{\Varid{encodeKey}\mathrel{=}\Varid{encode}\mathbin{\cdot}\Varid{symbolVal}}.

A final note: functions the \ensuremath{\Varid{encode}}, from the Haskell library {\sc cereal},
helps to convert certain datatype that are {\em serializable} into \ensuremath{\Conid{ByteString}}.
The function and its dual \ensuremath{\Varid{decode}} will be use more later.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{encode}{}\<[9]%
\>[9]{}\mathbin{::}\Conid{Serialize}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{ByteString}~~,{}\<[E]%
\\
\>[B]{}\Varid{decode}{}\<[9]%
\>[9]{}\mathbin{::}\Conid{Serialize}\;\Varid{a}\Rightarrow \Conid{ByteString}\to \Conid{Either}\;\Conid{String}\;\Varid{a}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\subsection{Storing Primitive Datatypes Other Than Strings}
\label{sec:polymorphic-redis}

As mentioned before, while \Redis{} provide a number of container types
including lists, sets, and hash, etc., the primitive type is string.
\Hedis{} programmers manually convert other types of data to strings before
saving them into the data store. In \Edis{}, we wish to save some of the
effort for the programmers, as well as keeping a careful record of the intended
types of the strings in the data store.

To keep track of intended types of strings in the data store, we define the
following types (that have no terms):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{StringOf}\;\Varid{x}~~,{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\Conid{ListOf}\;\Varid{x}~~,{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\Conid{SetOf}\;\Varid{x}~~...{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
If an key is associated with, for example, \ensuremath{\Conid{StringOf}\;\Conid{Int}} in
our dictionary, we mean that its associated value in the data store was
serialized from an \ensuremath{\Conid{Int}} and should be used as an \ensuremath{\Conid{Int}}. Types
\ensuremath{\Conid{ListOf}\;\Varid{x}} and \ensuremath{\Conid{SetOf}\;\Varid{x}}, respectively, denotes that the value is a list
or a set of the given type.

While the \ensuremath{\Varid{set}} command in \Hedis{} always writes a string to the data store,
the corresponding \ensuremath{\Varid{set}} in \Redis{} applies to any serializable type (those
in the class \ensuremath{\Conid{Serialize}}), and performs the encoding for the user:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{set}{}\<[6]%
\>[6]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{Serialize}\;\Varid{x}){}\<[E]%
\\
\>[6]{}\Rightarrow \Conid{Proxy}\;\Varid{s}\to \Varid{x}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Set}\;\Varid{xs}\;\Varid{s}\;(\Conid{StringOf}\;\Varid{x}))\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\\
\>[B]{}\Varid{set}\;\Varid{key}\;\Varid{val}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.set}\;(\Varid{encodeKey}\;\Varid{key})\;(\Varid{encode}\;\Varid{val})~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where the function \ensuremath{\Varid{encode}\mathbin{::}\Conid{Serialize}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{ByteString}}

For example, executing \ensuremath{\Varid{set}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 A\char34})\;\Conid{True}} updates the dictionary
with an entry \ensuremath{\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{StringOf}\;\Conid{Bool})}: if \ensuremath{\text{\tt \char34 A\char34}} is not in the dictionary,
the entry is added; otherwise the old type of \ensuremath{\text{\tt \char34 A\char34}} is updated to
\ensuremath{\Conid{StringOf}\;\Conid{Bool}}.

\Redis{} command \texttt{INCR} reads the string associated with the given key,
parse it as an integer, and increments it, before storing it back. The command
\texttt{INCRBYFLOAT} increments the floating number, associated to the given
key. They can be
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{incr}{}\<[7]%
\>[7]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{Get}\;\Varid{xs}\;\Varid{s}\mathord{\sim}\Conid{Just}\;(\Conid{StringOf}\;\Conid{Integer})){}\<[E]%
\\
\>[7]{}\Rightarrow \Conid{Proxy}\;\Varid{s}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{incr}\;\Varid{key}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.incr}\;(\Varid{encodeKey}\;\Varid{key})~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{incrbyfloat}{}\<[14]%
\>[14]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{Get}\;\Varid{xs}\;\Varid{s}\mathord{\sim}\Conid{Just}\;(\Conid{StringOf}\;\Conid{Double})){}\<[E]%
\\
\>[14]{}\Rightarrow \Conid{Proxy}\;\Varid{s}\to \Conid{Double}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Double}){}\<[E]%
\\
\>[B]{}\Varid{incrbyfloat}\;\Varid{key}\;\Varid{eps}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.incrbyfloat}\;(\Varid{encodeKey}\;\Varid{key})\;\Varid{eps}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Notice the use of (\ensuremath{\mathord{\sim}}), \emph{equality constraints}~\cite{typeeq}, to enforce
that the intended type of value associated with key \ensuremath{\Varid{s}} must respectively be
\ensuremath{\Conid{Integer}} and \ensuremath{\Conid{Double}}.

\todo{Some more interesting functions that uses IF and FromJust, etc.
What about \ensuremath{\Varid{lset}}?}

\subsection{Constraint Disjunctions}

In the following example, \texttt{LPUSH} appends a string \texttt{"bar"} to the
list associated to \texttt{some-list}, or creates a list if \texttt{some-list}
does not yet present. The command \texttt{LLEN} then returns the length of
the list. If the key passed to \texttt{LLEN} is not associated to a list,
it signals an error.
\begin{tabbing}\tt
~redis\char62{}~LPUSH~some\char45{}list~bar\\
\tt ~\char40{}integer\char41{}~1\\
\tt ~redis\char62{}~LLEN~some\char45{}list\\
\tt ~\char40{}integer\char41{}~1\\
\tt ~redis\char62{}~SET~some\char45{}string~foo\\
\tt ~OK\\
\tt ~redis\char62{}~LLEN~some\char45{}string\\
\tt ~\char40{}error\char41{}~WRONGTYPE~Operation~against~a~key~holding\\
\tt ~the~wrong~kind~of~value
\end{tabbing}
One may give \ensuremath{\Varid{llen}} this type:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{llen}{}\<[7]%
\>[7]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{Get}\;\Varid{xs}\;\Varid{s}\mathord{\sim}\Conid{Just}\;(\Conid{ListOf}\;\Varid{x})){}\<[E]%
\\
\>[7]{}\Rightarrow \Conid{Proxy}\;\Varid{s}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{llen}\;\Varid{key}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.llen}\;(\Varid{encodeKey}\;\Varid{key})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
However, this is not an accurate specification of \texttt{LLEN} in \Redis{} ---
\texttt{LLEN} also accepts keys that do not exist, and replies with \texttt{0}:
\begin{tabbing}\tt
~redis\char62{}~LLEN~nonexistent\\
\tt ~\char40{}integer\char41{}~0
\end{tabbing}
What we wish to have in the class constraint of \ensuremath{\Varid{llen}} is thus a predicate
equivalent to \ensuremath{\Conid{Get}\;\Varid{xs}\;\Varid{s}\doubleequals\Conid{Just}\;(\Conid{ListOf}\;\Varid{x})\mathrel{\vee}\neg \;(\Conid{Member}\;\Varid{xs}\;\Varid{s})}. The situation is the same with \ensuremath{\Varid{lpush}}: it returns normally either when
the key is associated to a list or does not exist at all, while signaling an
error when the associated value is not a list.

Unfortunately, expressing disjunctions in constraints is much more difficult
than expressing conjunctions. To impose a class constraint \ensuremath{\Conid{P}\mathrel{\wedge}\Conid{Q}}, one may
simply put them both in the type: \ensuremath{(\Conid{P},\Conid{Q})\Rightarrow \mathbin{...}}. To express top-level
constraints

\todo{why cite \cite{singletons} here?}

We could achieve this simply by translating the semantics we want to the
 domain of Boolean, with type-level boolean functions such as
\ensuremath{(\mathrel{\wedge})},
\ensuremath{(\mid )}, \ensuremath{\Conid{Not}},
\ensuremath{(\doubleequals)}, etc.\footnotemark To avoid

\footnotetext{Available in \text{Data.Type.Bool} and
 \text{Data.Type.Equality}}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{Get}\;\Varid{xs}\;\Varid{s}\doubleequals\Conid{Just}\;(\Conid{ListOf}\;\Varid{x})\mathrel{\vee}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{s}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

To avoid addressing the type of value (as it may not exist at all), we defined
 an auxiliary predicate \ensuremath{\Conid{IsList}\mathbin{::}\Conid{Maybe}\mathbin{*}\to \Conid{Bool}} to
 replace the former part.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{IsList}\;(\Conid{Get}\;\Varid{xs}\;\Varid{s})\mathrel{\vee}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{s}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The type expression above has kind \ensuremath{\Conid{Bool}}, we could make it
 a type constraint by asserting equality.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\Conid{IsList}\;(\Conid{Get}\;\Varid{xs}\;\Varid{s})\mathrel{\vee}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{s}))\mathord{\sim}\Conid{True}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

With \emph{constraint kind}, a recent addition to GHC, type constraints now has
 its own kind: \ensuremath{\Conid{Constraint}}. That means type constraints
 are not restricted to the left side of a \ensuremath{\Rightarrow } anymore,
 they could appear in anywhere that accepts something of kind
 \ensuremath{\Conid{Constraint}}, and any type that has kind
 \ensuremath{\Conid{Constraint}} can also be used as a type constraint.
 \footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}

As many other list-related commands also have this ``List or nothing'' semantics,
 we could abstract the lengthy type constraint above and give it an alias with
 type synonym.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{ListOrNX}\;\Varid{xs}\;\Varid{s}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}(\Conid{IsList}\;(\Conid{Get}\;\Varid{xs}\;\Varid{s})\mathrel{\vee}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{s}))\mathord{\sim}\Conid{True}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The complete implementation of \text{LLEN} with
\ensuremath{\Conid{ListOrNX}} would become:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{llen}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{ListOrNX}\;\Varid{xs}\;\Varid{s}){}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Rightarrow \Conid{Proxy}\;\Varid{s}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{llen}\;\Varid{key}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.llen}\;(\Varid{encodeKey}\;\Varid{key}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

### 2016-06-15 aplas16/sections/Embedding.tex
Up to assertions, and the final example.

@@
 \end{hscode}\resethooks
-For every type \ensuremath{\Varid{t}}, \ensuremath{\Conid{Proxy}\;\Varid{t}} is a type that has only one term: \ensuremath{\Conid{Proxy}}.
+For every type \ensuremath{\Varid{t}}, \ensuremath{\Conid{Proxy}\;\Varid{t}} is a type that has only one term: \ensuremath{\Conid{Proxy}}.%
 \footnote{While giving the same name to both the type and the term can be very
@@
 
-A final note: functions the \ensuremath{\Varid{encode}}, from the Haskell library {\sc cereal},
+A final note: the function \ensuremath{\Varid{encode}}, from the Haskell library {\sc cereal},
 helps to convert certain datatype that are {\em serializable} into \ensuremath{\Conid{ByteString}}.
@@
 \end{hscode}\resethooks
-where the function \ensuremath{\Varid{encode}\mathbin{::}\Conid{Serialize}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{ByteString}}
-
 For example, executing \ensuremath{\Varid{set}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 A\char34})\;\Conid{True}} updates the dictionary
-with an entry \ensuremath{\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{StringOf}\;\Conid{Bool})}: if \ensuremath{\text{\tt \char34 A\char34}} is not in the dictionary,
-the entry is added; otherwise the old type of \ensuremath{\text{\tt \char34 A\char34}} is updated to
+with an entry \ensuremath{\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{StringOf}\;\Conid{Bool})}. If \ensuremath{\text{\tt \char34 A\char34}} is not in the dictionary,
+this entry is added; otherwise the old type of \ensuremath{\text{\tt \char34 A\char34}} is updated to
 \ensuremath{\Conid{StringOf}\;\Conid{Bool}}.
@@
 that the intended type of value associated with key \ensuremath{\Varid{s}} must respectively be
-\ensuremath{\Conid{Integer}} and \ensuremath{\Conid{Double}}.
+\ensuremath{\Conid{Integer}} and \ensuremath{\Conid{Double}}. The function \ensuremath{\Varid{incr}} is allowed to be called only
+in a context where the type checker is able to reduce \ensuremath{\Conid{Get}\;\Varid{xs}\;\Varid{s}} to
+\ensuremath{\Conid{Just}\;(\Conid{StringOf}\;\Conid{Integer})}. Similarly with \ensuremath{\Varid{incrbyfloat}}.
 
@@
 
-In the following example, \texttt{LPUSH} appends a string \texttt{"bar"} to the
-list associated to \texttt{some-list}, or creates a list if \texttt{some-list}
-does not yet present. The command \texttt{LLEN} then returns the length of
-the list. If the key passed to \texttt{LLEN} is not associated to a list,
-it signals an error.
-\begin{tabbing}\tt
-~redis\char62{}~LPUSH~some\char45{}list~bar\\
-\tt ~\char40{}integer\char41{}~1\\
-\tt ~redis\char62{}~LLEN~some\char45{}list\\
-\tt ~\char40{}integer\char41{}~1\\
-\tt ~redis\char62{}~SET~some\char45{}string~foo\\
-\tt ~OK\\
-\tt ~redis\char62{}~LLEN~some\char45{}string\\
-\tt ~\char40{}error\char41{}~WRONGTYPE~Operation~against~a~key~holding\\
-\tt ~the~wrong~kind~of~value
-\end{tabbing}
-One may give \ensuremath{\Varid{llen}} this type:
+Recall, from Section \ref{sec:introduction}, that commands \texttt{LPUSH key
+val} and \texttt{LLEN key} returns normally either when \ensuremath{\Varid{key}} does not present
+in the data store, or when \ensuremath{\Varid{key}} presents and is associated to a list.
+What we wish to have in their constraint is thus a predicate equivalent to \ensuremath{\Conid{Get}\;\Varid{xs}\;\Varid{s}\doubleequals\Conid{Just}\;(\Conid{ListOf}\;\Varid{x})\mathrel{\vee}\neg \;(\Conid{Member}\;\Varid{xs}\;\Varid{s})}.
+
+To impose a conjunctive constraint \ensuremath{\Conid{P}\mathrel{\wedge}\Conid{Q}}, one may simply put them both in the
+type: \ensuremath{(\Conid{P},\Conid{Q})\Rightarrow \mathbin{...}}. Expressing disjunctive constraints is only slightly
+harder, thanks to our type-level functions. Various operators for type-level
+boolean and equality are defined in \text{Data.Type.Bool} and
+\text{Data.Type.Equality}, like how we defined \ensuremath{\Conid{Or}} in Section
+\ref{sec:type-fun}. We may thus write the predicate as:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Conid{Get}\;\Varid{xs}\;\Varid{s}\doubleequals\Conid{Just}\;(\Conid{ListOf}\;\Varid{x})\mathbin{`\Conid{Or}`}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{s})~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+To avoid referring to \ensuremath{\Varid{x}}, which might not exist, we define an auxiliary predicate \ensuremath{\Conid{IsList}\mathbin{::}\Conid{Maybe}\mathbin{*}\to \Conid{Bool}}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{IsList}\;(\Varid{x}\mathbin{::}\mathbin{*})\mathbin{::}\Conid{Bool}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{IsList}\;(\Conid{ListOf}\;\Varid{n})\mathrel{=}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{IsList}\;\Varid{x}{}\<[23]%
+\>[23]{}\mathrel{=}\mbox{\textquotesingle}\Conid{False}~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+As many other list-related commands also have this ``List or nothing'' semantics, we give the type constraint a name:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Conid{ListOrNX}\;\Varid{xs}\;\Varid{s}\mathrel{=}(\Conid{IsList}\;(\Conid{Get}\;\Varid{xs}\;\Varid{s})\mathrel{\vee}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{s}))\mathord{\sim}\Conid{True}~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\noindent The complete implementation of \text{LLEN} with
+\ensuremath{\Conid{ListOrNX}} is therefore:
 \begin{hscode}\SaveRestoreHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{lpush}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{Serialize}\;\Varid{x},\Conid{ListOrNX}\;\Varid{xs}\;\Varid{s}){}\<[E]%
+\\
+\>[B]{}\hsindent{7}{}\<[7]%
+\>[7]{}\Rightarrow \Conid{Proxy}\;\Varid{s}\to \Varid{x}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Set}\;\Varid{xs}\;\Varid{s}\;(\Conid{ListOf}\;\Varid{x}))\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
+\\
+\>[B]{}\Varid{lpush}\;\Varid{key}\;\Varid{val}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Redis}.lpush}\;(\Varid{encodeKey}\;\Varid{key})\;[\mskip1.5mu \Varid{encode}\;\Varid{val}\mskip1.5mu]~~,{}\<[E]%
+\\[\blanklineskip]%
 \>[B]{}\Varid{llen}{}\<[7]%
-\>[7]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{Get}\;\Varid{xs}\;\Varid{s}\mathord{\sim}\Conid{Just}\;(\Conid{ListOf}\;\Varid{x})){}\<[E]%
+\>[7]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{ListOrNX}\;\Varid{xs}\;\Varid{s}){}\<[E]%
 \\
@@
 \end{hscode}\resethooks
-However, this is not an accurate specification of \texttt{LLEN} in \Redis{} ---
-\texttt{LLEN} also accepts keys that do not exist, and replies with \texttt{0}:
-\begin{tabbing}\tt
-~redis\char62{}~LLEN~nonexistent\\
-\tt ~\char40{}integer\char41{}~0
-\end{tabbing}
-What we wish to have in the class constraint of \ensuremath{\Varid{llen}} is thus a predicate
-equivalent to \ensuremath{\Conid{Get}\;\Varid{xs}\;\Varid{s}\doubleequals\Conid{Just}\;(\Conid{ListOf}\;\Varid{x})\mathrel{\vee}\neg \;(\Conid{Member}\;\Varid{xs}\;\Varid{s})}.
-The situation is the same with \ensuremath{\Varid{lpush}}: it returns normally either when
-the key is associated to a list or does not exist at all, while signaling an
-error when the associated value is not a list.
-
-Unfortunately, expressing disjunctions in constraints is much more difficult
-than expressing conjunctions. To impose a class constraint \ensuremath{\Conid{P}\mathrel{\wedge}\Conid{Q}}, one may
-simply put them both in the type: \ensuremath{(\Conid{P},\Conid{Q})\Rightarrow \mathbin{...}}. To express top-level
-constraints
 
@@
 
-We could achieve this simply by translating the semantics we want to the
- domain of Boolean, with type-level boolean functions such as
-\ensuremath{(\mathrel{\wedge})},
-\ensuremath{(\mid )}, \ensuremath{\Conid{Not}},
-\ensuremath{(\doubleequals)}, etc.\footnotemark To avoid
+% The type expression above has kind |Bool|, we could make it
+%  a type constraint by asserting equality.
+%
+% With \emph{constraint kind}, a recent addition to GHC, type constraints now has
+%  its own kind: |Constraint|. That means type constraints
+%  are not restricted to the left side of a |=>| anymore,
+%  they could appear in anywhere that accepts something of kind
+%  |Constraint|, and any type that has kind
+%  |Constraint| can also be used as a type constraint.
+%  \footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}
 
-\footnotetext{Available in \text{Data.Type.Bool} and
- \text{Data.Type.Equality}}
+\subsection{Assertions}
+\label{sec:assertions}
+
+Finally, the creation/update behavior of \Redis{} functions is, in our opinion,
+very error-prone. It might be preferable if we can explicit declare some new
+keys, after ensure that they do not already exist (in our types), and renounce
+them and forbid further access when we are sure that they shall not be referred
+to anymore. This can be done below:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{declare}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{Member}\;\Varid{xs}\;\Varid{s}\mathord{\sim}\Conid{False}){}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Rightarrow \Conid{Proxy}\;\Varid{s}\to \Conid{Proxy}\;\Varid{x}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Set}\;\Varid{xs}\;\Varid{s}\;\Varid{x})\;(){}\<[E]%
+\\
+\>[B]{}\Varid{declare}\;\Varid{s}\;\Varid{x}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{return}\;()~~,{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{renounce}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{Member}\;\Varid{xs}\;\Varid{s}\mathord{\sim}\Conid{True}){}\<[E]%
+\\
+\>[B]{}\hsindent{9}{}\<[9]%
+\>[9]{}\Rightarrow \Conid{Proxy}\;\Varid{s}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\;\Varid{s})\;(){}\<[E]%
+\\
+\>[B]{}\Varid{renounce}\;\Varid{s}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{return}\;()~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The command \ensuremath{\Varid{declare}\;\Varid{s}\;\Varid{x}} adds a new key \ensuremath{\Varid{s}} with type \ensuremath{\Varid{x}} to the dictionary,
+if it does not already exist. Dually, \ensuremath{\Varid{renounce}} removes a key from the
+dictionary. Even though it may still exist in the data store, it is not allowed
+to be referred to. The command \ensuremath{\Varid{start}} initializes the dictionary to \ensuremath{\mbox{\textquotesingle}[~]}:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{start}\mathbin{::}\Conid{Edis}\;\mbox{\textquotesingle}[~]\;\mbox{\textquotesingle}[~]\;(){}\<[E]%
+\\
+\>[B]{}\Varid{start}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{return}\;()~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\subsection{A Larger Example}
+
+The following program increases the value of \ensuremath{\text{\tt \char34 A\char34}} as an integer, push the result of the increment to list \ensuremath{\text{\tt \char34 L\char34}}, and then pops it out:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{main}\mathbin{::}\Conid{IO}\;(){}\<[E]%
+\\
+\>[B]{}\Varid{main}\mathrel{=}\mathbf{do}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{conn}{}\<[13]%
+\>[13]{}\leftarrow \Varid{connect}\;\Varid{defaultConnectInfo}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{result}{}\<[13]%
+\>[13]{}\leftarrow \Varid{runRedis}\;\Varid{conn}\mathbin{\$}\Varid{unEdis}\mathbin{\$}\Varid{start}{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\mathbin{`\Varid{bind}`}\lambda \anonymous \to {}\<[24]%
+\>[24]{}\Varid{declare}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 A\char34})\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\Conid{Integer}){}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\mathbin{`\Varid{bind}`}\lambda \anonymous \to {}\<[24]%
+\>[24]{}\Varid{incr}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 A\char34}){}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\mathbin{`\Varid{bind}`}\lambda \Varid{n}\to {}\<[23]%
+\>[23]{}\mathbf{case}\;\Varid{n}\;\mathbf{of}{}\<[E]%
+\\
+\>[9]{}\hsindent{4}{}\<[13]%
+\>[13]{}\Conid{Left}\;{}\<[19]%
+\>[19]{}\Varid{err}{}\<[24]%
+\>[24]{}\to \Varid{lpush}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 L\char34})\;\mathrm{0}{}\<[E]%
+\\
+\>[9]{}\hsindent{4}{}\<[13]%
+\>[13]{}\Conid{Right}\;\Varid{n}{}\<[24]%
+\>[24]{}\to \Varid{lpush}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 L\char34})\;\Varid{n}{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\mathbin{`\Varid{bind}`}\lambda \anonymous \to {}\<[24]%
+\>[24]{}\Varid{lpop}\;{}\<[33]%
+\>[33]{}(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 L\char34}){}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{print}\;\Varid{result}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+The syntax is pretty heavy, like the old days when there's no
+ \emph{do-notation}\cite{history}. But if we don't need any variable bindings
+ between operations, we could compose these commands with a sequencing operator
+ \ensuremath{(\mathbin{>>>})}.
 
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Conid{Get}\;\Varid{xs}\;\Varid{s}\doubleequals\Conid{Just}\;(\Conid{ListOf}\;\Varid{x})\mathrel{\vee}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{s}){}\<[E]%
+\>[B]{}(\mathbin{>>>})\mathbin{::}\Conid{IMonad}\;\Varid{m}\Rightarrow \Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b}\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
-
-To avoid addressing the type of value (as it may not exist at all), we defined
- an auxiliary predicate \ensuremath{\Conid{IsList}\mathbin{::}\Conid{Maybe}\mathbin{*}\to \Conid{Bool}} to
- replace the former part.
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Conid{IsList}\;(\Conid{Get}\;\Varid{xs}\;\Varid{s})\mathrel{\vee}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{s}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-The type expression above has kind \ensuremath{\Conid{Bool}}, we could make it
- a type constraint by asserting equality.
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}(\Conid{IsList}\;(\Conid{Get}\;\Varid{xs}\;\Varid{s})\mathrel{\vee}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{s}))\mathord{\sim}\Conid{True}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-With \emph{constraint kind}, a recent addition to GHC, type constraints now has
- its own kind: \ensuremath{\Conid{Constraint}}. That means type constraints
- are not restricted to the left side of a \ensuremath{\Rightarrow } anymore,
- they could appear in anywhere that accepts something of kind
- \ensuremath{\Conid{Constraint}}, and any type that has kind
- \ensuremath{\Conid{Constraint}} can also be used as a type constraint.
- \footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}
-
-As many other list-related commands also have this ``List or nothing'' semantics,
- we could abstract the lengthy type constraint above and give it an alias with
- type synonym.
-
 \begin{hscode}\SaveRestoreHook
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Conid{ListOrNX}\;\Varid{xs}\;\Varid{s}\mathrel{=}{}\<[E]%
+\>[B]{}\Varid{program}\mathrel{=}\Varid{start}{}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}(\Conid{IsList}\;(\Conid{Get}\;\Varid{xs}\;\Varid{s})\mathrel{\vee}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{s}))\mathord{\sim}\Conid{True}{}\<[E]%
+\>[5]{}\mathbin{>>>}\Varid{declare}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 A\char34})\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\Conid{Integer}){}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\mathbin{>>>}\Varid{incr}\;{}\<[17]%
+\>[17]{}(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 A\char34}){}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\mathbin{>>>}\Varid{lpush}\;{}\<[17]%
+\>[17]{}(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 L\char34})\;\mathrm{0}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\mathbin{>>>}\Varid{lpop}\;{}\<[17]%
+\>[17]{}(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 L\char34}){}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
-
-The complete implementation of \text{LLEN} with
-\ensuremath{\Conid{ListOrNX}} would become:
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{llen}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{s},\Conid{ListOrNX}\;\Varid{xs}\;\Varid{s}){}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\Rightarrow \Conid{Proxy}\;\Varid{s}{}\<[E]%
-\\
-\>[B]{}\hsindent{9}{}\<[9]%
-\>[9]{}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
-\\
-\>[B]{}\Varid{llen}\;\Varid{key}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.llen}\;(\Varid{encodeKey}\;\Varid{key}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks

### 2016-06-16 aplas16/sections/Conclusions.tex
A very brief Conclusion.

@@
 
-We have demonstrated how a DSL with variable bindings could be embedded in types.
-With more and more extensions added to the language, Haskell is gradually
- becoming a dependently-typed language,\footnote{Why not be dependently typed? \url{http://stackoverflow.com/questions/12961651/why-not-be-dependently-typed}}
- but it's not that dreadful as many (including us) would have thought.
+By exploiting various recent extensions and type-level programming techniques,
+we have designed a domain-specific embedded language \Edis{} which, at
+term-level, is simply a wrapper of \Hedis{}, but enforces typing disciplines that keeps track of available keys and their types and allows a program to be
+constructed only if it does not throw a runtime type error. We believe that it
+is a neat case study of application of type-level programming.
+
+% With more and more extensions added to the language, Haskell is gradually
+%  becoming a dependently-typed language,\footnote{Why not be dependently typed? \url{http://stackoverflow.com/questions/12961651/why-not-be-dependently-typed}}
+%  but it's not that dreadful as many (including us) would have thought.

### 2016-06-17 aplas16/main-aplas.tex
Section 2 done.
Half of Section 3.3.
Revision up to incr and incrfloat.
Moved some examples to Introduction.
Up to assertions, and the final example.
Another round of revision from the beginning. Corrected some bibliography.

@@
 \usepackage{makeidx}  % allows for indexgeneration
+\usepackage{textcomp} % for single quote
+\usepackage{fancyvrb} % for indented verbatim
 
@@
 
-\title{Safe Redis Queries}
-\subtitle{A Case Study of Type-Level Programming}
+\title{Type Safe Redis Queries}
+\subtitle{A Case Study of Type-Level Programming in Haskell}
 \author{Ting-Yan Lai \inst{1}\inst{2}\and Tyng-Ruey Chuang \inst{1}
@@
 \begin{abstract}
-\Redis{} is an in-memory data strcture store, often used as a database, with a
-Haskell interface \Hedis{}. \Redis{} is dynamically typed --- a value associated
-with a key can be overwritten by a value of a different type, and a command,
+\Redis{} is an in-memory data structure store, often used as a database, with a
+Haskell interface \Hedis{}. \Redis{} is dynamically typed --- a key can be
+discarded and re-associated to a value of a different type, and a command,
 when fetching a value of a type it does not expect, signals a runtime error. We
@@
 type families, keeps track of types of values in the database and statically
-guarantees that type errors cannot happen.
+guarantees that type errors cannot happen for a class of \Redis{} programs.
 \end{abstract}
@@
 \input{sections/Introduction}
-%\input{sections/Motivation}
+\input{sections/IndexedMonads}
 \input{sections/TypeLevelDict}
-\input{sections/IndexedMonads}
-\input{sections/TypeLevelFun}
-\input{sections/PolyRedis}
-\input{sections/Constraints}
-\input{sections/Assertions}
+\input{sections/Embedding}
 \input{sections/Discussions}
-\input{sections/Related}
 \input{sections/Conclusions}

### 2016-06-17 aplas16/sections/Conclusions.tex
Another round of revision from the beginning. Corrected some bibliography.

--- 2016-06-11 aplas16/sections/Conclusions.tex
@@
-\section{Conclusions}
+\section{Conclusions and Related Work}
 \label{sec:conclusions}
 
-We have demonstrated how a DSL with variable bindings could be embedded in types.
-With more and more extensions added to the language, Haskell is gradually
- becoming a dependently-typed language,\footnote{Why not be dependently typed? \url{http://stackoverflow.com/questions/12961651/why-not-be-dependently-typed}}
- but it's not that dreadful as many (including us) would have thought.
+By exploiting various recent extensions and type-level programming techniques,
+we have designed a domain-specific embedded language \Edis{} which enforces
+typing disciplines that keeps track of available keys and their types. The
+type of a command clearly specifies which keys must or must not present in
+the data store, what their types ought to be, as well as how the keys and types are updated after the execution. A program can be constructed only if it does
+not throw a runtime type error, and the type also serves as documentation of the
+commands. We believe that it is a neat case study of application of type-level programming.
+
+\Redis{} identifies itself as a data structure store/server, rather than a
+database. Nevertheless, there has been attempts to design DSELs for relational
+databases that guarantee all queries made are safe.
+Among them, {\sc HaskellDB}~\cite{haskelldb,haskelldbimproved} dynamically generates, from monad comprehensions, SQL queries to be executed on ODBC database servers.
+With the expressiveness of dependent types, Oury and Swierstra~\cite{pi}
+build a DSEL in Agda for relational algebra. After GHC introduced more features
+faciliting type-level programming, Eisenberg and Weirich~\cite{singletons}
+ported these techniques to Haskell using singleton types.
+
+None of the type-level programming techniques we used in this paper are
+particularly new. Indexed monads (also called parameterized monads) have been
+introduced by Atkey~\cite{indexedmonad}.
+Kiselyov et al~\cite{typefun} used indexed monads to track the locks held among
+a given finite set. The same paper also demonstrated implementation of a
+variety of features including memoization, generic maps, session types, typed \texttt{printf} and \texttt{sprintf}, etc., by type-level programming.
+Before the introduction of type families, Kiselyov and Shan~\cite{staticresources} used type classes and functional dependencies to
+implement type-level functions, and showed that they are sufficient to track
+resources in device drivers.
+
+\cite{constraintskinds}
+
+Lindley and McBride~\cite{phasedistinction} provided a good tutorial, analysis, and summary of the dependent type like features currently in Haskell, and
+compared them with ``real'' dependently typed languages.

### 2016-06-17 aplas16/sections/Conclusions.tex
Finished a round of revision...

@@
 particularly new. Indexed monads (also called parameterized monads) have been
-introduced by Atkey~\cite{indexedmonad}.
+introduced by Atkey~\cite{indexedmonad}. McBride~\cite{kleisli} showed how to
+construct indexed free-monads from Hoare Logic specifications.
 Kiselyov et al~\cite{typefun} used indexed monads to track the locks held among
@@
 
-\cite{constraintskinds}
+% \cite{constraintskinds}
 
-Lindley and McBride~\cite{phasedistinction} provided a good tutorial, analysis, and summary of the dependent type like features currently in Haskell, and
-compared them with ``real'' dependently typed languages.
+Lindley and McBride~\cite{phasedistinction} provided a through analysis and summary of the dependent-type-like features currently in Haskell, and
+compared them with dependently typed languages without phase distinction, such as Agda. It turns out that GHC's constraint solver work surprisingly well as a theorem
+prover.

### 2016-06-19 aplas16/sections/Conclusions.tex
Tweaks.

--- 2016-06-17 aplas16/sections/Conclusions.tex
@@
 we have designed a domain-specific embedded language \Edis{} which enforces
-typing disciplines that keeps track of available keys and their types. The
+typing disciplines that keep track of available keys and their types. The
 type of a command clearly specifies which keys must or must not present in
@@
 With the expressiveness of dependent types, Oury and Swierstra~\cite{pi}
-build a DSEL in Agda for relational algebra. After GHC introduced more features
-faciliting type-level programming, Eisenberg and Weirich~\cite{singletons}
-ported these techniques to Haskell using singleton types.
+build a DSEL in Agda for relational algebra. Eisenberg and
+Weirich~\cite{singletons} ported the result to Haskell using singleton types, after GHC introduced more features facilitating type-level programming,
 
-None of the type-level programming techniques we used in this paper are
-particularly new. Indexed monads (also called parameterized monads) have been
-introduced by Atkey~\cite{indexedmonad}.
-Kiselyov et al~\cite{typefun} used indexed monads to track the locks held among
-a given finite set. The same paper also demonstrated implementation of a
-variety of features including memoization, generic maps, session types, typed \texttt{printf} and \texttt{sprintf}, etc., by type-level programming.
-Before the introduction of type families, Kiselyov and Shan~\cite{staticresources} used type classes and functional dependencies to
+None of the type-level programming techniques we used in this paper are new.
+Indexed monads (also called parameterized monads) have been introduced by
+Atkey~\cite{indexedmonad}. McBride~\cite{kleisli} showed how to construct
+indexed free-monads from Hoare Logic specifications. Kiselyov et
+al~\cite{typefun} used indexed monads to track the locks held among a given
+finite set. The same paper also demonstrated implementation of a variety of
+features including memoization, generic maps, session types, typed
+\texttt{printf} and \texttt{sprintf}, etc., by type-level programming.
+Before the introduction of type families, Kiselyov and
+Shan~\cite{staticresources} used type classes and functional dependencies to
 implement type-level functions, and showed that they are sufficient to track
@@
 
-\cite{constraintskinds}
-
-Lindley and McBride~\cite{phasedistinction} provided a good tutorial, analysis, and summary of the dependent type like features currently in Haskell, and
-compared them with ``real'' dependently typed languages.
+Lindley and McBride~\cite{phasedistinction} provided a through analysis and summary of the dependent-type-like features currently in Haskell, and
+compared them with dependently typed languages without phase distinction such as Agda. It turns out that GHC's constraint solver works surprisingly well as an
+automatic theorem prover.

### 2016-06-20 aplas16/sections/Conclusions.tex
Various editing.

@@
 build a DSEL in Agda for relational algebra. Eisenberg and
-Weirich~\cite{singletons} ported the result to Haskell using singleton types, after GHC introduced more features facilitating type-level programming,
+Weirich~\cite{singletons} ported the result to Haskell using singleton types, after GHC introduced more features facilitating type-level programming.
 

### 2016-06-20 aplas16/sections/Conclusions.tex
fix some minor typos

@@
 
-Lindley and McBride~\cite{phasedistinction} provided a through analysis and summary of the dependent-type-like features currently in Haskell, and
+Lindley and McBride~\cite{phasedistinction} provided a thorough analysis and summary of the dependent-type-like features currently in Haskell, and
 compared them with dependently typed languages without phase distinction such as Agda. It turns out that GHC's constraint solver works surprisingly well as an

### 2017-07-11 TyDe2017/2017-05-25/main-tyde.tex
Added files for TyDe 2017.

\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2017}
\acmMonth{1}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[TyDe'17]{2nd Workshop on Type-Driven Development}{September 3, 2017}{Oxford, UK}
\acmYear{2017}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

\setcopyright{none}     
\bibliographystyle{ACM-Reference-Format}

\usepackage{amsmath}
%\usepackage{amsfonts}
%\usepackage{amsthm}
%\usepackage{color}
\usepackage{xcolor}
\usepackage{url}
\usepackage{mathptmx}
\usepackage{doubleequals}
%\usepackage{makeidx}  % allows for indexgeneration
\usepackage{textcomp} % for single quote
\usepackage{fancyvrb} % for indented verbatim

\input{Preamble}
\input{macro}

\begin{document}

\title{Type Safe Redis Queries}
\subtitle{A Case Study of Type--Level Programming in Haskell}

\author{Ting--Yan Lai}
\authornote{Ting--Yan Lai was summer intern from 
the Institute of Computer Science and Engineering,
National Chiao Tung University, Taiwan, 
when part of this work was developed.}
%% \authornote is optional;
\affiliation{
%  \position{Summer Intern}
  \department{Institute of Information Science} 
  \institution{Academia Sinica} 
  \streetaddress{128, Academia Road, Section 2}
  \city{Nangang, Taipei City}
  \postcode{115}
  \country{Taiwan}
}
\email{banacorn@gmail.com}          %% \email is recommended

\author{Tyng--Ruey Chuang}
\affiliation{
%  \position{Associate Research Fellow}
  \department{Institute of Information Science} 
  \institution{Academia Sinica} 
  \streetaddress{128, Academia Road, Section 2}
  \city{Nangang, Taipei City}
  \postcode{115}
  \country{Taiwan}
}
\email{trc@iis.sinica.edu.tw}

\author{Shin--Cheng Mu}
\affiliation{
%  \position{Associate Research Fellow}
  \department{Institute of Information Science} 
  \institution{Academia Sinica} 
  \streetaddress{128, Academia Road, Section 2}
  \city{Nangang, Taipei City}
  \postcode{115}
  \country{Taiwan}
}
\email{scm@iis.sinica.edu.tw}

\begin{abstract}
\Redis{} is an in-memory data structure store, often used as a database, with a
Haskell interface \Hedis{}. \Redis{} is dynamically typed --- a key can be
discarded and re-associated to a value of a different type, and a command,
when fetching a value of a type it does not expect, signals a runtime error. We
develop a domain-specific language that, by exploiting Haskell type-level
programming techniques including indexed monad, type-level literals and closed
type families, keeps track of types of values in the database and statically
guarantees that type errors cannot happen for a class of \Redis{} programs.
\end{abstract}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
<concept_desc>Software and its engineering~Functional languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011050.10011017</concept_id>
<concept_desc>Software and its engineering~Domain specific languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10002951.10002952.10003197</concept_id>
<concept_desc>Information systems~Query languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Functional languages}
\ccsdesc[300]{Software and its engineering~Domain specific languages}
\ccsdesc[300]{Information systems~Query languages}

\keywords{Haskell, Redis, Dependent Type, Type Safety, Key-Value Store}

\maketitle

\input{sections/Introduction}
\input{sections/IndexedMonads}
\input{sections/TypeLevelDict}
\input{sections/Embedding}
\input{sections/Discussions}
\input{sections/Conclusions}

\bibliography{cites}

\end{document}

### 2017-07-11 TyDe2017/2017-05-25/sections/Conclusions.tex
Added files for TyDe 2017.

--- 2016-06-20 aplas16/sections/Conclusions.tex
@@
 indexed free-monads from Hoare Logic specifications. Kiselyov et
-al~\cite{typefun} used indexed monads to track the locks held among a given
+al.~\cite{typefun} used indexed monads to track the locks held among a given
 finite set. The same paper also demonstrated implementation of a variety of
-features including memoization, generic maps, session types, typed
+features including memorization, generic maps, session types, typed
 \texttt{printf} and \texttt{sprintf}, etc., by type-level programming.

### 2017-07-12 TyDe2017/sections/Conclusions.tex
Lightweight revisions to reflect some of the reviewers' comment.

--- 2017-07-11 TyDe2017/2017-05-25/sections/Conclusions.tex
@@
 the data store, what their types ought to be, as well as how the keys and types are updated after the execution. A program can be constructed only if it does
-not throw a runtime type error, and the type also serves as documentation of the
+not throw a runtime type error when it is run with a store whose status matches
+its precondition and when it is the sole client interacting with the store.
+The type also serves as documentation of the
 commands. We believe that it is a neat case study of application of type-level programming.

### 2017-07-13 TyDe2017/main-tyde.tex
Lightweight revisions to reflect some of the reviewers' comment.
Auto stash before merge of "master" and "origin/master"

--- 2017-07-11 TyDe2017/2017-05-25/main-tyde.tex
@@
-\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
+\documentclass[sigplan]{acmart}
+\settopmatter{printfolios=true}
 
@@
 \acmYear{2017}
-\acmMonth{1}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\else\makeatother
-%% Conference information (used by SIGPLAN proceedings format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmConference[TyDe'17]{2nd Workshop on Type-Driven Development}{September 3, 2017}{Oxford, UK}
-\acmYear{2017}
-\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\fi
+\copyrightyear{2017}
+\acmPrice{15.00}
 
-\setcopyright{none}     
+\setcopyright{rightsretained}
 \bibliographystyle{ACM-Reference-Format}
@@
 \input{macro}
+
+\setlength{\mathindent}{10pt}
 
@@
 
-\keywords{Haskell, Redis, Dependent Type, Type Safety, Key-Value Store}
+\keywords{Haskell, Redis, Type-Level Programming Type Safety, Key-Value Store}
 

### 2017-07-13 TyDe2017/sections/Discussions-orig.tex
Auto stash before merge of "master" and "origin/master"

--- 2016-06-12 aplas16/sections/TypeLevelDict.tex
@@
 
-\section{Type-Level Dictionaries}
-\label{sec:type-level-dict}
+\section{Disscussions}
+\label{sec:discussions}
 
-One of the challenges of statically ensuring type correctness of \Redis{},
-which also presents in other stateful languages, is that the type of the value
-associated to a key can be altered after updating. To ensure type correctness,
-we have to keep track of the (\Redis{}) types of all existing keys in a
-{\em dictionary} --- conceptually, a list of pairs of keys and \Redis{} types.
-Each \Redis{} command is embedded in \Popcorn{} as a monadic computation. The
-monad, to be presented in Section~\ref{sec:indexed-monads}, is indexed by
-the dictionaries before and after the computation. In a dependently typed
-programming language (without the so-called ``phase distinction'' ---
-separation between types and terms), this would pose no problem. In Haskell
-however, the dictionaries, to index a monad, has to be a Haskell type as well.
+% \paragraph{Syntax}
+% No one could ignore the glaring shortcoming of the syntax, which occurs mainly
+%  in two places: \emph{symbol singletons} and \emph{indexed monad}. We are hoping
+%  that these issues could be resolved with future syntactic extensions.
 
-In this section we describe how to construct a type-level dictionary, to be
-used with the indexed monad in Section~\ref{sec:indexed-monads}. More operations
-on the dictionary will be presented in Section~\ref{sec:type-level-fun}.
+\paragraph{Returning Inferable Types.} \texttt{GET} is yet another command that
+is invokable only under a ``well-typed or non-existent'' precondition,
+mentioned in Section~\ref{sec:disjunctive-constraints}. It fetches the value of
+a key and, if the key does not exist, returns a special value \texttt{nil}. An
+error is raised if the value is not a string. In \Edis{} the situation is made
+slightly complex, since we parse the string to the type it was supposed to have
+encoded from. The \Edis{} version of \ensuremath{\Varid{get}} could be typed:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{get}{}\<[6]%
+\>[6]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Serialize}\;\Varid{a},\Conid{StringOrNX}\;\Varid{xs}\;\Varid{k}){}\<[E]%
+\\
+\>[6]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Maybe}\;\Varid{a})~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+where \ensuremath{\Conid{StringOrNX}} is defined in Figure~\ref{fig:xxxOrNX}.
 
-Haskell maintains the distinction between values, types, and kinds: values are
-categorized by types, and types are categorized by kinds. The kinds are relatively simple: \ensuremath{\mathbin{*}} is the kind of all {\em lifted} types, while type
-constructors have kinds such as \ensuremath{\mathbin{*}\to \mathbin{*}}, \ensuremath{\mathbin{*}\to \mathbin{*}\to \mathbin{*}}, etc. Consider the
-datatype definitions below:
+The problem with such typing, however, is that \ensuremath{\Varid{a}} cannot be inferred from \ensuremath{\Varid{xs}}
+and \ensuremath{\Varid{k}} when \ensuremath{\Varid{k}} does not appear in \ensuremath{\Varid{xs}}. In such situations, to avoid Haskell
+complaining about ambiguous type, \ensuremath{\Varid{a}} has to be specified by the caller of
+\ensuremath{\Varid{get}}. The user will then be forced to spell out the complete type signature,
+only to make \ensuremath{\Varid{a}} explicit.
+
+We think it is more reasonable to enforce that, when \ensuremath{\Varid{get}} is called, the key
+should exist in the data store. Thus \ensuremath{\Varid{get}} in \Redis{} has the following type:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{get}{}\<[6]%
+\>[6]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Serialize}\;\Varid{a},\Conid{Get}\;\Varid{xs}\;\Varid{k}\mathrel{\sim}\Conid{StringOf}\;\Varid{a}){}\<[E]%
+\\
+\>[6]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Maybe}\;\Varid{a})~~,{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+which requires that \ensuremath{(\Varid{k},\Varid{a})} presents in \ensuremath{\Varid{xs}} and thus \ensuremath{\Varid{a}} is inferrable from
+\ensuremath{\Varid{xs}} and \ensuremath{\Varid{k}}.
+
+\paragraph{Variable Number of Input/Outputs.} Recall that, in
+Section~\ref{sec:proxy-key}, the \Redis{} command \texttt{DEL} takes a variable
+number of keys, while our \Edis{} counterpart takes only one. Some \Redis{}
+commands take a variable number of arguments as inputs, and some returns
+multiple results. Most of them are accurately implemented in \Hedis{}. For
+another example of a variable-number input command, the type of \ensuremath{\Varid{sinter}} in
+\Hedis{} is shown below:
 \begin{hscode}\SaveRestoreHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{data}\;\Conid{Nat}\mathrel{=}\Conid{Zero}\mid \Conid{Suc}\;\Conid{Nat}~~,\qquad\;\mathbf{data}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\mathrel{=}[\mskip1.5mu \mskip1.5mu]\mid \Varid{a}\mathbin{:}[\mskip1.5mu \Varid{a}\mskip1.5mu]~~.{}\<[E]%
+\>[B]{}\Varid{\Conid{Hedis}.sinter}\mathbin{::}[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Redis}\;(\Conid{R} \uplus [\mskip1.5mu \Conid{ByteString}\mskip1.5mu])~~.{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
-The lefthand side is usually seen as having defined a type \ensuremath{\Conid{Nat}\mathbin{::}\mathbin{*}},
-and two value constructors \ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}}. The righthand
-side is how Haskell lists are understood. The {\em kind} of \ensuremath{[\mskip1.5mu \mathbin{\cdot}\mskip1.5mu]} is \ensuremath{\mathbin{*}\to \mathbin{*}},
-since it takes a lifted type \ensuremath{\Varid{a}} to a lifted type \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]}.
-The two value constructors respectively have types \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{(\mathbin{:})\mathbin{::}\Varid{a}\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]}, for all type \ensuremath{\Varid{a}}.
+It takes a list of keys, values of which are all supposed to be sets, and
+computes their intersection (the returned list is the intersected set).
 
-The GHC extension \emph{data kinds}~\cite{promotion}, however, automatically
-promotes certain ``suitable'' types to kinds.\footnote{It is only informally
-described in the GHC manual what types are ``suitable''.} With the extension,
-the \ensuremath{\mathbf{data}} definitions above has an alternative reading: \ensuremath{\Conid{Nat}} is a new kind,
-\ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} is a type having kind \ensuremath{\Conid{Nat}}, and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}} is a type
-constructor, taking a type in kind \ensuremath{\Conid{Nat}} to another type in \ensuremath{\Conid{Nat}}. Whether a
-constructor is promoted can often be inferred from the context. To be more specific, prefixing a constructor with a single quote, such as in \ensuremath{\mbox{\textquotesingle}\Conid{Zero}} and
-\ensuremath{\mbox{\textquotesingle}\Conid{Suc}}, denotes that it is promoted.
-
-The situation of lists is similar: for all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]} is also a kind. For
-all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{k}\mskip1.5mu]} is a type. Given a type \ensuremath{\Varid{x}}
-of kind \ensuremath{\Varid{k}} and a type \ensuremath{\Varid{xs}} of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}, \ensuremath{\Varid{x}\mathbin{:}\Varid{xs}} is again a type of
-kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}.
-Formally, \ensuremath{(\mathbin{:})\mathbin{::}\Varid{k}\to [\mskip1.5mu \Varid{k}\mskip1.5mu]\to [\mskip1.5mu \Varid{k}\mskip1.5mu]}.
-For example,
-\ensuremath{\Conid{Int}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Char}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[]))} is a type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} --- a list of
-(lifted) types. The optional quote denotes that the constructors are promoted.
-The same list can be denoted by a syntax sugar \ensuremath{\mbox{\textquotesingle}[\Conid{Int},\Conid{Char},\Conid{Bool}]}.
-
-Tuples are also promoted. Thus we may put two types in a pair to form another
-type, such as in \ensuremath{\mbox{\textquotesingle}(\Conid{Int},\Conid{Char})}, a type having kind \ensuremath{(\mathbin{*},\mathbin{*})}.
-
-Strings in Haskell are nothing but lists of \ensuremath{\Conid{Char}}s. Regarding promotion,
-however, a \ensuremath{\Conid{String}} can be promoted to a type having kind \ensuremath{\Conid{Symbol}}. \ensuremath{\Conid{Symbol}} is
-a type without a constructor: \ensuremath{\mathbf{data}\;\Conid{Symbol}},
-intended to be used as a promoted kind. In the expression:
+In \Edis{}, for a function to accept a list of keys as input, we have to
+specify that all the keys are in the class \ensuremath{\Conid{KnownSymbol}}. It can be done by
+defining a datatype, indexed by the keys, serving as a witness that they
+are all in \ensuremath{\Conid{KnownSymbol}}. We currently have not implemented such feature and
+leave it as a possible future work. For now, we offer commands that take fixed
+numbers of inputs. The \Edis{} version of \ensuremath{\Varid{sinter}} has type:
 \begin{hscode}\SaveRestoreHook
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\text{\tt \char34 this~is~a~type-level~string~literal\char34}\mathbin{::}\Conid{Symbol}~~,{}\<[E]%
+\>[B]{}\Varid{sinter}\mathbin{::}{}\<[12]%
+\>[12]{}({}\<[15]%
+\>[15]{}\Conid{KnownSymbol}\;\Varid{k}_{1},\Conid{KnownSymbol}\;\Varid{k}_{2},\Conid{Serialize}\;\Varid{a},{}\<[E]%
+\\
+\>[15]{}\Conid{SetOf}\;\Varid{x}\mathrel{\sim}\Conid{Get}\;\Varid{xs}\;\Varid{k}_{1},\Conid{SetOf}\;\Varid{x}\mathrel{\sim}\Conid{Get}\;\Varid{xs}\;\Varid{k}_{2}){}\<[E]%
+\\
+\>[12]{}\Rightarrow \Conid{Proxy}\;\Varid{k}_{1}\to \Conid{Proxy}\;\Varid{k}_{2}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus [\mskip1.5mu \Varid{a}\mskip1.5mu])~~.{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
-the string on the lefthand side of \ensuremath{(\mathbin{::})} is a type, having kind \ensuremath{\Conid{Symbol}}.
 
-With all of these ingredients, we are ready to build our dictionaries, or
-type-level associate lists:
+The function \ensuremath{\Varid{hmset}} in \Hedis{} allows one to set the values of many fields
+in a hash, while \ensuremath{\Varid{hgetall}} returns all the field-value pairs of a hash. They
+have the following types:
 \begin{hscode}\SaveRestoreHook
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{65}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{98}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{type}\;\Conid{DictEmpty}\mathrel{=}\mbox{\textquotesingle}[]~~,{}\<[E]%
+\>[B]{}\Varid{\Conid{Hedis}.hmset}{}\<[16]%
+\>[16]{}\mathbin{::}\Conid{ByteString}{}\<[33]%
+\>[33]{}\to [\mskip1.5mu (\Conid{ByteString},\Conid{ByteString})\mskip1.5mu]{}\<[65]%
+\>[65]{}\to \Conid{Redis}\;(\Conid{R} \uplus \Conid{Status}){}\<[98]%
+\>[98]{}~~,{}\<[E]%
 \\
-\>[B]{}\mathbf{type}\;\Conid{Dict0}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 key\char34},\Conid{Bool})]~~,{}\<[E]%
-\\
-\>[B]{}\mathbf{type}\;\Conid{Dict1}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int}),\mbox{\textquotesingle}(\text{\tt \char34 B\char34},\text{\tt \char34 A\char34})]~~.{}\<[E]%
+\>[B]{}\Varid{\Conid{Hedis}.hgetall}{}\<[16]%
+\>[16]{}\mathbin{::}\Conid{ByteString}\to \Conid{Redis}\;(\Conid{R} \uplus [\mskip1.5mu (\Conid{ByteString},\Conid{ByteString})\mskip1.5mu])~~.{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
-All the entities defined above are types, where \ensuremath{\Conid{Dict0}} and \ensuremath{\Conid{Dict1}}
-have kind \ensuremath{[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}.
+Proper implementations of them in \Edis{} should accept or return a
+\emph{heterogeneous list}~\cite{hetero} --- a list whose elements can be of
+different types. We also leave such functions as a future work.
+
+\paragraph{Not All Safe Redis Programs Can Be Typechecked.}
+Enforcing a typing discipline rules out some programs that are likely to be erroneous, and reduces the number of programs that are allowed. Like all type
+systems, our type system takes a conservative estimation: there are bound to be
+some \Redis{} programs that are not typable in our type system, but do not
+actually throw a type error. We demand that elements in our lists must be
+of the same type, for example, while a \Redis{} program could store in a list
+different types of data, encoded as strings, and still works well.
+
+One innate limitation is that we cannot allow dynamic generation of keys. In
+\Hedis{}, the Haskell program is free to generate arbitrary sequence of keys
+to be used in the data store, which is in general not possible due to the
+static nature of \Edis{}.
+
+\paragraph{Transactions.} Commands in \Redis{} can be wrapped in
+\emph{transactions}. \Redis{} offers two promises regarding commands in a
+transaction. Firstly, all commands in a transaction are serialized and
+executed sequentially, without interruption from another client. Secondly,
+either all of the commands or none are processed.
+
+Support of transactions in \Edis{} is a future work.
+We expect that there would not be too much difficulty --- in an early
+experiment, we have implemented a runtime type checker specifically targeting
+\Redis{} transactions, and we believe that the experience should be applicable
+to static type checking as well.

### 2017-07-13 TyDe2017/sections/Embedding-orig.tex
Auto stash before merge of "master" and "origin/master"

%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%
%
%
%
%
%
% This package provides two environments suitable to take the place
% of hscode, called "plainhscode" and "arrayhscode". 
%
% The plain environment surrounds each code block by vertical space,
% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
% similar to formulas. Note that if these dimensions are changed,
% the spacing around displayed math formulas changes as well.
% All code is indented using \leftskip.
%
% Changed 19.08.2004 to reflect changes in colorcode. Should work with
% CodeGroup.sty.
%
\ReadOnlyOnce{polycode.fmt}%
\makeatletter

\newcommand{\hsnewpar}[1]%
  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}

% can be used, for instance, to redefine the code size, by setting the
% command to \small or something alike
\newcommand{\hscodestyle}{}

% The command \sethscode can be used to switch the code formatting
% behaviour by mapping the hscode environment in the subst directive
% to a new LaTeX environment.

\newcommand{\sethscode}[1]%
  {\expandafter\let\expandafter\hscode\csname #1\endcsname
   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}

% "compatibility" mode restores the non-polycode.fmt layout.

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

\newcommand{\compaths}{\sethscode{compathscode}}

% "plain" mode is the proposed default.
% It should now work with \centering.
% This required some changes. The old version
% is still available for reference as oldplainhscode.

\newenvironment{plainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newenvironment{oldplainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

% Here, we make plainhscode the default environment.

\newcommand{\plainhs}{\sethscode{plainhscode}}
\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
\plainhs

% The arrayhscode is like plain, but makes use of polytable's
% parray environment which disallows page breaks in code blocks.

\newenvironment{arrayhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\parray}%
  {\endparray\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newcommand{\arrayhs}{\sethscode{arrayhscode}}

% The mathhscode environment also makes use of polytable's parray 
% environment. It is supposed to be used only inside math mode 
% (I used it to typeset the type rules in my thesis).

\newenvironment{mathhscode}%
  {\parray}{\endparray}

\newcommand{\mathhs}{\sethscode{mathhscode}}

% texths is similar to mathhs, but works in text mode.

\newenvironment{texthscode}%
  {\(\parray}{\endparray\)}

\newcommand{\texths}{\sethscode{texthscode}}

% The framed environment places code in a framed box.

\def\codeframewidth{\arrayrulewidth}
\RequirePackage{calc}

\newenvironment{framedhscode}%
  {\parskip=\abovedisplayskip\par\noindent
   \hscodestyle
   \arrayrulewidth=\codeframewidth
   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
   \hline\framedhslinecorrect\\{-1.5ex}%
   \let\endoflinesave=\\
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \framedhslinecorrect\endoflinesave{.5ex}\hline
   \endtabular
   \parskip=\belowdisplayskip\par\noindent
   \ignorespacesafterend}

\newcommand{\framedhslinecorrect}[2]%
  {#1[#2]}

\newcommand{\framedhs}{\sethscode{framedhscode}}

% The inlinehscode environment is an experimental environment
% that can be used to typeset displayed code inline.

\newenvironment{inlinehscode}%
  {\(\def\column##1##2{}%
   \let\>\undefined\let\<\undefined\let\\\undefined
   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
   \def\fromto##1##2##3{##3}%
   \def\nextline{}}{\) }%

\newcommand{\inlinehs}{\sethscode{inlinehscode}}

% The joincode environment is a separate environment that
% can be used to surround and thereby connect multiple code
% blocks.

\newenvironment{joincode}%
  {\let\orighscode=\hscode
   \let\origendhscode=\endhscode
   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
   %\let\SaveRestoreHook=\empty
   %\let\ColumnHook=\empty
   %\let\resethooks=\empty
   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
  {\origendhscode
   \global\let\hscode=\orighscode
   \global\let\endhscode=\origendhscode}%

\makeatother
\EndFmtInput
%

\ReadOnlyOnce{Formatting.fmt}%
\makeatletter

\let\Varid\mathit
\let\Conid\mathsf

\def\commentbegin{\quad\{\ }
\def\commentend{\}}

\newcommand{\ty}[1]{\Conid{#1}}
\newcommand{\con}[1]{\Conid{#1}}
\newcommand{\id}[1]{\Varid{#1}}
\newcommand{\cl}[1]{\Varid{#1}}
\newcommand{\opsym}[1]{\mathrel{#1}}

\newcommand\Keyword[1]{\textbf{\textsf{#1}}}
\newcommand\Hide{\mathbin{\downarrow}}
\newcommand\Reveal{\mathbin{\uparrow}}

%% Paper-specific keywords

\makeatother
\EndFmtInput

\section{Embedding \Hedis{} Commands}
\label{sec:embedding-commands}

Having the indexed monads and type-level dictionaries, in this section we
present our embedding of \Hedis{} commands into \Edis{}, while introducing
necessary concepts when they are used.

\subsection{Proxies and Singleton Types}
\label{sec:proxy-key}

The \Hedis{} function \ensuremath{\Varid{del}\mathbin{::}[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Redis}\;(\Conid{R} \uplus \Conid{Integer})} takes a list of keys (encoded to \ensuremath{\Conid{ByteString}}) and removes the entries having those
keys in the database. For some reason to be explained later, we consider an \Edis{}
counterpart that takes only one key. A first attempt may lead to something like
the following:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{del}\mathbin{::}\Conid{String}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}~\mathbin{?})\;(\Conid{R} \uplus \Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.del}\;[\mskip1.5mu \Varid{encode}\;\Varid{key}\mskip1.5mu])~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where the function \ensuremath{\Varid{encode}} converts \ensuremath{\Conid{String}} to \ensuremath{\Conid{ByteString}}. At term-level,
our \ensuremath{\Varid{del}} merely calls \ensuremath{\Varid{\Conid{Hedis}.del}}. At type-level, if the status of the database
before \ensuremath{\Varid{del}} is called meets the constraint represented by the dictionary
\ensuremath{\Varid{xs}}, the status afterwards should meet the constraint \ensuremath{\Conid{Del}\;\Varid{xs}~\mathbin{?}}. The question, however, is what to fill in place of the question mark. It cannot be
\ensuremath{\Conid{Del}\;\Varid{xs}\;\Varid{key}}, since \ensuremath{\Varid{key}} is a runtime value and not a type. How do we smuggle
a runtime value to type-level?

In a language with phase distinction like Haskell, it is certainly impossible
to pass the value of \ensuremath{\Varid{key}} to the type checker if it truly is a runtime value,
for example, a string read from the user. If the value of \ensuremath{\Varid{key}} can be
determined statically, however, {\em singleton types} can be used to represent a
type as a value, thus build a connection between the two realms.

A singleton type is a type that has only one term. When the term is built, it
carries a type that can be inspected by the type checker. The term can be
thought of as a representative of its type at the realm of runtime values. For
our purpose, we will use the following type \ensuremath{\Conid{Proxy}}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Proxy}\;\Varid{t}\mathrel{=}\Conid{Proxy}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
For every type \ensuremath{\Varid{t}}, \ensuremath{\Conid{Proxy}\;\Varid{t}} is a type that has only one term: \ensuremath{\Conid{Proxy}}.%
\footnote{While giving the same name to both the type and the term can be very
confusing, it is unfortunately a common practice in the Haskell community.}
To call \ensuremath{\Varid{del}}, instead of passing a key as a \ensuremath{\Conid{String}} value, we give it a proxy
with a specified type:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{del}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 A\char34})~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where \ensuremath{\text{\tt \char34 A\char34}} is not a value, but a string lifted to a type (of kind \ensuremath{\Conid{Symbol}}).
Now that the type checker has access to the key, the type of \ensuremath{\Varid{del}} can be
\ensuremath{\Conid{Proxy}\;\Varid{k}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\;\Varid{k})\;(\Conid{R} \uplus \Conid{Integer})}.

The next problem is that, \ensuremath{\Varid{del}}, at term level, gets only a value constructor
\ensuremath{\Conid{Proxy}} without further information, while it needs to pass a \ensuremath{\Conid{ByteString}} key
to \ensuremath{\Varid{\Conid{Hedis}.del}}. Every concrete string literal lifted to a type, for example,
\ensuremath{\text{\tt \char34 A\char34}}, belongs to a type class \ensuremath{\Conid{KnownSymbol}}. For all type \ensuremath{\Varid{k}} in \ensuremath{\Conid{KnownSymbol}},
the function \ensuremath{\Varid{symbolVal}}:\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Varid{symbolVal}\mathbin{::}\Conid{KnownSymbol}\;\Varid{k}\Rightarrow \Varid{proxy}\;\Varid{k}\to \Conid{String}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
retrieves the string associated with a type-level literal that is known at
compile time. In summary, \ensuremath{\Varid{del}} can be implemented as:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{45}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{del}{}\<[6]%
\>[6]{}\mathbin{::}\Conid{KnownSymbol}\;\Varid{k}{}\<[E]%
\\
\>[6]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\;\Varid{k})\;(\Conid{R} \uplus \Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.del}\;[\mskip1.5mu \Varid{encodeKey}\;\Varid{key}\mskip1.5mu]){}\<[45]%
\>[45]{}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where \ensuremath{\Varid{encodeKey}\mathrel{=}\Varid{encode}\mathbin{\cdot}\Varid{symbolVal}}.

A final note: the function \ensuremath{\Varid{encode}}, from the Haskell library {\sc cereal},
helps to convert certain datatypes that are {\em serializable} into \ensuremath{\Conid{ByteString}}.
The function and its dual \ensuremath{\Varid{decode}} will be used more later.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{encode}{}\<[9]%
\>[9]{}\mathbin{::}\Conid{Serialize}\;\Varid{a}\Rightarrow \Varid{a}\to \Conid{ByteString}~~,{}\<[E]%
\\
\>[B]{}\Varid{decode}{}\<[9]%
\>[9]{}\mathbin{::}\Conid{Serialize}\;\Varid{a}\Rightarrow \Conid{ByteString}\to \Conid{Either}\;\Conid{String}\;\Varid{a}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\subsection{Automatic Serialization}
\label{sec:polymorphic-redis}

As mentioned before, while \Redis{} provide a number of container types
including lists, sets, and hash, etc., the primitive type is string. \Hedis{}
programmers manually convert data of other types to strings before saving them
into the data store. In \Edis{}, we wish to save some of the effort for the
programmers, as well as keeping a careful record of the intended types of the
strings in the data store.

To keep track of intended types of strings in the data store, we define the
following types (that have no terms):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{StringOf}{}\<[16]%
\>[16]{}\mathbin{::}\mathbin{*}\to \mathbin{*}~~,{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\Conid{ListOf}{}\<[16]%
\>[16]{}\mathbin{::}\mathbin{*}\to \mathbin{*}~~,{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\Conid{SetOf}{}\<[16]%
\>[16]{}\mathbin{::}\mathbin{*}\to \mathbin{*}~~...{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
If a key is associated with, for example, \ensuremath{\Conid{StringOf}\;\Conid{Int}} in our dictionary, we
mean that its value in the data store was serialized from an \ensuremath{\Conid{Int}} and should be
used as an \ensuremath{\Conid{Int}}. Types \ensuremath{\Conid{ListOf}\;\Varid{a}} and \ensuremath{\Conid{SetOf}\;\Varid{a}}, respectively, denotes that the
value is a list or a set of type \ensuremath{\Varid{a}}.

While the \ensuremath{\Varid{set}} command in \Hedis{} always writes a string to the data store,
the corresponding \ensuremath{\Varid{set}} in \Redis{} applies to any serializable type (those
in the class \ensuremath{\Conid{Serialize}}), and performs the encoding for the user:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{set}{}\<[6]%
\>[6]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Serialize}\;\Varid{a}){}\<[E]%
\\
\>[6]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Varid{a}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Set}\;\Varid{xs}\;\Varid{k}\;(\Conid{StringOf}\;\Varid{a}))\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\\
\>[B]{}\Varid{set}\;\Varid{key}\;\Varid{val}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.set}\;(\Varid{encodeKey}\;\Varid{key})\;(\Varid{encode}\;\Varid{val}))~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
For example, executing \ensuremath{\Varid{set}\;(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\text{\tt \char34 A\char34})\;\Conid{True}} updates the dictionary
with an entry \ensuremath{\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{StringOf}\;\Conid{Bool})}. If \ensuremath{\text{\tt \char34 A\char34}} is not in the dictionary,
this entry is added; otherwise the old type of \ensuremath{\text{\tt \char34 A\char34}} is updated to
\ensuremath{\Conid{StringOf}\;\Conid{Bool}}.

\Redis{} command \texttt{INCR} reads the (string) value of the given key, parses
it as an integer, and increments it by one, before storing it back. The command
\texttt{INCRBYFLOAT} increments the floating point value of a key by a given
amount. They are defined in \Edis{} below:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{incr}{}\<[7]%
\>[7]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Get}\;\Varid{xs}\;\Varid{k}\mathrel{\sim}\Conid{StringOf}\;\Conid{Integer}){}\<[E]%
\\
\>[7]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{incr}\;\Varid{key}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.incr}\;(\Varid{encodeKey}\;\Varid{key}))~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{incrbyfloat}{}\<[14]%
\>[14]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Get}\;\Varid{xs}\;\Varid{k}\mathrel{\sim}\Conid{StringOf}\;\Conid{Double}){}\<[E]%
\\
\>[14]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Double}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Double}){}\<[E]%
\\
\>[B]{}\Varid{incrbyfloat}\;\Varid{key}\;\Varid{eps}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.incrbyfloat}\;(\Varid{encodeKey}\;\Varid{key})\;\Varid{eps})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Notice the use of (\ensuremath{\mathrel{\sim}}), \emph{equality constraints}~\cite{typeeq}, to enforce
that the intended type of the value of \ensuremath{\Varid{k}} must respectively be \ensuremath{\Conid{Integer}} and
\ensuremath{\Conid{Double}}. The function \ensuremath{\Varid{incr}} is only allowed to be called in a context where
the type checker is able to reduce \ensuremath{\Conid{Get}\;\Varid{xs}\;\Varid{k}} to \ensuremath{\Conid{StringOf}\;\Conid{Integer}} ---
recall that when \ensuremath{\Varid{k}} is not in \ensuremath{\Varid{xs}}, \ensuremath{\Conid{Get}\;\Varid{xs}\;\Varid{k}} cannot be fully reduced. The
type of \ensuremath{\Varid{incrbyfloat}} works in a similar way.

\subsection{Disjunctive Constraints}
\label{sec:disjunctive-constraints}

Recall, from Section \ref{sec:introduction}, that commands \texttt{LPUSH key
val} and \texttt{LLEN key} return normally either when \ensuremath{\Varid{key}} presents in the
data store and is assigned a list, or when \ensuremath{\Varid{key}} does not present at all.
What we wish to have in their constraint is thus a predicate equivalent to \ensuremath{\Conid{Get}\;\Varid{xs}\;\Varid{k}\doubleequals\Conid{ListOf}\;\Varid{a}\mathrel{\vee}\neg \;(\Conid{Member}\;\Varid{xs}\;\Varid{k})}. In fact, many \Redis{} commands
are invokable under such ``well-typed, or non-existent'' precondition.

To impose a conjunctive constraint \ensuremath{\Conid{P}\mathrel{\wedge}\Conid{Q}}, one may simply put them both in the
type: \ensuremath{(\Conid{P},\Conid{Q})\Rightarrow \mathbin{...}}. Expressing disjunctive constraints is only slightly
harder, thanks to our type-level functions. We may thus write the predicate as:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Conid{Get}\;\Varid{xs}\;\Varid{k}\mathrel{\sim}\Conid{ListOf}\;\Varid{a}\mathrel{\Conid{`Or`}}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{k})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
To avoid referring to \ensuremath{\Varid{a}}, which might not exist, we define an auxiliary predicate \ensuremath{\Conid{IsList}\mathbin{::}\mathbin{*}\to \Conid{Bool}} such that \ensuremath{\Conid{IsList}\;\Varid{t}} reduces to \ensuremath{\mbox{\textquotesingle}\Conid{True}}
only if \ensuremath{\Varid{t}\mathrel{=}\Conid{ListOf}\;\Varid{a}}. As many \Redis{} commands are invokable only under such
``well-typed, or non-existent'' precondition, we give names to such constraints,
as seen in Figure~\ref{fig:xxxOrNX}.

\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{IsList}\;(\Varid{t}\mathbin{::}\mathbin{*})\mathbin{::}\Conid{Bool}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{IsList}\;(\Conid{ListOf}\;\Varid{a}){}\<[24]%
\>[24]{}\mathrel{=}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{IsList}\;\Varid{t}{}\<[24]%
\>[24]{}\mathrel{=}\mbox{\textquotesingle}\Conid{False}{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{IsSet}\;(\Varid{t}\mathbin{::}\mathbin{*})\mathbin{::}\Conid{Bool}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{IsSet}\;(\Conid{SetOf}\;\Varid{a}){}\<[22]%
\>[22]{}\mathrel{=}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{IsSet}\;\Varid{t}{}\<[22]%
\>[22]{}\mathrel{=}\mbox{\textquotesingle}\Conid{False}{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{IsString}\;(\Varid{t}\mathbin{::}\mathbin{*})\mathbin{::}\Conid{Bool}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{IsString}\;(\Conid{StringOf}\;\Varid{a}){}\<[28]%
\>[28]{}\mathrel{=}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{IsString}\;\Varid{t}{}\<[28]%
\>[28]{}\mathrel{=}\mbox{\textquotesingle}\Conid{False}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{type}\;\Conid{ListOrNX}\;{}\<[18]%
\>[18]{}\Varid{xs}\;\Varid{k}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}(\Conid{IsList}\;{}\<[14]%
\>[14]{}(\Conid{Get}\;\Varid{xs}\;\Varid{k})\mathrel{\Conid{`Or`}}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{k}))\mathrel{\sim}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Conid{SetOrNX}\;{}\<[18]%
\>[18]{}\Varid{xs}\;\Varid{k}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}(\Conid{IsSet}\;{}\<[14]%
\>[14]{}(\Conid{Get}\;\Varid{xs}\;\Varid{k})\mathrel{\Conid{`Or`}}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{k}))\mathrel{\sim}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Conid{StringOrNX}\;{}\<[18]%
\>[18]{}\Varid{xs}\;\Varid{k}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}(\Conid{IsString}\;{}\<[14]%
\>[14]{}(\Conid{Get}\;\Varid{xs}\;\Varid{k})\mathrel{\Conid{`Or`}}\Conid{Not}\;(\Conid{Member}\;\Varid{xs}\;\Varid{k}))\mathrel{\sim}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{The ``well-typed, or non-existent'' constraints.}
\label{fig:xxxOrNX}
\end{figure}

The \Edis{} counterpart of \texttt{LPUSH} and \texttt{LLEN} are therefore:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{lpush}{}\<[8]%
\>[8]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Serialize}\;\Varid{a},\Conid{ListOrNX}\;\Varid{xs}\;\Varid{k}){}\<[E]%
\\
\>[8]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Varid{a}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Set}\;\Varid{xs}\;\Varid{k}\;(\Conid{ListOf}\;\Varid{a}))\;(\Conid{R} \uplus \Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{lpush}\;\Varid{key}\;\Varid{val}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.lpush}\;(\Varid{encodeKey}\;\Varid{key})\;[\mskip1.5mu \Varid{encode}\;\Varid{val}\mskip1.5mu])~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{llen}{}\<[7]%
\>[7]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{ListOrNX}\;\Varid{xs}\;\Varid{k}){}\<[E]%
\\
\>[7]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{llen}\;\Varid{key}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.llen}\;(\Varid{encodeKey}\;\Varid{key}))~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Similarly, the type of \ensuremath{\Varid{sadd}}, a function we have talked about a lot,
is given below:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{sadd}{}\<[7]%
\>[7]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Serialize}\;\Varid{a},\Conid{SetOrNX}\;\Varid{xs}\;\Varid{k}){}\<[E]%
\\
\>[7]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Varid{a}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Set}\;\Varid{xs}\;\Varid{k}\;(\Conid{SetOf}\;\Varid{a}))\;(\Conid{R} \uplus \Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{sadd}\;\Varid{key}\;\Varid{val}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.sadd}\;(\Varid{encodeKey}\;\Varid{key})\;[\mskip1.5mu \Varid{encode}\;\Varid{val}\mskip1.5mu])~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

To see a command with a more complex type, consider \ensuremath{\Varid{setnx}}, which
uses the type-level function \ensuremath{\Conid{If}} defined in Section \ref{sec:type-fun}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{setnx}{}\<[8]%
\>[8]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Serialize}\;\Varid{a}){}\<[E]%
\\
\>[8]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Varid{a}\to \Conid{Edis}\;\Varid{xs}\;{}\<[36]%
\>[36]{}(\Conid{If}\;(\Conid{Member}\;\Varid{xs}\;\Varid{k})\;\Varid{xs}\;(\Conid{Set}\;\Varid{xs}\;\Varid{k}\;(\Conid{StringOf}\;\Varid{a})))\;{}\<[E]%
\\
\>[36]{}(\Conid{Either}\;\Conid{Reply}\;\Conid{Bool}){}\<[E]%
\\
\>[B]{}\Varid{setnx}\;\Varid{key}\;\Varid{val}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.setnx}\;(\Varid{encodeKey}\;\Varid{key})\;(\Varid{encode}\;\Varid{val}))~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
From the type one can see that \ensuremath{\Varid{setnx}\;\Varid{key}\;\Varid{val}} creates a new entry \ensuremath{(\Varid{key},\Varid{val})}
in the data store only if \ensuremath{\Varid{key}} is fresh. The type of \ensuremath{\Varid{setnx}} computes a
postcondition for static checking, as well as serving as a good documentation
for its semantics.

\subsection{Hashes}

{\em Hash} is a useful datatype supported by \Redis{}. While the \Redis{} data
store can be seen as a set of key/value pairs, a hash is itself a set of
field/value pairs. The following commands assigns a hash to key \texttt{user}.
The fields are \texttt{name}, \texttt{birthyear}, and \texttt{verified},
respectively with values \texttt{banacorn}, \texttt{1992}, and \texttt{1}.
\begin{Verbatim}[xleftmargin=.4in]
redis> hmset user name banacorn birthyear 1992 verified 1
OK
redis> hget user name
"banacorn"
redis> hget user birthyear
"1992"
\end{Verbatim}

For a hash to be useful, we should allow the fields to have different types. To
keep track of types of fields in a hash, \ensuremath{\Conid{HashOf}} takes a list of \ensuremath{(\Conid{Symbol},\mathbin{*})}
pairs:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{HashOf}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]\to \mathbin{*}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
By having an entry \ensuremath{(\Varid{k},\Conid{HashOF}\;\Varid{ys})} in a dictionary, we denote that the value of
key \ensuremath{\Varid{k}} is a hash whose fields and their types are specified by \ensuremath{\Varid{ys}}, which is
also a dictionary.

\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
\column{43}{@{}>{\hspre}l<{\hspost}@{}}%
\column{44}{@{}>{\hspre}l<{\hspost}@{}}%
\column{45}{@{}>{\hspre}l<{\hspost}@{}}%
\column{76}{@{}>{\hspre}l<{\hspost}@{}}%
\column{77}{@{}>{\hspre}l<{\hspost}@{}}%
\column{81}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{GetHash}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{k}\mathbin{::}\Conid{Symbol})\;(\Varid{f}\mathbin{::}\Conid{Symbol})\mathbin{::}\mathbin{*}\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{GetHash}\;(\mbox{\textquotesingle}(\Varid{k},\Conid{HashOf}\;\Varid{hs}){}\<[35]%
\>[35]{}\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[44]%
\>[44]{}\Varid{k}\;\Varid{f}\mathrel{=}\Conid{Get}\;\Varid{hs}\;\Varid{f}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{GetHash}\;(\mbox{\textquotesingle}(\Varid{l},\Varid{y}){}\<[35]%
\>[35]{}\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[44]%
\>[44]{}\Varid{k}\;\Varid{f}\mathrel{=}\Conid{GetHash}\;\Varid{xs}\;\Varid{k}\;\Varid{f}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{SetHash}\;{}\<[22]%
\>[22]{}(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{k}\mathbin{::}\Conid{Symbol})\;{}\<[E]%
\\
\>[22]{}(\Varid{f}\mathbin{::}\Conid{Symbol})\;(\Varid{a}\mathbin{::}\mathbin{*})\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{SetHash}\;\mbox{\textquotesingle}[~]\;{}\<[44]%
\>[44]{}\Varid{k}\;\Varid{f}\;\Varid{a}\mathrel{=}\mbox{\textquotesingle}(\Varid{k},\Conid{HashOf}\;(\Conid{Set}\;\mbox{\textquotesingle}[~]\;\Varid{f}\;\Varid{a}))\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[~]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{SetHash}\;(\mbox{\textquotesingle}(\Varid{k},\Conid{HashOf}\;\Varid{hs}){}\<[35]%
\>[35]{}\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[44]%
\>[44]{}\Varid{k}\;\Varid{f}\;\Varid{a}\mathrel{=}\mbox{\textquotesingle}(\Varid{k},\Conid{HashOf}\;(\Conid{Set}\;\Varid{hs}\;{}\<[77]%
\>[77]{}\Varid{f}\;\Varid{a}))\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{SetHash}\;(\mbox{\textquotesingle}(\Varid{l},\Varid{y}){}\<[35]%
\>[35]{}\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[44]%
\>[44]{}\Varid{k}\;\Varid{f}\;\Varid{a}\mathrel{=}\mbox{\textquotesingle}(\Varid{l},\Varid{y}){}\<[81]%
\>[81]{}\mathbin{\mbox{\textquotesingle}\!:}\Conid{SetHash}\;\Varid{xs}\;\Varid{k}\;\Varid{f}\;\Varid{a}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{DelHash}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{k}\mathbin{::}\Conid{Symbol})\;(\Varid{f}\mathbin{::}\Conid{Symbol})\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{DelHash}\;\mbox{\textquotesingle}[~]\;{}\<[43]%
\>[43]{}\Varid{k}\;\Varid{f}\mathrel{=}\mbox{\textquotesingle}[~]{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{DelHash}\;(\mbox{\textquotesingle}(\Varid{k},\Conid{HashOf}\;\Varid{hs})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[43]%
\>[43]{}\Varid{k}\;\Varid{f}\mathrel{=}\mbox{\textquotesingle}(\Varid{k},\Conid{HashOf}\;(\Conid{Del}\;\Varid{hs}\;\Varid{f}))\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{DelHash}\;(\mbox{\textquotesingle}(\Varid{l},\Varid{y}){}\<[34]%
\>[34]{}\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[43]%
\>[43]{}\Varid{k}\;\Varid{f}\mathrel{=}\mbox{\textquotesingle}(\Varid{l},\Varid{y}){}\<[76]%
\>[76]{}\mathbin{\mbox{\textquotesingle}\!:}\Conid{DelHash}\;\Varid{xs}\;\Varid{k}\;\Varid{f}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{MemHash}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{k}\mathbin{::}\Conid{Symbol})\;(\Varid{f}\mathbin{::}\Conid{Symbol})\mathbin{::}\Conid{Bool}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{MemHash}\;\mbox{\textquotesingle}[~]\;{}\<[45]%
\>[45]{}\Varid{k}\;\Varid{f}\mathrel{=}\mbox{\textquotesingle}\Conid{False}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{MemHash}\;(\mbox{\textquotesingle}(\Varid{k},\Conid{HashOf}\;\Varid{hs}){}\<[36]%
\>[36]{}\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[45]%
\>[45]{}\Varid{k}\;\Varid{f}\mathrel{=}\Conid{Member}\;\Varid{hs}\;\Varid{f}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{MemHash}\;(\mbox{\textquotesingle}(\Varid{k},\Varid{x}){}\<[36]%
\>[36]{}\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[45]%
\>[45]{}\Varid{k}\;\Varid{f}\mathrel{=}\mbox{\textquotesingle}\Conid{False}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{MemHash}\;(\mbox{\textquotesingle}(\Varid{l},\Varid{y}){}\<[36]%
\>[36]{}\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[45]%
\>[45]{}\Varid{k}\;\Varid{f}\mathrel{=}\Conid{MemHash}\;\Varid{xs}\;\Varid{k}\;\Varid{f}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{Type-level operations for dictionaries with hashes.}
\label{fig:xxxHash}
\end{figure}

Figure \ref{fig:xxxHash} presents some operations we need on dictionaries when
dealing with hashes. Let \ensuremath{\Varid{xs}} be a dictionary, \ensuremath{\Conid{GetHash}\;\Varid{xs}\;\Varid{k}\;\Varid{f}} returns the type
of field \ensuremath{\Varid{f}} in the hash assigned to key \ensuremath{\Varid{k}}, if both \ensuremath{\Varid{k}} and \ensuremath{\Varid{f}} exists.
\ensuremath{\Conid{SetHash}\;\Varid{xs}\;\Varid{k}\;\Varid{f}\;\Varid{a}} assigns the type \ensuremath{\Varid{a}} to the field \ensuremath{\Varid{f}} of hash \ensuremath{\Varid{k}}; if either
\ensuremath{\Varid{f}} or \ensuremath{\Varid{k}} does not exist, the hash/field is created. \ensuremath{\Conid{Del}\;\Varid{xs}\;\Varid{k}\;\Varid{f}} removes a
field, while \ensuremath{\Conid{MemHash}\;\Varid{xs}\;\Varid{k}\;\Varid{f}} checks whether the key \ensuremath{\Varid{k}} exists in \ensuremath{\Varid{xs}}, and its
value is a hash having field \ensuremath{\Varid{f}}. Their definitions make use of functions \ensuremath{\Conid{Get}},
\ensuremath{\Conid{Set}}, and \ensuremath{\Conid{Member}} defined for dictionaries.

Once those type-level functions are defined, embedding of \Hedis{} commands for
hashes is more or less routine. For example, functions \ensuremath{\Varid{hset}} and \ensuremath{\Varid{hget}}
are shown below. Note that, instead of \ensuremath{\Varid{hmset}} (available in \Hedis{}), we
provide a function \ensuremath{\Varid{hset}} that assigns fields and values one pair at a time.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{hset}{}\<[7]%
\>[7]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{KnownSymbol}\;\Varid{f},\Conid{Serialize}\;\Varid{a},\Conid{HashOrNX}\;\Varid{xs}\;\Varid{k}){}\<[E]%
\\
\>[7]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Proxy}\;\Varid{f}\to \Varid{a}{}\<[E]%
\\
\>[7]{}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{SetHash}\;\Varid{xs}\;\Varid{k}\;\Varid{f}\;(\Conid{StringOf}\;\Varid{a}))\;(\Conid{R} \uplus \Conid{Bool}){}\<[E]%
\\
\>[B]{}\Varid{hset}\;\Varid{key}\;\Varid{field}\;\Varid{val}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{Edis}\;(\Varid{\Conid{Hedis}.hset}\;(\Varid{encodeKey}\;\Varid{key})\;(\Varid{encodeKey}\;\Varid{field})\;(\Varid{encode}\;\Varid{val}))~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{hget}{}\<[7]%
\>[7]{}\mathbin{::}({}\<[13]%
\>[13]{}\Conid{KnownSymbol}\;\Varid{k},\Conid{KnownSymbol}\;\Varid{f},\Conid{Serialize}\;\Varid{a},{}\<[E]%
\\
\>[13]{}\Conid{StringOf}\;\Varid{a}\mathrel{\sim}\Conid{GetHash}\;\Varid{xs}\;\Varid{k}\;\Varid{f}){}\<[E]%
\\
\>[7]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Proxy}\;\Varid{f}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Maybe}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\Varid{hget}\;\Varid{key}\;\Varid{field}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{Edis}\;(\Varid{\Conid{Hedis}.hget}\;(\Varid{encodeKey}\;\Varid{key})\;(\Varid{encodeKey}\;\Varid{field})\bind \Varid{decodeAsMaybe})~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where \ensuremath{\Varid{decodeAsMaybe}\mathbin{::}\Conid{Serialize}\;\Varid{a}\Rightarrow (\Conid{R} \uplus \Conid{Maybe}\;\Conid{ByteString})\to \Conid{Redis}\;(\Conid{R} \uplus \Conid{Maybe}\;\Varid{a})}, using the function \ensuremath{\Varid{decode}}
mentioned in Section \ref{sec:proxy-key}, parses the \ensuremath{\Conid{ByteString}} in
\ensuremath{\Conid{R} \uplus \Conid{Maybe}\;\anonymous } to type \ensuremath{\Varid{a}}. The definition is a bit tedious but
routine.

Note that, instead of \ensuremath{\Varid{hmset}} (available in \Hedis{}), we
provide a function \ensuremath{\Varid{hset}} that assigns fields and values one pair at at time.
We will talk about difficulties of implementing \ensuremath{\Varid{hmset}} in
Section~\ref{sec:discussions}.

\subsection{Assertions}
\label{sec:assertions}

% Consider the following scenario: We want to retrieve the value of some existing
% key, say, |"ab initio"|, with the function\footnote{The semantics of |get| in
% \Redis{} is actually more forgiving. See Section~\ref{sec:discussions}.}:
% \begin{spec}
% get  :: (KnownSymbol k, Serialize a, StringOf a ~ Get xs k)
%      => Proxy k -> Edis xs xs (EitherReply (Maybe a)) {-"~~."-}
% \end{spec}
%
% However, Haskell would complain that, the key |"ab initio"| was not
% before seen in the dictionary. So we could never write programs to retrieve the
% value of |"ab initio"|, unless its creation was witnessed by the type checker.

Finally, the creation/update behavior of \Redis{} functions is, in our opinion,
very error-prone. It might be preferable if we can explicit declare some new
keys, after ensuring that they do not already exist (in our types). This can be done below:
% It might be preferable if we can explicitly declare some new keys, under the
%  precondition that they do not already exist in our types. This can be done as
%  follows:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{declare}{}\<[10]%
\>[10]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Member}\;\Varid{xs}\;\Varid{k}\mathrel{\sim}\Conid{False}){}\<[E]%
\\
\>[10]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Proxy}\;\Varid{a}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Set}\;\Varid{xs}\;\Varid{k}\;\Varid{a})\;(){}\<[E]%
\\
\>[B]{}\Varid{declare}\;\Varid{key}\;\Varid{typ}\mathrel{=}\Conid{Edis}\;(\Varid{return}\;())~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% renounce :: (KnownSymbol s, Member xs s ~ True)
%         => Proxy s -> Edis xs (Del xs s) ()
% renounce s = Edis $ return () {-"~~."-}
The command \ensuremath{\Varid{declare}\;\Varid{key}\;\Varid{typ}}, where \ensuremath{\Varid{typ}} is the proxy of \ensuremath{\Varid{a}}, adds a fresh
 \ensuremath{\Varid{key}} with type \ensuremath{\Varid{a}} into the dictionary. Notice that \ensuremath{\Varid{declare}} does nothing at
 term level, but simply returns \ensuremath{()}, since it only has effects on types.

% The key is not actually created yet.
% The declaration, however, ensures that if |key| is actually created, perhaps
% by one of those ``well-typed or non-existent'' command, its type must be |a|.
% The command |start| initializes the dictionary to the empty list:

Another command for type level assertion, \ensuremath{\Varid{start}}, initializes the dictionary to
 the empty list, comes in handy when starting a series of \Edis{} commands:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{start}\mathbin{::}\Conid{Edis}\;\mbox{\textquotesingle}[~]\;\mbox{\textquotesingle}[~]\;(){}\<[E]%
\\
\>[B]{}\Varid{start}\mathrel{=}\Conid{Edis}\;(\Varid{return}\;())~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\subsection{A Larger Example}

We present a larger example as a summary. The task is to store a queue of
messages in \Redis{}. Messages are represented by a \ensuremath{\Conid{ByteString}} and an
\ensuremath{\Conid{Integer}} identifier:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Message}\mathrel{=}\Conid{Msg}\;\Conid{ByteString}\;\Conid{Integer}\;\mathbf{deriving}\;(\Conid{Show},\Conid{Generic})~~,{}\<[E]%
\\
\>[B]{}\mathbf{instance}\;\Conid{Serialize}\;\Conid{Message}\;\mathbf{where}\;{}\<[35]%
\>[35]{}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

In the data store, the queue is represented by a list. Before pushing a message
into the queue, we increment \ensuremath{\Varid{counter}}, a key storing a counter, and use it as the
identifier of the message:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{push}{}\<[7]%
\>[7]{}\mathbin{::}(\Conid{StringOfIntegerOrNX}\;\Varid{xs}\;\text{\tt \char34 counter\char34},\Conid{ListOrNX}\;\Varid{xs}\;\text{\tt \char34 queue\char34}){}\<[E]%
\\
\>[7]{}\Rightarrow \Conid{ByteString}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Set}\;\Varid{xs}\;\text{\tt \char34 queue\char34}\;(\Conid{ListOf}\;\Conid{Message}))\;(\Conid{R} \uplus \Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{push}\;\Varid{msg}\mathrel{=}{}\<[13]%
\>[13]{}\Varid{incr}\;\Varid{kCounter}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\mathbin{`\Varid{bind}`}{}\<[13]%
\>[13]{}\lambda \Varid{i}\to \Varid{lpush}\;\Varid{kQueue}\;(\Conid{Msg}\;\Varid{msg}\;(\Varid{fromRight}\;\Varid{i}))~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%​
where \ensuremath{\Varid{fromRight}\mathbin{::}\Conid{Either}\;\Varid{a}\;\Varid{b}\to \Varid{b}} extracts the value wrapped by constructor
\ensuremath{\Conid{Right}}, and the constraint \ensuremath{\Conid{StringOfIntegerOrNX}\;\Varid{xs}\;\Varid{k}} holds if either \ensuremath{\Varid{k}}
presents in \ensuremath{\Varid{xs}} and is converted from an \ensuremath{\Conid{Integer}}, or \ensuremath{\Varid{k}} does not
present in \ensuremath{\Varid{xs}}. For brevity, the proxies are given names: \\
\noindent{\centering %\small
\begin{minipage}[b]{0.4\linewidth}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{kCounter}\mathbin{::}{}\<[14]%
\>[14]{}\Conid{Proxy}\;\text{\tt \char34 counter\char34}{}\<[E]%
\\
\>[B]{}\Varid{kCounter}\mathrel{=}{}\<[14]%
\>[14]{}\Conid{Proxy}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{minipage}
\begin{minipage}[b]{0.4\linewidth}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{kQueue}\mathbin{::}{}\<[12]%
\>[12]{}\Conid{Proxy}\;\text{\tt \char34 queue\char34}{}\<[E]%
\\
\>[B]{}\Varid{kQueue}\mathrel{=}{}\<[12]%
\>[12]{}\Conid{Proxy}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{minipage}}\\
To pop a message we use the function \ensuremath{\Varid{rpop}} which, given a key associated with
a list, extracts the rightmost element of the list
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{pop}{}\<[6]%
\>[6]{}\mathbin{::}(\Conid{Get}\;\Varid{xs}\;\text{\tt \char34 queue\char34}\mathrel{\sim}\Conid{ListOf}\;\Conid{Message}){}\<[E]%
\\
\>[6]{}\Rightarrow \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Maybe}\;\Conid{Message}){}\<[E]%
\\
\>[B]{}\Varid{pop}\mathrel{=}\Varid{rpop}\;\Varid{kQueue}~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{rpop}{}\<[7]%
\>[7]{}\mathbin{::}(\Conid{KnownSymbol}\;\Varid{k},\Conid{Serialize}\;\Varid{a},\Conid{Get}\;\Varid{xs}\;\Varid{k}\mathrel{\sim}\Conid{ListOf}\;\Varid{a}){}\<[E]%
\\
\>[7]{}\Rightarrow \Conid{Proxy}\;\Varid{k}\to \Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Maybe}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\Varid{rpop}\;\Varid{key}\mathrel{=}\Conid{Edis}\;(\Varid{\Conid{Hedis}.rpop}\;(\Varid{encodeKey}\;\Varid{key})\bind \Varid{decodeAsMaybe})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
To get things going, the main program builds a connection with the \Redis{}
server, runs an embedded program \ensuremath{\Varid{prog}}, and prints the result, which in this
example is \ensuremath{\text{\tt \char34 hello\char34}}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}c<{\hspost}@{}}%
\column{4E}{@{}l@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{main}\mathbin{::}\Conid{IO}\;(){}\<[E]%
\\
\>[B]{}\Varid{main}\mathrel{=}\mathbf{do}\;{}\<[12]%
\>[12]{}\Varid{conn}{}\<[20]%
\>[20]{}\leftarrow \Varid{connect}\;\Varid{defaultConnectInfo}{}\<[E]%
\\
\>[12]{}\Varid{result}{}\<[20]%
\>[20]{}\leftarrow \Varid{runRedis}\;\Varid{conn}\;(\Varid{unEdis}\;(\Varid{start}\mathbin{>\!\!\!>\!\!\!>}\Varid{prog})){}\<[E]%
\\
\>[12]{}\Varid{print}\;\Varid{result}~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{prog}\mathrel{=}{}\<[9]%
\>[9]{}\Varid{declare}\;\Varid{kCounter}\;{}\<[27]%
\>[27]{}(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbin{>\!\!\!>\!\!\!>}{}\<[4E]%
\>[9]{}\Varid{declare}\;\Varid{kQueue}\;{}\<[27]%
\>[27]{}(\Conid{Proxy}\mathbin{::}\Conid{Proxy}\;(\Conid{ListOf}\;\Conid{Message})){}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbin{>\!\!\!>\!\!\!>}{}\<[4E]%
\>[9]{}\Varid{push}\;\text{\tt \char34 hello\char34}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbin{>\!\!\!>\!\!\!>}{}\<[4E]%
\>[9]{}\Varid{push}\;\text{\tt \char34 world\char34}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\mathbin{>\!\!\!>\!\!\!>}{}\<[4E]%
\>[9]{}\Varid{pop}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where the monadic sequencing operator \ensuremath{(\mathbin{>\!\!\!>\!\!\!>})} is defined by:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathbin{>\!\!\!>\!\!\!>})\mathbin{::}\Conid{IMonad}\;\Varid{m}\Rightarrow \Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b}\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}{}\<[E]%
\\
\>[B]{}\Varid{m}_{1}\mathbin{>\!\!\!>\!\!\!>}\Varid{m}_{2}\mathrel{=}\Varid{m}_{1}\mathbin{`\Varid{bind}`}(\lambda \anonymous \to \Varid{m}_{2})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Use of \ensuremath{\Varid{declare}} in \ensuremath{\Varid{prog}} ensures that neither \ensuremath{\text{\tt \char34 counter\char34}} nor \ensuremath{\text{\tt \char34 queue\char34}} exist
before the execution of \ensuremath{\Varid{prog}}, while \ensuremath{\Varid{start}} in \ensuremath{\Varid{main}} guarantees that the
program is given a fresh run without previously defined keys at all. Haskell
is able to infer the type of \ensuremath{\Varid{prog}}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{prog}\mathbin{::}\Conid{Edis}\;{}\<[15]%
\>[15]{}\mbox{\textquotesingle}[~]\;\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 counter\char34},\Conid{Integer}),\mbox{\textquotesingle}(\text{\tt \char34 queue\char34},\Conid{ListOf}\;\Conid{Message})]\;{}\<[E]%
\\
\>[15]{}(\Conid{R} \uplus \Conid{Maybe}\;\Conid{Message})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

% The following program increases the value of |"A"| as an integer, push the result of the increment to list |"L"|, and then pops it out:
% \begin{spec}
% main :: IO ()
% main = do
%     conn    <- connect defaultConnectInfo
%     result  <- runRedis conn $ unEdis $ start
%         `bind` \ _ ->  declare (Proxy :: Proxy "A") (Proxy :: Proxy Integer)
%         `bind` \ _ ->  incr (Proxy :: Proxy "A")
%         `bind` \n ->  case n of
%             Left  err  -> lpush (Proxy :: Proxy "L") 0
%             Right n    -> lpush (Proxy :: Proxy "L") n
%         `bind` \ _ ->  lpop     (Proxy :: Proxy "L")
%     print result
% \end{spec}

% The syntax is pretty heavy, like the old days when there's no
%  \emph{do-notation}\cite{history}. But if we don't need any variable bindings
%  between operations, we could compose these commands with a sequencing operator
%  |(>>>)|.

% \begin{spec}
% program = start
%     >>> declare (Proxy :: Proxy "A") (Proxy :: Proxy Integer)
%     >>> incr    (Proxy :: Proxy "A")
%     >>> lpush   (Proxy :: Proxy "L") 0
%     >>> lpop    (Proxy :: Proxy "L")
% \end{spec}

### 2017-07-13 TyDe2017/sections/IndexedMonads-orig.tex
Auto stash before merge of "master" and "origin/master"

--- 2016-06-12 aplas16/sections/IndexedMonads.tex
@@
 Stateful computations are often reasoned using Hoare logic. A {\em Hoare triple}
-$\{P\} S \{Q\}$ denotes such a proposition: if the statement $S$ is executed in
-a state satisfying prediate $P$, when it terminates, the state must satisfy
-predicate $Q$. Predicates $P$ and $Q$ are respectively is called the
+$\{P\} S \{Q\}$ denotes the following proposition: if the statement $S$ is
+executed in a state satisfying predicate $P$, when it terminates, the state must
+satisfy predicate $Q$. Predicates $P$ and $Q$ are respectively called the
 \emph{precondition} and the \emph{postcondition} of the Hoare triple.
 
-In Haskell, stateful computations are represented by monads. In order to
-reason about their behaviors within the type system, we wish to label a state
-monad with its pre and postcondition. An \emph{indexed monad}~%
-\cite{indexedmonad} (also called \emph{monadish} or \emph{parameterised monad})
-is a monad that, in addition to the type of value it computes, takes two more
-type arguments representing an initial state and a final state, to be
-interpreted like a Hoare triple~\cite{kleisli}:
+In Haskell, stateful computations are represented by monads. To reason about
+them within the type system, we wish to label a state monad with its pre and
+postcondition. An \emph{indexed monad}~\cite{indexedmonad} (also called a
+\emph{parameterised monad} or \emph{monadish}) is a monad that, in addition to
+the type of its result, takes two more type parameters representing an initial
+state and a final state, to be interpreted like a Hoare triple:
 \begin{hscode}\SaveRestoreHook
@@
 the computation returned by \ensuremath{\Varid{f}}. The result is a monad \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}}.
-Indexed monads have been used ~\cite{typefun,staticresources} ... \todo{for what? Some discriptions here to properly cite them.}
 
-We define a new indexed monad \ensuremath{\Conid{Edis}} which, at term level, merely wraps
-\ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the
-pre/postconditions at type level:
+We define a new indexed monad \ensuremath{\Conid{Edis}}. At term level, the \ensuremath{\Varid{unit}} and \ensuremath{\Varid{bind}}
+methods are not interesting: they merely make calls to \ensuremath{\Varid{return}} and \ensuremath{(\bind )} of
+\ensuremath{\Conid{Redis}}, and extract and re-apply the constructor \ensuremath{\Conid{Edis}} when necessary.
+With \ensuremath{\Conid{Edis}} being a \ensuremath{\mathbf{newtype}}, they can be optimized away in runtime. The
+purpose is to add the pre/postconditions at type level:
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-At term level, the \ensuremath{\Varid{unit}} and \ensuremath{\Varid{bind}} methods are not interesting: they merely
-make calls to \ensuremath{\Varid{return}} and \ensuremath{(\bind )} of \ensuremath{\Conid{Redis}}, and extracts and re-apply the constructor \ensuremath{\Conid{Edis}} when necessary.
-With \ensuremath{\Conid{Edis}} being a \ensuremath{\mathbf{newtype}}, they
-can be optimized away in runtime. The interesting bits happen in compile type,
-on the added type information.
 
-The properties of the state we care about are the set of currently allocated
-keys and their associated types. We will present, in Section~\ref{sec:type-level-dict}, techniques that allow us to specify
-properties such as ``the keys in the database are \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}},
-respectively associated to values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and \ensuremath{\Conid{Bool}}.''
-For now, however, let us look at the simplest \Redis{} command.
+The properties of the state we care about are the set of currently
+allocated keys and types of their values. We will present, in Section~\ref{sec:type-level-dict}, techniques that allow us to specify
+properties such as ``the keys in the data store are \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}},
+respectively assigned values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and \ensuremath{\Conid{Bool}}.'' For now,
+however, let us look at the simplest \Redis{} command.
 
-The command \text{PING} in \Redis{} does nothing but replies a message
-\text{PONG} if the connection is alive. In \Hedis{}, \ensuremath{\Varid{ping}} has type
-\ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status})}. The \Edis{} version of \ensuremath{\Varid{ping}} simply
-applys an additional constructor (functions from \Hedis{} are qualified with
-\ensuremath{\Conid{Hedis}} to prevent name clashing):
+\Redis{} commands can be executed in two contexts: the normal context, or a \emph{transaction}. In \Hedis{}, a command yielding value of type \ensuremath{\Varid{a}} in the
+normal case is represented by \ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Varid{a})}, as mentioned in
+Section~\ref{sec:introduction}; in a transaction, the command is represented by
+two other datatypes \ensuremath{\Conid{RedisTx}\;(\Conid{Queued}\;\Varid{a})}. In this paper we focus on the former
+case. For brevity we abbreviate \ensuremath{\Conid{Either}\;\Conid{Reply}\;\Varid{a}} to \ensuremath{\Conid{R} \uplus \Varid{a}}.
+
+The command \texttt{PING} in \Redis{} does nothing but replying a message
+\texttt{PONG} if the connection is alive. In \Hedis{}, \ensuremath{\Varid{ping}} has type
+\ensuremath{\Conid{Redis}\;(\Conid{R} \uplus \Conid{Status})}.
+The \Edis{} version of \ensuremath{\Varid{ping}} simply applies an additional constructor (functions from \Hedis{} are qualified with \ensuremath{\Conid{Hedis}} to
+prevent name clashing):
 \begin{hscode}\SaveRestoreHook
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{ping}\mathbin{::}\Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
+\>[B]{}\Varid{ping}\mathbin{::}\Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{R} \uplus \Conid{Status}){}\<[E]%
 \\
@@
 \end{hscode}\resethooks
-Since \ensuremath{\Varid{ping}} does not alter the database, the postcondition and precondition
+Since \ensuremath{\Varid{ping}} does not alter the data store, the postcondition and precondition
 are the same. Commands that are more interesting will be introduced after

### 2017-07-13 TyDe2017/sections/Introduction-orig.tex
Auto stash before merge of "master" and "origin/master"

--- 2016-06-11 aplas16/sections/Introduction.tex
@@
 
-\subsection{Redis}
+\Redis{}\footnote{\url{https://redis.io}} is an open source, in-memory data
+structure store that can be used as database, cache, and message broker, etc.
+A \Redis{} data store can be thought of as a set of key-value pairs. The value
+can be a string, a list of strings, a set of strings, or a hash table of
+strings, etc. However, string is the only primitive datatype. Numbers, for
+example, are serialized to strings before being saved in the data store, and
+parsed back to numbers to be manipulated with. While the concept is simple,
+\Redis{} is used as an essential component in a number of popular, matured services, including Twitter, GitHub, Weibo, StackOverflow, and Flickr, etc.
 
-Redis\footnote{\url{https://redis.io}} is an open source, in-memory data structure store, often used as
-database, cache and message broker.
-A Redis datatype can be think of as a set of key-value pairs, where each value is associated with a binary-safe string key to identify and manipulate with.
-Redis supports many different kind of values, such as strings, hashes, lists, and sets, etc, and provides a collection of of atomic \emph{commands} to manipulate these values.
+For an example, consider the following sequence of commands, entered through the
+interactive interface of \Redis{}. The keys \texttt{some-set} and
+\texttt{another-set} are both assigned a set of strings. The two calls to
+command \texttt{SADD} respectively add three and two strings to the two sets,
+before \texttt{SINTER} takes their intersection:
+\begin{Verbatim}[xleftmargin=.4in]
+redis> SADD some-set a b c
+(integer) 3
+redis> SADD another-set a b
+(integer) 2
+redis> SINTER some-set another-set
+1) "a"
+2) "b"
+\end{Verbatim}
+\noindent Notice that the keys \texttt{some-set} and \texttt{another-set}, if
+not existing before the call to \texttt{SADD}, are created on site. The calls to
+\texttt{SADD} return the sizes of the resulting sets.
 
-For an example, consider the following sequence of commands, entered through the interactive interface of Redis. The keys
-\text{some-set} abd \text{another-set} are
-both associated to a set. The two call to command \text{SADD} respectively adds three and two values to the two sets, before \text{SINTER} takes their intersection:
-\begin{tabbing}\tt
-~redis\char62{}~SADD~some\char45{}set~a~b~c\\
-\tt ~\char40{}integer\char41{}~3\\
-\tt ~redis\char62{}~SADD~another\char45{}set~a~b\\
-\tt ~\char40{}integer\char41{}~2\\
-\tt ~redis\char62{}~SINTER~some\char45{}set~another\char45{}set\\
-\tt ~1\char41{}~\char34{}a\char34{}\\
-\tt ~2\char41{}~\char34{}b\char34{}
-\end{tabbing}
-
-Note that the keys \text{some-set} and \text{another-set}, if not existing before the call to \text{SADD}, are created on site. The call to
-\text{SADD} returns the size of the set after
-completion of the command.
-
-\subsection{Hedis}
-
-Many third party library exist to allow general purpose programmings
-to access Redis databases through its TCP protocol. The most popular
-such library for Haskell is Hedis\footnote{\url{https://hackage.haskell.org/package/hedis}}.
-
-The following program is how the previous example looks like in
-Hedis:
-
+Many third party libraries provide interfaces for general purpose programming
+languages to access \Redis{} through its TCP protocol. For Haskell, the most
+popular library is
+\Hedis{}.\footnote{\url{https://hackage.haskell.org/package/hedis}}
+A (normal) \Redis{} computation returning a value of type \ensuremath{\Varid{a}} is represented in
+Haskell by \ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Varid{a})}, where the type \ensuremath{\Conid{Redis}} is a monad, while
+\ensuremath{\Conid{Either}\;\Conid{Reply}\;\Varid{a}} indicates that the computation either returns a value of type
+\ensuremath{\Varid{a}}, or fails with an error message represented by type \ensuremath{\Conid{Reply}}. The following
+program implements the previous example:
 \begin{hscode}\SaveRestoreHook
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]{}\<[E]%
+\>[5]{}\Varid{sinter}\;[\mskip1.5mu \text{\tt \char34 some-set\char34},\text{\tt \char34 another-set\char34}\mskip1.5mu]~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+The function \ensuremath{\Varid{sadd}\mathbin{::}\Conid{ByteString}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer})} takes a key and a list of values as arguments, and returns a
+\Redis{} computation yielding \ensuremath{\Conid{Integer}}. Keys and values, being nothing but
+binary strings in \Redis{}, are represented using Haskell \ensuremath{\Conid{ByteString}}.
+
+\paragraph{The Problems} Most commands work only with data of certain types. In
+the following example, the key \texttt{some-string} is assigned a string
+\texttt{foo} --- the command \texttt{SET} always assigns a string to a key.
+The subsequent call to \texttt{SADD}, which adds a value to a set, thus raises a
+runtime error.
+\begin{Verbatim}[xleftmargin=.4in]
+redis> SET some-string foo
+OK
+redis> SADD some-string bar
+(error) WRONGTYPE Operation against a key holding
+the wrong kind of value
+\end{Verbatim}
+\noindent For another source of type error, the command \texttt{INCR key}
+increments the value associated with \texttt{key} by one. With strings being the
+only primitive type, \Redis{} parses the stored string to an integer and, after
+incrementation, stores a string back. If the string cannot be parsed as an
+integer, a runtime error is raised.
+
+The reader must have noticed the peculiar pattern of value creation and updating
+in \Redis{}: the same command is used both to create a key-value pair and to
+update them. Similar to \texttt{SADD}, the command \texttt{LPUSH} appends a
+value (a string) to a list, or creates one if the key does not exist:
+\begin{Verbatim}[xleftmargin=.4in]
+redis> LPUSH some-list bar
+(integer) 1
+\end{Verbatim}
+\noindent Another command \texttt{LLEN} returns the length of the list, and
+signals an error if the key is not associated with a list:
+\begin{Verbatim}[xleftmargin=.4in]
+redis> LLEN some-list
+(integer) 1
+redis> SET some-string foo
+OK
+redis> LLEN some-string
+(error) WRONGTYPE Operation against a key holding
+the wrong kind of value
+\end{Verbatim}
+\noindent Curiously, however, when applied to a key not yet created, \Redis{}
+designers chose to let \texttt{LLEN} return \texttt{0}:
+\begin{Verbatim}[xleftmargin=.4in]
+redis> LLEN nonexistent
+(integer) 0
+\end{Verbatim}
+
+Being a simple wrapper on top of the TCP protocol of \Redis{}, \Hedis{}
+inherits all the behaviors. Executing the following program yields the same
+error, but wrapped in a Haskell constructor: \ensuremath{\Conid{Left}\;(\Conid{Error}} \texttt{"WRONGTYPE
+Operation against a key holding the wrong kind of value"}\ensuremath{)}.
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{program}\mathbin{::}\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
+\\
+\>[B]{}\Varid{program}\mathrel{=}\mathbf{do}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{set}\;\text{\tt \char34 some-string\char34}\;\text{\tt \char34 foo\char34}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{sadd}\;\text{\tt \char34 some-string\char34}\;[\mskip1.5mu \text{\tt \char34 a\char34}\mskip1.5mu]~~.{}\<[E]%
 \ColumnHook
@@
 
-The function \ensuremath{\Varid{sadd}} takes a key and a list of values as
- arguments, and returns an \ensuremath{\Conid{Integer}} on success, or
- returns a \ensuremath{\Conid{Reply}}, a low-level representation of replies
- from the Redis server, in case of failures. All wrapped in
- \ensuremath{\Conid{Redis}}\footnotemark, the context of command execution.
+Such a programming model is certainly very error-prone. Working within Haskell,
+a host language with a strong typing system, one naturally wishes to build a
+domain-specific embedded language (DSEL) that exploits the rich type system
+of Haskell to not only ensure the absence of \Redis{} type errors, but also
+provides better documentations. We wish to be sure that a program calling
+\texttt{INCR}, for example, can be type checked only if we can statically
+guarantee that the value to be accessed is indeed an integer. We wish to see
+from the type of operators such as \texttt{LLEN} when it can be called, and
+allow it to be used only in contexts that are safe. We may even want to
+explicitly declare a fresh key and its type, to avoid reusing an existing key
+by accident, and to prevent it from unexpectedly being used as some other type.
 
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{sadd}{}\<[7]%
-\>[7]{}\mathbin{::}\Conid{ByteString}{}\<[27]%
-\>[27]{}\mbox{\onelinecomment  key}{}\<[E]%
-\\
-\>[7]{}\to [\mskip1.5mu \Conid{ByteString}\mskip1.5mu]{}\<[27]%
-\>[27]{}\mbox{\onelinecomment  values}{}\<[E]%
-\\
-\>[7]{}\to \Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
+This paper discusses the techniques we used and the experiences we learned from
+building such a language, \Edis{}. We constructed an {\em indexed monad}, on
+top of the monad \ensuremath{\Conid{Redis}}, which is indexed by a dictionary that maintains the
+set of currently defined keys and their types. To represent the dictionary, we
+need to encode variable bindings with {\em type-level} lists and strings. And
+to manipulate the dictionary, we applied various type-level programming
+techniques. To summarize our contributions:
+\begin{itemize}
+\item We present \Edis{}, a statically typed domain-specific language embedded in Haskell and built on \Hedis{}. Serializable Haskell datatypes are
+automatically converted before being written to \Redis{} data store. Available
+keys and their types are kept track of in type-level dictionaries. The types of
+embedded commands state clearly their preconditions and postconditions on the
+available keys and types, and a program is allowed to be constructed only if
+it is guaranteed not to fail with a type error.
+%
+\item We demonstrate the use of various type-level programming techniques,
+including data kinds, singleton types and proxies, closed type families, etc.,
+to define type-level lists and operations that observe and manipulate the lists.
+%
+\item This is (yet another) example of encoding effects and constraints of programs in types, using indexed monad~\cite{indexedmonad}, closed
+type-families~\cite{tfclosed} and constraint kinds~\cite{constraintskinds}.
+\end{itemize}
 
-Note that keys and values, being nothing but binary strings in Redis,
-are represented using Haskell \ensuremath{\Conid{ByteString}}.
-Values of other types must be encoded as \ensuremath{\Conid{ByteString}}s before being written to the database,
-and decoded after being read back.
-
-\footnotetext{Hedis provides another kind of context, \ensuremath{\Conid{RedisTx}}, for \emph{transactions}, united with \ensuremath{\Conid{Redis}} under the class of \ensuremath{\Conid{RedisCtx}}.
-For brevity, we demonstrate only \ensuremath{\Conid{Redis}} in this paper.
-}
+In Section~\ref{sec:indexed-monads} we introduce indexed monads, to reason about
+pre and postconditions of stateful programs, and in Section~\ref{sec:type-level-dict} we review the basics of type-level
+programming in Haskell that allows us to build type-level dictionaries to keep
+track of keys and types. Embeddings of \Redis{} commands are presented in
+Section~\ref{sec:embedding-commands}. In Section~\ref{sec:discussions} we
+discuss various issues regarding design choices, limitations of this approach,
+as well as possible future works, before we review related work in
+Section~\ref{sec:conclusions}.

### 2017-07-13 TyDe2017/sections/TypeLevelDict-orig.tex
Auto stash before merge of "master" and "origin/master"

--- 2016-06-12 aplas16/sections/TypeLevelDict.tex
@@
 
-One of the challenges of statically ensuring type correctness of \Redis{},
-which also presents in other stateful languages, is that the type of the value
-associated to a key can be altered after updating. To ensure type correctness,
-we have to keep track of the (\Redis{}) types of all existing keys in a
-{\em dictionary} --- conceptually, a list of pairs of keys and \Redis{} types.
-Each \Redis{} command is embedded in \Popcorn{} as a monadic computation. The
-monad, to be presented in Section~\ref{sec:indexed-monads}, is indexed by
-the dictionaries before and after the computation. In a dependently typed
-programming language (without the so-called ``phase distinction'' ---
+One of the challenges of statically ensuring type correctness of stateful
+languages is that the type of the value of a key can be altered by updating.
+In \Redis{}, one may delete an existing key and create it again by assigning to
+it a value of a different type. To ensure type correctness, we keep track of the
+types of all existing keys in a {\em dictionary} --- an associate list, or a
+list of pairs of keys and some encoding of types. For example, we may use the
+dictionary \ensuremath{[\mskip1.5mu (\text{\tt \char34 A\char34},\Conid{Int}),(\text{\tt \char34 B\char34},\Conid{Char}),(\text{\tt \char34 C\char34},\Conid{Bool})\mskip1.5mu]} to represent a
+predicate, or a constraint, stating that ``the keys in the data store are \ensuremath{\text{\tt \char34 A\char34}},
+\ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}}, respectively assigned values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and
+\ensuremath{\Conid{Bool}}.'' (This representation will be refined in the next section.)
+
+The dictionary above mixes values (strings such as \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}) and types.
+Furthermore, as mentioned in Section~\ref{sec:indexed-monads}, the
+dictionaries will be parameters to the indexed monad \ensuremath{\Conid{Edis}}. In a dependently
+typed programming language (without the so-called ``phase distinction'' ---
 separation between types and terms), this would pose no problem. In Haskell
-however, the dictionaries, to index a monad, has to be a Haskell type as well.
+however, the dictionaries, to index a monad, has to be a type as well.
 
 In this section we describe how to construct a type-level dictionary, to be
-used with the indexed monad in Section~\ref{sec:indexed-monads}. More operations
-on the dictionary will be presented in Section~\ref{sec:type-level-fun}.
+used with the indexed monad in Section~\ref{sec:indexed-monads}.
+
+\subsection{Datatype Promotion}
 
 Haskell maintains the distinction between values, types, and kinds: values are
-categorized by types, and types are categorized by kinds. The kinds are relatively simple: \ensuremath{\mathbin{*}} is the kind of all {\em lifted} types, while type
-constructors have kinds such as \ensuremath{\mathbin{*}\to \mathbin{*}}, \ensuremath{\mathbin{*}\to \mathbin{*}\to \mathbin{*}}, etc. Consider the
-datatype definitions below:
+categorized by types, and types are categorized by kinds. The kinds are
+relatively simple: \ensuremath{\mathbin{*}} is the kind of all {\em lifted} types, while type
+constructors have kinds such as \ensuremath{\mathbin{*}\to \mathbin{*}}, or \ensuremath{\mathbin{*}\to \mathbin{*}\to \mathbin{*}}, etc.\footnotemark\,
+Consider the datatype definitions below:
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-The lefthand side is usually seen as having defined a type \ensuremath{\Conid{Nat}\mathbin{::}\mathbin{*}},
-and two value constructors \ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}}. The righthand
+The left-hand side is usually seen as having defined a type \ensuremath{\Conid{Nat}\mathbin{::}\mathbin{*}},
+and two value constructors \ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}}. The right-hand
 side is how Haskell lists are understood. The {\em kind} of \ensuremath{[\mskip1.5mu \mathbin{\cdot}\mskip1.5mu]} is \ensuremath{\mathbin{*}\to \mathbin{*}},
-since it takes a lifted type \ensuremath{\Varid{a}} to a lifted type \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]}.
-The two value constructors respectively have types \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{(\mathbin{:})\mathbin{::}\Varid{a}\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]}, for all type \ensuremath{\Varid{a}}.
+since it takes a lifted type \ensuremath{\Varid{a}} to a lifted type \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]}. The two value
+constructors respectively have types \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{(\mathbin{:})\mathbin{::}\Varid{a}\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]}, for all type \ensuremath{\Varid{a}}.
+
+\footnotetext{In Haskell, the opposite of \emph{lifted} types are \emph{unboxed}
+types, which are not represented by a pointer to a heap object, and cannot be
+stored in a polymorphic data type.}
 
@@
 \ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} is a type having kind \ensuremath{\Conid{Nat}}, and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}} is a type
-constructor, taking a type in kind \ensuremath{\Conid{Nat}} to another type in \ensuremath{\Conid{Nat}}. Whether a
-constructor is promoted can often be inferred from the context. To be more specific, prefixing a constructor with a single quote, such as in \ensuremath{\mbox{\textquotesingle}\Conid{Zero}} and
-\ensuremath{\mbox{\textquotesingle}\Conid{Suc}}, denotes that it is promoted.
+constructor, taking a type in kind \ensuremath{\Conid{Nat}} to another type in \ensuremath{\Conid{Nat}}. When one
+sees a constructor in an expression, whether it is promoted can often be
+inferred from the context. When one needs to be more specific, prefixing a
+constructor with a single quote, such as in \ensuremath{\mbox{\textquotesingle}\Conid{Zero}} and \ensuremath{\mbox{\textquotesingle}\Conid{Suc}}, denotes that it
+is promoted.
 
 The situation of lists is similar: for all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]} is also a kind. For
-all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{k}\mskip1.5mu]} is a type. Given a type \ensuremath{\Varid{x}}
-of kind \ensuremath{\Varid{k}} and a type \ensuremath{\Varid{xs}} of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}, \ensuremath{\Varid{x}\mathbin{:}\Varid{xs}} is again a type of
-kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}.
-Formally, \ensuremath{(\mathbin{:})\mathbin{::}\Varid{k}\to [\mskip1.5mu \Varid{k}\mskip1.5mu]\to [\mskip1.5mu \Varid{k}\mskip1.5mu]}.
-For example,
-\ensuremath{\Conid{Int}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Char}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[]))} is a type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} --- a list of
-(lifted) types. The optional quote denotes that the constructors are promoted.
-The same list can be denoted by a syntax sugar \ensuremath{\mbox{\textquotesingle}[\Conid{Int},\Conid{Char},\Conid{Bool}]}.
+all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \mskip1.5mu]} is a type of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}.
+Given a type \ensuremath{\Varid{a}} of kind \ensuremath{\Varid{k}} and a
+type \ensuremath{\Varid{as}} of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}, \ensuremath{\Varid{a}\mathbin{:}\Varid{as}} is again a type of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}.
+Formally,
+\ensuremath{(\mathbin{:})\mathbin{::}\Varid{k}\to [\mskip1.5mu \Varid{k}\mskip1.5mu]\to [\mskip1.5mu \Varid{k}\mskip1.5mu]}.
+For example, \ensuremath{\Conid{Int}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Char}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[~]))} is a
+type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} --- it is a list of (lifted) types. The optional quote
+denotes that the constructors are promoted. The same list can be denoted by a
+syntax sugar \ensuremath{\mbox{\textquotesingle}[\Conid{Int},\Conid{Char},\Conid{Bool}]}.
 
@@
 Strings in Haskell are nothing but lists of \ensuremath{\Conid{Char}}s. Regarding promotion,
-however, a \ensuremath{\Conid{String}} can be promoted to a type having kind \ensuremath{\Conid{Symbol}}. \ensuremath{\Conid{Symbol}} is
-a type without a constructor: \ensuremath{\mathbf{data}\;\Conid{Symbol}},
-intended to be used as a promoted kind. In the expression:
+however, a string can be promoted to a type having kind \ensuremath{\Conid{Symbol}}. In the expression:
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-the string on the lefthand side of \ensuremath{(\mathbin{::})} is a type, having kind \ensuremath{\Conid{Symbol}}.
+the string on the left-hand side of \ensuremath{(\mathbin{::})} is a type whose kind is \ensuremath{\Conid{Symbol}}.
 
@@
 \end{hscode}\resethooks
-All the entities defined above are types, where \ensuremath{\Conid{Dict0}} and \ensuremath{\Conid{Dict1}}
-have kind \ensuremath{[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}.
+All the entities defined above are types, where \ensuremath{\Conid{Dict0}} has kind \ensuremath{[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}.
+In \ensuremath{\Conid{Dict1}}, while \ensuremath{\Conid{Int}} has kind \ensuremath{\mathbin{*}} and \ensuremath{\text{\tt \char34 A\char34}} has kind \ensuremath{\Conid{Symbol}}, the former kind subsumes the later.
+Thus \ensuremath{\Conid{Dict1}} also has kind \ensuremath{[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}.
+
+\subsection{Type-Level Functions}
+\label{sec:type-fun}
+
+Now that we can represent dictionaries as types, the next step is to define
+operations on them. A function that inserts an entry to a dictionary, for
+example, is a function from a type to a type. While it was shown that it is
+possible to simulate type-level functions using Haskell type
+classes~\cite{McBride:02:Faking}, in recent versions of GHC, {\em indexed type
+families}, or \emph{type families} for short, are considered a cleaner solution.
+
+For example, compare disjunction \ensuremath{(\mathrel{\vee})} and its type-level
+counterpart:\\
+\noindent{\centering %\small
+\begin{minipage}[b]{0.35\linewidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}c<{\hspost}@{}}%
+\column{7E}{@{}l@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}(\mathrel{\vee})\mathbin{::}\Conid{Bool}\to \Conid{Bool}\to \Conid{Bool}{}\<[E]%
+\\
+\>[B]{}\Conid{True}{}\<[7]%
+\>[7]{}\mathrel{\vee}{}\<[7E]%
+\>[11]{}\Varid{b}{}\<[14]%
+\>[14]{}\mathrel{=}\Conid{True}{}\<[E]%
+\\
+\>[B]{}\Varid{a}{}\<[7]%
+\>[7]{}\mathrel{\vee}{}\<[7E]%
+\>[11]{}\Varid{b}{}\<[14]%
+\>[14]{}\mathrel{=}\Varid{b}~~,{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\begin{minipage}[b]{0.55\linewidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Or}\;(\Varid{a}\mathbin{::}\Conid{Bool})\;(\Varid{b}\mathbin{::}\Conid{Bool})\mathbin{::}\Conid{Bool}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\mathbf{where}\;{}\<[10]%
+\>[10]{}\mbox{\textquotesingle}\Conid{True}{}\<[16]%
+\>[16]{}\mathrel{\Conid{`Or`}}\Varid{b}{}\<[24]%
+\>[24]{}\mathrel{=}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
+\\
+\>[10]{}\Varid{a}{}\<[16]%
+\>[16]{}\mathrel{\Conid{`Or`}}\Varid{b}{}\<[24]%
+\>[24]{}\mathrel{=}\Varid{b}~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+}\\
+The left-hand side is a typical definition of \ensuremath{(\mathrel{\vee})} by pattern matching.
+On the right-hand side, \ensuremath{\Conid{Bool}} is not a type, but a type lifted to a kind,
+while \ensuremath{\Conid{True}} and \ensuremath{\Conid{False}} are types of kind \ensuremath{\Conid{Bool}}. The declaration says that
+\ensuremath{\Conid{Or}} is a family of types, indexed by two parameters \ensuremath{\Varid{a}} and \ensuremath{\Varid{b}} of kind \ensuremath{\Conid{Bool}}.
+The type with index \ensuremath{\mbox{\textquotesingle}\Conid{True}} and \ensuremath{\Varid{b}} is \ensuremath{\mbox{\textquotesingle}\Conid{True}}, and all other indices lead to \ensuremath{\Varid{b}}.
+For our purpose, we can read \ensuremath{\Conid{Or}} as a function from types to types ---
+observe how it resembles the term-level \ensuremath{(\mathrel{\vee})}. We present two more type-level
+functions about \ensuremath{\Conid{Bool}} --- negation, and conditional, that we will use later:\\
+\noindent{\centering %\small
+\begin{minipage}[b]{0.35\linewidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Not}\;\Varid{a}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Not}\;\mbox{\textquotesingle}\Conid{False}{}\<[14]%
+\>[14]{}\mathrel{=}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Not}\;\mbox{\textquotesingle}\Conid{True}{}\<[14]%
+\>[14]{}\mathrel{=}\mbox{\textquotesingle}\Conid{False}~~,{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\begin{minipage}[b]{0.55\linewidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{If}\;(\Varid{c}\mathbin{::}\Conid{Bool})\;(\Varid{t}\mathbin{::}\Varid{a})\;(\Varid{f}\mathbin{::}\Varid{a})\mathbin{::}\Varid{a}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{If}\;\mbox{\textquotesingle}\Conid{True}\;{}\<[12]%
+\>[12]{}\Varid{tru}\;{}\<[17]%
+\>[17]{}\Varid{fls}\mathrel{=}\Varid{tru}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{If}\;\mbox{\textquotesingle}\Conid{False}\;\Varid{tru}\;{}\<[17]%
+\>[17]{}\Varid{fls}\mathrel{=}\Varid{fls}~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+}
+
+As a remark, type families in Haskell come in many flavors. One can define families of \ensuremath{\mathbf{data}}, as well as families of \ensuremath{\mathbf{type}} synonym. They can appear
+inside type classes~\cite{tfclass,tfsynonym} or at toplevel. Top-level type families can be open~\cite{tfopen} or closed~\cite{tfclosed}. The flavor we
+chose is top-level, closed type synonym family, since it allows overlapping
+instances, and since we need none of the extensibility provided by open type
+families. Notice that the instance \ensuremath{\mbox{\textquotesingle}\Conid{True}\mathrel{\Conid{`Or`}}\Varid{b}} could be subsumed under the
+more general instance, \ensuremath{\Varid{a}\mathrel{\Conid{`Or`}}\Varid{b}}. In a closed type family we may resolve the
+overlapping in order, just like how cases overlapping is resolved in term-level
+functions.
+
+Now we can define operations on type-level dictionaries. Let us begin
+with:
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}c<{\hspost}@{}}%
+\column{33E}{@{}l@{}}%
+\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Get}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{k}\mathbin{::}\Conid{Symbol})\mathbin{::}\mathbin{*}\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{k},{}\<[20]%
+\>[20]{}\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{k}{}\<[33]%
+\>[33]{}\mathrel{=}{}\<[33E]%
+\>[36]{}\Varid{x}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{t},{}\<[20]%
+\>[20]{}\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{k}{}\<[33]%
+\>[33]{}\mathrel{=}{}\<[33E]%
+\>[36]{}\Conid{Get}\;\Varid{xs}\;\Varid{k}~~.{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\ensuremath{\Conid{Get}\;\Varid{xs}\;\Varid{k}} returns the entry associated with key \ensuremath{\Varid{k}} in the
+dictionary \ensuremath{\Varid{xs}}. Notice, in the first case, how type-level equality can be
+expressed by unifying type variables with the same name. Note also that \ensuremath{\Conid{Get}} is
+a partial function on types: while \ensuremath{\Conid{Get}\;\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int})]\;\text{\tt \char34 A\char34}} evaluates
+to \ensuremath{\Conid{Int}}, when \ensuremath{\Conid{Get}\;\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int})]\;\text{\tt \char34 B\char34}} appears in a type expression,
+there are no applicable rules to reduce it. The expression thus stays un-reduced.
+
+% For our applications it is more convenient to make |Get| total, as we would at
+% the term level, by having it return a |Maybe|:
+% \begin{spec}
+% type family Get (xs :: [(Symbol, *)]) (k :: Symbol) :: Maybe * where
+%     Get NIL                 k = Nothing
+%     Get (TPar(k, x) :- xs)  k = Just x
+%     Get (TPar(t, x) :- xs)  k = Get xs k {-"~~."-}
+% \end{spec}
+%
+Some other dictionary-related functions are defined
+in Figure \ref{fig:dict-operations}. The function \ensuremath{\Conid{Set}} either updates an
+existing entry or inserts a new entry, \ensuremath{\Conid{Del}} removes an entry matching
+a given key, while \ensuremath{\Conid{Member}} checks whether a given key exists in the
+dictionary.
+
+\begin{figure}[t]
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{48}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\mbox{\onelinecomment  inserts or updates an entry}{}\<[E]%
+\\
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Set}\;{}\<[18]%
+\>[18]{}(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{k}\mathbin{::}\Conid{Symbol})\;(\Varid{x}\mathbin{::}\mathbin{*})\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Set}\;\mbox{\textquotesingle}[~]\;{}\<[29]%
+\>[29]{}\Varid{k}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\Varid{k},\Varid{x})]{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Set}\;(\mbox{\textquotesingle}(\Varid{k},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
+\>[29]{}\Varid{k}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}(\Varid{k},\Varid{x}){}\<[48]%
+\>[48]{}\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Set}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
+\>[29]{}\Varid{k}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}(\Varid{t},\Varid{y}){}\<[48]%
+\>[48]{}\mathbin{\mbox{\textquotesingle}\!:}\Conid{Set}\;\Varid{xs}\;\Varid{k}\;\Varid{x}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mbox{\onelinecomment  removes an entry}{}\<[E]%
+\\
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Del}\;{}\<[18]%
+\>[18]{}(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{k}\mathbin{::}\Conid{Symbol})\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Del}\;\Conid{Nil}\;{}\<[30]%
+\>[30]{}\Varid{k}\mathrel{=}\Conid{Nil}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Del}\;(\mbox{\textquotesingle}(\Varid{k},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[30]%
+\>[30]{}\Varid{k}\mathrel{=}\Varid{xs}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Del}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[30]%
+\>[30]{}\Varid{k}\mathrel{=}\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Conid{Del}\;\Varid{xs}\;\Varid{k}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\mbox{\onelinecomment  membership}{}\<[E]%
+\\
+\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Member}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{k}\mathbin{::}\Conid{Symbol})\mathbin{::}\Conid{Bool}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Member}\;\Conid{Nil}\;{}\<[33]%
+\>[33]{}\Varid{k}\mathrel{=}\Conid{False}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Member}\;(\mbox{\textquotesingle}(\Varid{k},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[33]%
+\>[33]{}\Varid{k}\mathrel{=}\Conid{True}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Conid{Member}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[33]%
+\>[33]{}\Varid{k}\mathrel{=}\Conid{Member}\;\Varid{xs}\;\Varid{k}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\caption{Some operations on type-level dictionaries.}
+\label{fig:dict-operations}
+\end{figure}

### 2017-07-14 TyDe2017/main-tyde.tex
Minor formatting/layout changes.
Public domain issues.

@@
 
-\makeatletter\if@ACM@journal\makeatother
+%\makeatletter\if@ACM@journal\makeatother
 %% Journal information (used by PACMPL format)
@@
 
-\setcopyright{rightsretained}
+%%% The following is specific to TyDe'17 and the paper
+%%% 'Type Safe Redis Queries: A Case Study of Type-Level Programming in Haskell'
+%%% by Ting-Yan Lai, Tyng-Ruey Chuang, and Shin-Cheng Mu.
+%%%
+\setcopyright{licensedothergov}
+\acmDOI{10.1145/3122975.3122978}
+\acmISBN{978-1-4503-5183-6/17/09}
+\acmConference[TyDe'17]{2st ACM SIGPLAN International Workshop on Type-Driven Development}{September 3, 2017}{Oxford, UK}
+\acmYear{2017}
+\copyrightyear{2017}
+\acmPrice{15.00}
+
 \bibliographystyle{ACM-Reference-Format}
@@
 \usepackage{url}
-\usepackage{mathptmx}
+%\usepackage{mathptmx}
 \usepackage{doubleequals}
@@
 \usepackage{fancyvrb} % for indented verbatim
+\usepackage{flushend} % balance columns on last page
 
@@
 
+\pagestyle{empty} %no page number
 \setlength{\mathindent}{10pt}
@@
 \author{Ting--Yan Lai}
-\authornote{Ting--Yan Lai was summer intern from
-the Institute of Computer Science and Engineering,
-National Chiao Tung University, Taiwan,
+\authornote{Ting--Yan Lai was a summer intern from
+the Inst. of Computer Science and Engineering,
+National Chiao Tung Univ., Taiwan,
 when part of this work was developed.}
@@
 
-\keywords{Haskell, Redis, Type-Level Programming Type Safety, Key-Value Store}
+\keywords{Haskell, Redis, Type-Level Programming, Type Safety, Key-Value Store}
 
 \maketitle
+\thispagestyle{empty}
 

### 2017-07-14 TyDe2017/sections/TypeLevelDict-orig.tex
Auto stash before merge of "master" and "origin/master"

@@
 \ensuremath{(\mathbin{:})\mathbin{::}\Varid{k}\to [\mskip1.5mu \Varid{k}\mskip1.5mu]\to [\mskip1.5mu \Varid{k}\mskip1.5mu]}.
-For example, \ensuremath{\Conid{Int}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Char}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[~]))} is a
+For example, \ensuremath{\Conid{Int}\mathrel{\,\mbox{\textquotesingle}\!:}(\Conid{Char}\mathrel{\,\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathrel{\,\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[~]))} is a
 type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} --- it is a list of (lifted) types. The optional quote
@@
 \>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{k},{}\<[20]%
-\>[20]{}\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{k}{}\<[33]%
+\>[20]{}\Varid{x})\mathrel{\,\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{k}{}\<[33]%
 \>[33]{}\mathrel{=}{}\<[33E]%
@@
 \>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{t},{}\<[20]%
-\>[20]{}\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{k}{}\<[33]%
+\>[20]{}\Varid{x})\mathrel{\,\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{k}{}\<[33]%
 \>[33]{}\mathrel{=}{}\<[33E]%
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Conid{Set}\;(\mbox{\textquotesingle}(\Varid{k},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
+\>[5]{}\Conid{Set}\;(\mbox{\textquotesingle}(\Varid{k},\Varid{y})\mathrel{\,\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
 \>[29]{}\Varid{k}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}(\Varid{k},\Varid{x}){}\<[48]%
-\>[48]{}\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs}{}\<[E]%
+\>[48]{}\mathrel{\,\mbox{\textquotesingle}\!:}\Varid{xs}{}\<[E]%
 \\
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Conid{Set}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
+\>[5]{}\Conid{Set}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathrel{\,\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[29]%
 \>[29]{}\Varid{k}\;\Varid{x}\mathrel{=}\mbox{\textquotesingle}(\Varid{t},\Varid{y}){}\<[48]%
-\>[48]{}\mathbin{\mbox{\textquotesingle}\!:}\Conid{Set}\;\Varid{xs}\;\Varid{k}\;\Varid{x}{}\<[E]%
+\>[48]{}\mathrel{\,\mbox{\textquotesingle}\!:}\Conid{Set}\;\Varid{xs}\;\Varid{k}\;\Varid{x}{}\<[E]%
 \\[\blanklineskip]%
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Conid{Del}\;(\mbox{\textquotesingle}(\Varid{k},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[30]%
+\>[5]{}\Conid{Del}\;(\mbox{\textquotesingle}(\Varid{k},\Varid{y})\mathrel{\,\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[30]%
 \>[30]{}\Varid{k}\mathrel{=}\Varid{xs}{}\<[E]%
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Conid{Del}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[30]%
-\>[30]{}\Varid{k}\mathrel{=}\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathbin{\mbox{\textquotesingle}\!:}\Conid{Del}\;\Varid{xs}\;\Varid{k}{}\<[E]%
+\>[5]{}\Conid{Del}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathrel{\,\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[30]%
+\>[30]{}\Varid{k}\mathrel{=}\mbox{\textquotesingle}(\Varid{t},\Varid{y})\mathrel{\,\mbox{\textquotesingle}\!:}\Conid{Del}\;\Varid{xs}\;\Varid{k}{}\<[E]%
 \\[\blanklineskip]%
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Conid{Member}\;(\mbox{\textquotesingle}(\Varid{k},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[33]%
+\>[5]{}\Conid{Member}\;(\mbox{\textquotesingle}(\Varid{k},\Varid{x})\mathrel{\,\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[33]%
 \>[33]{}\Varid{k}\mathrel{=}\Conid{True}{}\<[E]%
@@
 \>[B]{}\hsindent{5}{}\<[5]%
-\>[5]{}\Conid{Member}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[33]%
+\>[5]{}\Conid{Member}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{x})\mathrel{\,\mbox{\textquotesingle}\!:}\Varid{xs})\;{}\<[33]%
 \>[33]{}\Varid{k}\mathrel{=}\Conid{Member}\;\Varid{xs}\;\Varid{k}{}\<[E]%

### 2017-07-25 TyDe2017/main-tyde.tex
Minor fixes before submission.
Reflect changes required by editor.

@@
-\documentclass[sigplan]{acmart}
+\documentclass[sigplan, screen, 9pt]{acmart}
 \settopmatter{printfolios=true}
-
-%\makeatletter\if@ACM@journal\makeatother
-%% Journal information (used by PACMPL format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{1}
-\acmArticle{1}
-\acmYear{2017}
-\copyrightyear{2017}
-\acmPrice{15.00}
 
@@
 \setcopyright{licensedothergov}
+\acmPrice{15.00}
 \acmDOI{10.1145/3122975.3122978}
-\acmISBN{978-1-4503-5183-6/17/09}
-\acmConference[TyDe'17]{2st ACM SIGPLAN International Workshop on Type-Driven Development}{September 3, 2017}{Oxford, UK}
 \acmYear{2017}
 \copyrightyear{2017}
-\acmPrice{15.00}
+\acmISBN{978-1-4503-5183-6/17/09}
+\acmConference[TyDe'17]{2nd ACM SIGPLAN International Workshop on Type-Driven Development}{September 3, 2017}{Oxford, UK}
 
@@
 \usepackage{amsmath}
-%\usepackage{amsfonts}
-%\usepackage{amsthm}
-%\usepackage{color}
 \usepackage{xcolor}
 \usepackage{url}
-%\usepackage{mathptmx}
 \usepackage{doubleequals}
-%\usepackage{makeidx}  % allows for indexgeneration
 \usepackage{textcomp} % for single quote
@@
 
-\pagestyle{empty} %no page number
+\pagenumbering{gobble} % no page number
 \setlength{\mathindent}{10pt}
@@
 
-\title{Type Safe Redis Queries}
-\subtitle{A Case Study of Type--Level Programming in Haskell}
+\title{Type Safe Redis Queries: A Case Study of Type-Level Programming in Haskell}
 
-\author{Ting--Yan Lai}
-\authornote{Ting--Yan Lai was a summer intern from
+\author{Ting-Yan Lai}
+\authornote{Ting-Yan Lai was a summer intern from
 the Inst. of Computer Science and Engineering,
@@
 
-\author{Tyng--Ruey Chuang}
+\author{Tyng-Ruey Chuang}
 \affiliation{
@@
 
-\author{Shin--Cheng Mu}
+\author{Shin-Cheng Mu}
 \affiliation{

### 2017-08-19 TyDe2017/main-tyde.tex
New acmart.cls and acmthm.sty, to fix font problem.
Minor changes regarding font issues.
No-copyright footnote.

@@
-\documentclass[sigplan, screen, 9pt]{acmart}
+\documentclass[sigplan]{acmart}
 \settopmatter{printfolios=true}
@@
 
-\title{Type Safe Redis Queries: A Case Study of Type-Level Programming in Haskell}
+\title{Type Safe Redis Queries:\\A Case Study of Type-Level Programming in Haskell}
+
+\titlenote{\copyright\ \!\!\!\!\!\!\!\! $\slash$ ~~~ No Copyright. This work is in the public domain. For avoidance of
+doubt, this work is released under the CC0 Public Domain Dedication
+(\url{http://creativecommons.org/publicdomain/zero/1.0/}). Please search
+for free copy.}
 
